### 2020-09-11 README.Rmd
Make README a dynamic file, closes #21

---
title: "Code Club Project: Assessing whether intra and inter-genomic variation hinder utility of ASVs"
author: "Pat Schloss"
date: "9/11/2020"
output:
  github_document:
    html_preview: false
---

Developed over a series of *Code Club* episodes led by Pat Schloss to answer an important question in microbiology and develop comfort using tools to develop reproducible research practices.

## Questions
* Within a genome, how many distinct sequences of the 16S rRNA gene are present relative to the number of copies per genome? How far apart are these sequences from each other? How does this scale from a genome to kingdoms?
* Within a taxa (any level), how many ASVs from that taxa are shared with sister taxa? How does this change with taxonomic level? Variable region?

* Make sure we have taxonomic data for all of our genomes
* Read FASTA files into R (do it on our own)
* inner_join with tsv file
* group_by / summarize to count number of sequences and copies per genome

```{r library_calls, message=FALSE, echo=FALSE}
library(tidyverse)
library(data.table)
library(rmarkdown)
```

### Dependencies:  
* [mothur v.1.44.2](https://github.com/mothur/mothur/tree/v.1.44.2) - `code/install_mothur.sh` installs mothur
* `wget`
* `r R.version.string`
  * `tidyverse` (v. `r packageVersion("tidyverse")`)
  * `data.table` (v. `r packageVersion("data.table")`)
  * `rmarkdown` (v. `r packageVersion("rmarkdown")`)

### My computer

```{r sessioninfo, echo=FALSE}
sessionInfo()
```

### 2020-09-11 exploratory/2020-09-09-genome-sens-spec.Rmd
Convert Rscript to Rmd to view on GitHub, closes #22

---
title: "Analyzing the sensitivity and specificity of ASVs for discriminating between genomes"
author: "Pat Schloss"
date: "9/9/2020"
output:
  github_document:
    html_preview: false
---

```{r, message=FALSE}
library(tidyverse)
library(here)
```

### Need to determine the numbe of *rrn* operons across genomes

Our analysis will use full length sequences

```{r}
fl <- read_tsv(here("data/v19/rrnDB.count_tibble"))
```

We want to count and plot the number of copies per genome

```{r n_rrn}
fl %>%
	group_by(genome) %>%
	summarize(n_rrn = sum(count), .groups="drop") %>%
	ggplot(aes(x=n_rrn)) + geom_histogram(binwidth=1)

fl %>%
	group_by(genome) %>%
	summarize(n_rrn = sum(count), .groups="drop") %>%
	count(n_rrn) %>%
	mutate(fraction = n / sum(n))
```

We see that most genomes actually have more than one copy of the *rrn* operon. I wonder whether those different copies are the same sequence / ASV...

### Determine number of ASVs per genome

Considering most genomes have multiple copes of the *rrn* operon, we need to know whether they all have the same ASV. Otherwise we run the risk of splitting a single genome into multiple ASVs.

```{r}
fl %>% 
	group_by(genome) %>%
	summarize(n_asv = n(), n_rrn = sum(count), .groups="drop") %>%
	group_by(n_rrn) %>%
	summarize(med_n_asv = median(n_asv),
						lq_n_asv = quantile(n_asv, prob=0.25),
						uq_n_asv = quantile(n_asv, prob=0.75))

fl %>% 
	group_by(genome) %>%
	summarize(n_asv = n(), n_rrn = sum(count), .groups="drop") %>%
	ggplot(aes(x=n_rrn, y=n_asv)) + geom_smooth(method="lm")
```

Surprisingly (or not!) the number of ASVs increases at a rate of about 2 ASVs per 3 copies of *rrn* operon in the genome.

### Determine whether an ASV is unique to genomes they're found in

Instead of looking at the number of ASVs per genome, we want to see the number of genomes per ASV.

```{r}
fl %>%
	group_by(asv) %>%
	summarize(n_genomes = n()) %>%
	count(n_genomes) %>%
	mutate(fraction = n/sum(n))
```

We see that will full length sequences, that 82% of the ASVs were unique to a genome.

### Does the sensitivity and specificity change if we look at a shorter region of the gene?

We know that the V4 region is less diverse than the full length sequence. So does the number of ASVs per genome differ than for full length sequences? Are ASVs as specific when using the V4 region compared to full length sequences?

```{r}
v_four <- read_tsv(here("data/v4/rrnDB.count_tibble"))

v_four %>% 
	group_by(genome) %>%
	summarize(n_asv = n(), n_rrn = sum(count), .groups="drop") %>%
	group_by(n_rrn) %>%
	summarize(mean_n_asv = mean(n_asv),
						lq_n_asv = quantile(n_asv, prob=0.25),
						uq_n_asv = quantile(n_asv, prob=0.75))

v_four %>% 
	group_by(genome) %>%
	summarize(n_asv = n(), n_rrn = sum(count), .groups="drop") %>%
	ggplot(aes(x=n_rrn, y=n_asv)) + geom_smooth(method="lm")
```

The number of ASVs per copy of *rrn* operon is lower than for full length sequences. We find 1.5 ASV per 10 copies of the *rrn* operon.

Next, let's look at the specificity of an ASV for a genome

```{r}
v_four %>%
	group_by(asv) %>%
	summarize(n_genomes = n()) %>%
	count(n_genomes) %>%
	mutate(fraction = n/sum(n))
```

We find that about 76% of ASVs using the V4 region were specific to a genome.

### To be determined...
* Can correct for over representation? 
* Consider analysis at species, genus, family, etc. levels
* Consider looking at more broad definition of an ASV

### 2020-09-16 exploratory/2020-09-09-genome-sens-spec.Rmd
Combine the count_tibble files and update genome-level exploratory analysis, closes #23

@@
     html_preview: false
+editor_options: 
+  chunk_output_type: console
 ---
@@
 ```{r}
-fl <- read_tsv(here("data/v19/rrnDB.count_tibble"))
+count_tibble <- read_tsv(here("data/processed/rrnDB.count_tibble"),
+												 col_types = "cccd")
 ```
@@
 ```{r n_rrn}
-fl %>%
+count_tibble %>%
+	filter(region == "v19") %>%
 	group_by(genome) %>%
@@
 
-fl %>%
+count_tibble %>%
+	filter(region == "v19") %>%
 	group_by(genome) %>%
-	summarize(n_rrn = sum(count), .groups="drop") %>%
+	summarize(n_rrn = sum(count)) %>%
 	count(n_rrn) %>%
@@
 ```{r}
-fl %>% 
-	group_by(genome) %>%
+count_tibble %>% 
+	group_by(region, genome) %>%
 	summarize(n_asv = n(), n_rrn = sum(count), .groups="drop") %>%
-	group_by(n_rrn) %>%
+	group_by(region, n_rrn) %>%
 	summarize(med_n_asv = median(n_asv),
+						mean_n_asv = mean(n_asv),
 						lq_n_asv = quantile(n_asv, prob=0.25),
-						uq_n_asv = quantile(n_asv, prob=0.75))
+						uq_n_asv = quantile(n_asv, prob=0.75)) %>%
+	filter(n_rrn == 7)
 
-fl %>% 
-	group_by(genome) %>%
+count_tibble %>% 
+	group_by(region, genome) %>%
 	summarize(n_asv = n(), n_rrn = sum(count), .groups="drop") %>%
-	ggplot(aes(x=n_rrn, y=n_asv)) + geom_smooth(method="lm")
+	ggplot(aes(x=n_rrn, y=n_asv, color=region)) + geom_smooth(method="lm")
 ```
 
-Surprisingly (or not!) the number of ASVs increases at a rate of about 2 ASVs per 3 copies of *rrn* operon in the genome.
+Surprisingly (or not!) the number of ASVs increases at a rate of about 2 ASVs per 3 copies of *rrn* operon in the genome. The sub regions of the 16S rRNA region have few ASVs per *rrn* operon.
 
@@
 ```{r}
-fl %>%
-	group_by(asv) %>%
+count_tibble %>%
+	group_by(region, asv) %>%
 	summarize(n_genomes = n()) %>%
 	count(n_genomes) %>%
-	mutate(fraction = n/sum(n))
+	mutate(fraction = n/sum(n)) %>%
+	filter(n_genomes == 1)
 ```
 
-We see that will full length sequences, that 82% of the ASVs were unique to a genome.
-
-### Does the sensitivity and specificity change if we look at a shorter region of the gene?
-
-We know that the V4 region is less diverse than the full length sequence.
-So does the number of ASVs per genome differ than for full length sequences?
-Are ASVs as specific when using the V4 region compared to full length sequences?
-
-```{r}
-v_four <- read_tsv(here("data/v4/rrnDB.count_tibble"))
-
-v_four %>% 
-	group_by(genome) %>%
-	summarize(n_asv = n(), n_rrn = sum(count), .groups="drop") %>%
-	group_by(n_rrn) %>%
-	summarize(mean_n_asv = mean(n_asv),
-						lq_n_asv = quantile(n_asv, prob=0.25),
-						uq_n_asv = quantile(n_asv, prob=0.75))
-
-v_four %>% 
-	group_by(genome) %>%
-	summarize(n_asv = n(), n_rrn = sum(count), .groups="drop") %>%
-	ggplot(aes(x=n_rrn, y=n_asv)) + geom_smooth(method="lm")
-```
-
-The number of ASVs per copy of *rrn* operon is lower than for full length sequences. We find 1.5 ASV per 10 copies of the *rrn* operon.
-
-Next, let's look at the specificity of an ASV for a genome
-
-```{r}
-v_four %>%
-	group_by(asv) %>%
-	summarize(n_genomes = n()) %>%
-	count(n_genomes) %>%
-	mutate(fraction = n/sum(n))
-```
-
-We find that about 76% of ASVs using the V4 region were specific to a genome.
+We see that will full length sequences, that 82% of the ASVs were unique to a genome. For the subregions, about 76% of the ASVs were unique to a genome.
 

### 2020-09-16 code/combine_count_tibble_files.R

@@ -1,0 +1,19 @@
+#!/usr/bin/env Rscript --vanilla
+
+# name: combine_count_tibble_files.R
+#
+# input: tidy versions of count_files for each region
+# output: composite tidy count file with a column to designate the region
+# note: the input file names need to be 'data/<region>/rrnDB.count_tibble'
+
+
+library(tidyverse)
+
+tibble_files <- commandArgs(trailingOnly=TRUE)
+
+names(tibble_files) <- str_replace(string=tibble_files,
+																	 pattern="data/(.*)/rrnDB.count_tibble",
+																	 replacement="\\1")
+
+map_dfr(.x=tibble_files, .f=read_tsv, .id="region", col_types="ccd") %>%
+	write_tsv("data/processed/rrnDB.count_tibble")

### 2020-09-18 code/get_genome_id_taxonomy.R
Create genome to taxonomy mapping file, closes #24

@@ -1,0 +1,52 @@
+#!/usr/bin/env Rscript --vanilla
+
+# name: get_genome_id_taxonomy.R
+#
+# input: 
+# 	- data/raw/rrnDB-5.6.tsv
+#		- data/references/sp_spp_lookup.tsv
+#		- data/raw/rrnDB-5.6_pantaxa_stats_NCBI.tsv
+# output: tsv containing the genome id along with taxonomic information
+# 				data/references/genome_id_taxonomy.tsv
+
+library(tidyverse)
+
+
+	
+metadata <- read_tsv("data/raw/rrnDB-5.6.tsv") %>%
+		rename(genome_id = `Data source record id`, 
+					 subspecies_id = `NCBI tax id`,
+					 rdp = `RDP taxonomic lineage`,
+					 scientific_name = `NCBI scientific name`
+					 ) %>%
+		select(genome_id, subspecies_id, rdp, scientific_name)
+
+sp_spp_lookup <- read_tsv("data/references/sp_spp_lookup.tsv",
+				 col_names=c("domain", "species_id", "subspecies_id"))
+
+makeup_tax <- tibble(
+	taxid = c(299583, 328813, 412965, 693153),
+	species = c("Francisella orientalis", "Alistipes onderdonkii",
+							"Candidatus Vesicomyosocius okutanii", "Vibrio atlanticus"))
+
+tax <- read_tsv("data/raw/rrnDB-5.6_pantaxa_stats_NCBI.tsv") %>%
+	filter(rank == "species") %>%
+	rename(species = name) %>%
+	select(taxid, species) %>%
+	bind_rows(., makeup_tax)
+
+test <- inner_join(metadata, sp_spp_lookup, by="subspecies_id") %>%
+	select(genome_id, rdp, scientific_name, species_id) %>%
+	anti_join(., tax, by=c("species_id" = "taxid")) %>%
+	count(species_id) %>%
+	nrow(.) == 0
+
+stopifnot(test)
+
+
+inner_join(metadata, sp_spp_lookup, by="subspecies_id") %>%
+	select(genome_id, rdp, scientific_name, species_id) %>%
+	inner_join(., tax, by=c("species_id" = "taxid")) %>%
+	select(genome_id, rdp, species, scientific_name) %>% 
+	write_tsv("data/references/genome_id_taxonomy.tsv")
+	

### 2020-09-18 code/get_sp_spp_lookup.sh

@@ -1,0 +1,17 @@
+#!/usr/bin/env bash
+
+# author: Pat Schloss
+# inputs:
+# outputs: data/references/sp_spp_lookup.tsv
+#
+#
+
+wget -P data/references/ -nc https://ftp.ncbi.nlm.nih.gov/pub/taxonomy/taxcat.zip
+unzip -n -d data/references/ data/references/taxcat.zip
+
+if [[ $? -eq 0 ]]
+then
+	mv data/references/categories.dmp data/references/sp_spp_lookup.tsv
+else
+	echo "FAIL: were not able to successfully extract $filename"
+fi

### 2020-09-29 exploratory/2020-09-29-taxa-representation.Rmd
Determine number of genomes and taxa represented at each taxonomic level, closes #27

---
title: "Analyzing the distribution of genomes across different taxonomic ranks"
author: "Pat Schloss"
date: "9/29/2020"
output:
  github_document:
    html_preview: false
editor_options:
  chunk_output_type: console
---

```{r, message=FALSE}
library(tidyverse)
library(here)

metadata <- read_tsv(here("data/references/genome_id_taxonomy.tsv"),
										 col_types = cols(.default = col_character())) %>%
	mutate(strain = if_else(scientific_name == species,
													NA_character_,
													scientific_name)) %>%
	select(-scientific_name) %>% 
	pivot_longer(-genome_id, names_to="rank", values_to="taxon") %>%
	drop_na(taxon) %>%
	mutate(rank = factor(rank,
											 levels=c("kingdom", "phylum", "class", "order",
											 				 "family", "genus", "species", "strain")))

asv <- read_tsv(here("data/processed/rrnDB.count_tibble"),
								col_types = cols(.default = col_character(),
																 count = col_integer()))

metadata_asv <- inner_join(metadata, asv, by=c("genome_id" = "genome"))
```

### Find the number of taxa within each taxonomic rank

```{r}
n_taxa_per_rank <- metadata_asv %>%
	filter(region == "v19") %>% 
	group_by(rank, taxon) %>%
	summarize(N = n_distinct(genome_id)) %>%
	summarize(n_1 = n_distinct(taxon),
						n_2 = sum(N >= 2),
						n_3 = sum(N >= 3),
						n_4 = sum(N >= 4),
						n_5 = sum(N >= 5),
						n_10 = sum(N >= 10)
						) %>%
	pivot_longer(-rank, names_to = "n_genomes", values_to="n_taxa") %>%
	mutate(n_genomes = factor(n_genomes,
														levels=c("n_1", "n_2", "n_3", "n_4", "n_5", "n_10")))

n_taxa_per_rank %>% 
	ggplot(aes(x=rank, y=n_taxa, group=n_genomes, color=n_genomes)) +
		geom_line() + 
		theme_classic()
```

* Even if we require that ever taxonomic group had at least 5 genomes, there would be a few hundred species represented
* Not sure I trust the strain level data as being complete. There maybe taxa in the strain rank that are in the species but the people that deposited the data didn't indicate the correct strain name

### 2020-10-05 exploratory/2020-10-05-rrn-copy-number.Rmd
Construct stripchart of rrn copies per taxonomic group and level, closes #28

---
title: "Quantifying the number of *rrn* operons across taxonomic ranks"
author: "Pat Schloss"
date: "10/05/2020"
output:
  github_document:
    html_preview: false
editor_options:
  chunk_output_type: console
---

```{r, message=FALSE}
library(tidyverse)
library(here)

metadata <- read_tsv(here("data/references/genome_id_taxonomy.tsv"),
										 col_types = cols(.default = col_character())) %>%
	mutate(strain = if_else(scientific_name == species,
													NA_character_,
													scientific_name)) %>%
	select(-scientific_name)

asv <- read_tsv(here("data/processed/rrnDB.count_tibble"),
								col_types = cols(.default = col_character(),
																 count = col_integer()))

metadata_asv <- inner_join(metadata, asv, by=c("genome_id" = "genome"))
```

### Plot the number of *rrn* copies per taxonomic rank

Our analysis will use full length sequences. We want to count and plot the number of copies per *taxonomic rank*. Before calculating the averages of each taxonomic group, we should first get an average number of copies for each species. This will allow us to control for uneven number of genomes in each species.

```{r}
rank_taxon_rrns <- metadata_asv %>%
	filter(region == "v19") %>%
	group_by(kingdom, phylum, class, order, family, genus, species, genome_id) %>%
	summarize(n_rrns = sum(count), .groups="drop") %>% 
	group_by(kingdom, phylum, class, order, family, genus, species) %>%
	summarize(mean_rrns = mean(n_rrns), .groups="drop") %>% 
	pivot_longer(-mean_rrns, names_to="rank", values_to="taxon") %>%
	drop_na(taxon) %>%
	mutate(rank = factor(rank,
											 levels=c("kingdom", "phylum", "class", "order",
											 				 "family", "genus", "species", "strain"))) %>%
	group_by(rank, taxon) %>%
	summarize(mean_rrns = mean(mean_rrns), .groups="drop")

rank_taxon_rrns %>%
	ggplot(aes(x=rank, y=mean_rrns)) + 
	geom_jitter(width=0.3, alpha=0.3) +
	theme_classic() +
	labs(x=NULL, 
			 y="Mean number of rrn copies per genome",
			 title="There is wide variation at the species\nlevel in the number of rrn copies",
			 subtitle="Each point represents a single taxon\nwithin that rank, numbers based on\nan average species copy number")
```

* Bacteria have more copies than Archaea
* Even after correcting for number of genomes per species, there is wide variation in the number of rrn operons per taxonomic group

### 2020-10-05 exploratory/2020-10-05-rrn-copy-number.Rmd
Fix line breaks in plot title

@@
 			 y="Mean number of rrn copies per genome",
-			 title="There is wide variation at the species\nlevel in the number of rrn copies",
-			 subtitle="Each point represents a single taxon\nwithin that rank, numbers based on\nan average species copy number")
+			 title="There is wide variation at the species level in the number of rrn copies",
+			 subtitle="Each point represents a single taxon within that rank, numbers based on\nan average species copy number")
 ```

### 2020-10-07 exploratory/2020-10-05-rrn-copy-number.Rmd
Add line segment to cloud of points to indicate average, closes #29

@@
     html_preview: false
-editor_options:
-  chunk_output_type: console
 ---
@@
 
+mean_of_means <- rank_taxon_rrns %>%
+	group_by(rank) %>%
+	summarize(mean_mean_rrns = mean(mean_rrns), .groups="drop")
+
+jitter_width <- 0.3
+n_ranks <- nrow(mean_of_means)
+
 rank_taxon_rrns %>%
 	ggplot(aes(x=rank, y=mean_rrns)) + 
-	geom_jitter(width=0.3, alpha=0.3) +
+	geom_jitter(width=jitter_width, color="gray") +
+	geom_segment(data=mean_of_means,
+							 aes(x=1:n_ranks-jitter_width, xend=1:n_ranks+jitter_width,
+							 		y=mean_mean_rrns, yend=mean_mean_rrns),
+						 color="red", group=1, size=2, lineend="round",
+						 inherit.aes=FALSE) + 
 	theme_classic() +
@@
 			 title="There is wide variation at the species level in the number of rrn copies",
-			 subtitle="Each point represents a single taxon within that rank, numbers based on\nan average species copy number")
+			 subtitle="Each point represents a single taxon within that rank, numbers based on\nan average species copy number") +
+	scale_x_discrete(breaks=c("kingdom", "phylum", "class", "order", "family", "genus", "species"),
+									 labels=c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species"))
 ```

### 2020-10-08 exploratory/2020-10-05-rrn-copy-number.Rmd
Create ridgeline plot of average number of rrn copies per species, closes #29

--- 2020-10-05 exploratory/2020-10-05-rrn-copy-number.Rmd
@@
 
+median_of_means <- rank_taxon_rrns %>%
+	group_by(rank) %>%
+	summarize(median_mean_rrns = median(mean_rrns), .groups="drop")
+
+jitter_width <- 0.3
+n_ranks <- nrow(median_of_means)
+
 rank_taxon_rrns %>%
 	ggplot(aes(x=rank, y=mean_rrns)) + 
-	geom_jitter(width=0.3, alpha=0.3) +
+	geom_jitter(width=jitter_width, color="gray") +
+	geom_segment(data=median_of_means,
+							 aes(x=1:n_ranks-jitter_width, xend=1:n_ranks+jitter_width,
+							 		y=median_mean_rrns, yend=median_mean_rrns),
+						 color="red", group=1, size=2, lineend="round",
+						 inherit.aes=FALSE) + 
 	theme_classic() +
@@
 			 title="There is wide variation at the species level in the number of rrn copies",
-			 subtitle="Each point represents a single taxon within that rank, numbers based on\nan average species copy number")
+			 subtitle="Each point represents a single taxon within that rank, numbers based on\nan average species copy number") +
+	scale_x_discrete(breaks=c("kingdom", "phylum", "class", "order", "family", "genus", "species"),
+									 labels=c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species"))
+```
+
+Here's another way of looking at the data...
+
+```{r}
+library(ggridges)
+
+rank_taxon_rrns %>%
+	ggplot(aes(y=rank, x=mean_rrns)) + 
+	geom_density_ridges(stat="binline", binwidth=1, scale=0.9) +
+	geom_point(data=median_of_means, aes(x=median_mean_rrns, y=1:n_ranks+0.1)) +
+	theme_classic() +
+	labs(y=NULL, 
+			 x="Mean number of rrn copies per genome",
+			 title="The distribution of rrn copies per genome is pretty consistent across ranks",
+			 subtitle="Each point represents a single taxon within that rank, numbers based on\nan average species copy number. The dot represents the median for the rank") +
+	scale_y_discrete(breaks=c("kingdom", "phylum", "class", "order", "family", "genus", "species"),
+									 labels=c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species"))
+
 ```

### 2020-10-14 exploratory/2020-10-15-asv-species-coverage.Rmd
Build faceted plot of asv rate by number of genomes, addresses #30

---
title: "Quantifying the coverage of ASVs within a species"
author: "Pat Schloss"
date: "10/05/2020"
output:
  github_document:
    html_preview: false
editor_options: 
  chunk_output_type: console
---

```{r, message=FALSE}
library(tidyverse)
library(here)

metadata <- read_tsv(here("data/references/genome_id_taxonomy.tsv"),
										 col_types = cols(.default = col_character())) %>%
	mutate(strain = if_else(scientific_name == species,
													NA_character_,
													scientific_name)) %>%
	select(-scientific_name)

asv <- read_tsv(here("data/processed/rrnDB.count_tibble"),
								col_types = cols(.default = col_character(),
																 count = col_integer()))

metadata_asv <- inner_join(metadata, asv, by=c("genome_id" = "genome"))
```

### Does the number of ASVs per species increase with sampling effort? Does this vary by region within the 16S rRNA gene?

We've previously seen that the number of ASVs per copy of the 16S rRNA gene is about 0.6. This means that if a genome has 10 copies of the 16S rRNA gene, we would expect to see about 6 different versions of the gene within that genome. Across genomes from the same species, we know that the number of copies of the gene is pretty consistent. So, if we look at multiple genomes from the same species, will we see the same versions of the gene or will we see new versions of the gene? To answer this, we would like to look at the relationship between the number of ASVs found in a species per number of genomes in the species versus the number of genomes. Besides looking at the full length sequences, let's also look at the V4, V3-V4, and V4-V5 regions.

```{r}
#x = number of genomes for that species
#y = ratio of the number of ASVs per genome
# each point represents a different species
# each facet represents a different region

species_asvs <- metadata_asv %>% 
	select(genome_id, species, region, asv) %>%
	group_by(region, species) %>%
	summarize(n_genomes = n_distinct(genome_id),
						n_asvs = n_distinct(asv),
						asv_rate = n_asvs/n_genomes,
						.groups="drop")

region_labels <- c("V1-V9", "V4", "V3-V4", "V4-V5")
names(region_labels) <- c("v19", "v4", "v34", "v45")

species_asvs %>%
	ggplot(aes(x=n_genomes, y=asv_rate)) +
	geom_point(alpha=0.2) +
	geom_smooth() + 
	facet_wrap(facet="region",
						 nrow=2,
						 strip.position = "top",
						 scales="fixed",
						 labeller=labeller(region = region_labels)) +
	scale_x_log10()+
	scale_y_log10() +
	theme_classic()
```

### Conclusions

* V1-V9 continues to add significant number of ASVs as more genomes are sampled
from a species
* The sub regions seem to have plateaued indicating that perhaps we'll always
add more ASVs for a species
* Perhaps we really are splitting genomes and species too finely with ASVs
* Would like to look at individual genomes in more detail

### 2020-10-16 exploratory/2020-10-15-asv-species-coverage.Rmd
Add table showing counts of most abundant species, closes #30

@@
 author: "Pat Schloss"
-date: "10/05/2020"
+date: "10/15/2020"
 output:
@@
 library(tidyverse)
+library(knitr)
 library(here)
@@
 species_asvs <- metadata_asv %>% 
-	select(genome_id, species, region, asv) %>%
+	select(genome_id, species, region, asv, count) %>%
 	group_by(region, species) %>%
@@
 						n_asvs = n_distinct(asv),
+						n_rrns = sum(count) / n_genomes,
 						asv_rate = n_asvs/n_genomes,
@@
 
+```{r}
+# species name, number of genomes, average number of rrns, number of ASVs across
+# genomes for each region of 16S rRNA gene
+
+count_table <- species_asvs %>%
+	select(species, region, n_genomes, n_rrns, n_asvs) %>%	
+	group_by(species) %>%
+	mutate(n_genomes = max(n_genomes),
+				 n_rrns = max(n_rrns)) %>%
+	ungroup() %>% 
+	pivot_wider(names_from=region, values_from=n_asvs) %>% 
+	arrange(species)
+
+#see also kableExtra
+count_table %>%
+	arrange(desc(n_genomes)) %>%
+	top_n(n_genomes, n=10) %>%
+	kable(caption = "Ten most commonly sequence species", digits=2)
+```
+
 ### Conclusions

### 2020-10-21 exploratory/2020-10-21-asv-taxa-overlap.Rmd
Create plot of overlap of ASVs across taxa with a rank, addresses #31

---
title: "Quantifying the overlap of ASVs between taxa"
author: "Pat Schloss"
date: "10/21/2020"
output:
  github_document:
    html_preview: false
editor_options: 
  chunk_output_type: inline
---

```{r, message=FALSE}
library(tidyverse)
library(here)
library(knitr)

metadata <- read_tsv(here("data/references/genome_id_taxonomy.tsv"),
										 col_types = cols(.default = col_character())) %>%
	mutate(strain = if_else(scientific_name == species,
													NA_character_,
													scientific_name)) %>%
	select(-scientific_name)

asv <- read_tsv(here("data/processed/rrnDB.count_tibble"),
								col_types = cols(.default = col_character(),
																 count = col_integer()))

metadata_asv <- inner_join(metadata, asv, by=c("genome_id" = "genome"))
```

### How often is the same ASV found in multiple taxa from the same rank?

Across the analyses, we've already seen that a single bacterial genome can have multiple ASVs and that depending on the species we are looking at, there may be as many ASVs as there are genome sequences for that species. To me this says that ASVs pose a risk at unnaturally splitting a species and even a genome into multiple taxonomic groupings! In my book, that's not good.

Today, I want to ask an equally important question: If I have an ASV, what's the probability that it is also found in another taxonomic group from the same rank? In other words, if I have an ASV from *Bacillus subtilis*, what's the probability of finding it in *Bacillus cereus*? Of course, it's probably more likely to find a *Bacillus subtilis* ASV in a more closely related organism like *Bacillus cereus* than *E. coli*. Perhaps we can control for relatedness in a future episode, but for today, I want to answer this question for any two taxa from the same rank.

```{r}
# metadata_asv - input data
overlap_data <- metadata_asv %>%
# - focus on taxonomic ranks - kingdom to species, asv, and region
	select(-genome_id, -count, -strain) %>%
# - make data frame tidy
	pivot_longer(cols=c(-region, -asv), names_to="rank", values_to="taxon") %>%
# - remove lines from the data where we don't have a taxonomy
	drop_na(taxon) %>%
# - remove redundant lines 
	distinct() %>%
# for each region and taxonomic rank, group by asvs
	group_by(region, rank, asv) %>%
# - for each asv - count the number of taxa
	summarize(n_taxa = n(), .groups="drop_last") %>%
# - count the number of asvs that appear in more than one taxa
	count(n_taxa) %>%
# - find the ratio of the number of asvs in more than one taxa to the total number of asvs
	summarize(overlap = sum((n_taxa > 1) * n) / sum(n), .groups="drop") %>%
	mutate(rank = factor(rank, levels=c("kingdom", "phylum", "class", "order", "family", "genus", "species"))) %>%
	mutate(overlap = 100*overlap)

# create a plot showing specificity at each taxonomic rank for each region
# x = taxonomic rank
# y = specificity or % of asvs found in more than one taxa
# geom = line plot
# different lines for each region of 16S rRNA gene
overlap_data %>%
	ggplot(aes(x=rank, y=overlap, group=region, color=region)) +
	geom_line()

overlap_data %>%
	filter(rank == "species") %>% kable
```

### Conclusions
* Sub regions are less specific at the species level than full-length. Still full-length has 3.6% overlap whereas the subregions are between 8.8 and 12.6% overlap
* This analysis does not control for uneven sampling of different species

### 2020-10-22 exploratory/2020-10-21-asv-taxa-overlap.Rmd
Randomly subset N genomes from each species, addresses #31

@@
 editor_options: 
-  chunk_output_type: inline
+  chunk_output_type: console
 ---
@@
 ```{r}
+# set RNG seed to Pat's birthday!
+set.seed(19760620)
+threshold <- 3
+
+# metadata_asv - species - genome_id
+subsample_species <- metadata_asv %>%
+	select(genome_id, species) %>%
+# return the distinct/unique rows
+	distinct() %>%
+# group_by the species
+	group_by(species) %>%
+# slice_sample on each species for N genomes
+	slice_sample(n=threshold) %>%
+	ungroup()
+
+good_species <- subsample_species %>% 
+# count number of genoems in each species
+	count(species) %>%
+# return species that have N genomes / filter out species with fewer than N genomes
+	filter(n == threshold) %>%
+	select(species)
+
+# going back to original list of species/genomes and return genome_ids from
+# species with N genomes
+
+subsampled_genomes <- inner_join(subsample_species, good_species,  by="species") %>%
+	select(genome_id)
+
 # metadata_asv - input data
-overlap_data <- metadata_asv %>%
+overlap_data <- inner_join(metadata_asv, subsampled_genomes, by="genome_id") %>%
 # - focus on taxonomic ranks - kingdom to species, asv, and region

### 2020-10-28 exploratory/2020-10-21-asv-taxa-overlap.Rmd
Randomize subsampling 100 times using threshold of 1, closes #31

@@
 set.seed(19760620)
-threshold <- 3
 
+get_subsample_result <- function(threshold){
 	# metadata_asv - species - genome_id
@@
 
+	return(overlap_data)
+}
+
+subsample_iterations <- map_dfr(1:100, ~get_subsample_result(threshold=1), .id="iterations")
+
+summary_overlap_data <- subsample_iterations %>%
+	group_by(region, rank) %>%
+	summarize(mean_overlap = mean(overlap),
+						lci = quantile(overlap, prob=0.025),
+						uci = quantile(overlap, prob=0.975),
+						.groups = "drop")
+
 # create a plot showing specificity at each taxonomic rank for each region
@@
 # different lines for each region of 16S rRNA gene
-overlap_data %>%
-	ggplot(aes(x=rank, y=overlap, group=region, color=region)) +
+
+summary_overlap_data %>%
+	ggplot(aes(x=rank, y=mean_overlap, group=region, color=region, ymin=lci, ymax=uci)) +
+	geom_linerange(show.legend=FALSE) + 
 	geom_line()
 
-overlap_data %>%
-	filter(rank == "species") %>% kable
+summary_overlap_data %>%
+	filter(rank == "species") %>% kable(digits=1)
 ```
@@
 ### Conclusions
-* Sub regions are less specific at the species level than full-length. Still full-length has 3.6% overlap whereas the subregions are between 8.8 and 12.6% overlap
-* This analysis does not control for uneven sampling of different species
+* Sub regions are less specific at every taxonomic rank than full-length sequences. Still full-length has 3.6% overlap whereas the sub-regions are between 10.2 and 15.0% overlap

### 2020-10-30 exploratory/2020-11-02-dominance-commonness-of-asvs.Rmd
Calculate dominance and commonness of ASVs for each species and region, closes #32

---
title: "Quantifying the dominance and commoness of ASVs"
author: "Pat Schloss"
date: "11/02/2020"
output:
  github_document:
    html_preview: false
editor_options: 
  chunk_output_type: console
---

```{r, message=FALSE}
library(tidyverse)
library(here)
library(knitr)

metadata <- read_tsv(here("data/references/genome_id_taxonomy.tsv"),
										 col_types = cols(.default = col_character())) %>%
	mutate(strain = if_else(scientific_name == species,
													NA_character_,
													scientific_name)) %>%
	select(-scientific_name)

asv <- read_tsv(here("data/processed/rrnDB.count_tibble"),
								col_types = cols(.default = col_character(),
																 count = col_integer()))

metadata_asv <- inner_join(metadata, asv, by=c("genome_id" = "genome"))
```

maybe there are many ASVs but only a handful of them are dominant?

```{r}

# want the number of genomes and average number of 16S rRNA gene (rrn) copies per
# genome for each species

n_genomes_rrn_copies_per_species <- 
	metadata_asv %>%
	# need the counts by region and by genome for each species
  select(region, genome_id, species, asv, count) %>%
	# group our ASVs by region and species
  group_by(region, species) %>%
  summarize(
  	# number of distinct genomes per region/species
  	n_genomes = n_distinct(genome_id), 
  	# summing up the number of 16S rRNA copies across all genomes for
  	# a species and region and divide by number of genomes
  	mean_rrns = sum(count) / n_genomes,
  	.groups="drop"
  )

# set the minimum number of genomes per species
min_n_genomes_per_species <- 5

# What percentage of genomes have an ASV? This focuses on the most
# common ASV for each species. Dominance is the percentage of genomes that an
# ASV is found in. So if a ASV shows up in 20 of 25 genomes for a species, it
# dominance is 0.80.

asv_dominance_per_species <-

	metadata_asv %>%

	# want to do our analysis for the species level for each region of 16S will
	# want to group our genomes by ASVs and species to count the number of genomes
	# that each ASV appears in
  select(region, genome_id, species, asv, count) %>%
  group_by(region, species, asv) %>%

	# count the number of genomes that each ASV appears in for region and species
	summarize(n_genomes_found = n_distinct(genome_id), .groups="drop") %>%

	# bring in the number of genomes and rrn copies per species so we can scale
	# our number of genomes that each ASV appeared in to get percent dominance
  inner_join(., n_genomes_rrn_copies_per_species, by=c("region", "species")) %>%
  mutate(dominance = n_genomes_found / n_genomes) %>%

	# want the dominant ASV for each region and species
	# want the number of genomes for each species and region so that we can filter
	#   our data to focus on those species with more genomes than the value of
	#   threshold
  group_by(region, species) %>%
  summarize(
  	n_genomes=max(n_genomes),
  	max_dominance = max(dominance),
  	.groups="drop"
  	) %>%
  filter(n_genomes >= min_n_genomes_per_species)

asv_dominance_per_species %>%
  ggplot(aes(x=max_dominance, fill=region)) +
    geom_histogram(binwidth=0.05) +
    facet_wrap(facet="region") +
		labs(y="Number of species", x="Dominance of ASV per species")

# what percentage of species have a max dominance over 80%
asv_dominance_per_species %>%
	group_by(region) %>%
  summarize(fraction = sum(max_dominance > 0.80)/n())
```

```{r}
asv_commonness_per_species <- metadata_asv %>%

	# want to do our analysis for the species level for each region of 16S will
	# want to group our genomes by ASVs and species to count the number of genomes
	# that each ASV appears in
  select(region, genome_id, species, asv, count) %>%
  group_by(region, species, asv) %>%

	# instead of counting the number of genoems an ASV is found in, I want to
	# count the total number of times each ASV appears across the genomes within
	# a species
	summarize(n_asvs = sum(count), .groups="drop") %>%

	# bring in the number of genomes and rrn copies per species so we can scale
	# our number of times that each ASV appeared in each species to get percent
	# commonness
  inner_join(., n_genomes_rrn_copies_per_species, by=c("region", "species")) %>%
  mutate(commonness = n_asvs/n_genomes/mean_rrns) %>%

	# want the most common ASV for each region and species
	# want the number of genomes for each species and region so that we can filter
	#   our data to focus on those species with more genomes than the value of
	#   threshold
	group_by(region, species) %>%
  summarize(n_genomes=max(n_genomes),
    max_commonness = max(commonness),
    .groups="drop") %>%
  filter(n_genomes >= min_n_genomes_per_species)

asv_commonness_per_species %>%
  ggplot(aes(x=max_commonness, fill=region)) +
    geom_histogram(binwidth=0.05) +
    facet_wrap(facet="region") + 
		labs(y="Number of species", x="Commonness of ASV per species")

# what percentage of species have an ASV that has a commonness over 80%
asv_commonness_per_species %>%
  group_by(region) %>%
  summarize(fraction = sum(max_commonness > 0.80)/n())
```

### Conclusions...
* Among the sub-regions, a majority of species have an ASV that is found in
  more than 80% of the genomes. For full length sequences, only 25% of ASVs
  are found across more than 80% of the genomes
* The most common ASVs (i.e. more than 80% of the rrn copies) account for
  46-74% of the genomes in the sub-regions and only 12% for the full length
  ASVs
* Underscores the problem that a single ASV is unlikely to be representative
  of the diversity of ASVs within the species

### 2020-11-04 README.Rmd
Rename ASVs to ESV and regenerate all data, closes #33

@@
 ---
-title: "Code Club Project: Assessing whether intra and inter-genomic variation hinder utility of ASVs"
+title: "Code Club Project: Assessing whether intra and inter-genomic variation hinder utility of ESVs and ASVs"
 author: "Pat Schloss"
@@
 How does this scale from a genome to kingdoms?
-* Within a taxa (any level), how many ASVs from that taxa are shared with sister taxa? How does this change with taxonomic level? Variable region?
-
-* Make sure we have taxonomic data for all of our genomes
-* Read FASTA files into R (do it on our own)
-* inner_join with tsv file
-* group_by / summarize to count number of sequences and copies per genome
+* Within a taxa (any level), how many ESVs or ASVs from that taxa are shared with sister taxa? How does this change with taxonomic level? Variable region?
 

### 2020-11-04 exploratory/2020-09-09-genome-sens-spec.Rmd
Rename ASVs to ESV and regenerate all data, closes #33

@@
 ---
-title: "Analyzing the sensitivity and specificity of ASVs for discriminating between genomes"
+title: "Analyzing the sensitivity and specificity of ESVs for discriminating between genomes"
 author: "Pat Schloss"
@@
 ```{r}
-count_tibble <- read_tsv(here("data/processed/rrnDB.count_tibble"),
+count_tibble <- read_tsv(here("data/processed/rrnDB.esv.count_tibble"),
 												 col_types = "cccd")
@@
 
-We see that most genomes actually have more than one copy of the *rrn* operon. I wonder whether those different copies are the same sequence / ASV...
+We see that most genomes actually have more than one copy of the *rrn* operon. I wonder whether those different copies are the same sequence / ESV...
 
-### Determine number of ASVs per genome
+### Determine number of ESVs per genome
 
-Considering most genomes have multiple copes of the *rrn* operon, we need to know whether they all have the same ASV. Otherwise we run the risk of splitting a single genome into multiple ASVs.
+Considering most genomes have multiple copes of the *rrn* operon, we need to know whether they all have the same ESV. Otherwise we run the risk of splitting a single genome into multiple ESVs.
 
@@
 	group_by(region, genome) %>%
-	summarize(n_asv = n(), n_rrn = sum(count), .groups="drop") %>%
+	summarize(n_esv = n(), n_rrn = sum(count), .groups="drop") %>%
 	group_by(region, n_rrn) %>%
-	summarize(med_n_asv = median(n_asv),
-						mean_n_asv = mean(n_asv),
-						lq_n_asv = quantile(n_asv, prob=0.25),
-						uq_n_asv = quantile(n_asv, prob=0.75)) %>%
+	summarize(med_n_esv = median(n_esv),
+						mean_n_esv = mean(n_esv),
+						lq_n_esv = quantile(n_esv, prob=0.25),
+						uq_n_esv = quantile(n_esv, prob=0.75)) %>%
 	filter(n_rrn == 7)
@@
 	group_by(region, genome) %>%
-	summarize(n_asv = n(), n_rrn = sum(count), .groups="drop") %>%
-	ggplot(aes(x=n_rrn, y=n_asv, color=region)) + geom_smooth(method="lm")
+	summarize(n_esv = n(), n_rrn = sum(count), .groups="drop") %>%
+	ggplot(aes(x=n_rrn, y=n_esv, color=region)) + geom_smooth(method="lm")
 ```
 
-Surprisingly (or not!) the number of ASVs increases at a rate of about 2 ASVs per 3 copies of *rrn* operon in the genome. The sub regions of the 16S rRNA region have few ASVs per *rrn* operon.
+Surprisingly (or not!) the number of ESVs increases at a rate of about 2 ESVs per 3 copies of *rrn* operon in the genome. The sub regions of the 16S rRNA region have few ESVs per *rrn* operon.
 
-### Determine whether an ASV is unique to genomes they're found in
+### Determine whether an ESV is unique to genomes they're found in
 
-Instead of looking at the number of ASVs per genome, we want to see the number of genomes per ASV.
+Instead of looking at the number of ESVs per genome, we want to see the number of genomes per ESV.
 
@@
 count_tibble %>%
-	group_by(region, asv) %>%
+	group_by(region, esv) %>%
 	summarize(n_genomes = n()) %>%
@@
 
-We see that will full length sequences, that 82% of the ASVs were unique to a genome. For the subregions, about 76% of the ASVs were unique to a genome.
+We see that will full length sequences, that 82% of the ESVs were unique to a genome. For the subregions, about 76% of the ESVs were unique to a genome.
 
@@
 * Consider analysis at species, genus, family, etc. levels
-* Consider looking at more broad definition of an ASV
+* Consider looking at more broad definition of an ESV

### 2020-11-04 exploratory/2020-09-29-taxa-representation.Rmd
Rename ASVs to ESV and regenerate all data, closes #33

@@
 
-asv <- read_tsv(here("data/processed/rrnDB.count_tibble"),
+esv <- read_tsv(here("data/processed/rrnDB.esv.count_tibble"),
 								col_types = cols(.default = col_character(),
@@
 
-metadata_asv <- inner_join(metadata, asv, by=c("genome_id" = "genome"))
+metadata_esv <- inner_join(metadata, esv, by=c("genome_id" = "genome"))
 ```
@@
 ```{r}
-n_taxa_per_rank <- metadata_asv %>%
+n_taxa_per_rank <- metadata_esv %>%
 	filter(region == "v19") %>%

### 2020-11-04 exploratory/2020-10-05-rrn-copy-number.Rmd
Rename ASVs to ESV and regenerate all data, closes #33

@@
 ```{r}
-rank_taxon_rrns <- metadata_asv %>%
+rank_taxon_rrns <- metadata_esv %>%
 	filter(region == "v19") %>%

### 2020-11-04 exploratory/2020-10-15-esv-species-coverage.Rmd
Rename ASVs to ESV and regenerate all data, closes #33

---
title: "Quantifying the coverage of ESVs within a species"
author: "Pat Schloss"
date: "10/15/2020"
output:
  github_document:
    html_preview: false
editor_options:
  chunk_output_type: console
---

```{r, message=FALSE}
library(tidyverse)
library(knitr)
library(here)

metadata <- read_tsv(here("data/references/genome_id_taxonomy.tsv"),
										 col_types = cols(.default = col_character())) %>%
	mutate(strain = if_else(scientific_name == species,
													NA_character_,
													scientific_name)) %>%
	select(-scientific_name)

esv <- read_tsv(here("data/processed/rrnDB.esv.count_tibble"),
								col_types = cols(.default = col_character(),
																 count = col_integer()))

metadata_esv <- inner_join(metadata, esv, by=c("genome_id" = "genome"))
```

### Does the number of ESVs per species increase with sampling effort? Does this vary by region within the 16S rRNA gene?

We've previously seen that the number of ESVs per copy of the 16S rRNA gene is about 0.6. This means that if a genome has 10 copies of the 16S rRNA gene, we would expect to see about 6 different versions of the gene within that genome. Across genomes from the same species, we know that the number of copies of the gene is pretty consistent. So, if we look at multiple genomes from the same species, will we see the same versions of the gene or will we see new versions of the gene? To answer this, we would like to look at the relationship between the number of ESVs found in a species per number of genomes in the species versus the number of genomes. Besides looking at the full length sequences, let's also look at the V4, V3-V4, and V4-V5 regions.

```{r}
#x = number of genomes for that species
#y = ratio of the number of ESVs per genome
# each point represents a different species
# each facet represents a different region

species_esvs <- metadata_esv %>%
	select(genome_id, species, region, esv, count) %>%
	group_by(region, species) %>%
	summarize(n_genomes = n_distinct(genome_id),
						n_esvs = n_distinct(esv),
						n_rrns = sum(count) / n_genomes,
						esv_rate = n_esvs/n_genomes,
						.groups="drop")

region_labels <- c("V1-V9", "V4", "V3-V4", "V4-V5")
names(region_labels) <- c("v19", "v4", "v34", "v45")

species_esvs %>%
	ggplot(aes(x=n_genomes, y=esv_rate)) +
	geom_point(alpha=0.2) +
	geom_smooth() +
	facet_wrap(facet="region",
						 nrow=2,
						 strip.position = "top",
						 scales="fixed",
						 labeller=labeller(region = region_labels)) +
	scale_x_log10()+
	scale_y_log10() +
	theme_classic()
```

```{r}
# species name, number of genomes, average number of rrns, number of ESVs across
# genomes for each region of 16S rRNA gene

count_table <- species_esvs %>%
	select(species, region, n_genomes, n_rrns, n_esvs) %>%
	group_by(species) %>%
	mutate(n_genomes = max(n_genomes),
				 n_rrns = max(n_rrns)) %>%
	ungroup() %>%
	pivot_wider(names_from=region, values_from=n_esvs) %>%
	arrange(species)

#see also kableExtra
count_table %>%
	arrange(desc(n_genomes)) %>%
	top_n(n_genomes, n=10) %>%
	kable(caption = "Ten most commonly sequence species", digits=2)
```

### Conclusions

* V1-V9 continues to add significant number of ESVs as more genomes are sampled
from a species
* The sub regions seem to have plateaued indicating that perhaps we'll always
add more ESVs for a species
* Perhaps we really are splitting genomes and species too finely with ESVs
* Would like to look at individual genomes in more detail

### 2020-11-04 exploratory/2020-10-21-esv-taxa-overlap.Rmd
Rename ASVs to ESV and regenerate all data, closes #33

---
title: "Quantifying the overlap of ESVs between taxa"
author: "Pat Schloss"
date: "10/21/2020"
output:
  github_document:
    html_preview: false
editor_options:
  chunk_output_type: console
---

```{r, message=FALSE}
library(tidyverse)
library(here)
library(knitr)

metadata <- read_tsv(here("data/references/genome_id_taxonomy.tsv"),
										 col_types = cols(.default = col_character())) %>%
	mutate(strain = if_else(scientific_name == species,
													NA_character_,
													scientific_name)) %>%
	select(-scientific_name)

esv <- read_tsv(here("data/processed/rrnDB.esv.count_tibble"),
								col_types = cols(.default = col_character(),
																 count = col_integer()))

metadata_esv <- inner_join(metadata, esv, by=c("genome_id" = "genome"))
```

### How often is the same ESV found in multiple taxa from the same rank?

Across the analyses, we've already seen that a single bacterial genome can have multiple ESVs and that depending on the species we are looking at, there may be as many ESVs as there are genome sequences for that species. To me this says that ESVs pose a risk at unnaturally splitting a species and even a genome into multiple taxonomic groupings! In my book, that's not good.

Today, I want to ask an equally important question: If I have an ESV, what's the probability that it is also found in another taxonomic group from the same rank? In other words, if I have an ESV from *Bacillus subtilis*, what's the probability of finding it in *Bacillus cereus*? Of course, it's probably more likely to find a *Bacillus subtilis* ESV in a more closely related organism like *Bacillus cereus* than *E. coli*. Perhaps we can control for relatedness in a future episode, but for today, I want to answer this question for any two taxa from the same rank.

```{r}
# set RNG seed to Pat's birthday!
set.seed(19760620)

get_subsample_result <- function(threshold){
	# metadata_esv - species - genome_id
	subsample_species <- metadata_esv %>%
		select(genome_id, species) %>%
	# return the distinct/unique rows
		distinct() %>%
	# group_by the species
		group_by(species) %>%
	# slice_sample on each species for N genomes
		slice_sample(n=threshold) %>%
		ungroup()

	good_species <- subsample_species %>%
	# count number of genoems in each species
		count(species) %>%
	# return species that have N genomes / filter out species with fewer than N genomes
		filter(n == threshold) %>%
		select(species)

	# going back to original list of species/genomes and return genome_ids from
	# species with N genomes

	subsampled_genomes <- inner_join(subsample_species, good_species,  by="species") %>%
		select(genome_id)

	# metadata_esv - input data
	overlap_data <- inner_join(metadata_esv, subsampled_genomes, by="genome_id") %>%
	# - focus on taxonomic ranks - kingdom to species, esv, and region
		select(-genome_id, -count, -strain) %>%
	# - make data frame tidy
		pivot_longer(cols=c(-region, -esv), names_to="rank", values_to="taxon") %>%
	# - remove lines from the data where we don't have a taxonomy
		drop_na(taxon) %>%
	# - remove redundant lines
		distinct() %>%
	# for each region and taxonomic rank, group by esvs
		group_by(region, rank, esv) %>%
	# - for each esv - count the number of taxa
		summarize(n_taxa = n(), .groups="drop_last") %>%
	# - count the number of esvs that appear in more than one taxa
		count(n_taxa) %>%
	# - find the ratio of the number of esvs in more than one taxa to the total number of esvs
		summarize(overlap = sum((n_taxa > 1) * n) / sum(n), .groups="drop") %>%
		mutate(rank = factor(rank, levels=c("kingdom", "phylum", "class", "order", "family", "genus", "species"))) %>%
		mutate(overlap = 100*overlap)

	return(overlap_data)
}

subsample_iterations <- map_dfr(1:100, ~get_subsample_result(threshold=1), .id="iterations")

summary_overlap_data <- subsample_iterations %>%
	group_by(region, rank) %>%
	summarize(mean_overlap = mean(overlap),
						lci = quantile(overlap, prob=0.025),
						uci = quantile(overlap, prob=0.975),
						.groups = "drop")

# create a plot showing specificity at each taxonomic rank for each region
# x = taxonomic rank
# y = specificity or % of esvs found in more than one taxa
# geom = line plot
# different lines for each region of 16S rRNA gene

summary_overlap_data %>%
	ggplot(aes(x=rank, y=mean_overlap, group=region, color=region, ymin=lci, ymax=uci)) +
	geom_linerange(show.legend=FALSE) +
	geom_line()

summary_overlap_data %>%
	filter(rank == "species") %>% kable(digits=1)
```

### Conclusions
* Sub regions are less specific at every taxonomic rank than full-length sequences. Still full-length has 3.6% overlap whereas the sub-regions are between 10.2 and 15.0% overlap

### 2020-11-04 exploratory/2020-11-02-dominance-commonness-of-esvs.Rmd
Rename ASVs to ESV and regenerate all data, closes #33

---
title: "Quantifying the dominance and commoness of ESVs"
author: "Pat Schloss"
date: "11/02/2020"
output:
  github_document:
    html_preview: false
editor_options:
  chunk_output_type: console
---

```{r, message=FALSE}
library(tidyverse)
library(here)
library(knitr)

metadata <- read_tsv(here("data/references/genome_id_taxonomy.tsv"),
										 col_types = cols(.default = col_character())) %>%
	mutate(strain = if_else(scientific_name == species,
													NA_character_,
													scientific_name)) %>%
	select(-scientific_name)

esv <- read_tsv(here("data/processed/rrnDB.esv.count_tibble"),
								col_types = cols(.default = col_character(),
																 count = col_integer()))

metadata_esv <- inner_join(metadata, esv, by=c("genome_id" = "genome"))
```

maybe there are many ESVs but only a handful of them are dominant?

```{r}

# want the number of genomes and average number of 16S rRNA gene (rrn) copies per
# genome for each species

n_genomes_rrn_copies_per_species <-
	metadata_esv %>%
	# need the counts by region and by genome for each species
  select(region, genome_id, species, esv, count) %>%
	# group our ESVs by region and species
  group_by(region, species) %>%
  summarize(
  	# number of distinct genomes per region/species
  	n_genomes = n_distinct(genome_id),
  	# summing up the number of 16S rRNA copies across all genomes for
  	# a species and region and divide by number of genomes
  	mean_rrns = sum(count) / n_genomes,
  	.groups="drop"
  )

# set the minimum number of genomes per species
min_n_genomes_per_species <- 5

# What percentage of genomes have an ESV? This focuses on the most
# common ESV for each species. Dominance is the percentage of genomes that an
# ESV is found in. So if a ESV shows up in 20 of 25 genomes for a species, it
# dominance is 0.80.

esv_dominance_per_species <-

	metadata_esv %>%

	# want to do our analysis for the species level for each region of 16S will
	# want to group our genomes by ESVs and species to count the number of genomes
	# that each ESV appears in
  select(region, genome_id, species, esv, count) %>%
  group_by(region, species, esv) %>%

	# count the number of genomes that each ESV appears in for region and species
	summarize(n_genomes_found = n_distinct(genome_id), .groups="drop") %>%

	# bring in the number of genomes and rrn copies per species so we can scale
	# our number of genomes that each ESV appeared in to get percent dominance
  inner_join(., n_genomes_rrn_copies_per_species, by=c("region", "species")) %>%
  mutate(dominance = n_genomes_found / n_genomes) %>%

	# want the dominant ESV for each region and species
	# want the number of genomes for each species and region so that we can filter
	#   our data to focus on those species with more genomes than the value of
	#   threshold
  group_by(region, species) %>%
  summarize(
  	n_genomes=max(n_genomes),
  	max_dominance = max(dominance),
  	.groups="drop"
  	) %>%
  filter(n_genomes >= min_n_genomes_per_species)

esv_dominance_per_species %>%
  ggplot(aes(x=max_dominance, fill=region)) +
    geom_histogram(binwidth=0.05) +
    facet_wrap(facet="region") +
		labs(y="Number of species", x="Dominance of ESV per species")

# what percentage of species have a max dominance over 80%
esv_dominance_per_species %>%
	group_by(region) %>%
  summarize(fraction = sum(max_dominance > 0.80)/n())
```

```{r}
esv_commonness_per_species <- metadata_esv %>%

	# want to do our analysis for the species level for each region of 16S will
	# want to group our genomes by ESVs and species to count the number of genomes
	# that each ESV appears in
  select(region, genome_id, species, esv, count) %>%
  group_by(region, species, esv) %>%

	# instead of counting the number of genoems an ESV is found in, I want to
	# count the total number of times each ESV appears across the genomes within
	# a species
	summarize(n_esvs = sum(count), .groups="drop") %>%

	# bring in the number of genomes and rrn copies per species so we can scale
	# our number of times that each ESV appeared in each species to get percent
	# commonness
  inner_join(., n_genomes_rrn_copies_per_species, by=c("region", "species")) %>%
  mutate(commonness = n_esvs/n_genomes/mean_rrns) %>%

	# want the most common ESV for each region and species
	# want the number of genomes for each species and region so that we can filter
	#   our data to focus on those species with more genomes than the value of
	#   threshold
	group_by(region, species) %>%
  summarize(n_genomes=max(n_genomes),
    max_commonness = max(commonness),
    .groups="drop") %>%
  filter(n_genomes >= min_n_genomes_per_species)

esv_commonness_per_species %>%
  ggplot(aes(x=max_commonness, fill=region)) +
    geom_histogram(binwidth=0.05) +
    facet_wrap(facet="region") +
		labs(y="Number of species", x="Commonness of ESV per species")

# what percentage of species have an ESV that has a commonness over 80%
esv_commonness_per_species %>%
  group_by(region) %>%
  summarize(fraction = sum(max_commonness > 0.80)/n())
```

### Conclusions...
* Among the sub-regions, a majority of species have an ESV that is found in
  more than 80% of the genomes. For full length sequences, only 25% of ESVs
  are found across more than 80% of the genomes
* The most common ESVs (i.e. more than 80% of the rrn copies) account for
  46-74% of the genomes in the sub-regions and only 12% for the full length
  ESVs
* Underscores the problem that a single ESV is unlikely to be representative
  of the diversity of ESVs within the species

### 2020-11-04 code/combine_count_tibble_files.R

@@ -12,8 +12,8 @@ library(tidyverse)
 tibble_files <- commandArgs(trailingOnly=TRUE)
 
 names(tibble_files) <- str_replace(string=tibble_files,
-																	 pattern="data/(.*)/rrnDB.count_tibble",
+																	 pattern="data/(.*)/rrnDB.esv.count_tibble",
 																	 replacement="\\1")
 
 map_dfr(.x=tibble_files, .f=read_tsv, .id="region", col_types="ccd") %>%
-	write_tsv("data/processed/rrnDB.count_tibble")
+	write_tsv("data/processed/rrnDB.esv.count_tibble")

### 2020-11-04 code/convert_count_table_to_tibble.R

@@ -3,7 +3,7 @@
 # name: convert_count_table_to_tibble.R
 #
 # input: mothur-formatted count file
-# output: tidy data frame with asv, genome, count as columns
+# output: tidy data frame with esv, genome, count as columns
 # note: we expect input to be in order of input , output
 
 library(tidyverse)
@@ -15,8 +15,8 @@ input_file <- args[1]
 output_file <- args[2]
 
 fread(input_file) %>%
-	rename(asv=Representative_Sequence) %>%
+	rename(esv=Representative_Sequence) %>%
 	select(-total) %>%
-	melt(id.vars="asv", variable.name="genome", value.name="count") %>%
+	melt(id.vars="esv", variable.name="genome", value.name="count") %>%
 	filter(count != 0) %>%
 	write_tsv(output_file)

### 2020-11-04 code/count_unique_seqs.sh

@@ -3,7 +3,7 @@
 # name: count_unique_seqs.sh
 #
 # takes in fasta file and uniques and counts the sequences. it then converts count_table to a
-# tibble with columns asv, genome, and count to indicate the number of times each asv appears
+# tibble with columns esv, genome, and count to indicate the number of times each esv appears
 # in each genome
 #
 # input: target - either the count_tibble or align file - also the output!
@@ -24,7 +24,7 @@ grep ">" $TEMP_ALIGN | sed -E "s/>((.*)\|.*\|.*)/\1 \2/" > $TEMP_GROUPS
 code/mothur/mothur "#unique.seqs(fasta=$TEMP_ALIGN);
 	count.seqs(group=$TEMP_GROUPS, compress=FALSE)"
 
-code/convert_count_table_to_tibble.R $STUB_TEMP.count_table $STUB.count_tibble
+code/convert_count_table_to_tibble.R $STUB_TEMP.count_table $STUB.esv.count_tibble
 
 mv $STUB_TEMP.unique.align $STUB.unique.align
 

### 2020-11-06 code/get_asvs.sh
Break apart ESVs, start making ASVs

@@ -1,0 +1,27 @@
+#
+# takes in distance file, count file, and cutoff value and generate a
+# count_tibble file to indicate the number of times each asv appears
+# in each genome
+#
+# input: target - file name in the format of data/v4/rrnDB.01.count_tibble
+#               - this file name is also the output!
+#               - the "01" in this example is threshold (i.e. 0.01)
+
+TARGET=$1 # data/v4/rrnDB.01.count_tibble
+STUB=`echo $TARGET | sed -E "s/(.*rrnDB).*/\1/"`
+THRESHOLD=`echo $TARGET | sed -E "s/.*rrnDB\.(.*).count_tibble/\1/"`
+
+DISTANCES=$STUB.unique.dist
+COUNT=$STUB.count_table
+CUTOFF=0.$THRESHOLD
+
+code/mothur/mothur "#cluster(column=$DISTANCES, count=$COUNT, cutoff=$CUTOFF);
+  make.shared()"
+
+
+# Garbage collection
+rm $STUB.unique.opti_mcc.list
+rm $STUB.unique.opti_mcc.steps
+rm $STUB.unique.opti_mcc.sensspec
+
+#data/v4/rrnDB.unique.opti_mcc.shared

### 2020-11-06 code/get_distances.sh

@@ -1,0 +1,11 @@
+#
+# takes in fasta file and calcualtes a distance file in mothur's column format
+# the default is to use a cutoff of 0.05
+#
+# input: target - the distance matrix name, which is also the output!
+
+TARGET=$1 # data/v4/rrnDB.unique.dist
+
+ALIGN=`echo $TARGET | sed -E "s/dist/align/"`
+
+code/mothur/mothur "#dist.seqs(fasta=$ALIGN, cutoff=0.05)"

### 2020-11-06 code/get_unique_seqs.sh

@@ -8,7 +8,7 @@
 #
 # input: target - either the count_tibble or align file - also the output!
 
-TARGET=$1 # data/v19/rrnDB.unique.align
+TARGET=$1 # data/v4/rrnDB.unique.align
 
 STUB=`echo $TARGET | sed -E "s/(.*rrnDB).*/\1/"`
 STUB_TEMP=$STUB.temp
@@ -24,11 +24,9 @@ grep ">" $TEMP_ALIGN | sed -E "s/>((.*)\|.*\|.*)/\1 \2/" > $TEMP_GROUPS
 code/mothur/mothur "#unique.seqs(fasta=$TEMP_ALIGN);
 	count.seqs(group=$TEMP_GROUPS, compress=FALSE)"
 
-code/convert_count_table_to_tibble.R $STUB_TEMP.count_table $STUB.esv.count_tibble
-
 mv $STUB_TEMP.unique.align $STUB.unique.align
+mv $STUB_TEMP.count_table $STUB.count_table
 
-rm $STUB_TEMP.count_table
 rm $STUB_TEMP.names
 rm $STUB_TEMP.groups
 rm $STUB_TEMP.align

### 2020-11-19 exploratory/2020-09-09-genome-sens-spec.Rmd
Combine count tibble files, closes #34

@@
 ```{r}
-count_tibble <- read_tsv(here("data/processed/rrnDB.esv.count_tibble"),
-												 col_types = "cccd")
+count_tibble <- read_tsv(here("data/processed/rrnDB.easv.count_tibble"),
+								col_types = cols(.default = col_character(),
+																 count = col_integer())) %>%
+	filter(threshold == "esv") %>%
+	select(-threshold)
 ```
@@
 count_tibble %>%
-	group_by(region, esv) %>%
+	group_by(region, easv) %>%
 	summarize(n_genomes = n()) %>%

### 2020-11-19 exploratory/2020-09-29-taxa-representation.Rmd
Combine count tibble files, closes #34

@@
 
-esv <- read_tsv(here("data/processed/rrnDB.esv.count_tibble"),
+esv <- read_tsv(here("data/processed/rrnDB.easv.count_tibble"),
 								col_types = cols(.default = col_character(),
-																 count = col_integer()))
+																 count = col_integer())) %>%
+	filter(threshold == "esv") %>%
+	select(-threshold)
 

### 2020-11-19 exploratory/2020-10-15-esv-species-coverage.Rmd
Combine count tibble files, closes #34

@@
 species_esvs <- metadata_esv %>%
-	select(genome_id, species, region, esv, count) %>%
+	select(genome_id, species, region, easv, count) %>%
 	group_by(region, species) %>%
 	summarize(n_genomes = n_distinct(genome_id),
-						n_esvs = n_distinct(esv),
+						n_esvs = n_distinct(easv),
 						n_rrns = sum(count) / n_genomes,

### 2020-11-19 exploratory/2020-10-21-esv-taxa-overlap.Rmd
Combine count tibble files, closes #34

@@
 	# - make data frame tidy
-		pivot_longer(cols=c(-region, -esv), names_to="rank", values_to="taxon") %>%
+		pivot_longer(cols=c(-region, -easv), names_to="rank", values_to="taxon") %>%
 	# - remove lines from the data where we don't have a taxonomy
@@
 	# for each region and taxonomic rank, group by esvs
-		group_by(region, rank, esv) %>%
-	# - for each esv - count the number of taxa
+		group_by(region, rank, easv) %>%
+	# - for each easv - count the number of taxa
 		summarize(n_taxa = n(), .groups="drop_last") %>%

### 2020-11-19 exploratory/2020-11-02-dominance-commonness-of-esvs.Rmd
Combine count tibble files, closes #34

@@
 	# need the counts by region and by genome for each species
-  select(region, genome_id, species, esv, count) %>%
+  select(region, genome_id, species, easv, count) %>%
 	# group our ESVs by region and species
@@
 	# that each ESV appears in
-  select(region, genome_id, species, esv, count) %>%
-  group_by(region, species, esv) %>%
+  select(region, genome_id, species, easv, count) %>%
+  group_by(region, species, easv) %>%
 

### 2020-11-19 code/combine_count_tibble_files.R

@@ -9,11 +9,35 @@
 
 library(tidyverse)
 
+# rename_function <- function(x){
+# 	"easv"
+# }
+
+read_count_tibble <- function(count_tibble_file){
+
+	region <- str_replace(string=count_tibble_file,
+											 pattern="data/(.*)/rrnDB\\..*\\.count_tibble",
+											 replacement="\\1")
+
+	type <- if_else(str_detect(count_tibble_file, "esv"), "esv", "asv")
+
+	threshold <- if_else(type == "esv", "esv",
+											 str_replace(string=count_tibble_file,
+													pattern="data/.*/rrnDB\\.(.*)\\.count_tibble",
+													replacement="0.\\1"))
+
+	# rename asv or esv to be easv for column name
+	read_tsv(count_tibble_file, col_types="ccd") %>%
+		mutate(region = region,
+					 threshold = threshold) %>%
+		rename_with(function(x){"easv"}, ends_with("sv"))
+
+
+}
+
+
 tibble_files <- commandArgs(trailingOnly=TRUE)
 
-names(tibble_files) <- str_replace(string=tibble_files,
-																	 pattern="data/(.*)/rrnDB.esv.count_tibble",
-																	 replacement="\\1")
 
-map_dfr(.x=tibble_files, .f=read_tsv, .id="region", col_types="ccd") %>%
-	write_tsv("data/processed/rrnDB.esv.count_tibble")
+map_dfr(.x=tibble_files, .f=read_count_tibble) %>%
+	write_tsv("data/processed/rrnDB.easv.count_tibble")

### 2020-11-24 exploratory/2020-11-24-threshold-to-drop-n-asvs.Rmd
Generate plots looking at E/ASVs per genome, closes #35

---
title: "Determine the threshold needed to get 1 ASV per genome"
author: "Pat Schloss"
date: "11/24/2020"
output:
  github_document:
    html_preview: false
editor_options:
  chunk_output_type: console
---

```{r, message=FALSE}
library(tidyverse)
library(here)
library(knitr)

set.seed(19760620)

metadata <- read_tsv(here("data/references/genome_id_taxonomy.tsv"),
										 col_types = cols(.default = col_character())) %>%
	select(genome_id, species) %>%
	group_by(species) %>% # Get one genome per species
	slice_sample(n=1) %>%
	ungroup()

easv <- read_tsv(here("data/processed/rrnDB.easv.count_tibble"),
								col_types = cols(.default = col_character(),
																 count = col_integer()))

metadata_easv <- inner_join(metadata, easv, by=c("genome_id" = "genome")) %>%
	mutate(threshold = recode(threshold, "esv" = "0.000"),
				 threshold = as.numeric(threshold))
```

### Overivew

We saw earlier that a problem with E/ASVs is the risk of splitting the operons from the same genome into multiple taxonomic groups. I would like to determine the threshold we should use so that 95% of the genomes are represented by a single E/ASV.
* Create a line plot of the 95th percentile for each threshold at different number of operons per genome. This should be faceted by region within the 16S rRNA gene.
* Create a line plot showing the 95th percentile line for number of E/ASVs per genome for each region when we only consider those genomes with 7 copies of the *rrn* operon

* Notes:
  * Should sample 1 genome per species
  * Is 95th percentile too strict?

```{r}
easvs_by_threshold_region <- metadata_easv %>%

# Aggregate data by species/genome, region, threshold, 
	group_by(region, threshold, genome_id) %>%

# Count the number of E/ASVs per grouping and the # rrns per genome
	summarize(n_rrns = sum(count), n_easvs = n(), .groups="drop") %>%

# Determine the 95th percentile for each region, threshold, and # rrns
	group_by(region, threshold, n_rrns) %>%
	summarize(upper_bound = quantile(n_easvs, prob=0.95), .groups="drop")

# Plot the...
#   * 95th percentile as a function of region,  threshold, #rrns
easvs_by_threshold_region %>%
	filter(threshold %in% c(0, 0.01, 0.02, 0.03)) %>%
	mutate(threshold = as.character(threshold)) %>%
	ggplot(aes(x=n_rrns, y=upper_bound, color=threshold)) +
	geom_line() +
	facet_wrap(~region, nrow=4)

#   * 95th percentile for the number of E/ASVs for those genomes with 7 copies 
#     of the rrn operon
easvs_by_threshold_region %>%
	filter(n_rrns == 4) %>%
	ggplot(aes(x=threshold, y=upper_bound, color=region)) +
	geom_line()
```

### Conclusions...
* Analysis depends on our comfort with uncertainty (prob=90%? 95?) and the number of rrn copies per genome
* Regardless of the the number of operons or region, we need a significantly higher threshold than ESV or even "traditional" ASVs are defined at. Frankly, 3% doesn't look so bad for this type of analysis.

### 2020-11-25 exploratory/2020-11-30-lumping-and-splitting.Rmd
Compare rate of lumping and splitting by threshold and region, closes #36

---
title: "What is the effect of distance thrshold on the lumping and splitting of bacterial species and genomes?"
author: "Pat Schloss"
date: "11/30/2020"
output:
  github_document:
    html_preview: false
editor_options:
  chunk_output_type: console
---

```{r, message=FALSE}
library(tidyverse)
library(here)
library(knitr)

set.seed(19760620)

metadata <- read_tsv(here("data/references/genome_id_taxonomy.tsv"),
										 col_types = cols(.default = col_character())) %>%
	select(genome_id, species) %>%
	group_by(species) %>% # Get one genome per species
	slice_sample(n=1) %>%
	ungroup()

easv <- read_tsv(here("data/processed/rrnDB.easv.count_tibble"),
								col_types = cols(.default = col_character(),
																 count = col_integer()))

metadata_easv <- inner_join(metadata, easv, by=c("genome_id" = "genome")) %>%
	mutate(threshold = recode(threshold, "esv" = "0.000"),
				 threshold = as.numeric(threshold))
```

### Overivew
Besides the risk of splitting a genome into multiple taxonomic groups, there's also the possibility that the same E/ASV can appear in multiple species. In otherwords, by using too broad of a threshold to define an ASV, we run the risk of lumping different species together (e.g. *Bacillus cereus*, *B. anthracis*, and *B. thuringiensis*). I would like to determine...

* How often is the same E/ASV found in the multiple species?
* What fraction of species have multiple E/ASVs?
* Create a plot showing the fraction of species with multiple E/ASVs and the fraction of E/ASVs that appear in multiple species as a function of the threshold used to define E/ASVs.

Notes:
* Determine for each region of the 16S rRNA gene
* Select one genome per species

```{r}
# Measuring degree of splitting...
# Determine fraction of genomes with more than one E/ASV by region and threshold
splitting_data <- metadata_easv %>%

# - group data by region, threshold, and genome_id
	group_by(region, threshold, genome_id) %>%
# - determine whether each genome has more than 1 E/ASV
	summarize(n_easvs = n_distinct(easv),
						is_split = n_easvs > 1, .groups="drop") %>%

# - for each region and threshold, determine the fraction of genomes with more
#   than one E/ASV
	group_by(region, threshold) %>%
	summarize(f_split = sum(is_split) / n(), .groups="drop")

# Measure degree of lumping...
# Determine the fraction of E/ASVs that are found in more than one
# genome/species
lumping_data <- metadata_easv %>%

# - group data by region, threshold, and E/ASV
	group_by(region, threshold, easv) %>%

# - determine whether each E/ASV is observed in more than 1 genome
	summarize(n_genomes = n_distinct(genome_id),
						is_lumped = n_genomes > 1, .groups="drop")  %>%

# - for each region and threshold, determine the fraction of E/ASVs that appear
#   in more than one genome
	group_by(region, threshold) %>%
	summarize(f_lumped = sum(is_lumped)/n(), .groups="drop")

# Join lumping and splitting data
lumping_splitting_data <- inner_join(splitting_data, lumping_data, by=c("region", "threshold")) %>%

# * Tidy so that we have columns for region, threshold, lumping/splitting, and
#   the degree of lumping and splitting
	pivot_longer(cols=c(f_split, f_lumped), names_to="method", values_to="fraction")

# * Plot degree of lumping and splitting as a function of threshold for each
#   region. Try faceting...
#   - By region with lumping/splitting as separate lines
lumping_splitting_data %>%
	ggplot(aes(x=threshold, y=fraction, color=method)) +
	geom_line() +
	facet_wrap(~region) # i like this

#   - By lumping/splitting with region as separate lines
lumping_splitting_data %>%
	ggplot(aes(x=threshold, y=fraction, color=region)) + 
	geom_line() +
	facet_wrap(~method)

#   - By lumping/splitting and by region
lumping_splitting_data %>%
	ggplot(aes(x=threshold, y=fraction)) +
	geom_line() +
	facet_grid(region~method)

lumping_splitting_data %>%
	ggplot(aes(x=threshold, y=fraction)) +
	geom_line() +
	facet_grid(method~region) # i like this
```

### Conclusions...
* As we increase the threshold, splitting drops and lumping increases
* Again, we need a decent sized threshold (>0.01) to reduce the level of splitting
* Would prefer to limit splitting over lumping because species designations are too squishy - human made rather bacterial made

### 2020-11-27 code/get_roc_data.R
Generate ROC curve using expand_grid

@@ -1,0 +1,121 @@
+#!/usr/bin/env Rscript --vanilla
+
+# name: get_roc_data.R
+#
+# Trade off analysis of ASV threshold impact on lumping and splitting taxa
+#
+# input:
+# 	- data/processed/rrnDB.easv.count_tibble
+# output:
+#   - data/processed/rrnDB.roc.tsv
+#
+# Generate a confusion matrix, visualization, and associated statistics (e.g.
+# sensitivity, specificity, Matthew's correlation coefficient) for each
+# threshold and region represented in the pooled count_tibble file. Will
+# look at the trade-off between thresholds and risk of splitting a single
+# genome or species into multiple bins and the risk of merging species into
+# a single E/ASV. Let's think of it as a classification problem.
+#
+# Confusion matrix...
+# * same taxa, same easv = true positive
+# * different taxa, different easv = true negative
+# * different taxa, same easv = false positive (merging different taxa)
+# * same taxa, different easv = false negative (splitting single taxa)
+#
+# Could then look at measures of sensitivity and specificity, plot receiver
+# operator characteristic curve over thresholds, and calculate Matthew's
+# correlation coefficient for each threshold
+#
+# To calculate confusion matrix, we need to create a data frame that
+# provides an all vs. all comparison for each operon retaining the genome_id
+# and easv for each operon. We could then easily compare across columns
+# whether all possible pairs of easvs represent each cell of the confusion
+# matrix. Because some taxa are over represented, we would
+# need to grab representative genomes from each taxa, calculate metrics and
+# repeat a decent number of times.
+#
+# Considerations...
+# * May need to generate data for additional thresholds depending on shape of
+#   curves
+# * Might be ideal to look into how to parallelize steps to speed things up
+
+library(tidyverse)
+
+set.seed(19760620)
+
+desired_threshold <- 0.01
+desired_region <- "v4"
+
+# Plan of attack...
+# 1. Read in easvs and subset to get separate *region* and easv *threshold*
+#    combinations
+
+easv <- read_tsv("data/processed/rrnDB.easv.count_tibble",
+								 col_types = cols(.default = col_character(),
+								 								 count = col_integer())) %>%
+	mutate(threshold = recode(threshold, "esv" = "0.000"),
[…]

### 2020-11-27 code/get_roc_data.R
For loop implementation

@@ -81,13 +81,49 @@ metadata_easv <- inner_join(metadata, easv, by=c("genome_id" = "genome")) %>%
 n_operons <- nrow(metadata_easv)
 
 # 4. Create a data frame that compares each operon to every other operon
+# expand_grid
+# real	1m25.596s
+# user	1m11.007s
+# sys	0m14.223s
 
-all_v_all_comparison <- expand_grid(x=1:n_operons, y=1:n_operons) %>%
-	filter(x < y) %>%
-	inner_join(., metadata_easv, by=c("x"="index")) %>%
-	inner_join(., metadata_easv, by=c("y"="index")) %>%
-	select(-x, -y)
+# all_v_all_comparison <- expand_grid(x=1:n_operons, y=1:n_operons) %>%
+# 	filter(x < y) %>%
+# 	inner_join(., metadata_easv, by=c("x"="index")) %>%
+# 	inner_join(., metadata_easv, by=c("y"="index")) %>%
+# 	select(-x, -y)
 	
+library(pryr)
+object_size(all_v_all_comparison)
+
+# 	tp = (genome_id.x == genome_id.y) & (easv.x == easv.y),
+# 	tn = (genome_id.x != genome_id.y) & (easv.x != easv.y),
+# 	fn = (genome_id.x == genome_id.y) & (easv.x != easv.y),
+# 	fp = (genome_id.x != genome_id.y) & (easv.x == easv.y)) %>%
+
+# iteration 1: with confusion as vector
+# tp     tn     fp     fn 
+# 63 217705    348      0 
+# iteration 2: with confusion as a list
+#> confusion %>% unlist()
+# tp     tn     fp     fn 
+# 72 219658    348      0 
+
+confusion <- as.list(c(tp = 0, tn = 0, fp = 0, fn = 0))
+
+for(i in 1:(n_operons-1)){
+	for(j in 2:n_operons){
+		same_genome <- metadata_easv[i, "genome_id"] == metadata_easv[j, "genome_id"]
+		same_easv <- metadata_easv[i, "easv"] == metadata_easv[j, "easv"] 
+		
+		if(same_genome & same_easv) { confusion[["tp"]] <- confusion[["tp"]] + 1}
+		else if(!same_genome & !same_easv) { confusion[["tn"]] <- confusion[["tn"]] + 1}
+		else if(same_genome & !same_easv) { confusion[["fn"]] <- confusion[["fn"]] + 1}
[…]

### 2020-12-02 code/get_roc_data.R
Use Rcpp to speed up all v all comparison

@@ -40,6 +40,7 @@
 # * Might be ideal to look into how to parallelize steps to speed things up
 
 library(tidyverse)
+library(Rcpp)
 
 set.seed(19760620)
 
@@ -75,54 +76,41 @@ metadata <- read_tsv("data/references/genome_id_taxonomy.tsv",
 
 metadata_easv <- inner_join(metadata, easv, by=c("genome_id" = "genome")) %>%
 	uncount(count) %>%
-	mutate(index = row_number()) %>%
 	select(-species)
 
-n_operons <- nrow(metadata_easv)
 
 # 4. Create a data frame that compares each operon to every other operon
-# expand_grid
-# real	1m25.596s
-# user	1m11.007s
-# sys	0m14.223s
 
-# all_v_all_comparison <- expand_grid(x=1:n_operons, y=1:n_operons) %>%
-# 	filter(x < y) %>%
-# 	inner_join(., metadata_easv, by=c("x"="index")) %>%
-# 	inner_join(., metadata_easv, by=c("y"="index")) %>%
-# 	select(-x, -y)
-	
-library(pryr)
-object_size(all_v_all_comparison)
+cppFunction('DataFrame get_confusion_matrix(DataFrame genomes_easvs){
 
-# 	tp = (genome_id.x == genome_id.y) & (easv.x == easv.y),
-# 	tn = (genome_id.x != genome_id.y) & (easv.x != easv.y),
-# 	fn = (genome_id.x == genome_id.y) & (easv.x != easv.y),
-# 	fp = (genome_id.x != genome_id.y) & (easv.x == easv.y)) %>%
+						CharacterVector genomes = genomes_easvs[0];
+						CharacterVector easvs = genomes_easvs[1];
+						
+						int n_operons = genomes.length();
+						
+						int tp = 0;
+						int fp = 0;
+						int tn = 0;
+						int fn = 0;
+						bool same_genome, same_easv;
+						
+						for(int i=0;i<(n_operons-1);i++){
+							for(int j=1;j<n_operons;j++){
 
-# iteration 1: with confusion as vector
-# tp     tn     fp     fn 
-# 63 217705    348      0 
-# iteration 2: with confusion as a list
-#> confusion %>% unlist()
-# tp     tn     fp     fn 
-# 72 219658    348      0 
+								same_genome = genomes[i] == genomes[j];
+								same_easv = easvs[i] == easvs[j];
+							
[…]

### 2020-12-03 code/get_roc_data.R
Apply get_confusion_matrix to all regions and thresholds

@@ -42,45 +42,7 @@
 library(tidyverse)
 library(Rcpp)
 
-set.seed(19760620)
-
-desired_threshold <- 0.01
-desired_region <- "v4"
-
-# Plan of attack...
-# 1. Read in easvs and subset to get separate *region* and easv *threshold*
-#    combinations
-
-easv <- read_tsv("data/processed/rrnDB.easv.count_tibble",
-								 col_types = cols(.default = col_character(),
-								 								 count = col_integer())) %>%
-	mutate(threshold = recode(threshold, "esv" = "0.000"),
-				 threshold = as.numeric(threshold)) %>%
-	filter(threshold == desired_threshold & region == desired_region) %>%
-	select(easv, genome, count)
-
-
-# 2. Read in metadata and sub-sample to get 1 genome per species so that we have
-#    the genome_id, easv, and count columns for each taxon
-
-metadata <- read_tsv("data/references/genome_id_taxonomy.tsv",
-										 col_types = cols(.default = col_character())) %>%
-	select(genome_id, species) %>%
-	group_by(species) %>% # Get one genome per species
-	slice_sample(n=1) %>%
-	ungroup()
-
-
-# 3. Generate data frame that contains separate rows for each operon in the
-#		 dataset by replicating each genome_id/easv row count times
-
-metadata_easv <- inner_join(metadata, easv, by=c("genome_id" = "genome")) %>%
-	uncount(count) %>%
-	select(-species)
-
-
 # 4. Create a data frame that compares each operon to every other operon
-
 cppFunction('DataFrame get_confusion_matrix(DataFrame genomes_easvs){
 
 						CharacterVector genomes = genomes_easvs[0];
@@ -107,9 +69,47 @@ cppFunction('DataFrame get_confusion_matrix(DataFrame genomes_easvs){
 							}
 						}
 						
-						DataFrame confusion = DataFrame::create(Named("tp") = tp, Named("fp") = fp, Named("tn") = tn, Named("fn") = fn);
+						DataFrame confusion = DataFrame::create(
+							Named("true_pos") = tp,
+							Named("false_pos") = fp,
+							Named("true_neg") = tn,
+							Named("false_neg") = fn);
 						return(confusion);
 				}')
+
+
+set.seed(19760620)
+
+
+# Plan of attack...
+# 1. Read in easvs and subset to get separate *region* and easv *threshold*
+#    combinations
+
[…]

### 2020-12-12 code/get_roc_data.R
Deploy generation of confusion matrix data for 100 iterations

@@ -41,34 +41,35 @@
 
 library(tidyverse)
 library(Rcpp)
+library(furrr)
 
 # 4. Create a data frame that compares each operon to every other operon
 cppFunction('DataFrame get_confusion_matrix(DataFrame genomes_easvs){
 
 						CharacterVector genomes = genomes_easvs[0];
 						CharacterVector easvs = genomes_easvs[1];
-						
+
 						int n_operons = genomes.length();
-						
+
 						int tp = 0;
 						int fp = 0;
 						int tn = 0;
 						int fn = 0;
 						bool same_genome, same_easv;
-						
+
 						for(int i=0;i<(n_operons-1);i++){
 							for(int j=1;j<n_operons;j++){
 
 								same_genome = genomes[i] == genomes[j];
 								same_easv = easvs[i] == easvs[j];
-							
+
 									if(same_genome & same_easv) 			{ tp++; }
 									else if(!same_genome & !same_easv){ tn++; }
 									else if(same_genome & !same_easv) { fn++;	}
 									else if(!same_genome & same_easv) { fp++; }
 							}
 						}
-						
+
 						DataFrame confusion = DataFrame::create(
 							Named("true_pos") = tp,
 							Named("false_pos") = fp,
@@ -77,8 +78,6 @@ cppFunction('DataFrame get_confusion_matrix(DataFrame genomes_easvs){
 						return(confusion);
 				}')
 
-
-set.seed(19760620)
 
 
 # Plan of attack...
@@ -96,41 +95,48 @@ easv <- read_tsv("data/processed/rrnDB.easv.count_tibble",
 # 2. Read in metadata and sub-sample to get 1 genome per species so that we have
 #    the genome_id, easv, and count columns for each taxon
 
-metadata <- read_tsv("data/references/genome_id_taxonomy.tsv",
-										 col_types = cols(.default = col_character())) %>%
-	select(genome_id, species) %>%
-	group_by(species) %>% # Get one genome per species
-	slice_sample(n=1) %>%
-	ungroup()
+get_iteration <- function() {
+	metadata <- read_tsv("data/references/genome_id_taxonomy.tsv",
+											 col_types = cols(.default = col_character())) %>%
+		select(genome_id, species) %>%
+		group_by(species) %>% # Get one genome per species
+		slice_sample(n=1) %>%
+		ungroup()
 
 
-# 3. Generate data frame that contains separate rows for each operon in the
-#		 dataset by replicating each genome_id/easv row count times
[…]

### 2020-12-16 exploratory/2020-12-21-roc-curve.Rmd
Build ROC curves and update data, closes #37

---
title: "Balancing sensitivity and specificity by varying the ASV definition"
author: "Pat Schloss"
date: "12/21/2020"
output:
  github_document:
    html_preview: false
editor_options:
  chunk_output_type: console
---

```{r, message=FALSE}
library(tidyverse)
library(here)
library(knitr)

sensitivity_specificity <- read_tsv(here("data/processed/rrnDB.roc.tsv"),
													 col_types = cols(.default = col_double(),
													 								 region = col_character()
													 								 )
													 ) %>%
	mutate(sensitivity = true_pos / (true_pos + false_neg),
				 specificity = true_neg / (true_neg + false_pos)) %>%
	group_by(region, threshold) %>%
	summarize(sensitivity = median(sensitivity),
						specificity = median(specificity),
						.groups="drop") %>%
	select(region, threshold, sensitivity, specificity)
```

### Overivew

The sensitivity and specificity of assigning 16S rRNA gene sequences to ASVs relative to the genome/species have tradeoffs. We see these tradeoffs in whether ASVs from the same genome are split apart (i.e. sensitivity) or whether genomes from different species have the same ASV (i.e. specificity). As we've discussed, we'd like to prioritize minimizing the likelihood that we split genomes apart (i.e. high specificity), but minimizing lumping (i.e. low specificity). We can look at how both measures vary with the threshold used to define ASVs.

* sensitivity (true positive rate): true_pos / (true_pos + false_neg)
* specificity (true negative rate): true_neg / (true_neg + false_pos)

### Receiver operator characteristic curve (ROC curve)

```{r}
sensitivity_specificity %>%
	ggplot(aes(x=1-specificity, y=sensitivity, color=region)) + 
	geom_line() +
	theme_classic()
```

### What are the sensitivity and specificity at a 3% difference?

```{r}
three <- sensitivity_specificity %>%
	filter(threshold == 0.03)

three

sensitivity_specificity %>%
	ggplot(aes(x=1-specificity, y=sensitivity, color=region)) + 
	geom_abline(intercept=1, slope=-1, color="gray") +
	geom_line() +
	geom_point(data=three, aes(x=1-specificity, y=sensitivity, color=region)) +
	theme_classic()
```

### What threshold provides the equal balance of sensitivity and specificity?

```{r}
balance <- sensitivity_specificity %>%
	mutate(diff=abs(sensitivity - specificity)) %>%
	group_by(region) %>%
	summarize(min_diff = min(diff),
						threshold = threshold[which.min(diff)],
						sensitivity = sensitivity[which.min(diff)],
						specificity = specificity[which.min(diff)],
						.groups="drop"
						)

balance

sensitivity_specificity %>%
	ggplot(aes(x=1-specificity, y=sensitivity, color=region)) + 
	geom_abline(intercept=1, slope=-1, color="gray") +
	geom_line() +
	geom_point(data=balance, aes(x=1-specificity, y=sensitivity, color=region)) +
	theme_classic()

```

### What threshold provides a sensitivity and specificity closest to perfect classification?

```{r}
distance <- sensitivity_specificity %>%
	mutate(distance = sqrt((specificity - 1)^2 + (sensitivity - 1)^2)) %>%
	group_by(region) %>%
	summarize(min_distance = min(distance),
						threshold = threshold[which.min(distance)],
						sensitivity = sensitivity[which.min(distance)],
						specificity = specificity[which.min(distance)],
						.groups="drop"
						)

distance

sensitivity_specificity %>%
	ggplot(aes(x=1-specificity, y=sensitivity, color=region)) + 
	geom_abline(intercept=1, slope=-1, color="gray") +
	geom_line() +
	geom_point(data=distance, aes(x=1-specificity, y=sensitivity, color=region)) +
	theme_classic()
```

### Conclusions...
* We need more points going out to 0.10
* We need more points between percent thresholds (e.g. 0.025)
* Appears that "larger" thresholds provide the best balance and overall classification of ASVs relative to genomes/species

### 2020-12-16 code/get_distances.sh
Build ROC curves and update data, closes #37

@@ -8,4 +8,4 @@ TARGET=$1 # data/v4/rrnDB.unique.dist
 
 ALIGN=`echo $TARGET | sed -E "s/dist/align/"`
 
-code/mothur/mothur "#dist.seqs(fasta=$ALIGN, cutoff=0.05)"
+code/mothur/mothur "#dist.seqs(fasta=$ALIGN, cutoff=0.10)"

### 2020-12-21 submission/manuscript.Rmd
Initial outline

## Abstract (250 words)

## Importance (150 words)

## Introduction
* 16S rRNA gene sequencing is a very powerful technique for describing and comparing microbial communities
* How do we analyze them (classification, clustering)?
* What has changed in recent years? ASVs

* Efforts to link 16S rRNA gene sequences to taxonomic levels based on distance thresholds go back a long way
* ESVs/ASVs have been an attempt to adopt the thresholds suggested by genome sequencing to microbial community analysis using 16S rRNA gene sequences

* Most bacterial genomes have more than 1 copy of the rrn operon and those copies are not identical
* Using too fine a threshold to create taxonomic groups runs risk of splitting single genome into multiple bins
* For example, E. coli K-12 has 7 copies of the 16S rRNA gene with 5 variants
* Using too broad a threshold to define ASVs or OTUs risks lumping together bacterial species into the same grouping
* For example, B. cereus, thuringiensis, anthracis share the same 16S rRNA gene sequences

* Goal of this study

## Results
* ESVs/ASVs
  - copy number varies by taxonomy
  - more copies, more variants per genome
  - full length sequences have more variants than sub-regions
  - as more sequences are added to a species, the number of variants increases
* OTUs
  - increasing a threshold decreases the number of variants
  - this limits the splitting of a single genome into multiple bins
  - this increases the lumping of species into single bin

## Conclusions
* Briefly synthesize results
  - Unlikely that the unit of inference should be an ASV

* No biological argument to split a genome into multiple bins
* This analysis has allowed some splitting to balance with lumping
* To reduce splitting further, you would need larger thresholds

* There is general agreement in the field that if you want to classify something to a bacterial species, you need more than the 16S rRNA gene
* Furthermore, using only a few hundred bases of that gene are even more limited.
* We are asking too much of a short section of sequence

* Surprisingly, 3% performs pretty well for an operational definition that limits splitting of bacterial genomes and avoiding the lumping of bacterial species

## Materials and Methods
* rrnDB
* NCBI taxonomy
* R and R packages
* GitHub / YouTube

## Acknowledgements

## Figures

### 2020-12-22 submission/manuscript.tex
Create title page

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1.0in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\usepackage{helvet}
\renewcommand*\familydefault{\sfdefault}
\usepackage{setspace}
\doublespacing
\usepackage[left]{lineno}
\linenumbers
\ifluatex
  \usepackage{selnolig}  % disable illegal ligatures
\fi

\author{}
\date{\vspace{-2.5em}}

\begin{document}

\hypertarget{amplicon-sequence-variants-should-not-replace-operational-taxonomic-units-in-marker-gene-data-analysis}{%
\section{Amplicon sequence variants should not replace operational
taxonomic units in marker-gene data
analysis}\label{amplicon-sequence-variants-should-not-replace-operational-taxonomic-units-in-marker-gene-data-analysis}}

\vspace{20mm}

\textbf{Running title:} ASVs vs.~OTUs

\vspace{20mm}

Patrick D. Schloss\({^\dagger}\)

\vspace{40mm}

\({\dagger}\) To whom corresponsdence should be addressed:

\href{mailto:pschloss@umich.edu}{pschloss@umich.edu}

Department of Microbiology \& Immunology

University of Michigan

Ann Arbor, MI 48109

\vspace{20mm}

\textbf{Observation Format}

\newpage

\hypertarget{abstract-250-words}{%
\subsection{Abstract (250 words)}\label{abstract-250-words}}

\hypertarget{importance-150-words}{%
\subsection{Importance (150 words)}\label{importance-150-words}}

\newpage

\hypertarget{introduction}{%
\subsection{Introduction}\label{introduction}}

\begin{itemize}
\item
  16S rRNA gene sequencing is a very powerful technique for describing
  and comparing microbial communities
\item
  How do we analyze them (classification, clustering)?
\item
  What has changed in recent years? ASVs
\item
  Efforts to link 16S rRNA gene sequences to taxonomic levels based on
  distance thresholds go back a long way
\item
  ESVs/ASVs have been an attempt to adopt the thresholds suggested by
  genome sequencing to microbial community analysis using 16S rRNA gene
  sequences
\item
  Most bacterial genomes have more than 1 copy of the rrn operon and
  those copies are not identical
\item
  Using too fine a threshold to create taxonomic groups runs risk of
  splitting single genome into multiple bins
\item
  For example, E. coli K-12 has 7 copies of the 16S rRNA gene with 5
  variants
\item
  Using too broad a threshold to define ASVs or OTUs risks lumping
  together bacterial species into the same grouping
\item
  For example, B. cereus, thuringiensis, anthracis share the same 16S
  rRNA gene sequences
\item
  Goal of this study
\end{itemize}

\newpage

\hypertarget{results}{%
\subsection{Results}\label{results}}

\begin{itemize}
\tightlist
\item
  ESVs/ASVs

  \begin{itemize}
  \tightlist
  \item
    copy number varies by taxonomy
  \item
    more copies, more variants per genome
  \item
    full length sequences have more variants than sub-regions
  \item
    as more sequences are added to a species, the number of variants
    increases
  \end{itemize}
\item
  OTUs

  \begin{itemize}
  \tightlist
  \item
    increasing a threshold decreases the number of variants
  \item
    this limits the splitting of a single genome into multiple bins
  \item
    this increases the lumping of species into single bin
  \end{itemize}
\end{itemize}

\newpage

\hypertarget{conclusions}{%
\subsection{Conclusions}\label{conclusions}}

\begin{itemize}
\item
  Briefly synthesize results

  \begin{itemize}
  \tightlist
  \item
    Unlikely that the unit of inference should be an ASV
  \end{itemize}
\item
  No biological argument to split a genome into multiple bins
\item
  This analysis has allowed some splitting to balance with lumping
\item
  To reduce splitting further, you would need larger thresholds
\item
  There is general agreement in the field that if you want to classify
  something to a bacterial species, you need more than the 16S rRNA gene
\item
  Furthermore, using only a few hundred bases of that gene are even more
  limited.
\item
  We are asking too much of a short section of sequence
\item
  Surprisingly, 3\% performs pretty well for an operational definition
  that limits splitting of bacterial genomes and avoiding the lumping of
  bacterial species
\end{itemize}

\newpage

\hypertarget{materials-and-methods}{%
\subsection{Materials and Methods}\label{materials-and-methods}}

\begin{itemize}
\tightlist
\item
  rrnDB
\item
  NCBI taxonomy
\item
  R and R packages
\item
  GitHub / YouTube
\end{itemize}

\newpage

\hypertarget{acknowledgements}{%
\subsection{Acknowledgements}\label{acknowledgements}}

\newpage

\hypertarget{figures}{%
\subsection{Figures}\label{figures}}

\end{document}

### 2020-12-22 submission/manuscript.Rmd
Create title page

@@
+---
+output:
+  word_document: default
+  pdf_document:
+    keep_tex: true
+geometry: margin=1.0in
+font-size: 11pt
+header-includes:
+  - \usepackage{helvet}
+  - \renewcommand*\familydefault{\sfdefault}
+  - \usepackage{setspace}
+  - \doublespacing
+  - \usepackage[left]{lineno}
+  - \linenumbers
+---
+
+# Amplicon sequence variants should not replace operational taxonomic units in marker-gene data analysis
+
+\vspace{20mm}
+
+**Running title:** ASVs vs. OTUs
+
+\vspace{20mm}
+
+Patrick D. Schloss${^\dagger}$
+
+\vspace{40mm}
+
+${\dagger}$ To whom corresponsdence should be addressed:
+
+\href{mailto:pschloss@umich.edu}{pschloss@umich.edu}
+
+Department of Microbiology & Immunology
+
+University of Michigan
+
+Ann Arbor, MI 48109
+
+\vspace{20mm}
+
+**Observation Format**
+
+\newpage
+
 ## Abstract (250 words)
@@
 ## Importance (150 words)
+
+\newpage
 
@@
 
+\newpage
+
 ## Results
@@
   - this increases the lumping of species into single bin
+
+\newpage
 
@@
 
+\newpage
+
 ## Materials and Methods
@@
 
+\newpage
+
 ## Acknowledgements
 
+\newpage
+
 ## Figures

### 2021-01-05 submission/manuscript.Rmd
First rough draft

@@
 
-## Introduction
-* 16S rRNA gene sequencing is a very powerful technique for describing and comparing microbial communities
-* How do we analyze them (classification, clustering)?
-* What has changed in recent years? ASVs
+16S rRNA gene sequencing is a very powerful technique for describing and comparing microbial communities.
+Efforts to link 16S rRNA gene sequences to taxonomic levels based on distance thresholds go back to at least the 1990s.
+The distance-based thresholds that were developed and are now widely used (3%) were based on DNA-DNA hybridization approaches that are not as precise as genome sequencing.
+Instead, genome sequencing technologies have suggested that the widely used 3% distance threshold to operationally define bacterial taxa is too coarse.
+As an alternative to OTUs, amplicon sequencing variants (ASVs) have been proposed as a way to adopt the thresholds suggested by genome sequencing to microbial community analysis using 16S rRNA gene sequences.
+ASVs are a unit of microbial community inference that do not cluster sequences based on a distance-based threshold.
+However, most bacterial genomes have more than 1 copy of the rrn operon and those copies are not identical.
+Therefore, using too fine a threshold to identify OTUs creates the risk of splitting a single genome into multiple bins and using too broad of a threshold to define OTUs creates the risk of lumping together bacterial species into the same OTU.
+An example of both is seen in the comparison of Staphylococcus aureus (NCTC 8325) and S.
+epidermidis (ATCC 12228) where each genome has 5 copies of the 16S rRNA gene.
+The 10 copies of the 16S rRNA gene each have a different sequence and so if OTUs are defined based on ASVs, each genome would be split into 5 OTUs.
+Conversely, if the copies were clustered using a 3% distance threshold all 10 copies would cluster into the same OTU.
+The goal of this study was to quantify the risk of splitting a single genome into multiple bins and the risk of lumping together different bacterial species into the same bin.
 
-* Efforts to link 16S rRNA gene sequences to taxonomic levels based on distance thresholds go back a long way
-* ESVs/ASVs have been an attempt to adopt the thresholds suggested by genome sequencing to microbial community analysis using 16S rRNA gene sequences
+To investigate the variation in the number of copies of the 16S rRNA gene per genome as well as the intragenomic variation among copies of the 16S rRNA gene, I obtained reference 16S rRNA sequences from the rrn copy number database (rrnDB; CITATION).
+Among the XX,XXX genomes represented in the rrnDB, each genome had between X (e.g.
+XXXXXXX) and XX (e.g.
+XXXXXXX) copies of the rrn operon.
+As the number of copies of the operon in a genome increased, the number of variants of the 16S rRNA gene in each genome also increased (**FIGURE**).
+On average, there were XXX variants per copy of the full length 16S rRNA gene and an average of XXX, XXX, and XXX variants when considering the V4, V3-V4, and V4-V5 regions of the gene.
+Although a species tended to have a consistent number of 16S rRNA gene copies per genome, the number of total variants increased with the number of genomes that were sampled (**FIGURE**).
+For example, M.
+tuberculosis generally only had one copy of the gene per genome, but across the XXX genomes that have been sequenced there were XX versions of the gene.
+Similarly, a E.
+coli genome typically had 7 copies of the 16S rRNA gene with between X and X distinct full length sequences per genome.
+Across the XXX genomes that have been sequenced, there were XXX different variants of the gene.
+These observations highlight the risk of selecting a threshold for defining units of inference that is too narrow because it is possible to split a single genome into multiple units.
 
-* Most bacterial genomes have more than 1 copy of the rrn operon and those copies are not identical
-* Using too fine a threshold to create taxonomic groups runs risk of splitting single genome into multiple bins
-* For example, E. coli K-12 has 7 copies of the 16S rRNA gene with 5 variants
-* Using too broad a threshold to define ASVs or OTUs risks lumping together bacterial species into the same grouping
-* For example, B. cereus, thuringiensis, anthracis share the same 16S rRNA gene sequences
+A method to avoid splitting a single genome into multiple units of inference is to cluster 16S rRNA gene sequences together that are similar.
+However, this also increases the risk of lumping together genes from different species that are similar to each other.
+Therefore, I assessed the impact of the threshold used to define clusters of 16S rRNA genes on the propopensity to lump species together and split genome apart.
+I identified the threshold where 90% of bacterial species would be represented by a single OTU.
+For full length 16S rRNA gene sequences, I found that at a threshold of XX%, 90% of the species would be represented by a single OTU.
+Similarly, thresholds of XX, XX, and XX% were observed for the V4, V3-V4, and V4-V5 regions.
+However, at these thresholds, multiple species could be represented by the same OTU.
+At the highest level of resolution, XX% of the species shared a 16S rRNA gene sequence variant with another species.
+Given the risk of splitting a genome into multiple OTUs is more biologically problematic than lumping species together, larger thresholds are advisable.
 
-* Goal of this study
+To provide a more nuanced approach to selecting a threshold, it would be useful to to quantify the sensitivity and specificity of characterizing bacterial species using OTUs defined at different thresholds.
+I created confusion matrices for multiple regions of the 16S rRNA gene: true positives were those cases where two ASVs were joined in the same OTU and the same species; true negatives were those cases where two ASVs from different OTUs came from different species; false positives were those ASVs that joined the same OTU, but were from different species; and false negatives were those ASVs that joined different OTUs, but were from the same species.
+By calculating the sensitivity and specificity for each threshold and each region of the 16S rRNA gene, I was able to constuct a receiver operator characteristic curve (ROC).
+Because the ROC curve represents a range of possible thresholds and sensitivities and specificities, I used two metrics to select the best threshold for defining an OTU.
+First, I identified the thresholds where the sensitivity and specificity were most similar to each other.
+For this criterion, the best thresholds were XX (V1-V9), XX (V4), XX (V3V4), and XX (V4V5).
+Second, I identified the threshold that resulted in the point on the ROC curve that was closest to perfect classification.
+For this criterion, the best thresholds were XX (V1-V9), XX (V4), XX (V3V4), and XX (V4V5).
+Surprisingly, these analyses revealed that thresholds near 3% balance the risks of splitting genomes into separate OTUs and lumping species into the same OTU.
+
+The results of this analysis demonstrate that there is a significant risk of splitting single genomes into multiple bins if too fine of a threshold is applied to defining an OTU.
+An ongoing problem for amplicon-based studies is defining a meaningful taxonomic unit of inference.
+Since there is no consensus definition for a biological species concept, microbiologists must accept that how we have named bacterial species is biased and that taxonomic rules are not applied in a consistent manner.
+This makes it more challenging to attempt to fit a distance threshold to define an OTU definition that matches a set of species names.
+Furthermore, it is unlikely that the 16S rRNA gene evolves at the same rate across all bacterial lineages, which limits the biological interpretation of a common OTU definition.
+At best, a distance-based definition of a taxonomic unit is operational.
+There is general agreement in bacterial systematics that to classify something to a bacterial species, you need phenotypic and genome sequence data (CITATION).
+We are asking too much of a short section of a bacterial genome to be able to differentiate between species.
+It is difficult to defend a unit of inference that would split a single genome into multiple taxonomic units.
+It is not biologically plausible to entertain the possability that parts of a genome would have different ecologies.
+Although there are multiple reasons that proponents of ASVs encourage their use, the significant risk of splitting genomes is too high to warrant their use.
+
+**Materials and Methods.
+(i) Data availability.** The 16S rRNA gene sequences used in this study were  obtained from the *rrn*DB (https://rrndb.umms.med.umich.edu; version 5.6, released November 8, 2019).
+At the time of submission, this is the most current version of the database.
+The *rrn*DB obtained the curated 16S rRNA gene sequences from the KEGG database, which ultimately obtained them from NCBI's non-redundant RefSeq database.
+The *rrn*DB provides downloadable versions of the sequences with their taxonomy as determined using the naive Bayesian classifier trained on the RDP reference taxonomy.
+For some genomes this resulted in multiple classifications since a genome's 16S rRNA gene sequences were not identical.
+Instead, I mapped the RefSeq accession number for each genome in the database to obtain a single taxonomy for each genome.
+Because strain names were not consistently given to genomes across bacterial species, the strain level designations were ignored.
+
+**(ii) Definition of regions within 16S rRNA gene.** The full length 16S rRNA gene sequences were aligned to a SILVA reference alignment of the 16S rRNA gene (v138) using the mothur software package (v.
+1.XX).
+Regions of the 16S rRNA gene were selected because of their use in the microbial ecology literature.
+Full length sequences corresponded to *E.
+coli* positions XX through XXXX, V4 to positions XXX through XXX, V3-V4 to positions XXX through XXX, and V4-V5 to positions XXX through XXX.
+
+**(iii) Controlling for uneven sampling of genomes by species.** Because of the uneven distribution of genome sequences across species, for the analysis of splitting genomes and lumping species I randomly selected one genome for each species.
+The random selection was repeated 100 times.
+Analyses based on this randomization report the median of the 100 randomizations.
+The intraquartile range between randomizations was typically less than XXXX.
+Because it was so small, confidence intervals are not included in Figure 2.
+
+**(iv) Reproducible data analysis.** The code to perform the analysis in this manuscript and its hisotry are available as a git-based version control repository on GitHub (https://github.com/pschloss/Schloss_rrnAnalysis_XXXX_2020).
+The analysis can be regenerated using a GNU Make-based workflow that made use of built-in bash tools (v.
+3.2.57), mothur (v.
+1.XX), and R (v.
+4.X.X).
+Within R, I used the tidyverse (v.
+4.X.X), data.table (v.
+4.X.X), Rcpp (v.
+4.X.X), furrr (v.
+4.X.X), and rmarkdown (v.
+4.X.X) packages.
+The conception and development of this analysis is available as a playlist on the Riffomonas YouTube channel (https://www.youtube.com/playlist?list=PLmNrK_nkqBpKY3SZiivlIGvcLX-KHmfR8).
+
+**Acknowledgements.** I am grateful to Robert Hein and Thomas Schmidt who maintain the rrnDB for their help in understanding the curation of the database and for making the 16S rRNA gene sequences and related metadata publicly available.
+I am also grateful to community members who watched the serialized version of this analysis on YouTube and provided their suggestions and questions.
+
+This work was supported, in part, through grants from the NIH to PDS (P30DK034933, U01AI124255, and R01CA215574).
 
@@
 
-## Results
-* ESVs/ASVs
-  - copy number varies by taxonomy
-  - more copies, more variants per genome
-  - full length sequences have more variants than sub-regions
-  - as more sequences are added to a species, the number of variants increases
-* OTUs
-  - increasing a threshold decreases the number of variants
-  - this limits the splitting of a single genome into multiple bins
-  - this increases the lumping of species into single bin
-
-\newpage
-
-## Conclusions
-* Briefly synthesize results
-  - Unlikely that the unit of inference should be an ASV
-
-* No biological argument to split a genome into multiple bins
-* This analysis has allowed some splitting to balance with lumping
-* To reduce splitting further, you would need larger thresholds
-
-* There is general agreement in the field that if you want to classify something to a bacterial species, you need more than the 16S rRNA gene
-* Furthermore, using only a few hundred bases of that gene are even more limited.
-* We are asking too much of a short section of sequence
-
-* Surprisingly, 3% performs pretty well for an operational definition that limits splitting of bacterial genomes and avoiding the lumping of bacterial species
-
-\newpage
-
-## Materials and Methods
-* rrnDB
-* NCBI taxonomy
-* R and R packages
-* GitHub / YouTube
-
-\newpage
-
-## Acknowledgements
+## References
 

### 2021-01-11 submission/manuscript.Rmd
Insert code for first paragraph of results

@@
   - \linenumbers
+editor_options: 
+  chunk_output_type: console
 ---
+
+```{r, echo=FALSE}
+library(tidyverse)
+library(knitr)
+library(here)
+```
 
@@
 
+```{r, echo=FALSE}
+metadata <- read_tsv(here("data/references/genome_id_taxonomy.tsv"),
+										 col_types = cols(.default = col_character()))
+
+easv <- read_tsv(here("data/processed/rrnDB.easv.count_tibble"),
+								col_types = cols(.default = col_character(),
+																 count = col_integer()))
+
+metadata_easv <- inner_join(metadata, easv, by=c("genome_id" = "genome")) %>%
+	mutate(threshold = recode(threshold, "esv" = "0.000"),
+				 threshold = as.numeric(threshold))
+
+rrn_copy_number <- metadata_easv %>% 
+	filter(region == "v19", threshold == 0) %>%
+	group_by(species, genome_id) %>%
+	summarize(n_rrn = sum(count), .groups="drop") %>%
+	group_by(species) %>%
+	summarize(min = min(n_rrn), max = max(n_rrn), median = median(n_rrn), N=n(), .groups="drop")
+
+single_copy <- rrn_copy_number %>% filter(median == 1) %>% top_n(n=1, N)
+max_copy <- rrn_copy_number %>% top_n(n=1, median)
+
+single_copy_N <- rrn_copy_number %>% filter(species == single_copy$species) %>% pull(N)
+max_copy_N <- rrn_copy_number %>% filter(species == max_copy$species) %>% pull(N)
+ecoli_copy <- rrn_copy_number %>% filter(species == "Escherichia coli")
+
+rates <- metadata_easv %>%
+	filter(threshold == 0) %>%
+	group_by(region, genome_id) %>%
+	summarize(n_rrns = sum(count), n_asvs = n_distinct(easv), .groups="drop") %>%
+	group_by(region) %>%
+	summarize(rate = sum(n_asvs)/sum(n_rrns), .groups="drop")
+
+rate_v19 <- rates %>% filter(region == "v19") %>% pull(rate)
+rate_v4 <- rates %>% filter(region == "v4") %>% pull(rate)
+rate_v34 <- rates %>% filter(region == "v34") %>% pull(rate)
+rate_v45 <- rates %>% filter(region == "v45") %>% pull(rate)
+
+asvs_per_species <- metadata_easv %>%
+	filter(threshold == 0 & region == "v19") %>%
+	group_by(species) %>%
+	summarize(n_asvs = n_distinct(easv), .groups="drop")
+
+single_nasvs <- asvs_per_species %>% filter(species == single_copy$species) %>% pull(n_asvs)
+ecoli_nasvs <- asvs_per_species %>% filter(species == "Escherichia coli") %>% pull(n_asvs)
+```
+
 To investigate the variation in the number of copies of the 16S rRNA gene per genome as well as the intragenomic variation among copies of the 16S rRNA gene, I obtained reference 16S rRNA sequences from the rrn copy number database (rrnDB; CITATION).
-Among the XX,XXX genomes represented in the rrnDB, each genome had between X (e.g.
-XXXXXXX) and XX (e.g.
-XXXXXXX) copies of the rrn operon.
+Among the `r n_distinct(metadata$species)` species represented in the rrnDB there were `r nrow(metadata)` genomes.
+The median number of rrn operson per species ranged between `r single_copy$median` (e.g.
+`r single_copy$species`) and `r max_copy$median` (e.g.
+`r max_copy$species`) copies of the rrn operon.
 As the number of copies of the operon in a genome increased, the number of variants of the 16S rRNA gene in each genome also increased (**FIGURE**).
-On average, there were XXX variants per copy of the full length 16S rRNA gene and an average of XXX, XXX, and XXX variants when considering the V4, V3-V4, and V4-V5 regions of the gene.
+On average, there were `r rate_v19` variants per copy of the full length 16S rRNA gene and an average of `r rate_v4`, `r rate_v34`, and `r rate_v45` variants when considering the V4, V3-V4, and V4-V5 regions of the gene, respectively.
 Although a species tended to have a consistent number of 16S rRNA gene copies per genome, the number of total variants increased with the number of genomes that were sampled (**FIGURE**).
-For example, M.
-tuberculosis generally only had one copy of the gene per genome, but across the XXX genomes that have been sequenced there were XX versions of the gene.
+For example, `r single_copy$species` generally only had `r single_copy$median` copy of the gene per genome, but across the `r single_copy_N` genomes that have been sequenced there were `r single_nasvs` versions of the gene.
 Similarly, a E.
-coli genome typically had 7 copies of the 16S rRNA gene with between X and X distinct full length sequences per genome.
-Across the XXX genomes that have been sequenced, there were XXX different variants of the gene.
+coli genome typically had `r ecoli_copy$median` copies of the 16S rRNA gene with between `r ecoli_copy$min` and `r ecoli_copy$max` distinct full length sequences per genome.
+Across the `r ecoli_copy$N` E.
+coli genomes that have been sequenced, there were `r ecoli_nasvs` different variants of the gene.
 These observations highlight the risk of selecting a threshold for defining units of inference that is too narrow because it is possible to split a single genome into multiple units.

### 2021-01-11 submission/manuscript.tex
Insert code for first paragraph of results

@@
 
-\hypertarget{introduction}{%
-\subsection{Introduction}\label{introduction}}
+16S rRNA gene sequencing is a very powerful technique for describing and
+comparing microbial communities. Efforts to link 16S rRNA gene sequences
+to taxonomic levels based on distance thresholds go back to at least the
+1990s. The distance-based thresholds that were developed and are now
+widely used (3\%) were based on DNA-DNA hybridization approaches that
+are not as precise as genome sequencing. Instead, genome sequencing
+technologies have suggested that the widely used 3\% distance threshold
+to operationally define bacterial taxa is too coarse. As an alternative
+to OTUs, amplicon sequencing variants (ASVs) have been proposed as a way
+to adopt the thresholds suggested by genome sequencing to microbial
+community analysis using 16S rRNA gene sequences. ASVs are a unit of
+microbial community inference that do not cluster sequences based on a
+distance-based threshold. However, most bacterial genomes have more than
+1 copy of the rrn operon and those copies are not identical. Therefore,
+using too fine a threshold to identify OTUs creates the risk of
+splitting a single genome into multiple bins and using too broad of a
+threshold to define OTUs creates the risk of lumping together bacterial
+species into the same OTU. An example of both is seen in the comparison
+of Staphylococcus aureus (NCTC 8325) and S. epidermidis (ATCC 12228)
+where each genome has 5 copies of the 16S rRNA gene. The 10 copies of
+the 16S rRNA gene each have a different sequence and so if OTUs are
+defined based on ASVs, each genome would be split into 5 OTUs.
+Conversely, if the copies were clustered using a 3\% distance threshold
+all 10 copies would cluster into the same OTU. The goal of this study
+was to quantify the risk of splitting a single genome into multiple bins
+and the risk of lumping together different bacterial species into the
+same bin.
 
-\begin{itemize}
-\item
-  16S rRNA gene sequencing is a very powerful technique for describing
-  and comparing microbial communities
-\item
-  How do we analyze them (classification, clustering)?
-\item
-  What has changed in recent years? ASVs
-\item
-  Efforts to link 16S rRNA gene sequences to taxonomic levels based on
-  distance thresholds go back a long way
-\item
-  ESVs/ASVs have been an attempt to adopt the thresholds suggested by
-  genome sequencing to microbial community analysis using 16S rRNA gene
-  sequences
-\item
-  Most bacterial genomes have more than 1 copy of the rrn operon and
-  those copies are not identical
-\item
-  Using too fine a threshold to create taxonomic groups runs risk of
-  splitting single genome into multiple bins
-\item
-  For example, E. coli K-12 has 7 copies of the 16S rRNA gene with 5
-  variants
-\item
-  Using too broad a threshold to define ASVs or OTUs risks lumping
-  together bacterial species into the same grouping
-\item
-  For example, B. cereus, thuringiensis, anthracis share the same 16S
-  rRNA gene sequences
-\item
-  Goal of this study
-\end{itemize}
+To investigate the variation in the number of copies of the 16S rRNA
+gene per genome as well as the intragenomic variation among copies of
+the 16S rRNA gene, I obtained reference 16S rRNA sequences from the rrn
+copy number database (rrnDB; CITATION). Among the 4774 species
+represented in the rrnDB there were 15614 genomes. The median number of
+rrn operson per species ranged between 1 (e.g.~Mycobacterium
+tuberculosis) and 19 (e.g.~Metabacillus litoralis) copies of the rrn
+operon. As the number of copies of the operon in a genome increased, the
+number of variants of the 16S rRNA gene in each genome also increased
+(\textbf{FIGURE}). On average, there were 0.5970303 variants per copy of
+the full length 16S rRNA gene and an average of 0.2551013, 0.3265188,
+and 0.2744228 variants when considering the V4, V3-V4, and V4-V5 regions
+of the gene, respectively. Although a species tended to have a
+consistent number of 16S rRNA gene copies per genome, the number of
+total variants increased with the number of genomes that were sampled
+(\textbf{FIGURE}). For example, Mycobacterium tuberculosis generally
+only had 1 copy of the gene per genome, but across the 180 genomes that
+have been sequenced there were 11 versions of the gene. Similarly, a E.
+coli genome typically had 7 copies of the 16S rRNA gene with between 6
+and 10 distinct full length sequences per genome. Across the 958 E. coli
+genomes that have been sequenced, there were 1013 different variants of
+the gene. These observations highlight the risk of selecting a threshold
+for defining units of inference that is too narrow because it is
+possible to split a single genome into multiple units.
+
+A method to avoid splitting a single genome into multiple units of
+inference is to cluster 16S rRNA gene sequences together that are
+similar. However, this also increases the risk of lumping together genes
+from different species that are similar to each other. Therefore, I
+assessed the impact of the threshold used to define clusters of 16S rRNA
+genes on the propopensity to lump species together and split genome
+apart. I identified the threshold where 90\% of bacterial species would
+be represented by a single OTU. For full length 16S rRNA gene sequences,
+I found that at a threshold of XX\%, 90\% of the species would be
+represented by a single OTU. Similarly, thresholds of XX, XX, and XX\%
+were observed for the V4, V3-V4, and V4-V5 regions. However, at these
+thresholds, multiple species could be represented by the same OTU. At
+the highest level of resolution, XX\% of the species shared a 16S rRNA
+gene sequence variant with another species. Given the risk of splitting
+a genome into multiple OTUs is more biologically problematic than
+lumping species together, larger thresholds are advisable.
+
+To provide a more nuanced approach to selecting a threshold, it would be
+useful to to quantify the sensitivity and specificity of characterizing
+bacterial species using OTUs defined at different thresholds. I created
+confusion matrices for multiple regions of the 16S rRNA gene: true
+positives were those cases where two ASVs were joined in the same OTU
+and the same species; true negatives were those cases where two ASVs
+from different OTUs came from different species; false positives were
+those ASVs that joined the same OTU, but were from different species;
+and false negatives were those ASVs that joined different OTUs, but were
+from the same species. By calculating the sensitivity and specificity
+for each threshold and each region of the 16S rRNA gene, I was able to
+constuct a receiver operator characteristic curve (ROC). Because the ROC
+curve represents a range of possible thresholds and sensitivities and
+specificities, I used two metrics to select the best threshold for
+defining an OTU. First, I identified the thresholds where the
+sensitivity and specificity were most similar to each other. For this
+criterion, the best thresholds were XX (V1-V9), XX (V4), XX (V3V4), and
+XX (V4V5). Second, I identified the threshold that resulted in the point
+on the ROC curve that was closest to perfect classification. For this
+criterion, the best thresholds were XX (V1-V9), XX (V4), XX (V3V4), and
+XX (V4V5). Surprisingly, these analyses revealed that thresholds near
+3\% balance the risks of splitting genomes into separate OTUs and
+lumping species into the same OTU.
+
+The results of this analysis demonstrate that there is a significant
+risk of splitting single genomes into multiple bins if too fine of a
+threshold is applied to defining an OTU. An ongoing problem for
+amplicon-based studies is defining a meaningful taxonomic unit of
+inference. Since there is no consensus definition for a biological
+species concept, microbiologists must accept that how we have named
+bacterial species is biased and that taxonomic rules are not applied in
+a consistent manner. This makes it more challenging to attempt to fit a
+distance threshold to define an OTU definition that matches a set of
+species names. Furthermore, it is unlikely that the 16S rRNA gene
+evolves at the same rate across all bacterial lineages, which limits the
+biological interpretation of a common OTU definition. At best, a
+distance-based definition of a taxonomic unit is operational. There is
+general agreement in bacterial systematics that to classify something to
+a bacterial species, you need phenotypic and genome sequence data
+(CITATION). We are asking too much of a short section of a bacterial
+genome to be able to differentiate between species. It is difficult to
+defend a unit of inference that would split a single genome into
+multiple taxonomic units. It is not biologically plausible to entertain
+the possability that parts of a genome would have different ecologies.
+Although there are multiple reasons that proponents of ASVs encourage
+their use, the significant risk of splitting genomes is too high to
+warrant their use.
+
+\textbf{Materials and Methods. (i) Data availability.} The 16S rRNA gene
+sequences used in this study were obtained from the \emph{rrn}DB
+(\url{https://rrndb.umms.med.umich.edu}; version 5.6, released November
+8, 2019). At the time of submission, this is the most current version of
+the database. The \emph{rrn}DB obtained the curated 16S rRNA gene
+sequences from the KEGG database, which ultimately obtained them from
+NCBI's non-redundant RefSeq database. The \emph{rrn}DB provides
+downloadable versions of the sequences with their taxonomy as determined
+using the naive Bayesian classifier trained on the RDP reference
+taxonomy. For some genomes this resulted in multiple classifications
+since a genome's 16S rRNA gene sequences were not identical. Instead, I
+mapped the RefSeq accession number for each genome in the database to
+obtain a single taxonomy for each genome. Because strain names were not
+consistently given to genomes across bacterial species, the strain level
+designations were ignored.
+
+\textbf{(ii) Definition of regions within 16S rRNA gene.} The full
+length 16S rRNA gene sequences were aligned to a SILVA reference
+alignment of the 16S rRNA gene (v138) using the mothur software package
+(v. 1.XX). Regions of the 16S rRNA gene were selected because of their
+use in the microbial ecology literature. Full length sequences
+corresponded to \emph{E. coli} positions XX through XXXX, V4 to
+positions XXX through XXX, V3-V4 to positions XXX through XXX, and V4-V5
+to positions XXX through XXX.
+
+\textbf{(iii) Controlling for uneven sampling of genomes by species.}
+Because of the uneven distribution of genome sequences across species,
+for the analysis of splitting genomes and lumping species I randomly
+selected one genome for each species. The random selection was repeated
+100 times. Analyses based on this randomization report the median of the
+100 randomizations. The intraquartile range between randomizations was
+typically less than XXXX. Because it was so small, confidence intervals
+are not included in Figure 2.
+
+\textbf{(iv) Reproducible data analysis.} The code to perform the
+analysis in this manuscript and its hisotry are available as a git-based
+version control repository on GitHub
+(\url{https://github.com/pschloss/Schloss_rrnAnalysis_XXXX_2020}). The
+analysis can be regenerated using a GNU Make-based workflow that made
+use of built-in bash tools (v. 3.2.57), mothur (v. 1.XX), and R (v.
+4.X.X). Within R, I used the tidyverse (v. 4.X.X), data.table (v.
+4.X.X), Rcpp (v. 4.X.X), furrr (v. 4.X.X), and rmarkdown (v. 4.X.X)
+packages. The conception and development of this analysis is available
+as a playlist on the Riffomonas YouTube channel
+(\url{https://www.youtube.com/playlist?list=PLmNrK_nkqBpKY3SZiivlIGvcLX-KHmfR8}).
+
+\textbf{Acknowledgements.} I am grateful to Robert Hein and Thomas
+Schmidt who maintain the rrnDB for their help in understanding the
+curation of the database and for making the 16S rRNA gene sequences and
+related metadata publicly available. I am also grateful to community
+members who watched the serialized version of this analysis on YouTube
+and provided their suggestions and questions.
+
+This work was supported, in part, through grants from the NIH to PDS
+(P30DK034933, U01AI124255, and R01CA215574).
 
@@
 
-\hypertarget{results}{%
-\subsection{Results}\label{results}}
-
-\begin{itemize}
-\tightlist
-\item
-  ESVs/ASVs
-
-  \begin{itemize}
-  \tightlist
-  \item
-    copy number varies by taxonomy
-  \item
-    more copies, more variants per genome
-  \item
-    full length sequences have more variants than sub-regions
-  \item
-    as more sequences are added to a species, the number of variants
-    increases
-  \end{itemize}
-\item
-  OTUs
-
-  \begin{itemize}
-  \tightlist
-  \item
-    increasing a threshold decreases the number of variants
-  \item
-    this limits the splitting of a single genome into multiple bins
-  \item
-    this increases the lumping of species into single bin
-  \end{itemize}
-\end{itemize}
-
-\newpage
-
-\hypertarget{conclusions}{%
-\subsection{Conclusions}\label{conclusions}}
-
-\begin{itemize}
-\item
-  Briefly synthesize results
-
-  \begin{itemize}
-  \tightlist
-  \item
-    Unlikely that the unit of inference should be an ASV
-  \end{itemize}
-\item
-  No biological argument to split a genome into multiple bins
-\item
-  This analysis has allowed some splitting to balance with lumping
-\item
-  To reduce splitting further, you would need larger thresholds
-\item
-  There is general agreement in the field that if you want to classify
-  something to a bacterial species, you need more than the 16S rRNA gene
-\item
-  Furthermore, using only a few hundred bases of that gene are even more
-  limited.
-\item
-  We are asking too much of a short section of sequence
-\item
-  Surprisingly, 3\% performs pretty well for an operational definition
-  that limits splitting of bacterial genomes and avoiding the lumping of
-  bacterial species
-\end{itemize}
-
-\newpage
-
-\hypertarget{materials-and-methods}{%
-\subsection{Materials and Methods}\label{materials-and-methods}}
-
-\begin{itemize}
-\tightlist
-\item
-  rrnDB
-\item
-  NCBI taxonomy
-\item
-  R and R packages
-\item
-  GitHub / YouTube
-\end{itemize}
-
-\newpage
-
-\hypertarget{acknowledgements}{%
-\subsection{Acknowledgements}\label{acknowledgements}}
+\hypertarget{references}{%
+\subsection{References}\label{references}}
 

### 2021-01-12 submission/manuscript.Rmd
Format inline code

@@
 library(here)
+
+opts_chunk$set(echo = FALSE)
+
+inline_hook <- function(x){
+
+	if(is.numeric(x)){
+		if(abs(x - round(x) < .Machine$double.eps)){
+			# treat as integer
+			formatted <- format(x, digits=0, big.mark = ",")
+
+		} else {
+			# treat as floating point number
+			formatted <- format(x, digits=2, nsmall=2, big.mark = ",")
+		}
+	} else {
+		formatted <-x
+	}
+
+	paste0("**", formatted, "**")
+
+}
+
+knit_hooks$set(inline=inline_hook)
+
+format_pct <- function(x, my_digits=1){
+
+	format(x, digits=my_digits, nsmall=my_digits)
+
+}
 ```
@@
 Therefore, using too fine a threshold to identify OTUs creates the risk of splitting a single genome into multiple bins and using too broad of a threshold to define OTUs creates the risk of lumping together bacterial species into the same OTU.
-An example of both is seen in the comparison of Staphylococcus aureus (NCTC 8325) and S.
-epidermidis (ATCC 12228) where each genome has 5 copies of the 16S rRNA gene.
+An example of both is seen in the comparison of *Staphylococcus aureus* (NCTC 8325) and *S.
+epidermidis* (ATCC 12228) where each genome has 5 copies of the 16S rRNA gene.
 The 10 copies of the 16S rRNA gene each have a different sequence and so if OTUs are defined based on ASVs, each genome would be split into 5 OTUs.
@@
 
-```{r, echo=FALSE}
+```{r}
 metadata <- read_tsv(here("data/references/genome_id_taxonomy.tsv"),
@@
 The median number of rrn operson per species ranged between `r single_copy$median` (e.g.
-`r single_copy$species`) and `r max_copy$median` (e.g.
-`r max_copy$species`) copies of the rrn operon.
+*`r single_copy$species`*) and `r max_copy$median` (*`r max_copy$species`*) copies of the rrn operon.
 As the number of copies of the operon in a genome increased, the number of variants of the 16S rRNA gene in each genome also increased (**FIGURE**).
@@
 Although a species tended to have a consistent number of 16S rRNA gene copies per genome, the number of total variants increased with the number of genomes that were sampled (**FIGURE**).
-For example, `r single_copy$species` generally only had `r single_copy$median` copy of the gene per genome, but across the `r single_copy_N` genomes that have been sequenced there were `r single_nasvs` versions of the gene.
-Similarly, a E.
-coli genome typically had `r ecoli_copy$median` copies of the 16S rRNA gene with between `r ecoli_copy$min` and `r ecoli_copy$max` distinct full length sequences per genome.
-Across the `r ecoli_copy$N` E.
-coli genomes that have been sequenced, there were `r ecoli_nasvs` different variants of the gene.
+For example, *`r single_copy$species`* generally only had `r single_copy$median` copy of the gene per genome, but across the `r single_copy_N` genomes that have been sequenced there were `r single_nasvs` versions of the gene.
+Similarly, a *E.
+coli* genome typically had `r ecoli_copy$median` copies of the 16S rRNA gene with between `r ecoli_copy$min` and `r ecoli_copy$max` distinct full length sequences per genome.
+Across the `r ecoli_copy$N` *E.
+coli* genomes that have been sequenced, there were `r ecoli_nasvs` different variants of the gene.
 These observations highlight the risk of selecting a threshold for defining units of inference that is too narrow because it is possible to split a single genome into multiple units.
@@
 
+```{r}
+sensitivity_specificity <- read_tsv(here("data/processed/rrnDB.roc.tsv"),
+													 col_types = cols(.default = col_double(),
+													 								 region = col_character()
+													 								 )
+													 ) %>%
+	mutate(sensitivity = true_pos / (true_pos + false_neg),
+				 specificity = true_neg / (true_neg + false_pos)) %>%
+	group_by(region, threshold) %>%
+	summarize(sensitivity = median(sensitivity),
+						specificity = median(specificity),
+						.groups="drop") %>%
+	select(region, threshold, sensitivity, specificity)
+
+balance <- sensitivity_specificity %>%
+	mutate(diff=abs(sensitivity - specificity)) %>%
+	group_by(region) %>%
+	summarize(min_diff = min(diff),
+						threshold = threshold[which.min(diff)],
+						sensitivity = sensitivity[which.min(diff)],
+						specificity = specificity[which.min(diff)],
+						.groups="drop"
+						)
+
+balance_v19 <- balance %>% filter(region == "v19") %>% pull(threshold) * 100
+balance_v4 <- balance %>% filter(region == "v4") %>% pull(threshold) * 100
+balance_v34 <- balance %>% filter(region == "v34") %>% pull(threshold) * 100
+balance_v45 <- balance %>% filter(region == "v45") %>% pull(threshold) * 100
+
+distance <- sensitivity_specificity %>%
+	mutate(distance = sqrt((specificity - 1)^2 + (sensitivity - 1)^2)) %>%
+	group_by(region) %>%
+	summarize(min_distance = min(distance),
+						threshold = threshold[which.min(distance)],
+						sensitivity = sensitivity[which.min(distance)],
+						specificity = specificity[which.min(distance)],
+						.groups="drop"
+						)
+
+distance_v19 <- distance %>% filter(region == "v19") %>% pull(threshold) * 100
+distance_v4 <- distance %>% filter(region == "v4") %>% pull(threshold) * 100
+distance_v34 <- distance %>% filter(region == "v34") %>% pull(threshold) * 100
+distance_v45 <- distance %>% filter(region == "v45") %>% pull(threshold) * 100
+
+```
+
 To provide a more nuanced approach to selecting a threshold, it would be useful to to quantify the sensitivity and specificity of characterizing bacterial species using OTUs defined at different thresholds.
@@
 First, I identified the thresholds where the sensitivity and specificity were most similar to each other.
-For this criterion, the best thresholds were XX (V1-V9), XX (V4), XX (V3V4), and XX (V4V5).
-Second, I identified the threshold that resulted in the point on the ROC curve that was closest to perfect classification.
-For this criterion, the best thresholds were XX (V1-V9), XX (V4), XX (V3V4), and XX (V4V5).
-Surprisingly, these analyses revealed that thresholds near 3% balance the risks of splitting genomes into separate OTUs and lumping species into the same OTU.
+For this criterion, the best distance thresholds were `r format_pct(balance_v19)`% (V1-V9), `r format_pct(balance_v4)`% (V4), `r format_pct(balance_v34)`% (V3-V4), and `r format_pct(balance_v45)`% (V4-V5).
+Second, I identified the distance threshold that resulted in the point on the ROC curve that was closest to perfect classification.
+For this criterion, the best distance thresholds were `r format_pct(distance_v19)`% (V1-V9), `r format_pct(distance_v4)`% (V4), `r format_pct(distance_v34)`% (V3-V4), and `r format_pct(distance_v45)`% (V4-V5).
+Surprisingly, these analyses revealed that thresholds near 3% distance balance the risks of splitting genomes into separate OTUs and lumping species into the same OTU.
 

### 2021-01-12 submission/manuscript.tex
Format inline code

@@
 species into the same OTU. An example of both is seen in the comparison
-of Staphylococcus aureus (NCTC 8325) and S. epidermidis (ATCC 12228)
-where each genome has 5 copies of the 16S rRNA gene. The 10 copies of
-the 16S rRNA gene each have a different sequence and so if OTUs are
-defined based on ASVs, each genome would be split into 5 OTUs.
+of \emph{Staphylococcus aureus} (NCTC 8325) and \emph{S. epidermidis}
+(ATCC 12228) where each genome has 5 copies of the 16S rRNA gene. The 10
+copies of the 16S rRNA gene each have a different sequence and so if
+OTUs are defined based on ASVs, each genome would be split into 5 OTUs.
 Conversely, if the copies were clustered using a 3\% distance threshold
@@
 the 16S rRNA gene, I obtained reference 16S rRNA sequences from the rrn
-copy number database (rrnDB; CITATION). Among the 4774 species
-represented in the rrnDB there were 15614 genomes. The median number of
-rrn operson per species ranged between 1 (e.g.~Mycobacterium
-tuberculosis) and 19 (e.g.~Metabacillus litoralis) copies of the rrn
-operon. As the number of copies of the operon in a genome increased, the
-number of variants of the 16S rRNA gene in each genome also increased
-(\textbf{FIGURE}). On average, there were 0.5970303 variants per copy of
-the full length 16S rRNA gene and an average of 0.2551013, 0.3265188,
-and 0.2744228 variants when considering the V4, V3-V4, and V4-V5 regions
-of the gene, respectively. Although a species tended to have a
-consistent number of 16S rRNA gene copies per genome, the number of
-total variants increased with the number of genomes that were sampled
-(\textbf{FIGURE}). For example, Mycobacterium tuberculosis generally
-only had 1 copy of the gene per genome, but across the 180 genomes that
-have been sequenced there were 11 versions of the gene. Similarly, a E.
-coli genome typically had 7 copies of the 16S rRNA gene with between 6
-and 10 distinct full length sequences per genome. Across the 958 E. coli
-genomes that have been sequenced, there were 1013 different variants of
-the gene. These observations highlight the risk of selecting a threshold
-for defining units of inference that is too narrow because it is
-possible to split a single genome into multiple units.
+copy number database (rrnDB; CITATION). Among the \textbf{4,774} species
+represented in the rrnDB there were \textbf{15,614} genomes. The median
+number of rrn operson per species ranged between \textbf{1}
+(e.g.~\textbf{\emph{Mycobacterium tuberculosis}}) and \textbf{19}
+(\textbf{\emph{Metabacillus litoralis}}) copies of the rrn operon. As
+the number of copies of the operon in a genome increased, the number of
+variants of the 16S rRNA gene in each genome also increased
+(\textbf{FIGURE}). On average, there were \textbf{1} variants per copy
+of the full length 16S rRNA gene and an average of \textbf{0.26},
+\textbf{0.33}, and \textbf{0.27} variants when considering the V4,
+V3-V4, and V4-V5 regions of the gene, respectively. Although a species
+tended to have a consistent number of 16S rRNA gene copies per genome,
+the number of total variants increased with the number of genomes that
+were sampled (\textbf{FIGURE}). For example, \textbf{\emph{Mycobacterium
+tuberculosis}} generally only had \textbf{1} copy of the gene per
+genome, but across the \textbf{180} genomes that have been sequenced
+there were \textbf{11} versions of the gene. Similarly, a \emph{E. coli}
+genome typically had \textbf{7} copies of the 16S rRNA gene with between
+\textbf{6} and \textbf{10} distinct full length sequences per genome.
+Across the \textbf{958} \emph{E. coli} genomes that have been sequenced,
+there were \textbf{1,013} different variants of the gene. These
+observations highlight the risk of selecting a threshold for defining
+units of inference that is too narrow because it is possible to split a
+single genome into multiple units.
 
@@
 sensitivity and specificity were most similar to each other. For this
-criterion, the best thresholds were XX (V1-V9), XX (V4), XX (V3V4), and
-XX (V4V5). Second, I identified the threshold that resulted in the point
+criterion, the best distance thresholds were \textbf{6.0}\% (V1-V9),
+\textbf{4.5}\% (V4), \textbf{5.5}\% (V3-V4), and \textbf{4.0}\% (V4-V5).
+Second, I identified the distance threshold that resulted in the point
 on the ROC curve that was closest to perfect classification. For this
-criterion, the best thresholds were XX (V1-V9), XX (V4), XX (V3V4), and
-XX (V4V5). Surprisingly, these analyses revealed that thresholds near
-3\% balance the risks of splitting genomes into separate OTUs and
-lumping species into the same OTU.
+criterion, the best distance thresholds were \textbf{5.5}\% (V1-V9),
+\textbf{3.5}\% (V4), \textbf{4.5}\% (V3-V4), and \textbf{3.5}\% (V4-V5).
+Surprisingly, these analyses revealed that thresholds near 3\% distance
+balance the risks of splitting genomes into separate OTUs and lumping
+species into the same OTU.
 

### 2021-01-12 submission/manuscript.Rmd
Format inline code

@@
 Therefore, using too fine a threshold to identify OTUs creates the risk of splitting a single genome into multiple bins and using too broad of a threshold to define OTUs creates the risk of lumping together bacterial species into the same OTU.
-An example of both is seen in the comparison of *Staphylococcus aureus* (NCTC 8325) and *S.
-epidermidis* (ATCC 12228) where each genome has 5 copies of the 16S rRNA gene.
+An example of both is seen in the comparison of Staphylococcus aureus (NCTC 8325) and S.
+epidermidis (ATCC 12228) where each genome has 5 copies of the 16S rRNA gene.
 The 10 copies of the 16S rRNA gene each have a different sequence and so if OTUs are defined based on ASVs, each genome would be split into 5 OTUs.
@@
 The median number of rrn operson per species ranged between `r single_copy$median` (e.g.
-*`r single_copy$species`*) and `r max_copy$median` (*`r max_copy$species`*) copies of the rrn operon.
+`r single_copy$species`) and `r max_copy$median` (e.g.
+`r max_copy$species`) copies of the rrn operon.
 As the number of copies of the operon in a genome increased, the number of variants of the 16S rRNA gene in each genome also increased (**FIGURE**).
@@
 Although a species tended to have a consistent number of 16S rRNA gene copies per genome, the number of total variants increased with the number of genomes that were sampled (**FIGURE**).
-For example, *`r single_copy$species`* generally only had `r single_copy$median` copy of the gene per genome, but across the `r single_copy_N` genomes that have been sequenced there were `r single_nasvs` versions of the gene.
-Similarly, a *E.
-coli* genome typically had `r ecoli_copy$median` copies of the 16S rRNA gene with between `r ecoli_copy$min` and `r ecoli_copy$max` distinct full length sequences per genome.
-Across the `r ecoli_copy$N` *E.
-coli* genomes that have been sequenced, there were `r ecoli_nasvs` different variants of the gene.
+For example, `r single_copy$species` generally only had `r single_copy$median` copy of the gene per genome, but across the `r single_copy_N` genomes that have been sequenced there were `r single_nasvs` versions of the gene.
+Similarly, a E.
+coli genome typically had `r ecoli_copy$median` copies of the 16S rRNA gene with between `r ecoli_copy$min` and `r ecoli_copy$max` distinct full length sequences per genome.
+Across the `r ecoli_copy$N` E.
+coli genomes that have been sequenced, there were `r ecoli_nasvs` different variants of the gene.
 These observations highlight the risk of selecting a threshold for defining units of inference that is too narrow because it is possible to split a single genome into multiple units.

### 2021-01-12 submission/manuscript.tex
Format inline code

@@
 species into the same OTU. An example of both is seen in the comparison
-of \emph{Staphylococcus aureus} (NCTC 8325) and \emph{S. epidermidis}
-(ATCC 12228) where each genome has 5 copies of the 16S rRNA gene. The 10
-copies of the 16S rRNA gene each have a different sequence and so if
-OTUs are defined based on ASVs, each genome would be split into 5 OTUs.
+of Staphylococcus aureus (NCTC 8325) and S. epidermidis (ATCC 12228)
+where each genome has 5 copies of the 16S rRNA gene. The 10 copies of
+the 16S rRNA gene each have a different sequence and so if OTUs are
+defined based on ASVs, each genome would be split into 5 OTUs.
 Conversely, if the copies were clustered using a 3\% distance threshold
@@
 number of rrn operson per species ranged between \textbf{1}
-(e.g.~\textbf{\emph{Mycobacterium tuberculosis}}) and \textbf{19}
-(\textbf{\emph{Metabacillus litoralis}}) copies of the rrn operon. As
-the number of copies of the operon in a genome increased, the number of
+(e.g.~\textbf{Mycobacterium tuberculosis}) and \textbf{19}
+(e.g.~\textbf{Metabacillus litoralis}) copies of the rrn operon. As the
+number of copies of the operon in a genome increased, the number of
 variants of the 16S rRNA gene in each genome also increased
@@
 the number of total variants increased with the number of genomes that
-were sampled (\textbf{FIGURE}). For example, \textbf{\emph{Mycobacterium
-tuberculosis}} generally only had \textbf{1} copy of the gene per
-genome, but across the \textbf{180} genomes that have been sequenced
-there were \textbf{11} versions of the gene. Similarly, a \emph{E. coli}
-genome typically had \textbf{7} copies of the 16S rRNA gene with between
-\textbf{6} and \textbf{10} distinct full length sequences per genome.
-Across the \textbf{958} \emph{E. coli} genomes that have been sequenced,
-there were \textbf{1,013} different variants of the gene. These
-observations highlight the risk of selecting a threshold for defining
-units of inference that is too narrow because it is possible to split a
-single genome into multiple units.
+were sampled (\textbf{FIGURE}). For example, \textbf{Mycobacterium
+tuberculosis} generally only had \textbf{1} copy of the gene per genome,
+but across the \textbf{180} genomes that have been sequenced there were
+\textbf{11} versions of the gene. Similarly, a E. coli genome typically
+had \textbf{7} copies of the 16S rRNA gene with between \textbf{6} and
+\textbf{10} distinct full length sequences per genome. Across the
+\textbf{958} E. coli genomes that have been sequenced, there were
+\textbf{1,013} different variants of the gene. These observations
+highlight the risk of selecting a threshold for defining units of
+inference that is too narrow because it is possible to split a single
+genome into multiple units.
 

### 2021-01-15 submission/manuscript.Rmd
Replicate threshold to get one OTU per genome/species 100 times

--- 2021-01-12 submission/manuscript.Rmd
@@
 
+```{r}
+thresholds <- read_tsv(here("data/processed/thresholds_for_single_otu.tsv"),
+											 col_types=cols(region = col_character(),
+											 							 .default = col_double())) %>%
+	filter(n_rrns == 7)
+
+threshold_v19 <- thresholds %>% filter(region == "v19") %>% pull(threshold) * 100
+threshold_v4 <- thresholds %>% filter(region == "v4") %>% pull(threshold) * 100
+threshold_v34 <- thresholds %>% filter(region == "v34") %>% pull(threshold) * 100
+threshold_v45 <- thresholds %>% filter(region == "v45") %>% pull(threshold) * 100
+```
+
 A method to avoid splitting a single genome into multiple units of inference is to cluster 16S rRNA gene sequences together that are similar.
 However, this also increases the risk of lumping together genes from different species that are similar to each other.
-Therefore, I assessed the impact of the threshold used to define clusters of 16S rRNA genes on the propopensity to lump species together and split genome apart.
-I identified the threshold where 90% of bacterial species would be represented by a single OTU.
-For full length 16S rRNA gene sequences, I found that at a threshold of XX%, 90% of the species would be represented by a single OTU.
-Similarly, thresholds of XX, XX, and XX% were observed for the V4, V3-V4, and V4-V5 regions.
+Therefore, I assessed the impact of the threshold used to define clusters of 16S rRNA genes on the propensity to split a genome apart or to lump species together.
+For full length 16S rRNA gene sequences, I found that at a threshold of `r format_pct(threshold_v19)`%, 95% of the species with 7 copies of the rrn operon would be represented by a single OTU.
+Similarly, thresholds of `r format_pct(threshold_v4)`, `r format_pct(threshold_v34)`, and `r format_pct(threshold_v45)`% were observed for the V4, V3-V4, and V4-V5 regions, respectively.
 However, at these thresholds, multiple species could be represented by the same OTU.

### 2021-01-15 submission/manuscript.tex
Replicate threshold to get one OTU per genome/species 100 times

--- 2021-01-12 submission/manuscript.tex
@@
 assessed the impact of the threshold used to define clusters of 16S rRNA
-genes on the propopensity to lump species together and split genome
-apart. I identified the threshold where 90\% of bacterial species would
-be represented by a single OTU. For full length 16S rRNA gene sequences,
-I found that at a threshold of XX\%, 90\% of the species would be
-represented by a single OTU. Similarly, thresholds of XX, XX, and XX\%
-were observed for the V4, V3-V4, and V4-V5 regions. However, at these
+genes on the propensity to split a genome apart or to lump species
+together. For full length 16S rRNA gene sequences, I found that at a
+threshold of \textbf{5.5}\%, 95\% of the species with 7 copies of the
+rrn operon would be represented by a single OTU. Similarly, thresholds
+of \textbf{2.5}, \textbf{4.0}, and \textbf{3.5}\% were observed for the
+V4, V3-V4, and V4-V5 regions, respectively. However, at these
 thresholds, multiple species could be represented by the same OTU. At

### 2021-01-15 code/get_threshold_for_single_otu.R
Replicate threshold to get one OTU per genome/species 100 times

@@ -1,0 +1,48 @@
+#!/usr/bin/env Rscript --vanilla
+
+library(tidyverse)
+library(furrr)
+
+
+metadata <- read_tsv("data/references/genome_id_taxonomy.tsv",
+										 col_types = cols(.default = col_character())) %>%
+	select(genome_id, species)
+
+asv <- read_tsv("data/processed/rrnDB.easv.count_tibble",
+								 col_types = cols(.default = col_character(),
+								 								 count = col_integer())) %>%
+				mutate(threshold = recode(threshold, "esv" = "0.000"),
+							 threshold = as.numeric(threshold))
+
+
+get_threshold_for_single_otu <- function(prob = 0.95){
+
+		metadata %>%
+		group_by(species) %>% # Get one genome per species
+		slice_sample(n=1) %>%
+		ungroup() %>%
+		inner_join(., asv, by=c("genome_id" = "genome")) %>%
+		group_by(region, threshold, genome_id) %>%
+		summarize(n_rrns = sum(count), n_easvs = n(), .groups="drop") %>%
+		group_by(region, threshold, n_rrns) %>%
+		summarize(upper_bound = quantile(n_easvs, prob=prob),
+							n_genomes = n(),
+							.groups="drop") %>%
+		filter(upper_bound == 1) %>%
+		group_by(region, n_rrns) %>%
+		summarize(threshold = min(threshold), n_genomes = median(n_genomes), .groups="drop")
+
+}
+
+plan(multicore)
+
+future_map_dfr(1:100,
+				~get_threshold_for_single_otu(),
+				.options = furrr_options(seed = 19760620)
+				) %>%
+	group_by(region, n_rrns) %>%
+	summarize(threshold = median(threshold),
+						iqr = IQR(threshold),
+						n_genomes = median(n_genomes),
+						.groups="drop") %>%
+	write_tsv("data/processed/thresholds_for_single_otu.tsv")		

### 2021-01-15 submission/manuscript.tex
Replicate lump/split calculations 100 times

@@
 thresholds, multiple species could be represented by the same OTU. At
-the highest level of resolution, XX\% of the species shared a 16S rRNA
-gene sequence variant with another species. Given the risk of splitting
+the highest level of resolution, \textbf{3.6}\% of the species shared a
+16S rRNA gene sequence variant with another species when considering
+full length sequences and \textbf{14.9}, \textbf{10.2}, and
+\textbf{12.0}\% when considering the V4, V3-V4, and V4-V5 regions,
+respectively. At the commonly used 3\% threshold, \textbf{25.2}\% of the
+species shared an OTU when considering full length sequences and
+\textbf{33.0}, \textbf{29.4}, and \textbf{32.2}\% when considering the
+V4, V3-V4, and V4-V5 regions, respectively. Given the risk of splitting
 a genome into multiple OTUs is more biologically problematic than

### 2021-01-15 code/get_lump_split_rates.R

@@ -1,0 +1,68 @@
+#!/usr/bin/env Rscript --vanilla
+
+
+library(tidyverse)
+library(furrr)
+
+metadata <- read_tsv("data/references/genome_id_taxonomy.tsv",
+										 col_types = cols(.default = col_character())) %>%
+	select(genome_id, species)
+
+asv <- read_tsv("data/processed/rrnDB.easv.count_tibble",
+								 col_types = cols(.default = col_character(),
+								 								 count = col_integer())) %>%
+	mutate(threshold = recode(threshold, "esv" = "0.000"),
+				 threshold = as.numeric(threshold))
+
+
+get_rate_of_splitting <- function(){
+	
+	metadata %>% 
+		group_by(species) %>% # Get one genome per species
+		slice_sample(n=1) %>%
+		ungroup() %>%
+		inner_join(., asv, by=c("genome_id" = "genome")) %>%
+		group_by(region, threshold, genome_id) %>%
+		summarize(n_asvs = n_distinct(easv),
+							is_split = n_asvs > 1, .groups="drop") %>%
+		group_by(region, threshold) %>%
+		summarize(f_split = sum(is_split) / n(), .groups="drop")
+}
+
+get_rate_of_lumping <- function(){
+	
+	metadata %>% 
+		group_by(species) %>% # Get one genome per species
+		slice_sample(n=1) %>%
+		ungroup() %>%
+		inner_join(., asv, by=c("genome_id" = "genome")) %>%
+		group_by(region, threshold, easv) %>%
+		summarize(n_genomes = n_distinct(genome_id),
+							is_lumped = n_genomes > 1, .groups="drop")  %>%
+		group_by(region, threshold) %>%
+		summarize(f_lumped = sum(is_lumped)/n(), .groups="drop")
+}
+
+plan(multicore)
+
+rate_of_splitting <- future_map_dfr(1:100,
+																		~get_rate_of_splitting(),
+																		.options = furrr_options(seed = 19760620)
+																		) %>%
+	group_by(region, threshold) %>%
+	summarize(split_rate = median(f_split), split_iqr = IQR(f_split), .groups="drop")
+
+
+rate_of_lumping <- future_map_dfr(1:100,
+																	~get_rate_of_lumping(),
+																	.options = furrr_options(seed = 19760620)
+																	) %>%
+	group_by(region, threshold) %>%
[…]

### 2021-01-15 submission/manuscript.Rmd
Replicate lump/split calculations 100 times

@@
 threshold_v45 <- thresholds %>% filter(region == "v45") %>% pull(threshold) * 100
+
+lump_split <- read_tsv(here("data/processed/lumped_split_rate.tsv"),
+											 col_types = cols(region = col_character(),
+											 								 .default = col_double()))
+
+lump_v19 <- lump_split %>% filter(threshold ==0 &region == "v19") %>% pull(lump_rate) * 100
+lump_v4 <- lump_split %>% filter(threshold ==0 &region == "v4") %>% pull(lump_rate) * 100
+lump_v34 <- lump_split %>% filter(threshold ==0 &region == "v34") %>% pull(lump_rate) * 100
+lump_v45 <- lump_split %>% filter(threshold ==0 &region == "v45") %>% pull(lump_rate) * 100
+
+lump3_v19 <- lump_split %>% filter(threshold ==0.03 & region == "v19") %>% pull(lump_rate) * 100
+lump3_v4 <- lump_split %>% filter(threshold ==0.03 & region == "v4") %>% pull(lump_rate) * 100
+lump3_v34 <- lump_split %>% filter(threshold ==0.03 & region == "v34") %>% pull(lump_rate) * 100
+lump3_v45 <- lump_split %>% filter(threshold ==0.03 & region == "v45") %>% pull(lump_rate) * 100
+
 ```
@@
 However, at these thresholds, multiple species could be represented by the same OTU.
-At the highest level of resolution, XX% of the species shared a 16S rRNA gene sequence variant with another species.
+At the highest level of resolution, `r format_pct(lump_v19)`% of the species shared a 16S rRNA gene sequence variant with another species when considering full length sequences and `r format_pct(lump_v4)`, `r format_pct(lump_v34)`, and `r format_pct(lump_v45)`% when considering the V4, V3-V4, and V4-V5 regions, respectively.
+At the commonly used 3% threshold, `r format_pct(lump3_v19)`% of the species shared an OTU when considering full length sequences and `r format_pct(lump3_v4)`, `r format_pct(lump3_v34)`, and `r format_pct(lump3_v45)`% when considering the V4, V3-V4, and V4-V5 regions, respectively.
 Given the risk of splitting a genome into multiple OTUs is more biologically problematic than lumping species together, larger thresholds are advisable.

### 2021-01-20 submission/manuscript.Rmd
Plot number of ASVs per rrn copy for each species by number of genomes

@@
 On average, there were `r rate_v19` variants per copy of the full length 16S rRNA gene and an average of `r rate_v4`, `r rate_v34`, and `r rate_v45` variants when considering the V4, V3-V4, and V4-V5 regions of the gene, respectively.
-Although a species tended to have a consistent number of 16S rRNA gene copies per genome, the number of total variants increased with the number of genomes that were sampled (**FIGURE**).
+Although a species tended to have a consistent number of 16S rRNA gene copies per genome, the number of total variants increased with the number of genomes that were sampled (**Figure 1**).
 For example, *`r single_copy$species`* generally only had `r single_copy$median` copy of the gene per genome, but across the `r single_copy_N` genomes that have been sequenced there were `r single_nasvs` versions of the gene.
@@
 
-## Figures
+\includegraphics{../figures/esv_rate.pdf}
+
+**Figure 1. ESV rate increases as the number of genomes sampled per species increases**

### 2021-01-20 submission/manuscript.tex
Plot number of ASVs per rrn copy for each species by number of genomes

@@
 the number of total variants increased with the number of genomes that
-were sampled (\textbf{FIGURE}). For example, \textbf{\emph{Mycobacterium
-tuberculosis}} generally only had \textbf{1} copy of the gene per
-genome, but across the \textbf{180} genomes that have been sequenced
-there were \textbf{11} versions of the gene. Similarly, a \emph{E. coli}
-genome typically had \textbf{7} copies of the 16S rRNA gene with between
-\textbf{6} and \textbf{10} distinct full length sequences per genome.
-Across the \textbf{958} \emph{E. coli} genomes that have been sequenced,
-there were \textbf{1,013} different variants of the gene. These
-observations highlight the risk of selecting a threshold for defining
-units of inference that is too narrow because it is possible to split a
-single genome into multiple units.
+were sampled (\textbf{Figure 1}). For example,
+\textbf{\emph{Mycobacterium tuberculosis}} generally only had \textbf{1}
+copy of the gene per genome, but across the \textbf{180} genomes that
+have been sequenced there were \textbf{11} versions of the gene.
+Similarly, a \emph{E. coli} genome typically had \textbf{7} copies of
+the 16S rRNA gene with between \textbf{6} and \textbf{10} distinct full
+length sequences per genome. Across the \textbf{958} \emph{E. coli}
+genomes that have been sequenced, there were \textbf{1,013} different
+variants of the gene. These observations highlight the risk of selecting
+a threshold for defining units of inference that is too narrow because
+it is possible to split a single genome into multiple units.
 
@@
 
-\hypertarget{figures}{%
-\subsection{Figures}\label{figures}}
+\includegraphics{../figures/esv_rate.pdf}
+
+\textbf{Figure 1. ESV rate increases as the number of genomes sampled
+per species increases}
 

### 2021-01-20 code/plot_esv_rate.R

@@ -1,0 +1,60 @@
+#!/usr/bin/env Rscript --vanilla
+
+library(tidyverse)
+library(knitr)
+library(here)
+
+metadata <- read_tsv(here("data/references/genome_id_taxonomy.tsv"),
+										 col_types = cols(.default = col_character())) %>%
+	mutate(strain = if_else(scientific_name == species,
+													NA_character_,
+													scientific_name)) %>%
+	select(-scientific_name)
+
+esv <- read_tsv(here("data/processed/rrnDB.easv.count_tibble"),
+								col_types = cols(.default = col_character(),
+																 count = col_integer())) %>%
+	filter(threshold == "esv") %>%
+	select(-threshold)
+
+metadata_esv <- inner_join(metadata, esv, by=c("genome_id" = "genome"))
+
+region_labels <- c("V1-V9", "V4", "V3-V4", "V4-V5")
+names(region_labels) <- c("v19", "v4", "v34", "v45")
+
+species_esvs <- metadata_esv %>%
+	select(genome_id, species, region, easv, count) %>%
+	group_by(region, species) %>%
+	summarize(n_genomes = n_distinct(genome_id),
+						n_esvs = n_distinct(easv),
+						n_rrns = sum(count) / n_genomes,
+						esv_rate = n_esvs/n_rrns,
+						.groups="drop") %>%
+	mutate(region = region_labels[region])
+
+facet_label <- species_esvs %>%
+	distinct(region) %>%
+	mutate(x = 1, y=100)
+	
+
+species_esvs %>%
+	ggplot(aes(x=n_genomes, y=esv_rate)) +
+	geom_point(alpha=0.2) +
+	geom_smooth(se=FALSE) +
+	geom_text(data=facet_label, aes(label=region, x=x, y=y), hjust=0, fontface=2) +
+	facet_wrap(facet="region",
+						 nrow=4,
+						 strip.position = "top",
+						 scales="fixed",
+						 labeller=labeller(region = region_labels)) +
+	scale_x_log10()+
+	scale_y_log10(breaks=c(0.1, 1, 10, 100), labels = c("0.1", "1", "10", "100")) +
+	labs(x="Number of genomes",
+			 y="Number of ASVs per rrn copy") +
+	theme_classic() +
+	theme(
+		strip.text = element_blank()
+	)
+
+ggsave("figures/esv_rate.tiff", width=3.5, height=8, compression="lzw")
+ggsave("figures/esv_rate.pdf", width=3.5, height=8)

### 2021-01-20 code/plot_lump_split.R
Generate lumping and splitting rate figure

@@ -1,0 +1,35 @@
+#!/usr/bin/env Rscript --vanilla
+
+library(tidyverse)
+
+read_tsv("data/processed/lumped_split_rate.tsv",
+				 col_types = cols(region = col_character(),
+				 								 .default = col_double())) %>%
+	select(-ends_with("iqr")) %>%
+	pivot_longer(cols=c(split_rate, lump_rate), names_to="method", values_to="fraction") %>%
+	mutate(method = ifelse(method == "lump_rate",
+												 "Fraction of species\nmerged together",
+												 "Fraction of genomes\nsplit apart")) %>%
+	ggplot(aes(x=threshold, y=fraction, color=region)) + 
+	geom_line() +
+	facet_wrap(~method, nrow=2, strip.position = "left") +
+	scale_x_continuous(
+		breaks = c(0, 0.025, 0.05, 0.075, 0.1),
+		labels=c("0", "2.5", "5", "7.5", "10")
+		) + 
+	scale_color_manual(name = NULL,
+										 breaks = c("v19", "v34", "v4", "v45"),
+										 values = c("black", "blue", "green", "red"),
+										 labels = c("V1-V9", "V3-V4", "V4", "V4-V5")) +
+	labs(x="Distance theshold\nto define ASV/OTU (%)", y=NULL) +
+	theme_classic() +
+	theme(
+		strip.placement="outside",
+		strip.background =element_rect(color=NA),
+		strip.text = element_text(size=11),
+		legend.position = c(0.8, 0.9),
+		legend.key.size = unit(0.4, "cm")
+	)
+
+ggsave("figures/lump_split.tiff", height =5, width =4, compression="lzw")
+ggsave("figures/lump_split.pdf", height =5, width =4)

### 2021-01-20 code/plot_roc_curve.R
Generate ROC curve figure

@@ -1,0 +1,81 @@
+#!/usr/bin/env Rscript --vanilla
+
+library(tidyverse)
+library(here)
+library(knitr)
+
+sensitivity_specificity <- read_tsv(here("data/processed/rrnDB.roc.tsv"),
+																		col_types = cols(.default = col_double(),
+																										 region = col_character()
+																		)) %>%
+	mutate(sensitivity = true_pos / (true_pos + false_neg),
+				 specificity = true_neg / (true_neg + false_pos)) %>%
+	group_by(region, threshold) %>%
+	summarize(sensitivity = median(sensitivity),
+						specificity = median(specificity),
+						.groups="drop") %>%
+	select(region, threshold, sensitivity, specificity)
+
+
+three <- sensitivity_specificity %>%
+	filter(threshold == 0.03) %>%
+	select(region, sensitivity, specificity)
+
+
+balance <- sensitivity_specificity %>%
+	mutate(diff=abs(sensitivity - specificity)) %>%
+	group_by(region) %>%
+	summarize(min_diff = min(diff),
+						threshold = threshold[which.min(diff)],
+						sensitivity = sensitivity[which.min(diff)],
+						specificity = specificity[which.min(diff)],
+						.groups="drop"
+	) %>%
+	select(region, sensitivity, specificity)
+
+distance <- sensitivity_specificity %>%
+	mutate(distance = sqrt((specificity - 1)^2 + (sensitivity - 1)^2)) %>%
+	group_by(region) %>%
+	summarize(min_distance = min(distance),
+						threshold = threshold[which.min(distance)],
+						sensitivity = sensitivity[which.min(distance)],
+						specificity = specificity[which.min(distance)],
+						.groups="drop"
+	) %>%
+	select(region, sensitivity, specificity)
+
+annotation <- bind_rows(three = three, balance=balance, distance=distance,
+												.id="method") %>%
+	mutate(xend = case_when(method == "three" ~ 0.015,
+													method == "balance" ~ 0.04,
+													method == "distance" ~ 0.025),
+				 yend = case_when(method == "three" ~ 0.925,
+	 											method == "balance" ~ 0.975,
+	 											method == "distance" ~ 0.950),
+				 pretty = case_when(method == "three" ~ "3% distance\nthreshold",
[…]

### 2021-01-20 code/plot_esv_rate.R
Use colors that are friendly for red/green color defficient people

@@ -3,6 +3,7 @@
 library(tidyverse)
 library(knitr)
 library(here)
+library(ggtext)
 
 metadata <- read_tsv(here("data/references/genome_id_taxonomy.tsv"),
 										 col_types = cols(.default = col_character())) %>%
@@ -50,10 +51,11 @@ species_esvs %>%
 	scale_x_log10()+
 	scale_y_log10(breaks=c(0.1, 1, 10, 100), labels = c("0.1", "1", "10", "100")) +
 	labs(x="Number of genomes",
-			 y="Number of ASVs per rrn copy") +
+			 y="Number of ASVs per *rrn* copy") +
 	theme_classic() +
 	theme(
-		strip.text = element_blank()
+		strip.text = element_blank(),
+		axis.title.y = element_markdown()
 	)
 
 ggsave("figures/esv_rate.tiff", width=3.5, height=8, compression="lzw")

### 2021-01-20 code/plot_lump_split.R

@@ -1,6 +1,7 @@
 #!/usr/bin/env Rscript --vanilla
 
 library(tidyverse)
+source('code/colors.R')
 
 read_tsv("data/processed/lumped_split_rate.tsv",
 				 col_types = cols(region = col_character(),
@@ -17,10 +18,7 @@ read_tsv("data/processed/lumped_split_rate.tsv",
 		breaks = c(0, 0.025, 0.05, 0.075, 0.1),
 		labels=c("0", "2.5", "5", "7.5", "10")
 		) + 
-	scale_color_manual(name = NULL,
-										 breaks = c("v19", "v34", "v4", "v45"),
-										 values = c("black", "blue", "green", "red"),
-										 labels = c("V1-V9", "V3-V4", "V4", "V4-V5")) +
+	custom_color_scale() + 
 	labs(x="Distance theshold\nto define ASV/OTU (%)", y=NULL) +
 	theme_classic() +
 	theme(

### 2021-01-20 code/plot_roc_curve.R

@@ -2,7 +2,7 @@
 
 library(tidyverse)
 library(here)
-library(knitr)
+source(here("code/colors.R"))
 
 sensitivity_specificity <- read_tsv(here("data/processed/rrnDB.roc.tsv"),
 																		col_types = cols(.default = col_double(),
@@ -66,11 +66,8 @@ sensitivity_specificity %>%
 	geom_segment(data=annotation, aes(x=1-specificity, y=sensitivity, xend=xend, yend=yend)) +
 	geom_label(data=annotation, aes(label=pretty, x=xend, y=yend), color="black") +
 	coord_cartesian(ylim=c(0.9, 1)) +
-	scale_color_manual(name = NULL,
-										 breaks = c("v19", "v34", "v4", "v45"),
-										 values = c("black", "blue", "green", "red"),
-										 labels = c("V1-V9", "V3-V4", "V4", "V4-V5")) +
-		labs(x="1-Specificity", y="Sensitivity") +
+	custom_color_scale() + 
+	labs(x="1-Specificity", y="Sensitivity") +
 	theme_classic() +
 	theme(
 		legend.position = c(0.9, 0.2),

### 2021-01-20 code/colors.R
Use colors that are friendly for red/green color defficient people

@@ -1,0 +1,8 @@
+custom_color_scale <- function(){
+	
+	scale_color_manual(name = NULL,
+										 breaks = c("v19", "v34", "v4", "v45"),
+										 values = c("black", "#1b9e77", "#d95f02", "#7570b3"),
+										 labels = c("V1-V9", "V3-V4", "V4", "V4-V5"))
+	
+}

### 2021-01-20 submission/manuscript.Rmd
Generate lumping and splitting rate figure
Generate ROC curve figure
Create figure captions and integrate numbers into manuscript

@@
 *`r single_copy$species`*) and `r max_copy$median` (*`r max_copy$species`*) copies of the rrn operon.
-As the number of copies of the operon in a genome increased, the number of variants of the 16S rRNA gene in each genome also increased (**FIGURE**).
+As the number of copies of the operon in a genome increased, the number of variants of the 16S rRNA gene in each genome also increased.
 On average, there were `r rate_v19` variants per copy of the full length 16S rRNA gene and an average of `r rate_v4`, `r rate_v34`, and `r rate_v45` variants when considering the V4, V3-V4, and V4-V5 regions of the gene, respectively.
@@
 Similarly, thresholds of `r format_pct(threshold_v4)`, `r format_pct(threshold_v34)`, and `r format_pct(threshold_v45)`% were observed for the V4, V3-V4, and V4-V5 regions, respectively.
-However, at these thresholds, multiple species could be represented by the same OTU.
+However, at these thresholds, multiple species could be represented by the same OTU (**Figure 2**).
 At the highest level of resolution, `r format_pct(lump_v19)`% of the species shared a 16S rRNA gene sequence variant with another species when considering full length sequences and `r format_pct(lump_v4)`, `r format_pct(lump_v34)`, and `r format_pct(lump_v45)`% when considering the V4, V3-V4, and V4-V5 regions, respectively.
@@
 I created confusion matrices for multiple regions of the 16S rRNA gene: true positives were those cases where two ASVs were joined in the same OTU and the same species; true negatives were those cases where two ASVs from different OTUs came from different species; false positives were those ASVs that joined the same OTU, but were from different species; and false negatives were those ASVs that joined different OTUs, but were from the same species.
-By calculating the sensitivity and specificity for each threshold and each region of the 16S rRNA gene, I was able to constuct a receiver operator characteristic curve (ROC).
+By calculating the sensitivity and specificity for each threshold and each region of the 16S rRNA gene, I was able to constuct a receiver operator characteristic curve (ROC; **Figure S1**).
 Because the ROC curve represents a range of possible thresholds and sensitivities and specificities, I used two metrics to select the best threshold for defining an OTU.
@@
 
-**Figure 1. ESV rate increases as the number of genomes sampled per species increases**
+**Figure 1.
+The ratio of number of distinct ASVs per copy of the *rrn* operon increases for a species as the number of genomes sampled increases.** Each point represents a different species and is shaded to be 80% transparent so that when points overlap they become darker.
+The blue line represents a smoothed fit through the data.
+
+\newpage
+
+\includegraphics{../figures/lump_split.pdf}
+
+**Figure 2.
+As the distance threshold used to define an OTU increases, the fraction of genomes split into separate OTUs decreases while the fraction of species that are merged into the same OTU increases.** These data represent the median fractions for both measurements across 100 randomizations.
+In each randomization, one genome was sampled from each species.
+
+\newpage
+
+\includegraphics{../figures/roc_curve.pdf}
+
+```{r}
+asv_sens_spec <- sensitivity_specificity %>%
+	filter(threshold == 0) %>% 
+	mutate(sensitivity = format(sensitivity, digits=3, nsmall=3),
+				 specificity = format(specificity, digits=3, nsmall=3),
+				 pretty = paste(sensitivity, specificity, sep="/")
+				 )
+v19 <- asv_sens_spec %>% filter(region == "v19") %>% pull(pretty)
+v34 <- asv_sens_spec %>% filter(region == "v34") %>% pull(pretty)
+v4 <- asv_sens_spec %>% filter(region == "v4") %>% pull(pretty)
+v45 <- asv_sens_spec %>% filter(region == "v45") %>% pull(pretty)
+
+```
+
+**Figure S1.
+Distance thresholds larger than 3% provide better sensitivity and specificity when assigning ASVs to OTUs when trying to represent species-level classifications.** The sensitivities and specificities for ASVs were `r v19`, `r v34`, `r v4`, and `r v45` for the V1-V9, V3-V4, V4, V4-V5 regions, respectively.
+The gray diagonal line represents the position of where the sensitivity and specificity were equal.
+These data represent the median fractions for sensitivity and specificity across 100 randomizations.
+In each randomization, one genome was sampled from each species.

### 2021-01-20 submission/manuscript.tex
Generate lumping and splitting rate figure
Generate ROC curve figure
Create figure captions and integrate numbers into manuscript

@@
 the number of copies of the operon in a genome increased, the number of
-variants of the 16S rRNA gene in each genome also increased
-(\textbf{FIGURE}). On average, there were \textbf{1} variants per copy
-of the full length 16S rRNA gene and an average of \textbf{0.26},
-\textbf{0.33}, and \textbf{0.27} variants when considering the V4,
-V3-V4, and V4-V5 regions of the gene, respectively. Although a species
-tended to have a consistent number of 16S rRNA gene copies per genome,
-the number of total variants increased with the number of genomes that
-were sampled (\textbf{Figure 1}). For example,
-\textbf{\emph{Mycobacterium tuberculosis}} generally only had \textbf{1}
-copy of the gene per genome, but across the \textbf{180} genomes that
-have been sequenced there were \textbf{11} versions of the gene.
-Similarly, a \emph{E. coli} genome typically had \textbf{7} copies of
-the 16S rRNA gene with between \textbf{6} and \textbf{10} distinct full
-length sequences per genome. Across the \textbf{958} \emph{E. coli}
+variants of the 16S rRNA gene in each genome also increased. On average,
+there were \textbf{1} variants per copy of the full length 16S rRNA gene
+and an average of \textbf{0.26}, \textbf{0.33}, and \textbf{0.27}
+variants when considering the V4, V3-V4, and V4-V5 regions of the gene,
+respectively. Although a species tended to have a consistent number of
+16S rRNA gene copies per genome, the number of total variants increased
+with the number of genomes that were sampled (\textbf{Figure 1}). For
+example, \textbf{\emph{Mycobacterium tuberculosis}} generally only had
+\textbf{1} copy of the gene per genome, but across the \textbf{180}
+genomes that have been sequenced there were \textbf{11} versions of the
+gene. Similarly, a \emph{E. coli} genome typically had \textbf{7} copies
+of the 16S rRNA gene with between \textbf{6} and \textbf{10} distinct
+full length sequences per genome. Across the \textbf{958} \emph{E. coli}
 genomes that have been sequenced, there were \textbf{1,013} different
@@
 V4, V3-V4, and V4-V5 regions, respectively. However, at these
-thresholds, multiple species could be represented by the same OTU. At
-the highest level of resolution, \textbf{3.6}\% of the species shared a
-16S rRNA gene sequence variant with another species when considering
-full length sequences and \textbf{14.9}, \textbf{10.2}, and
-\textbf{12.0}\% when considering the V4, V3-V4, and V4-V5 regions,
-respectively. At the commonly used 3\% threshold, \textbf{25.2}\% of the
-species shared an OTU when considering full length sequences and
-\textbf{33.0}, \textbf{29.4}, and \textbf{32.2}\% when considering the
-V4, V3-V4, and V4-V5 regions, respectively. Given the risk of splitting
-a genome into multiple OTUs is more biologically problematic than
-lumping species together, larger thresholds are advisable.
+thresholds, multiple species could be represented by the same OTU
+(\textbf{Figure 2}). At the highest level of resolution, \textbf{3.6}\%
+of the species shared a 16S rRNA gene sequence variant with another
+species when considering full length sequences and \textbf{14.9},
+\textbf{10.2}, and \textbf{12.0}\% when considering the V4, V3-V4, and
+V4-V5 regions, respectively. At the commonly used 3\% threshold,
+\textbf{25.2}\% of the species shared an OTU when considering full
+length sequences and \textbf{33.0}, \textbf{29.4}, and \textbf{32.2}\%
+when considering the V4, V3-V4, and V4-V5 regions, respectively. Given
+the risk of splitting a genome into multiple OTUs is more biologically
+problematic than lumping species together, larger thresholds are
+advisable.
 
@@
 for each threshold and each region of the 16S rRNA gene, I was able to
-constuct a receiver operator characteristic curve (ROC). Because the ROC
-curve represents a range of possible thresholds and sensitivities and
-specificities, I used two metrics to select the best threshold for
-defining an OTU. First, I identified the thresholds where the
-sensitivity and specificity were most similar to each other. For this
-criterion, the best distance thresholds were \textbf{6.0}\% (V1-V9),
-\textbf{4.5}\% (V4), \textbf{5.5}\% (V3-V4), and \textbf{4.0}\% (V4-V5).
-Second, I identified the distance threshold that resulted in the point
-on the ROC curve that was closest to perfect classification. For this
-criterion, the best distance thresholds were \textbf{5.5}\% (V1-V9),
-\textbf{3.5}\% (V4), \textbf{4.5}\% (V3-V4), and \textbf{3.5}\% (V4-V5).
-Surprisingly, these analyses revealed that thresholds near 3\% distance
-balance the risks of splitting genomes into separate OTUs and lumping
-species into the same OTU.
+constuct a receiver operator characteristic curve (ROC; \textbf{Figure
+S1}). Because the ROC curve represents a range of possible thresholds
+and sensitivities and specificities, I used two metrics to select the
+best threshold for defining an OTU. First, I identified the thresholds
+where the sensitivity and specificity were most similar to each other.
+For this criterion, the best distance thresholds were \textbf{6.0}\%
+(V1-V9), \textbf{4.5}\% (V4), \textbf{5.5}\% (V3-V4), and \textbf{4.0}\%
+(V4-V5). Second, I identified the distance threshold that resulted in
+the point on the ROC curve that was closest to perfect classification.
+For this criterion, the best distance thresholds were \textbf{5.5}\%
+(V1-V9), \textbf{3.5}\% (V4), \textbf{4.5}\% (V3-V4), and \textbf{3.5}\%
+(V4-V5). Surprisingly, these analyses revealed that thresholds near 3\%
+distance balance the risks of splitting genomes into separate OTUs and
+lumping species into the same OTU.
 
@@
 
-\textbf{Figure 1. ESV rate increases as the number of genomes sampled
-per species increases}
+\textbf{Figure 1. The ratio of number of distinct ASVs per copy of the
+\emph{rrn} operon increases for a species as the number of genomes
+sampled increases.} Each point represents a different species and is
+shaded to be 80\% transparent so that when points overlap they become
+darker. The blue line represents a smoothed fit through the data.
+
+\newpage
+
+\includegraphics{../figures/lump_split.pdf}
+
+\textbf{Figure 2. As the distance threshold used to define an OTU
+increases, the fraction of genomes split into separate OTUs decreases
+while the fraction of species that are merged into the same OTU
+increases.} These data represent the median fractions for both
+measurements across 100 randomizations. In each randomization, one
+genome was sampled from each species.
+
+\newpage
+
+\includegraphics{../figures/roc_curve.pdf}
+
+\textbf{Figure S1. Distance thresholds larger than 3\% provide better
+sensitivity and specificity when assigning ASVs to OTUs when trying to
+represent species-level classifications.} The sensitivities and
+specificities for ASVs were \textbf{0.435/1.000}, \textbf{0.748/1.000},
+\textbf{0.887/0.999}, and \textbf{0.853/0.999} for the V1-V9, V3-V4, V4,
+V4-V5 regions, respectively. The gray diagonal line represents the
+position of where the sensitivity and specificity were equal. These data
+represent the median fractions for sensitivity and specificity across
+100 randomizations. In each randomization, one genome was sampled from
+each species.
 

### 2021-02-05 submission/manuscript.Rmd
Remove ROC curve analysis and replace with threshold analysis

@@
 											 							 .default = col_double())) %>%
-	filter(n_rrns == 7)
+								filter(n_genomes > 100)
 
-threshold_v19 <- thresholds %>% filter(region == "v19") %>% pull(threshold) * 100
-threshold_v4 <- thresholds %>% filter(region == "v4") %>% pull(threshold) * 100
-threshold_v34 <- thresholds %>% filter(region == "v34") %>% pull(threshold) * 100
-threshold_v45 <- thresholds %>% filter(region == "v45") %>% pull(threshold) * 100
+threshold_v19 <- thresholds %>% filter(n_rrns == 7 & region == "v19") %>% pull(threshold) * 100
+threshold_v4 <- thresholds %>% filter(n_rrns == 7 & region == "v4") %>% pull(threshold) * 100
+threshold_v34 <- thresholds %>% filter(n_rrns == 7 & region == "v34") %>% pull(threshold) * 100
+threshold_v45 <- thresholds %>% filter(n_rrns == 7 & region == "v45") %>% pull(threshold) * 100
 
+three_v19 <- thresholds %>% filter(threshold <= 0.03 & region == "v19") %>% tail(n=1) %>% pull(n_rrns)
+three_v4 <- thresholds %>% filter(threshold <= 0.03 & region == "v4") %>% tail(n=1) %>% pull(n_rrns)
+three_v34 <- thresholds %>% filter(threshold <= 0.03 & region == "v34") %>% tail(n=1) %>% pull(n_rrns)
+three_v45 <- thresholds %>% filter(threshold <= 0.03 & region == "v45") %>% tail(n=1) %>% pull(n_rrns)
+```
+
+A method to avoid splitting a single genome into multiple units of inference is to cluster together similar 16S rRNA gene sequences.
+Therefore, I assessed the impact of the distance threshold used to define clusters of 16S rRNA genes on the propensity to split a genome into separate clusters.
+I observed that as the number of copies of the *rrn* operon increased, the distance threshold required to reduce the ASVs in each genome to a single OTU increased (Figure 1).
+Among species with 7 copies of the *rrn* operon (e.g., *E.
+coli*), I found that a threshold of `r format_pct(threshold_v19)`% was required to reduce full length ASVs to a single OTU in 95% of the species.
+Similarly, thresholds of `r format_pct(threshold_v4)`, `r format_pct(threshold_v34)`, and `r format_pct(threshold_v45)`% were required for the V4, V3-V4, and V4-V5 regions, respectively.
+But, if a 3% distance threshold was used, then ASVs from genomes containing fewer than `r three_v19`, `r three_v4`, `r three_v34`, and `r three_v45` copies of the *rrn* operon would reliably be clustered into a single OTU for ASVs from the V1-V9, V4, V3-V4, and V4-V5 regions, respectively.
+Consequently, these results demonstrate that broad thresholds must be used to avoid splitting different operons from the same genome into separate clusters.
+
+```{r}
 lump_split <- read_tsv(here("data/processed/lumped_split_rate.tsv"),
@@
 lump3_v45 <- lump_split %>% filter(threshold ==0.03 & region == "v45") %>% pull(lump_rate) * 100
-
 ```
 
-A method to avoid splitting a single genome into multiple units of inference is to cluster 16S rRNA gene sequences together that are similar.
-However, this also increases the risk of lumping together genes from different species that are similar to each other.
-Therefore, I assessed the impact of the threshold used to define clusters of 16S rRNA genes on the propensity to split a genome apart or to lump species together.
-For full length 16S rRNA gene sequences, I found that at a threshold of `r format_pct(threshold_v19)`%, 95% of the species with 7 copies of the rrn operon would be represented by a single OTU.
-Similarly, thresholds of `r format_pct(threshold_v4)`, `r format_pct(threshold_v34)`, and `r format_pct(threshold_v45)`% were observed for the V4, V3-V4, and V4-V5 regions, respectively.
-However, at these thresholds, multiple species could be represented by the same OTU (**Figure 2**).
-At the highest level of resolution, `r format_pct(lump_v19)`% of the species shared a 16S rRNA gene sequence variant with another species when considering full length sequences and `r format_pct(lump_v4)`, `r format_pct(lump_v34)`, and `r format_pct(lump_v45)`% when considering the V4, V3-V4, and V4-V5 regions, respectively.
+At broad thresholds thresholds multiple species could be represented by the same OTU (**Figure 2**).
+Using ASVs, `r format_pct(lump_v19)`% of the species shared a 16S rRNA gene sequence variant with another species when considering full length sequences and `r format_pct(lump_v4)`, `r format_pct(lump_v34)`, and `r format_pct(lump_v45)`% when considering the V4, V3-V4, and V4-V5 regions, respectively.
 At the commonly used 3% threshold, `r format_pct(lump3_v19)`% of the species shared an OTU when considering full length sequences and `r format_pct(lump3_v4)`, `r format_pct(lump3_v34)`, and `r format_pct(lump3_v45)`% when considering the V4, V3-V4, and V4-V5 regions, respectively.
-Given the risk of splitting a genome into multiple OTUs is more biologically problematic than lumping species together, larger thresholds are advisable.
-
-```{r}
-sensitivity_specificity <- read_tsv(here("data/processed/rrnDB.roc.tsv"),
-													 col_types = cols(.default = col_double(),
-													 								 region = col_character()
-													 								 )
-													 ) %>%
-	mutate(sensitivity = true_pos / (true_pos + false_neg),
-				 specificity = true_neg / (true_neg + false_pos)) %>%
-	group_by(region, threshold) %>%
-	summarize(sensitivity = median(sensitivity),
-						specificity = median(specificity),
-						.groups="drop") %>%
-	select(region, threshold, sensitivity, specificity)
-
-balance <- sensitivity_specificity %>%
-	mutate(diff=abs(sensitivity - specificity)) %>%
-	group_by(region) %>%
-	summarize(min_diff = min(diff),
-						threshold = threshold[which.min(diff)],
-						sensitivity = sensitivity[which.min(diff)],
-						specificity = specificity[which.min(diff)],
-						.groups="drop"
-						)
-
-balance_v19 <- balance %>% filter(region == "v19") %>% pull(threshold) * 100
-balance_v4 <- balance %>% filter(region == "v4") %>% pull(threshold) * 100
-balance_v34 <- balance %>% filter(region == "v34") %>% pull(threshold) * 100
-balance_v45 <- balance %>% filter(region == "v45") %>% pull(threshold) * 100
-
-distance <- sensitivity_specificity %>%
-	mutate(distance = sqrt((specificity - 1)^2 + (sensitivity - 1)^2)) %>%
-	group_by(region) %>%
-	summarize(min_distance = min(distance),
-						threshold = threshold[which.min(distance)],
-						sensitivity = sensitivity[which.min(distance)],
-						specificity = specificity[which.min(distance)],
-						.groups="drop"
-						)
-
-distance_v19 <- distance %>% filter(region == "v19") %>% pull(threshold) * 100
-distance_v4 <- distance %>% filter(region == "v4") %>% pull(threshold) * 100
-distance_v34 <- distance %>% filter(region == "v34") %>% pull(threshold) * 100
-distance_v45 <- distance %>% filter(region == "v45") %>% pull(threshold) * 100
-
-```
-
-To provide a more nuanced approach to selecting a threshold, it would be useful to to quantify the sensitivity and specificity of characterizing bacterial species using OTUs defined at different thresholds.
-I created confusion matrices for multiple regions of the 16S rRNA gene: true positives were those cases where two ASVs were joined in the same OTU and the same species; true negatives were those cases where two ASVs from different OTUs came from different species; false positives were those ASVs that joined the same OTU, but were from different species; and false negatives were those ASVs that joined different OTUs, but were from the same species.
-By calculating the sensitivity and specificity for each threshold and each region of the 16S rRNA gene, I was able to constuct a receiver operator characteristic curve (ROC; **Figure S1**).
-Because the ROC curve represents a range of possible thresholds and sensitivities and specificities, I used two metrics to select the best threshold for defining an OTU.
-First, I identified the thresholds where the sensitivity and specificity were most similar to each other.
-For this criterion, the best distance thresholds were `r format_pct(balance_v19)`% (V1-V9), `r format_pct(balance_v4)`% (V4), `r format_pct(balance_v34)`% (V3-V4), and `r format_pct(balance_v45)`% (V4-V5).
-Second, I identified the distance threshold that resulted in the point on the ROC curve that was closest to perfect classification.
-For this criterion, the best distance thresholds were `r format_pct(distance_v19)`% (V1-V9), `r format_pct(distance_v4)`% (V4), `r format_pct(distance_v34)`% (V3-V4), and `r format_pct(distance_v45)`% (V4-V5).
-Surprisingly, these analyses revealed that thresholds near 3% distance balance the risks of splitting genomes into separate OTUs and lumping species into the same OTU.
+Considering that species designations are unevenly applied and reflect multiple biases, the risk of splitting a genome into multiple OTUs more problematic than clustering species together.
+Therefore, larger thresholds are advisable.
 
@@
 
-\includegraphics{../figures/esv_rate.pdf}
+\includegraphics{../figures/copy_number_threshold_plot.pdf}
 
 **Figure 1.
-The ratio of number of distinct ASVs per copy of the *rrn* operon increases for a species as the number of genomes sampled increases.** Each point represents a different species and is shaded to be 80% transparent so that when points overlap they become darker.
-The blue line represents a smoothed fit through the data.
+The distance threshold required to prevent the splitting of genomes into multiple OTUs increases as the number of *rrn* operons in the genome increases.** Each line represents the median distance threshold for each region of the 16S rRNA gene that is required for 95% of the species with the indicated numbrer of *rrn* operons to cluster their ASVs to a single OTU.
+The median distance threshold was calculated across 100 randomizations in which one genome was sampled from each species.
+Only those number of *rrn* operons that were found in more than 100 species are included.
 
@@
 
-\includegraphics{../figures/roc_curve.pdf}
-
-```{r}
-asv_sens_spec <- sensitivity_specificity %>%
-	filter(threshold == 0) %>% 
-	mutate(sensitivity = format(sensitivity, digits=3, nsmall=3),
-				 specificity = format(specificity, digits=3, nsmall=3),
-				 pretty = paste(sensitivity, specificity, sep="/")
-				 )
-v19 <- asv_sens_spec %>% filter(region == "v19") %>% pull(pretty)
-v34 <- asv_sens_spec %>% filter(region == "v34") %>% pull(pretty)
-v4 <- asv_sens_spec %>% filter(region == "v4") %>% pull(pretty)
-v45 <- asv_sens_spec %>% filter(region == "v45") %>% pull(pretty)
-
-```
+\includegraphics{../figures/esv_rate.pdf}
 
 **Figure S1.
-Distance thresholds larger than 3% provide better sensitivity and specificity when assigning ASVs to OTUs when trying to represent species-level classifications.** The sensitivities and specificities for ASVs were `r v19`, `r v34`, `r v4`, and `r v45` for the V1-V9, V3-V4, V4, V4-V5 regions, respectively.
-The gray diagonal line represents the position of where the sensitivity and specificity were equal.
-These data represent the median fractions for sensitivity and specificity across 100 randomizations.
-In each randomization, one genome was sampled from each species.
+The ratio of number of distinct ASVs per copy of the *rrn* operon increases for a species as the number of genomes sampled increases.** Each point represents a different species and is shaded to be 80% transparent so that when points overlap they become darker.
+The blue line represents a smoothed fit through the data.

### 2021-02-05 submission/manuscript.tex
Remove ROC curve analysis and replace with threshold analysis

@@
 A method to avoid splitting a single genome into multiple units of
-inference is to cluster 16S rRNA gene sequences together that are
-similar. However, this also increases the risk of lumping together genes
-from different species that are similar to each other. Therefore, I
-assessed the impact of the threshold used to define clusters of 16S rRNA
-genes on the propensity to split a genome apart or to lump species
-together. For full length 16S rRNA gene sequences, I found that at a
-threshold of \textbf{5.5}\%, 95\% of the species with 7 copies of the
-rrn operon would be represented by a single OTU. Similarly, thresholds
-of \textbf{2.5}, \textbf{4.0}, and \textbf{3.5}\% were observed for the
-V4, V3-V4, and V4-V5 regions, respectively. However, at these
-thresholds, multiple species could be represented by the same OTU
-(\textbf{Figure 2}). At the highest level of resolution, \textbf{3.6}\%
-of the species shared a 16S rRNA gene sequence variant with another
-species when considering full length sequences and \textbf{14.9},
-\textbf{10.2}, and \textbf{12.0}\% when considering the V4, V3-V4, and
-V4-V5 regions, respectively. At the commonly used 3\% threshold,
-\textbf{25.2}\% of the species shared an OTU when considering full
-length sequences and \textbf{33.0}, \textbf{29.4}, and \textbf{32.2}\%
-when considering the V4, V3-V4, and V4-V5 regions, respectively. Given
-the risk of splitting a genome into multiple OTUs is more biologically
-problematic than lumping species together, larger thresholds are
-advisable.
+inference is to cluster together similar 16S rRNA gene sequences.
+Therefore, I assessed the impact of the distance threshold used to
+define clusters of 16S rRNA genes on the propensity to split a genome
+into separate clusters. I observed that as the number of copies of the
+\emph{rrn} operon increased, the distance threshold required to reduce
+the ASVs in each genome to a single OTU increased (Figure 1). Among
+species with 7 copies of the \emph{rrn} operon (e.g., \emph{E. coli}), I
+found that a threshold of \textbf{5.5}\% was required to reduce full
+length ASVs to a single OTU in 95\% of the species. Similarly,
+thresholds of \textbf{2.5}, \textbf{4.0}, and \textbf{3.5}\% were
+required for the V4, V3-V4, and V4-V5 regions, respectively. But, if a
+3\% distance threshold was used, then ASVs from genomes containing fewer
+than \textbf{5}, \textbf{8}, \textbf{6}, and \textbf{6} copies of the
+\emph{rrn} operon would reliably be clustered into a single OTU for ASVs
+from the V1-V9, V4, V3-V4, and V4-V5 regions, respectively.
+Consequently, these results demonstrate that broad thresholds must be
+used to avoid splitting different operons from the same genome into
+separate clusters.
 
-To provide a more nuanced approach to selecting a threshold, it would be
-useful to to quantify the sensitivity and specificity of characterizing
-bacterial species using OTUs defined at different thresholds. I created
-confusion matrices for multiple regions of the 16S rRNA gene: true
-positives were those cases where two ASVs were joined in the same OTU
-and the same species; true negatives were those cases where two ASVs
-from different OTUs came from different species; false positives were
-those ASVs that joined the same OTU, but were from different species;
-and false negatives were those ASVs that joined different OTUs, but were
-from the same species. By calculating the sensitivity and specificity
-for each threshold and each region of the 16S rRNA gene, I was able to
-constuct a receiver operator characteristic curve (ROC; \textbf{Figure
-S1}). Because the ROC curve represents a range of possible thresholds
-and sensitivities and specificities, I used two metrics to select the
-best threshold for defining an OTU. First, I identified the thresholds
-where the sensitivity and specificity were most similar to each other.
-For this criterion, the best distance thresholds were \textbf{6.0}\%
-(V1-V9), \textbf{4.5}\% (V4), \textbf{5.5}\% (V3-V4), and \textbf{4.0}\%
-(V4-V5). Second, I identified the distance threshold that resulted in
-the point on the ROC curve that was closest to perfect classification.
-For this criterion, the best distance thresholds were \textbf{5.5}\%
-(V1-V9), \textbf{3.5}\% (V4), \textbf{4.5}\% (V3-V4), and \textbf{3.5}\%
-(V4-V5). Surprisingly, these analyses revealed that thresholds near 3\%
-distance balance the risks of splitting genomes into separate OTUs and
-lumping species into the same OTU.
+At broad thresholds thresholds multiple species could be represented by
+the same OTU (\textbf{Figure 2}). Using ASVs, \textbf{3.6}\% of the
+species shared a 16S rRNA gene sequence variant with another species
+when considering full length sequences and \textbf{14.9}, \textbf{10.2},
+and \textbf{12.0}\% when considering the V4, V3-V4, and V4-V5 regions,
+respectively. At the commonly used 3\% threshold, \textbf{25.2}\% of the
+species shared an OTU when considering full length sequences and
+\textbf{33.0}, \textbf{29.4}, and \textbf{32.2}\% when considering the
+V4, V3-V4, and V4-V5 regions, respectively. Considering that species
+designations are unevenly applied and reflect multiple biases, the risk
+of splitting a genome into multiple OTUs more problematic than
+clustering species together. Therefore, larger thresholds are advisable.
 
@@
 
-\includegraphics{../figures/esv_rate.pdf}
+\includegraphics{../figures/copy_number_threshold_plot.pdf}
 
-\textbf{Figure 1. The ratio of number of distinct ASVs per copy of the
-\emph{rrn} operon increases for a species as the number of genomes
-sampled increases.} Each point represents a different species and is
-shaded to be 80\% transparent so that when points overlap they become
-darker. The blue line represents a smoothed fit through the data.
+\textbf{Figure 1. The distance threshold required to prevent the
+splitting of genomes into multiple OTUs increases as the number of
+\emph{rrn} operons in the genome increases.} Each line represents the
+median distance threshold for each region of the 16S rRNA gene that is
+required for 95\% of the species with the indicated numbrer of
+\emph{rrn} operons to cluster their ASVs to a single OTU. The median
+distance threshold was calculated across 100 randomizations in which one
+genome was sampled from each species. Only those number of \emph{rrn}
+operons that were found in more than 100 species are included.
 
@@
 
-\includegraphics{../figures/roc_curve.pdf}
+\includegraphics{../figures/esv_rate.pdf}
 
-\textbf{Figure S1. Distance thresholds larger than 3\% provide better
-sensitivity and specificity when assigning ASVs to OTUs when trying to
-represent species-level classifications.} The sensitivities and
-specificities for ASVs were \textbf{0.435/1.000}, \textbf{0.748/1.000},
-\textbf{0.887/0.999}, and \textbf{0.853/0.999} for the V1-V9, V3-V4, V4,
-V4-V5 regions, respectively. The gray diagonal line represents the
-position of where the sensitivity and specificity were equal. These data
-represent the median fractions for sensitivity and specificity across
-100 randomizations. In each randomization, one genome was sampled from
-each species.
+\textbf{Figure S1. The ratio of number of distinct ASVs per copy of the
+\emph{rrn} operon increases for a species as the number of genomes
+sampled increases.} Each point represents a different species and is
+shaded to be 80\% transparent so that when points overlap they become
+darker. The blue line represents a smoothed fit through the data.
 

### 2021-02-05 code/plot_copy_number_threshold.R

@@ -1,0 +1,25 @@
+#!/usr/bin/env Rscript --vanilla
+
+library(tidyverse)
+library(ggtext)
+source("code/colors.R")
+
+read_tsv("data/processed/thresholds_for_single_otu.tsv") %>%
+	filter(n_genomes > 100) %>%
+	ggplot(aes(x=n_rrns, y=threshold, color=region)) +
+	geom_line(size=1) +
+	custom_color_scale() +
+	scale_x_continuous(breaks=seq(1,9,2), labels=seq(1,9,2)) +
+	scale_y_continuous(breaks=seq(0,0.06,0.02), labels=seq(0,6,2), limits =c(0,0.06)) +
+	labs(x="Number of copies of *rrn* operon",
+			 y="Distance threshold where ASVs from\n95% of species have one OTU (%)") +
+	theme_classic() +
+	theme(
+		legend.position = c(0.8, 0.2),
+		legend.key.height = unit(0.4, "cm"),
+		axis.title.x = element_markdown(),
+		axis.title.y = element_text(margin=margin(t=0, r=10, b=0, l=0))
+	)
+
+ggsave("figures/copy_number_threshold_plot.pdf", width=5, height=5)
+ggsave("figures/copy_number_threshold_plot.tiff", width=5, height=5)

### 2021-02-05 code/plot_lump_split.R

@@ -12,7 +12,7 @@ read_tsv("data/processed/lumped_split_rate.tsv",
 												 "Fraction of species\nmerged together",
 												 "Fraction of genomes\nsplit apart")) %>%
 	ggplot(aes(x=threshold, y=fraction, color=region)) + 
-	geom_line() +
+	geom_line(size=1) +
 	facet_wrap(~method, nrow=2, strip.position = "left") +
 	scale_x_continuous(
 		breaks = c(0, 0.025, 0.05, 0.075, 0.1),
@@ -26,7 +26,7 @@ read_tsv("data/processed/lumped_split_rate.tsv",
 		strip.background =element_rect(color=NA),
 		strip.text = element_text(size=11),
 		legend.position = c(0.8, 0.9),
-		legend.key.size = unit(0.4, "cm")
+		legend.key.height = unit(0.4, "cm")
 	)
 
 ggsave("figures/lump_split.tiff", height =5, width =4, compression="lzw")

### 2021-02-05 code/plot_roc_curve.R

@@ -1,78 +1,0 @@
-#!/usr/bin/env Rscript --vanilla
-
-library(tidyverse)
-library(here)
-source(here("code/colors.R"))
-
-sensitivity_specificity <- read_tsv(here("data/processed/rrnDB.roc.tsv"),
-																		col_types = cols(.default = col_double(),
-																										 region = col_character()
-																		)) %>%
-	mutate(sensitivity = true_pos / (true_pos + false_neg),
-				 specificity = true_neg / (true_neg + false_pos)) %>%
-	group_by(region, threshold) %>%
-	summarize(sensitivity = median(sensitivity),
-						specificity = median(specificity),
-						.groups="drop") %>%
-	select(region, threshold, sensitivity, specificity)
-
-
-three <- sensitivity_specificity %>%
-	filter(threshold == 0.03) %>%
-	select(region, sensitivity, specificity)
-
-
-balance <- sensitivity_specificity %>%
-	mutate(diff=abs(sensitivity - specificity)) %>%
-	group_by(region) %>%
-	summarize(min_diff = min(diff),
-						threshold = threshold[which.min(diff)],
-						sensitivity = sensitivity[which.min(diff)],
-						specificity = specificity[which.min(diff)],
-						.groups="drop"
-	) %>%
-	select(region, sensitivity, specificity)
-
-distance <- sensitivity_specificity %>%
-	mutate(distance = sqrt((specificity - 1)^2 + (sensitivity - 1)^2)) %>%
-	group_by(region) %>%
-	summarize(min_distance = min(distance),
-						threshold = threshold[which.min(distance)],
-						sensitivity = sensitivity[which.min(distance)],
-						specificity = specificity[which.min(distance)],
-						.groups="drop"
-	) %>%
-	select(region, sensitivity, specificity)
-
-annotation <- bind_rows(three = three, balance=balance, distance=distance,
-												.id="method") %>%
-	mutate(xend = case_when(method == "three" ~ 0.015,
-													method == "balance" ~ 0.04,
-													method == "distance" ~ 0.025),
-				 yend = case_when(method == "three" ~ 0.925,
-	 											method == "balance" ~ 0.975,
-	 											method == "distance" ~ 0.950),
-				 pretty = case_when(method == "three" ~ "3% distance\nthreshold",
[…]

### 2021-02-05 submission/manuscript.Rmd
Insert references, closes #1

@@
   chunk_output_type: console
+bibliography: references.bib
+csl: asm.csl
 ---
@@
 
-16S rRNA gene sequencing is a very powerful technique for describing and comparing microbial communities.
+16S rRNA gene sequencing is a powerful technique for describing and comparing microbial communities [@Lane1985].
 Efforts to link 16S rRNA gene sequences to taxonomic levels based on distance thresholds go back to at least the 1990s.
-The distance-based thresholds that were developed and are now widely used (3%) were based on DNA-DNA hybridization approaches that are not as precise as genome sequencing.
-Instead, genome sequencing technologies have suggested that the widely used 3% distance threshold to operationally define bacterial taxa is too coarse.
-As an alternative to OTUs, amplicon sequencing variants (ASVs) have been proposed as a way to adopt the thresholds suggested by genome sequencing to microbial community analysis using 16S rRNA gene sequences.
-ASVs are a unit of microbial community inference that do not cluster sequences based on a distance-based threshold.
-However, most bacterial genomes have more than 1 copy of the rrn operon and those copies are not identical.
+The distance-based thresholds that were developed and are now widely used (3%) were based on DNA-DNA hybridization approaches that are not as precise as genome sequencing [@Stackebrandt1994; @Goris2007].
+Instead, genome sequencing technologies have suggested that the widely used 3% distance threshold to operationally define bacterial taxa is too coarse [@RodriguezR2018; @Stackebrandt2006; @Edgar2018].
+As an alternative to OTUs, amplicon sequencing variants (ASVs) have been proposed as a way to adopt the thresholds suggested by genome sequencing to microbial community analysis using 16S rRNA gene sequences [@Edgar2016; @Amir2017; @Callahan2016; @Eren2014].
+ASVs are a unit of microbial community inference that do not cluster sequences based on a distance-based threshold [@Callahan2017].
+However, most bacterial genomes have more than 1 copy of the rrn operon and those copies are not identical [@Pei2010; @Sun2013].
 Therefore, using too fine a threshold to identify OTUs creates the risk of splitting a single genome into multiple bins and using too broad of a threshold to define OTUs creates the risk of lumping together bacterial species into the same OTU.
@@
 
-To investigate the variation in the number of copies of the 16S rRNA gene per genome as well as the intragenomic variation among copies of the 16S rRNA gene, I obtained reference 16S rRNA sequences from the rrn copy number database (rrnDB; CITATION).
+To investigate the variation in the number of copies of the 16S rRNA gene per genome as well as the intragenomic variation among copies of the 16S rRNA gene, I obtained reference 16S rRNA sequences from the rrn copy number database (rrnDB)[@Stoddard2014].
 Among the `r n_distinct(metadata$species)` species represented in the rrnDB there were `r nrow(metadata)` genomes.
-The median number of rrn operson per species ranged between `r single_copy$median` (e.g.
-*`r single_copy$species`*) and `r max_copy$median` (*`r max_copy$species`*) copies of the rrn operon.
+The median number of rrn operson per species ranged between `r single_copy$median` (e.g., *`r single_copy$species`*) and `r max_copy$median` (*`r max_copy$species`*) copies of the rrn operon.
 As the number of copies of the operon in a genome increased, the number of variants of the 16S rRNA gene in each genome also increased.
@@
 
-At broad thresholds thresholds multiple species could be represented by the same OTU (**Figure 2**).
+At broad thresholds multiple species could be represented by the same OTU (**Figure 2**).
 Using ASVs, `r format_pct(lump_v19)`% of the species shared a 16S rRNA gene sequence variant with another species when considering full length sequences and `r format_pct(lump_v4)`, `r format_pct(lump_v34)`, and `r format_pct(lump_v45)`% when considering the V4, V3-V4, and V4-V5 regions, respectively.
@@
 The results of this analysis demonstrate that there is a significant risk of splitting single genomes into multiple bins if too fine of a threshold is applied to defining an OTU.
-An ongoing problem for amplicon-based studies is defining a meaningful taxonomic unit of inference.
-Since there is no consensus definition for a biological species concept, microbiologists must accept that how we have named bacterial species is biased and that taxonomic rules are not applied in a consistent manner.
-This makes it more challenging to attempt to fit a distance threshold to define an OTU definition that matches a set of species names.
-Furthermore, it is unlikely that the 16S rRNA gene evolves at the same rate across all bacterial lineages, which limits the biological interpretation of a common OTU definition.
-At best, a distance-based definition of a taxonomic unit is operational.
-There is general agreement in bacterial systematics that to classify something to a bacterial species, you need phenotypic and genome sequence data (CITATION).
+An ongoing problem for amplicon-based studies is defining a meaningful taxonomic unit of inference [@Callahan2017; @Schloss2011; @Johnson2019].
+Since there is no consensus definition for a biological species concept [@Staley2006; @Oren2013], microbiologists must accept that how we have named bacterial species is biased and that taxonomic rules are not applied in a consistent manner (e.g., [@Baltrus2016]).
+This makes it more challenging to attempt to fit a distance threshold to define an OTU definition that matches a set of species names [@Konstantinidis2005].
+Furthermore, the 16S rRNA gene does not evolve at the same rate across all bacterial lineages [@Schloss2011], which limits the biological interpretation of a common OTU definition.
+A distance-based definition of a taxonomic unit based on 16S rRNA gene or full genome sequences is, at best, operational and not grounded in biological theory [@Schloss2011; @Barco2020; @Parks2020; @Yarza2014].
+There is general agreement in bacterial systematics that to classify something to a bacterial species, you need phenotypic and genome sequence data [@Staley2006; @Oren2013; @Baltrus2016].
 We are asking too much of a short section of a bacterial genome to be able to differentiate between species.
 It is difficult to defend a unit of inference that would split a single genome into multiple taxonomic units.
-It is not biologically plausible to entertain the possability that parts of a genome would have different ecologies.
+It is not biologically plausible to entertain the possibility that parts of a genome would have different ecologies.
 Although there are multiple reasons that proponents of ASVs encourage their use, the significant risk of splitting genomes is too high to warrant their use.
@@
 **Materials and Methods.
-(i) Data availability.** The 16S rRNA gene sequences used in this study were  obtained from the *rrn*DB (https://rrndb.umms.med.umich.edu; version 5.6, released November 8, 2019).
+(i) Data availability.** The 16S rRNA gene sequences used in this study were obtained from the *rrn*DB (https://rrndb.umms.med.umich.edu; version 5.6, released November 8, 2019) [@Stoddard2014].
 At the time of submission, this is the most current version of the database.
@@
 **(ii) Definition of regions within 16S rRNA gene.** The full length 16S rRNA gene sequences were aligned to a SILVA reference alignment of the 16S rRNA gene (v138) using the mothur software package (v.
-1.XX).
+1.XX) [@Schloss2009; @Quast2012].
 Regions of the 16S rRNA gene were selected because of their use in the microbial ecology literature.
@@
 
+\setlength{\parindent}{-0.25in}
+\setlength{\leftskip}{0.25in}
+\noindent
+
+<div id="refs"></div>
+
+\setlength{\parindent}{0in}
+\setlength{\leftskip}{0in}
+
 \newpage

### 2021-02-05 submission/manuscript.tex
Insert references, closes #1

@@
 \fi
+\newlength{\cslhangindent}
+\setlength{\cslhangindent}{1.5em}
+\newenvironment{cslreferences}%
+  {}%
+  {\par}
 
@@
 
-16S rRNA gene sequencing is a very powerful technique for describing and
-comparing microbial communities. Efforts to link 16S rRNA gene sequences
-to taxonomic levels based on distance thresholds go back to at least the
-1990s. The distance-based thresholds that were developed and are now
-widely used (3\%) were based on DNA-DNA hybridization approaches that
-are not as precise as genome sequencing. Instead, genome sequencing
-technologies have suggested that the widely used 3\% distance threshold
-to operationally define bacterial taxa is too coarse. As an alternative
-to OTUs, amplicon sequencing variants (ASVs) have been proposed as a way
-to adopt the thresholds suggested by genome sequencing to microbial
-community analysis using 16S rRNA gene sequences. ASVs are a unit of
-microbial community inference that do not cluster sequences based on a
-distance-based threshold. However, most bacterial genomes have more than
-1 copy of the rrn operon and those copies are not identical. Therefore,
-using too fine a threshold to identify OTUs creates the risk of
-splitting a single genome into multiple bins and using too broad of a
-threshold to define OTUs creates the risk of lumping together bacterial
-species into the same OTU. An example of both is seen in the comparison
-of \emph{Staphylococcus aureus} (NCTC 8325) and \emph{S. epidermidis}
-(ATCC 12228) where each genome has 5 copies of the 16S rRNA gene. The 10
-copies of the 16S rRNA gene each have a different sequence and so if
-OTUs are defined based on ASVs, each genome would be split into 5 OTUs.
-Conversely, if the copies were clustered using a 3\% distance threshold
-all 10 copies would cluster into the same OTU. The goal of this study
-was to quantify the risk of splitting a single genome into multiple bins
-and the risk of lumping together different bacterial species into the
-same bin.
+16S rRNA gene sequencing is a powerful technique for describing and
+comparing microbial communities (1). Efforts to link 16S rRNA gene
+sequences to taxonomic levels based on distance thresholds go back to at
+least the 1990s. The distance-based thresholds that were developed and
+are now widely used (3\%) were based on DNA-DNA hybridization approaches
+that are not as precise as genome sequencing (2, 3). Instead, genome
+sequencing technologies have suggested that the widely used 3\% distance
+threshold to operationally define bacterial taxa is too coarse (4--6).
+As an alternative to OTUs, amplicon sequencing variants (ASVs) have been
+proposed as a way to adopt the thresholds suggested by genome sequencing
+to microbial community analysis using 16S rRNA gene sequences (7--10).
+ASVs are a unit of microbial community inference that do not cluster
+sequences based on a distance-based threshold (11). However, most
+bacterial genomes have more than 1 copy of the rrn operon and those
+copies are not identical (12, 13). Therefore, using too fine a threshold
+to identify OTUs creates the risk of splitting a single genome into
+multiple bins and using too broad of a threshold to define OTUs creates
+the risk of lumping together bacterial species into the same OTU. An
+example of both is seen in the comparison of \emph{Staphylococcus
+aureus} (NCTC 8325) and \emph{S. epidermidis} (ATCC 12228) where each
+genome has 5 copies of the 16S rRNA gene. The 10 copies of the 16S rRNA
+gene each have a different sequence and so if OTUs are defined based on
+ASVs, each genome would be split into 5 OTUs. Conversely, if the copies
+were clustered using a 3\% distance threshold all 10 copies would
+cluster into the same OTU. The goal of this study was to quantify the
+risk of splitting a single genome into multiple bins and the risk of
+lumping together different bacterial species into the same bin.
 
@@
 the 16S rRNA gene, I obtained reference 16S rRNA sequences from the rrn
-copy number database (rrnDB; CITATION). Among the \textbf{4,774} species
+copy number database (rrnDB)(14). Among the \textbf{4,774} species
 represented in the rrnDB there were \textbf{15,614} genomes. The median
-number of rrn operson per species ranged between \textbf{1}
-(e.g.~\textbf{\emph{Mycobacterium tuberculosis}}) and \textbf{19}
+number of rrn operson per species ranged between \textbf{1} (e.g.,
+\textbf{\emph{Mycobacterium tuberculosis}}) and \textbf{19}
 (\textbf{\emph{Metabacillus litoralis}}) copies of the rrn operon. As
@@
 
-At broad thresholds thresholds multiple species could be represented by
-the same OTU (\textbf{Figure 2}). Using ASVs, \textbf{3.6}\% of the
-species shared a 16S rRNA gene sequence variant with another species
-when considering full length sequences and \textbf{14.9}, \textbf{10.2},
-and \textbf{12.0}\% when considering the V4, V3-V4, and V4-V5 regions,
+At broad thresholds multiple species could be represented by the same
+OTU (\textbf{Figure 2}). Using ASVs, \textbf{3.6}\% of the species
+shared a 16S rRNA gene sequence variant with another species when
+considering full length sequences and \textbf{14.9}, \textbf{10.2}, and
+\textbf{12.0}\% when considering the V4, V3-V4, and V4-V5 regions,
 respectively. At the commonly used 3\% threshold, \textbf{25.2}\% of the
@@
 amplicon-based studies is defining a meaningful taxonomic unit of
-inference. Since there is no consensus definition for a biological
-species concept, microbiologists must accept that how we have named
-bacterial species is biased and that taxonomic rules are not applied in
-a consistent manner. This makes it more challenging to attempt to fit a
-distance threshold to define an OTU definition that matches a set of
-species names. Furthermore, it is unlikely that the 16S rRNA gene
-evolves at the same rate across all bacterial lineages, which limits the
-biological interpretation of a common OTU definition. At best, a
-distance-based definition of a taxonomic unit is operational. There is
-general agreement in bacterial systematics that to classify something to
-a bacterial species, you need phenotypic and genome sequence data
-(CITATION). We are asking too much of a short section of a bacterial
-genome to be able to differentiate between species. It is difficult to
-defend a unit of inference that would split a single genome into
-multiple taxonomic units. It is not biologically plausible to entertain
-the possability that parts of a genome would have different ecologies.
-Although there are multiple reasons that proponents of ASVs encourage
-their use, the significant risk of splitting genomes is too high to
-warrant their use.
+inference (11, 15, 16). Since there is no consensus definition for a
+biological species concept (17, 18), microbiologists must accept that
+how we have named bacterial species is biased and that taxonomic rules
+are not applied in a consistent manner (e.g., (19)). This makes it more
+challenging to attempt to fit a distance threshold to define an OTU
+definition that matches a set of species names (20). Furthermore, the
+16S rRNA gene does not evolve at the same rate across all bacterial
+lineages (15), which limits the biological interpretation of a common
+OTU definition. A distance-based definition of a taxonomic unit based on
+16S rRNA gene or full genome sequences is, at best, operational and not
+grounded in biological theory (15, 21--23). There is general agreement
+in bacterial systematics that to classify something to a bacterial
+species, you need phenotypic and genome sequence data (17--19). We are
+asking too much of a short section of a bacterial genome to be able to
+differentiate between species. It is difficult to defend a unit of
+inference that would split a single genome into multiple taxonomic
+units. It is not biologically plausible to entertain the possibility
+that parts of a genome would have different ecologies. Although there
+are multiple reasons that proponents of ASVs encourage their use, the
+significant risk of splitting genomes is too high to warrant their use.
 
@@
 (\url{https://rrndb.umms.med.umich.edu}; version 5.6, released November
-8, 2019). At the time of submission, this is the most current version of
-the database. The \emph{rrn}DB obtained the curated 16S rRNA gene
-sequences from the KEGG database, which ultimately obtained them from
-NCBI's non-redundant RefSeq database. The \emph{rrn}DB provides
+8, 2019) (14). At the time of submission, this is the most current
+version of the database. The \emph{rrn}DB obtained the curated 16S rRNA
+gene sequences from the KEGG database, which ultimately obtained them
+from NCBI's non-redundant RefSeq database. The \emph{rrn}DB provides
 downloadable versions of the sequences with their taxonomy as determined
@@
 alignment of the 16S rRNA gene (v138) using the mothur software package
-(v. 1.XX). Regions of the 16S rRNA gene were selected because of their
-use in the microbial ecology literature. Full length sequences
+(v. 1.XX) (24, 25). Regions of the 16S rRNA gene were selected because
+of their use in the microbial ecology literature. Full length sequences
 corresponded to \emph{E. coli} positions XX through XXXX, V4 to
@@
 
+\setlength{\parindent}{-0.25in}
+\setlength{\leftskip}{0.25in}
+
+\noindent
+
+\hypertarget{refs}{}
+\begin{cslreferences}
+\leavevmode\hypertarget{ref-Lane1985}{}%
+1. Lane DJ, Pace B, Olsen GJ, Stahl DA, Sogin ML, Pace NR. 1985. Rapid
+determination of 16S ribosomal RNA sequences for phylogenetic analyses.
+Proceedings of the National Academy of Sciences 82:6955--6959
+\url{https://doi.org/10.1073/pnas.82.20.6955}.
+
+\leavevmode\hypertarget{ref-Stackebrandt1994}{}%
+2. Stackebrandt E, Goebel BM. 1994. Taxonomic note: A place for DNA-DNA
+reassociation and 16S rRNA sequence analysis in the present species
+definition in bacteriology. International Journal of Systematic and
+Evolutionary Microbiology 44:846--849
+\url{https://doi.org/10.1099/00207713-44-4-846}.
+
+\leavevmode\hypertarget{ref-Goris2007}{}%
+3. Goris J, Konstantinidis KT, Klappenbach JA, Coenye T, Vandamme P,
+Tiedje JM. 2007. DNA-DNA hybridization values and their relationship to
+whole-genome sequence similarities. International Journal of Systematic
+and Evolutionary Microbiology 57:81--91
+\url{https://doi.org/10.1099/ijs.0.64483-0}.
+
+\leavevmode\hypertarget{ref-RodriguezR2018}{}%
+4. Rodriguez-R LM, Castro JC, Kyrpides NC, Cole JR, Tiedje JM,
+Konstantinidis KT. 2018. How much do rRNA gene surveys underestimate
+extant bacterial diversity? Applied and Environmental Microbiology
+84:e00014--18 \url{https://doi.org/10.1128/aem.00014-18}.
+
+\leavevmode\hypertarget{ref-Stackebrandt2006}{}%
+5. Stackebrandt E, Ebers J. 2006. Taxonomic parameters revisited:
+Tarnished gold standards. Microbiol Today 33:152--155.
+
+\leavevmode\hypertarget{ref-Edgar2018}{}%
+6. Edgar RC. 2018. Updating the 97\% identity threshold for 16S
+ribosomal RNA OTUs. Bioinformatics 34:2371--2375
+\url{https://doi.org/10.1093/bioinformatics/bty113}.
+
+\leavevmode\hypertarget{ref-Edgar2016}{}%
+7. Edgar RC. 2016. UNOISE2: Improved error-correction for illumina 16S
+and its amplicon sequencing. bioRxiv
+\url{https://doi.org/10.1101/081257}.
+
+\leavevmode\hypertarget{ref-Amir2017}{}%
+8. Amir A, McDonald D, Navas-Molina JA, Kopylova E, Morton JT, Zech Xu
+Z, Kightley EP, Thompson LR, Hyde ER, Gonzalez A, Knight R. 2017. Deblur
+rapidly resolves single-nucleotide community sequence patterns. mSystems
+2:e00191--16 \url{https://doi.org/10.1128/mSystems.00191-16}.
+
+\leavevmode\hypertarget{ref-Callahan2016}{}%
+9. Callahan BJ, McMurdie PJ, Rosen MJ, Han AW, Johnson AJA, Holmes SP.
+2016. DADA2: High-resolution sample inference from illumina amplicon
+data. Nature Methods 13:581--583
+\url{https://doi.org/10.1038/nmeth.3869}.
+
+\leavevmode\hypertarget{ref-Eren2014}{}%
+10. Eren AM, Morrison HG, Lescault PJ, Reveillaud J, Vineis JH, Sogin
+ML. 2014. Minimum entropy decomposition: Unsupervised oligotyping for
+sensitive partitioning of high-throughput marker gene sequences. The
+ISME Journal 9:968--979 \url{https://doi.org/10.1038/ismej.2014.195}.
+
+\leavevmode\hypertarget{ref-Callahan2017}{}%
+11. Callahan BJ, McMurdie PJ, Holmes SP. 2017. Exact sequence variants
+should replace operational taxonomic units in marker-gene data analysis.
+The ISME Journal 11:2639--2643
+\url{https://doi.org/10.1038/ismej.2017.119}.
+
+\leavevmode\hypertarget{ref-Pei2010}{}%
+12. Pei AY, Oberdorf WE, Nossa CW, Agarwal A, Chokshi P, Gerz EA, Jin Z,
+Lee P, Yang L, Poles M, Brown SM, Sotero S, DeSantis T, Brodie E, Nelson
+K, Pei Z. 2010. Diversity of 16S rRNA genes within individual
+prokaryotic genomes. Applied and Environmental Microbiology
+76:3886--3897 \url{https://doi.org/10.1128/aem.02953-09}.
+
+\leavevmode\hypertarget{ref-Sun2013}{}%
+13. Sun D-L, Jiang X, Wu QL, Zhou N-Y. 2013. Intragenomic heterogeneity
+of 16S rRNA genes causes overestimation of prokaryotic diversity.
+Applied and Environmental Microbiology 79:5962--5969
+\url{https://doi.org/10.1128/aem.01282-13}.
+
+\leavevmode\hypertarget{ref-Stoddard2014}{}%
+14. Stoddard SF, Smith BJ, Hein R, Roller BRK, Schmidt TM. 2014. rrnDB:
+Improved tools for interpreting rRNA gene abundance in bacteria and
+archaea and a new foundation for future development. Nucleic Acids
+Research 43:D593--D598 \url{https://doi.org/10.1093/nar/gku1201}.
+
+\leavevmode\hypertarget{ref-Schloss2011}{}%
+15. Schloss PD, Westcott SL. 2011. Assessing and improving methods used
+in operational taxonomic unit-based approaches for 16S rRNA gene
+sequence analysis. Applied and Environmental Microbiology 77:3219--3226
+\url{https://doi.org/10.1128/aem.02810-10}.
+
+\leavevmode\hypertarget{ref-Johnson2019}{}%
+16. Johnson JS, Spakowicz DJ, Hong B-Y, Petersen LM, Demkowicz P, Chen
+L, Leopold SR, Hanson BM, Agresta HO, Gerstein M, Sodergren E, Weinstock
+GM. 2019. Evaluation of 16S rRNA gene sequencing for species and
+strain-level microbiome analysis. Nature Communications 10:5029
+\url{https://doi.org/10.1038/s41467-019-13036-1}.
+
+\leavevmode\hypertarget{ref-Staley2006}{}%
+17. Staley JT. 2006. The bacterial species dilemma and the
+genomicphylogenetic species concept. Philosophical Transactions of the
+Royal Society B: Biological Sciences 361:1899--1909
+\url{https://doi.org/10.1098/rstb.2006.1914}.
+
+\leavevmode\hypertarget{ref-Oren2013}{}%
+18. Oren A, Garrity GM. 2013. Then and now: A systematic review of the
+systematics of prokaryotes in the last 80~years. Antonie van Leeuwenhoek
+106:43--56 \url{https://doi.org/10.1007/s10482-013-0084-1}.
+
+\leavevmode\hypertarget{ref-Baltrus2016}{}%
+19. Baltrus DA, McCann HC, Guttman DS. 2016. Evolution, genomics and
+epidemiology ofPseudomonas syringae. Molecular Plant Pathology
+18:152--168 \url{https://doi.org/10.1111/mpp.12506}.
+
+\leavevmode\hypertarget{ref-Konstantinidis2005}{}%
+20. Konstantinidis KT, Tiedje JM. 2005. Towards a genome-based taxonomy
+for prokaryotes. Journal of Bacteriology 187:6258--6264
+\url{https://doi.org/10.1128/jb.187.18.6258-6264.2005}.
+
+\leavevmode\hypertarget{ref-Barco2020}{}%
+21. Barco RA, Garrity GM, Scott JJ, Amend JP, Nealson KH, Emerson D.
+2020. A genus definition for bacteria and archaea based on a standard
+genome relatedness index. mBio 11:02475--19
+\url{https://doi.org/10.1128/mbio.02475-19}.
+
+\leavevmode\hypertarget{ref-Parks2020}{}%
+22. Parks DH, Chuvochina M, Chaumeil P-A, Rinke C, Mussig AJ, Hugenholtz
+P. 2020. A complete domain-to-species taxonomy for bacteria and archaea.
+Nature Biotechnology 38:1079--1086
+\url{https://doi.org/10.1038/s41587-020-0501-8}.
+
+\leavevmode\hypertarget{ref-Yarza2014}{}%
+23. Yarza P, Yilmaz P, Pruesse E, Glöckner FO, Ludwig W, Schleifer K-H,
+Whitman WB, Euzéby J, Amann R, Rosselló-Móra R. 2014. Uniting the
+classification of cultured and uncultured bacteria and archaea using 16S
+rRNA gene sequences. Nature Reviews Microbiology 12:635--645
+\url{https://doi.org/10.1038/nrmicro3330}.
+
+\leavevmode\hypertarget{ref-Schloss2009}{}%
+24. Schloss PD, Westcott SL, Ryabin T, Hall JR, Hartmann M, Hollister
+EB, Lesniewski RA, Oakley BB, Parks DH, Robinson CJ, Sahl JW, Stres B,
+Thallinger GG, Horn DJV, Weber CF. 2009. Introducing mothur:
+Open-source, platform-independent, community-supported software for
+describing and comparing microbial communities. Applied and
+Environmental Microbiology 75:7537--7541
+\url{https://doi.org/10.1128/aem.01541-09}.
+
+\leavevmode\hypertarget{ref-Quast2012}{}%
+25. Quast C, Pruesse E, Yilmaz P, Gerken J, Schweer T, Yarza P, Peplies
+J, Glöckner FO. 2012. The SILVA ribosomal RNA gene database project:
+Improved data processing and web-based tools. Nucleic Acids Research
+41:D590--D596 \url{https://doi.org/10.1093/nar/gks1219}.
+\end{cslreferences}
+
+\setlength{\parindent}{0in}
+\setlength{\leftskip}{0in}
+
 \newpage

### 2021-02-17 submission/manuscript.Rmd
Create title, abstract, and importance sections

@@
 	if(is.numeric(x)){
-		if(abs(x - round(x) < .Machine$double.eps)){
+		if(abs(x - round(x)) < .Machine$double.eps){
 			# treat as integer
@@
 
-# Amplicon sequence variants should not replace operational taxonomic units in marker-gene data analysis
+# Amplicon sequence variants artificially split bacterial genomes into separate units of inference
 
@@
 
-**Running title:** ASVs vs. OTUs
+**Running title:** ASVs artificially split bacterial genomes
 
@@
 
-## Abstract (250 words)
-
-## Importance (150 words)
-
-\newpage
-
-16S rRNA gene sequencing is a powerful technique for describing and comparing microbial communities [@Lane1985].
-Efforts to link 16S rRNA gene sequences to taxonomic levels based on distance thresholds go back to at least the 1990s.
-The distance-based thresholds that were developed and are now widely used (3%) were based on DNA-DNA hybridization approaches that are not as precise as genome sequencing [@Stackebrandt1994; @Goris2007].
-Instead, genome sequencing technologies have suggested that the widely used 3% distance threshold to operationally define bacterial taxa is too coarse [@RodriguezR2018; @Stackebrandt2006; @Edgar2018].
-As an alternative to OTUs, amplicon sequencing variants (ASVs) have been proposed as a way to adopt the thresholds suggested by genome sequencing to microbial community analysis using 16S rRNA gene sequences [@Edgar2016; @Amir2017; @Callahan2016; @Eren2014].
-ASVs are a unit of microbial community inference that do not cluster sequences based on a distance-based threshold [@Callahan2017].
-However, most bacterial genomes have more than 1 copy of the rrn operon and those copies are not identical [@Pei2010; @Sun2013].
-Therefore, using too fine a threshold to identify OTUs creates the risk of splitting a single genome into multiple bins and using too broad of a threshold to define OTUs creates the risk of lumping together bacterial species into the same OTU.
-An example of both is seen in the comparison of *Staphylococcus aureus* (NCTC 8325) and *S.
-epidermidis* (ATCC 12228) where each genome has 5 copies of the 16S rRNA gene.
-The 10 copies of the 16S rRNA gene each have a different sequence and so if OTUs are defined based on ASVs, each genome would be split into 5 OTUs.
-Conversely, if the copies were clustered using a 3% distance threshold all 10 copies would cluster into the same OTU.
-The goal of this study was to quantify the risk of splitting a single genome into multiple bins and the risk of lumping together different bacterial species into the same bin.
-
 ```{r}
+## Paragraph 1
 metadata <- read_tsv(here("data/references/genome_id_taxonomy.tsv"),
@@
 ecoli_nasvs <- asvs_per_species %>% filter(species == "Escherichia coli") %>% pull(n_asvs)
+
+## Paragraph 2
+thresholds <- read_tsv(here("data/processed/thresholds_for_single_otu.tsv"),
+											 col_types=cols(region = col_character(),
+											 							 .default = col_double())) %>%
+								filter(n_genomes > 100)
+
+threshold_v19 <- thresholds %>% filter(n_rrns == 7 & region == "v19") %>% pull(threshold) * 100
+threshold_v4 <- thresholds %>% filter(n_rrns == 7 & region == "v4") %>% pull(threshold) * 100
+threshold_v34 <- thresholds %>% filter(n_rrns == 7 & region == "v34") %>% pull(threshold) * 100
+threshold_v45 <- thresholds %>% filter(n_rrns == 7 & region == "v45") %>% pull(threshold) * 100
+
+three_v19 <- thresholds %>% filter(threshold <= 0.03 & region == "v19") %>% tail(n=1) %>% pull(n_rrns)
+three_v4 <- thresholds %>% filter(threshold <= 0.03 & region == "v4") %>% tail(n=1) %>% pull(n_rrns)
+three_v34 <- thresholds %>% filter(threshold <= 0.03 & region == "v34") %>% tail(n=1) %>% pull(n_rrns)
+three_v45 <- thresholds %>% filter(threshold <= 0.03 & region == "v45") %>% tail(n=1) %>% pull(n_rrns)
+
+## Paragraph 3
+lump_split <- read_tsv(here("data/processed/lumped_split_rate.tsv"),
+											 col_types = cols(region = col_character(),
+											 								 .default = col_double()))
+
+lump_v19 <- lump_split %>% filter(threshold ==0 &region == "v19") %>% pull(lump_rate) * 100
+lump_v4 <- lump_split %>% filter(threshold ==0 &region == "v4") %>% pull(lump_rate) * 100
+lump_v34 <- lump_split %>% filter(threshold ==0 &region == "v34") %>% pull(lump_rate) * 100
+lump_v45 <- lump_split %>% filter(threshold ==0 &region == "v45") %>% pull(lump_rate) * 100
+
+lump3_v19 <- lump_split %>% filter(threshold ==0.03 & region == "v19") %>% pull(lump_rate) * 100
+lump3_v4 <- lump_split %>% filter(threshold ==0.03 & region == "v4") %>% pull(lump_rate) * 100
+lump3_v34 <- lump_split %>% filter(threshold ==0.03 & region == "v34") %>% pull(lump_rate) * 100
+lump3_v45 <- lump_split %>% filter(threshold ==0.03 & region == "v45") %>% pull(lump_rate) * 100
 ```
+
+## Abstract
+
+Amplicon sequencing variants (ASVs) have been proposed as an alternative to operational taxonomic units (OTUs) for analyzing microbiomes.
+ASVs have grown in popularity, in part, because of a desire to reflect a more refined level of taxonomy because they do not cluster sequences based on a distance-based threshold.
+However, ASVs and the use of overly narrow thresholds to identify OTUs increases the risk of splitting a single genome into separate clusters.
+I addressed this problem by analyzing the intragenomic variation of 16S rRNA genes from the bacterial genomes represented in the rrn copy number database, which contained `r nrow(metadata)` genomes from `r n_distinct(metadata$species)` species.
+The analysis confirmed the severity of this risk.
+As the number of copies of the 16S rRNA gene increased in a genome, the number of ASVs also increased.
+For full length 16S rRNA genes, there was an average of `r rate_v19` ASVs per copy of the 16S rRNA gene.
+Among genomes with 7 copies of the 16S rRNA, such as *E.
+coli*, it was necessary to use a distance threshold of `r format_pct(threshold_v19)`% to cluster the ASVs from the same genome into a single OTU.
+This research highlights the risk of splitting a single bacterial genome into separate clusters when ASVs are used to analyze 16S rRNA gene sequence data.
+Although there is also a risk of clustering different species into the same OTU, the effects of those risks on biological inferences are less than those from artificially splitting a genome into separate ASVs and OTUs.
+
+## Importance
+
+16S rRNA gene sequencing has propelled significant interest into host-associated and environmental microbiomes.
+There is a tension between trying to classify 16S rRNA gene sequences to increasingly lower taxonomic levels and the reality that those levels were defined using more sequence and physiological information than is available from a fragment of the 16S rRNA gene.
+Furthermore, naming of bacterial taxa reflects the biases of those who name them.
+One motivation for the recent push to adopt ASVs in place of OTUs in microbiome analyses is to allow researchers to perform their analyes at the finest possible level that reflects species-level taxonomy.
+The current research is significant because it quantifies the risk of artificially splitting bacterial genomes into separate clusters.
+Far from providing a better represenation of microbiomes, the ASV approach could lead to conflicting inferences about the ecology of different clusters from the same genome.
+
+\newpage
+
+16S rRNA gene sequencing is a powerful technique for describing and comparing microbial communities [@Lane1985].
+Efforts to link 16S rRNA gene sequences to taxonomic levels based on distance thresholds go back to at least the 1990s.
+The distance-based thresholds that were developed and are now widely used (3%) were based on DNA-DNA hybridization approaches that are not as precise as genome sequencing [@Stackebrandt1994; @Goris2007].
+Instead, genome sequencing technologies have suggested that the widely used 3% distance threshold to operationally define bacterial taxa is too coarse [@RodriguezR2018; @Stackebrandt2006; @Edgar2018].
+As an alternative to OTUs, amplicon sequencing variants (ASVs) have been proposed as a way to adopt the thresholds suggested by genome sequencing to microbial community analysis using 16S rRNA gene sequences [@Edgar2016; @Amir2017; @Callahan2016; @Eren2014].
+ASVs are a unit of microbial community inference that do not cluster sequences based on a distance-based threshold [@Callahan2017].
+However, most bacterial genomes have more than 1 copy of the rrn operon and those copies are not identical [@Pei2010; @Sun2013].
+Therefore, using too fine a threshold to identify OTUs creates the risk of splitting a single genome into multiple bins and using too broad of a threshold to define OTUs creates the risk of lumping together bacterial species into the same OTU.
+An example of both is seen in the comparison of *Staphylococcus aureus* (NCTC 8325) and *S.
+epidermidis* (ATCC 12228) where each genome has 5 copies of the 16S rRNA gene.
+The 10 copies of the 16S rRNA gene each have a different sequence and so if OTUs are defined based on ASVs, each genome would be split into 5 OTUs.
+Conversely, if the copies were clustered using a 3% distance threshold all 10 copies would cluster into the same OTU.
+The goal of this study was to quantify the risk of splitting a single genome into multiple bins and the risk of lumping together different bacterial species into the same bin.
 
@@
 
-```{r}
-thresholds <- read_tsv(here("data/processed/thresholds_for_single_otu.tsv"),
-											 col_types=cols(region = col_character(),
-											 							 .default = col_double())) %>%
-								filter(n_genomes > 100)
-
-threshold_v19 <- thresholds %>% filter(n_rrns == 7 & region == "v19") %>% pull(threshold) * 100
-threshold_v4 <- thresholds %>% filter(n_rrns == 7 & region == "v4") %>% pull(threshold) * 100
-threshold_v34 <- thresholds %>% filter(n_rrns == 7 & region == "v34") %>% pull(threshold) * 100
-threshold_v45 <- thresholds %>% filter(n_rrns == 7 & region == "v45") %>% pull(threshold) * 100
-
-three_v19 <- thresholds %>% filter(threshold <= 0.03 & region == "v19") %>% tail(n=1) %>% pull(n_rrns)
-three_v4 <- thresholds %>% filter(threshold <= 0.03 & region == "v4") %>% tail(n=1) %>% pull(n_rrns)
-three_v34 <- thresholds %>% filter(threshold <= 0.03 & region == "v34") %>% tail(n=1) %>% pull(n_rrns)
-three_v45 <- thresholds %>% filter(threshold <= 0.03 & region == "v45") %>% tail(n=1) %>% pull(n_rrns)
-```
-
 A method to avoid splitting a single genome into multiple units of inference is to cluster together similar 16S rRNA gene sequences.
@@
 Consequently, these results demonstrate that broad thresholds must be used to avoid splitting different operons from the same genome into separate clusters.
-
-```{r}
-lump_split <- read_tsv(here("data/processed/lumped_split_rate.tsv"),
-											 col_types = cols(region = col_character(),
-											 								 .default = col_double()))
-
-lump_v19 <- lump_split %>% filter(threshold ==0 &region == "v19") %>% pull(lump_rate) * 100
-lump_v4 <- lump_split %>% filter(threshold ==0 &region == "v4") %>% pull(lump_rate) * 100
-lump_v34 <- lump_split %>% filter(threshold ==0 &region == "v34") %>% pull(lump_rate) * 100
-lump_v45 <- lump_split %>% filter(threshold ==0 &region == "v45") %>% pull(lump_rate) * 100
-
-lump3_v19 <- lump_split %>% filter(threshold ==0.03 & region == "v19") %>% pull(lump_rate) * 100
-lump3_v4 <- lump_split %>% filter(threshold ==0.03 & region == "v4") %>% pull(lump_rate) * 100
-lump3_v34 <- lump_split %>% filter(threshold ==0.03 & region == "v34") %>% pull(lump_rate) * 100
-lump3_v45 <- lump_split %>% filter(threshold ==0.03 & region == "v45") %>% pull(lump_rate) * 100
-```
 

### 2021-02-17 submission/manuscript.tex
Create title, abstract, and importance sections

@@
 
-\hypertarget{amplicon-sequence-variants-should-not-replace-operational-taxonomic-units-in-marker-gene-data-analysis}{%
-\section{Amplicon sequence variants should not replace operational
-taxonomic units in marker-gene data
-analysis}\label{amplicon-sequence-variants-should-not-replace-operational-taxonomic-units-in-marker-gene-data-analysis}}
+\hypertarget{amplicon-sequence-variants-artificially-split-bacterial-genomes-into-separate-units-of-inference}{%
+\section{Amplicon sequence variants artificially split bacterial genomes
+into separate units of
+inference}\label{amplicon-sequence-variants-artificially-split-bacterial-genomes-into-separate-units-of-inference}}
 
@@
 
-\textbf{Running title:} ASVs vs.~OTUs
+\textbf{Running title:} ASVs artificially split bacterial genomes
 
@@
 
-\hypertarget{abstract-250-words}{%
-\subsection{Abstract (250 words)}\label{abstract-250-words}}
+\hypertarget{abstract}{%
+\subsection{Abstract}\label{abstract}}
 
-\hypertarget{importance-150-words}{%
-\subsection{Importance (150 words)}\label{importance-150-words}}
+Amplicon sequencing variants (ASVs) have been proposed as an alternative
+to operational taxonomic units (OTUs) for analyzing microbiomes. ASVs
+have grown in popularity, in part, because of a desire to reflect a more
+refined level of taxonomy because they do not cluster sequences based on
+a distance-based threshold. However, ASVs and the use of overly narrow
+thresholds to identify OTUs increases the risk of splitting a single
+genome into separate clusters. I addressed this problem by analyzing the
+intragenomic variation of 16S rRNA genes from the bacterial genomes
+represented in the rrn copy number database, which contained
+\textbf{15,614} genomes from \textbf{4,774} species. The analysis
+confirmed the severity of this risk. As the number of copies of the 16S
+rRNA gene increased in a genome, the number of ASVs also increased. For
+full length 16S rRNA genes, there was an average of \textbf{0.60} ASVs
+per copy of the 16S rRNA gene. Among genomes with 7 copies of the 16S
+rRNA, such as \emph{E. coli}, it was necessary to use a distance
+threshold of \textbf{5.5}\% to cluster the ASVs from the same genome
+into a single OTU. This research highlights the risk of splitting a
+single bacterial genome into separate clusters when ASVs are used to
+analyze 16S rRNA gene sequence data. Although there is also a risk of
+clustering different species into the same OTU, the effects of those
+risks on biological inferences are less than those from artificially
+splitting a genome into separate ASVs and OTUs.
+
+\hypertarget{importance}{%
+\subsection{Importance}\label{importance}}
+
+16S rRNA gene sequencing has propelled significant interest into
+host-associated and environmental microbiomes. There is a tension
+between trying to classify 16S rRNA gene sequences to increasingly lower
+taxonomic levels and the reality that those levels were defined using
+more sequence and physiological information than is available from a
+fragment of the 16S rRNA gene. Furthermore, naming of bacterial taxa
+reflects the biases of those who name them. One motivation for the
+recent push to adopt ASVs in place of OTUs in microbiome analyses is to
+allow researchers to perform their analyes at the finest possible level
+that reflects species-level taxonomy. The current research is
+significant because it quantifies the risk of artificially splitting
+bacterial genomes into separate clusters. Far from providing a better
+represenation of microbiomes, the ASV approach could lead to conflicting
+inferences about the ecology of different clusters from the same genome.
 
@@
 variants of the 16S rRNA gene in each genome also increased. On average,
-there were \textbf{1} variants per copy of the full length 16S rRNA gene
-and an average of \textbf{0.26}, \textbf{0.33}, and \textbf{0.27}
+there were \textbf{0.60} variants per copy of the full length 16S rRNA
+gene and an average of \textbf{0.26}, \textbf{0.33}, and \textbf{0.27}
 variants when considering the V4, V3-V4, and V4-V5 regions of the gene,

### 2021-02-19 submission/manuscript.Rmd
First round of revisions

@@
 
-# Amplicon sequence variants artificially split bacterial genomes into separate units of inference
+# Amplicon sequence variants artificially split bacterial genomes into separate clusters
 
@@
 However, ASVs and the use of overly narrow thresholds to identify OTUs increases the risk of splitting a single genome into separate clusters.
-I addressed this problem by analyzing the intragenomic variation of 16S rRNA genes from the bacterial genomes represented in the rrn copy number database, which contained `r nrow(metadata)` genomes from `r n_distinct(metadata$species)` species.
-The analysis confirmed the severity of this risk.
+I assessed this problem by analyzing the intragenomic variation of 16S rRNA genes from the bacterial genomes represented in a *rrn* copy number database, which contained `r nrow(metadata)` genomes from `r n_distinct(metadata$species)` species.
 As the number of copies of the 16S rRNA gene increased in a genome, the number of ASVs also increased.
-For full length 16S rRNA genes, there was an average of `r rate_v19` ASVs per copy of the 16S rRNA gene.
-Among genomes with 7 copies of the 16S rRNA, such as *E.
-coli*, it was necessary to use a distance threshold of `r format_pct(threshold_v19)`% to cluster the ASVs from the same genome into a single OTU.
+There was an average of `r rate_v19` ASVs per copy of the 16S rRNA gene, for full length 16S rRNA genes.
+It was necessary to use a distance threshold of `r format_pct(threshold_v19)`% to cluster the ASVs from the same genome into a single OTU, for genomes with 7 copies of the 16S rRNA, such as *E.
+coli*.
 This research highlights the risk of splitting a single bacterial genome into separate clusters when ASVs are used to analyze 16S rRNA gene sequence data.
@@
 Efforts to link 16S rRNA gene sequences to taxonomic levels based on distance thresholds go back to at least the 1990s.
-The distance-based thresholds that were developed and are now widely used (3%) were based on DNA-DNA hybridization approaches that are not as precise as genome sequencing [@Stackebrandt1994; @Goris2007].
+The distance-based thresholds that were developed and are now widely used (i.e., 3%) were based on DNA-DNA hybridization approaches that are not as precise as genome sequencing [@Stackebrandt1994; @Goris2007].
 Instead, genome sequencing technologies have suggested that the widely used 3% distance threshold to operationally define bacterial taxa is too coarse [@RodriguezR2018; @Stackebrandt2006; @Edgar2018].
@@
 ASVs are a unit of microbial community inference that do not cluster sequences based on a distance-based threshold [@Callahan2017].
-However, most bacterial genomes have more than 1 copy of the rrn operon and those copies are not identical [@Pei2010; @Sun2013].
+However, most bacterial genomes have more than 1 copy of the *rrn* operon and those copies are not identical [@Pei2010; @Sun2013].
 Therefore, using too fine a threshold to identify OTUs creates the risk of splitting a single genome into multiple bins and using too broad of a threshold to define OTUs creates the risk of lumping together bacterial species into the same OTU.
@@
 epidermidis* (ATCC 12228) where each genome has 5 copies of the 16S rRNA gene.
-The 10 copies of the 16S rRNA gene each have a different sequence and so if OTUs are defined based on ASVs, each genome would be split into 5 OTUs.
-Conversely, if the copies were clustered using a 3% distance threshold all 10 copies would cluster into the same OTU.
-The goal of this study was to quantify the risk of splitting a single genome into multiple bins and the risk of lumping together different bacterial species into the same bin.
+Each of the 10 copies of the 16S rRNA gene in these two genomes were distinct and would yield 10 ASVs.
+Conversely, if the copies were clustered using a 3% distance threshold, then all 10 ASVs would cluster into the same OTU.
+The goal of this study was to quantify the risk of splitting a single genome into multiple clusters and the risk of lumping together different bacterial species into the same cluster.
 
-To investigate the variation in the number of copies of the 16S rRNA gene per genome as well as the intragenomic variation among copies of the 16S rRNA gene, I obtained reference 16S rRNA sequences from the rrn copy number database (rrnDB)[@Stoddard2014].
-Among the `r n_distinct(metadata$species)` species represented in the rrnDB there were `r nrow(metadata)` genomes.
-The median number of rrn operson per species ranged between `r single_copy$median` (e.g., *`r single_copy$species`*) and `r max_copy$median` (*`r max_copy$species`*) copies of the rrn operon.
+To investigate the variation in the number of copies of the 16S rRNA gene per genome as well as the intragenomic variation among copies of the 16S rRNA gene, I obtained reference 16S rRNA sequences from the *rrn* copy number database (*rrn*DB)[@Stoddard2014].
+Among the `r n_distinct(metadata$species)` species represented in the *rrn*DB there were `r nrow(metadata)` genomes.
+The median number of *rrn* operons per species ranged between `r single_copy$median` (e.g., *`r single_copy$species`*) and `r max_copy$median` (*`r max_copy$species`*) copies of the *rrn* operon.
 As the number of copies of the operon in a genome increased, the number of variants of the 16S rRNA gene in each genome also increased.
@@
 Although a species tended to have a consistent number of 16S rRNA gene copies per genome, the number of total variants increased with the number of genomes that were sampled (**Figure 1**).
-For example, *`r single_copy$species`* generally only had `r single_copy$median` copy of the gene per genome, but across the `r single_copy_N` genomes that have been sequenced there were `r single_nasvs` versions of the gene.
+For example, *`r single_copy$species`* generally only had `r single_copy$median` copy of the gene per genome, but across the `r single_copy_N` genomes in the *rrn*DB there were `r single_nasvs` versions of the gene.
 Similarly, a *E.
@@
 Across the `r ecoli_copy$N` *E.
-coli* genomes that have been sequenced, there were `r ecoli_nasvs` different variants of the gene.
-These observations highlight the risk of selecting a threshold for defining units of inference that is too narrow because it is possible to split a single genome into multiple units.
+coli* genomes in the *rrn*DB, there were `r ecoli_nasvs` different variants of the gene.
+These observations highlight the risk of selecting a threshold for defining clusters that is too narrow because it is possible to split a single genome into multiple clusters.
 
-A method to avoid splitting a single genome into multiple units of inference is to cluster together similar 16S rRNA gene sequences.
+A method to avoid splitting a single genome into multiple clusters is to cluster 16S rRNA gene sequences together based on their similarity to each other.
 Therefore, I assessed the impact of the distance threshold used to define clusters of 16S rRNA genes on the propensity to split a genome into separate clusters.
@@
 Consequently, these results demonstrate that broad thresholds must be used to avoid splitting different operons from the same genome into separate clusters.
-
 At broad thresholds multiple species could be represented by the same OTU (**Figure 2**).
 Using ASVs, `r format_pct(lump_v19)`% of the species shared a 16S rRNA gene sequence variant with another species when considering full length sequences and `r format_pct(lump_v4)`, `r format_pct(lump_v34)`, and `r format_pct(lump_v45)`% when considering the V4, V3-V4, and V4-V5 regions, respectively.
-At the commonly used 3% threshold, `r format_pct(lump3_v19)`% of the species shared an OTU when considering full length sequences and `r format_pct(lump3_v4)`, `r format_pct(lump3_v34)`, and `r format_pct(lump3_v45)`% when considering the V4, V3-V4, and V4-V5 regions, respectively.
-Considering that species designations are unevenly applied and reflect multiple biases, the risk of splitting a genome into multiple OTUs more problematic than clustering species together.
+At the commonly used 3% threshold, `r format_pct(lump3_v19)`% of the species shared an OTU with another speecies when considering full length sequences and `r format_pct(lump3_v4)`, `r format_pct(lump3_v34)`, and `r format_pct(lump3_v45)`% when considering the V4, V3-V4, and V4-V5 regions, respectively.
+Considering that species designations are unevenly applied and reflect multiple human-imposed biases, the risk of splitting a genome into multiple OTUs more problematic than clustering species together.
 Therefore, larger thresholds are advisable.
 
-The results of this analysis demonstrate that there is a significant risk of splitting single genomes into multiple bins if too fine of a threshold is applied to defining an OTU.
-An ongoing problem for amplicon-based studies is defining a meaningful taxonomic unit of inference [@Callahan2017; @Schloss2011; @Johnson2019].
+The results of this analysis demonstrate that there is a significant risk of splitting a single genome into multiple clusters if too fine of a threshold is applied to defining an OTU.
+An ongoing problem for amplicon-based studies is defining a meaningful taxonomic unit [@Callahan2017; @Schloss2011; @Johnson2019].
 Since there is no consensus definition for a biological species concept [@Staley2006; @Oren2013], microbiologists must accept that how we have named bacterial species is biased and that taxonomic rules are not applied in a consistent manner (e.g., [@Baltrus2016]).
-This makes it more challenging to attempt to fit a distance threshold to define an OTU definition that matches a set of species names [@Konstantinidis2005].
+This makes it impossible to fit a distance threshold to define an OTU definition that matches a set of species names [@Konstantinidis2005].
 Furthermore, the 16S rRNA gene does not evolve at the same rate across all bacterial lineages [@Schloss2011], which limits the biological interpretation of a common OTU definition.
 A distance-based definition of a taxonomic unit based on 16S rRNA gene or full genome sequences is, at best, operational and not grounded in biological theory [@Schloss2011; @Barco2020; @Parks2020; @Yarza2014].
-There is general agreement in bacterial systematics that to classify something to a bacterial species, you need phenotypic and genome sequence data [@Staley2006; @Oren2013; @Baltrus2016].
-We are asking too much of a short section of a bacterial genome to be able to differentiate between species.
-It is difficult to defend a unit of inference that would split a single genome into multiple taxonomic units.
+There is general agreement in bacterial systematics that to classify something to a bacterial species, phenotypic and genome sequence data are needed [@Staley2006; @Oren2013; @Baltrus2016].
+A short section of a bacterial genome simply cannot differentiate between species.
+Moreover, it is difficult to defend a clustering approach that would split a single genome into multiple taxonomic units.
 It is not biologically plausible to entertain the possibility that parts of a genome would have different ecologies.
-Although there are multiple reasons that proponents of ASVs encourage their use, the significant risk of splitting genomes is too high to warrant their use.
+Although there are multiple reasons that proponents favor ASVs, the significant risk of artificially splitting genomes into separate clusters is too high to warrant their use.
 
@@
 (i) Data availability.** The 16S rRNA gene sequences used in this study were obtained from the *rrn*DB (https://rrndb.umms.med.umich.edu; version 5.6, released November 8, 2019) [@Stoddard2014].
-At the time of submission, this is the most current version of the database.
+At the time of submission, this was the most current version of the database.
 The *rrn*DB obtained the curated 16S rRNA gene sequences from the KEGG database, which ultimately obtained them from NCBI's non-redundant RefSeq database.
-The *rrn*DB provides downloadable versions of the sequences with their taxonomy as determined using the naive Bayesian classifier trained on the RDP reference taxonomy.
+The *rrn*DB provided downloadable versions of the sequences with their taxonomy as determined using the naive Bayesian classifier trained on the RDP reference taxonomy.
 For some genomes this resulted in multiple classifications since a genome's 16S rRNA gene sequences were not identical.
@@
 
-**(ii) Definition of regions within 16S rRNA gene.** The full length 16S rRNA gene sequences were aligned to a SILVA reference alignment of the 16S rRNA gene (v138) using the mothur software package (v.
-1.XX) [@Schloss2009; @Quast2012].
+**(ii) Definition of regions within 16S rRNA gene.** The full length 16S rRNA gene sequences were aligned to a SILVA reference alignment of the 16S rRNA gene (v.
+138) using the mothur software package (v.
+1.44.2) [@Schloss2009; @Quast2012].
 Regions of the 16S rRNA gene were selected because of their use in the microbial ecology literature.
 Full length sequences corresponded to *E.
-coli* positions XX through XXXX, V4 to positions XXX through XXX, V3-V4 to positions XXX through XXX, and V4-V5 to positions XXX through XXX.
+coli* str.
+K-12 substr.
+MG1655 (NC_000913) positions 28 through 1491, V4 to positions 534 through 786, V3-V4 to positions 358 through 786, and V4-V5 to positions 534 through 908.
 
@@
 The random selection was repeated 100 times.
-Analyses based on this randomization report the median of the 100 randomizations.
-The intraquartile range between randomizations was typically less than XXXX.
-Because it was so small, confidence intervals are not included in Figure 2.
+Analyses based on this randomization reported the median of the 100 randomizations.
+The intraquartile range between randomizations was less than XXXX.
+Because the range was so small, the confidence intervals were smaller than the thickness of the lines in Figures 1 and 2 and were not included.
 
-**(iv) Reproducible data analysis.** The code to perform the analysis in this manuscript and its hisotry are available as a git-based version control repository on GitHub (https://github.com/pschloss/Schloss_rrnAnalysis_XXXX_2020).
+**(iv) Reproducible data analysis.** The code to perform the analysis in this manuscript and its history are available as a git-based version control repository on GitHub (https://github.com/SchlossLab/Schloss_rrnAnalysis_mSphere_2021).
 The analysis can be regenerated using a GNU Make-based workflow that made use of built-in bash tools (v.
 3.2.57), mothur (v.
-1.XX), and R (v.
-4.X.X).
+1.44.2), and R (v.
+`r paste(R.version$major, R.version$minor, sep=".")`).
 Within R, I used the tidyverse (v.
-4.X.X), data.table (v.
-4.X.X), Rcpp (v.
-4.X.X), furrr (v.
-4.X.X), and rmarkdown (v.
-4.X.X) packages.
-The conception and development of this analysis is available as a playlist on the Riffomonas YouTube channel (https://www.youtube.com/playlist?list=PLmNrK_nkqBpKY3SZiivlIGvcLX-KHmfR8).
+`r packageVersion("tidyverse")`), data.table (v.
+`r packageVersion("data.table")`), Rcpp (v.
+packageVersion("Rcpp")), furrr (v.
+packageVersion("furrr")), and rmarkdown (v.
+packageVersion("rmarkdown")) packages.
+The conception and development of this analysis is available as a playlist on the Riffomonas YouTube channel (https://youtube.com/playlist?list=PLmNrK_nkqBpL7m_tyWdQgdyurerttCsPY).
 
-**Acknowledgements.** I am grateful to Robert Hein and Thomas Schmidt who maintain the rrnDB for their help in understanding the curation of the database and for making the 16S rRNA gene sequences and related metadata publicly available.
-I am also grateful to community members who watched the serialized version of this analysis on YouTube and provided their suggestions and questions.
+**Acknowledgements.** I am grateful to Robert Hein and Thomas Schmidt, who maintain the *rrn*DB, for their help in understanding the curation of the database and for making the 16S rRNA gene sequences and related metadata publicly available.
+I am also grateful to community members who watched the serialized version of this analysis on YouTube and provided suggestions and questions over the course of the development of this project.
 
@@
 **Figure 1.
-The distance threshold required to prevent the splitting of genomes into multiple OTUs increases as the number of *rrn* operons in the genome increases.** Each line represents the median distance threshold for each region of the 16S rRNA gene that is required for 95% of the species with the indicated numbrer of *rrn* operons to cluster their ASVs to a single OTU.
+The distance threshold required to prevent the splitting of genomes into multiple OTUs increased as the number of *rrn* operons in the genome increased.** Each line represents the median distance threshold for each region of the 16S rRNA gene that is required for 95% of the species with the indicated numbrer of *rrn* operons to cluster their ASVs to a single OTU.
 The median distance threshold was calculated across 100 randomizations in which one genome was sampled from each species.
@@
 **Figure 2.
-As the distance threshold used to define an OTU increases, the fraction of genomes split into separate OTUs decreases while the fraction of species that are merged into the same OTU increases.** These data represent the median fractions for both measurements across 100 randomizations.
+As the distance threshold used to define an OTU increased, the fraction of genomes split into separate OTUs decreased while the fraction of species that were merged into the same OTU increases.** These data represent the median fractions for both measurements across 100 randomizations.
 In each randomization, one genome was sampled from each species.
@@
 **Figure S1.
-The ratio of number of distinct ASVs per copy of the *rrn* operon increases for a species as the number of genomes sampled increases.** Each point represents a different species and is shaded to be 80% transparent so that when points overlap they become darker.
+The ratio of number of distinct ASVs per copy of the *rrn* operon increased for a species as the number of genomes sampled increased.** Each point represents a different species and was shaded to be 80% transparent so that when points overlap they become darker.
 The blue line represents a smoothed fit through the data.

### 2021-02-19 submission/manuscript.tex
First round of revisions

@@
 
-\hypertarget{amplicon-sequence-variants-artificially-split-bacterial-genomes-into-separate-units-of-inference}{%
+\hypertarget{amplicon-sequence-variants-artificially-split-bacterial-genomes-into-separate-clusters}{%
 \section{Amplicon sequence variants artificially split bacterial genomes
-into separate units of
-inference}\label{amplicon-sequence-variants-artificially-split-bacterial-genomes-into-separate-units-of-inference}}
+into separate
+clusters}\label{amplicon-sequence-variants-artificially-split-bacterial-genomes-into-separate-clusters}}
 
@@
 thresholds to identify OTUs increases the risk of splitting a single
-genome into separate clusters. I addressed this problem by analyzing the
+genome into separate clusters. I assessed this problem by analyzing the
 intragenomic variation of 16S rRNA genes from the bacterial genomes
-represented in the rrn copy number database, which contained
-\textbf{15,614} genomes from \textbf{4,774} species. The analysis
-confirmed the severity of this risk. As the number of copies of the 16S
-rRNA gene increased in a genome, the number of ASVs also increased. For
-full length 16S rRNA genes, there was an average of \textbf{0.60} ASVs
-per copy of the 16S rRNA gene. Among genomes with 7 copies of the 16S
-rRNA, such as \emph{E. coli}, it was necessary to use a distance
-threshold of \textbf{5.5}\% to cluster the ASVs from the same genome
-into a single OTU. This research highlights the risk of splitting a
-single bacterial genome into separate clusters when ASVs are used to
-analyze 16S rRNA gene sequence data. Although there is also a risk of
-clustering different species into the same OTU, the effects of those
-risks on biological inferences are less than those from artificially
-splitting a genome into separate ASVs and OTUs.
+represented in a \emph{rrn} copy number database, which contained
+\textbf{15,614} genomes from \textbf{4,774} species. As the number of
+copies of the 16S rRNA gene increased in a genome, the number of ASVs
+also increased. There was an average of \textbf{0.60} ASVs per copy of
+the 16S rRNA gene, for full length 16S rRNA genes. It was necessary to
+use a distance threshold of \textbf{5.5}\% to cluster the ASVs from the
+same genome into a single OTU, for genomes with 7 copies of the 16S
+rRNA, such as \emph{E. coli}. This research highlights the risk of
+splitting a single bacterial genome into separate clusters when ASVs are
+used to analyze 16S rRNA gene sequence data. Although there is also a
+risk of clustering different species into the same OTU, the effects of
+those risks on biological inferences are less than those from
+artificially splitting a genome into separate ASVs and OTUs.
 
@@
 least the 1990s. The distance-based thresholds that were developed and
-are now widely used (3\%) were based on DNA-DNA hybridization approaches
-that are not as precise as genome sequencing (2, 3). Instead, genome
-sequencing technologies have suggested that the widely used 3\% distance
-threshold to operationally define bacterial taxa is too coarse (4--6).
-As an alternative to OTUs, amplicon sequencing variants (ASVs) have been
-proposed as a way to adopt the thresholds suggested by genome sequencing
-to microbial community analysis using 16S rRNA gene sequences (7--10).
-ASVs are a unit of microbial community inference that do not cluster
-sequences based on a distance-based threshold (11). However, most
-bacterial genomes have more than 1 copy of the rrn operon and those
-copies are not identical (12, 13). Therefore, using too fine a threshold
-to identify OTUs creates the risk of splitting a single genome into
-multiple bins and using too broad of a threshold to define OTUs creates
-the risk of lumping together bacterial species into the same OTU. An
-example of both is seen in the comparison of \emph{Staphylococcus
-aureus} (NCTC 8325) and \emph{S. epidermidis} (ATCC 12228) where each
-genome has 5 copies of the 16S rRNA gene. The 10 copies of the 16S rRNA
-gene each have a different sequence and so if OTUs are defined based on
-ASVs, each genome would be split into 5 OTUs. Conversely, if the copies
-were clustered using a 3\% distance threshold all 10 copies would
-cluster into the same OTU. The goal of this study was to quantify the
-risk of splitting a single genome into multiple bins and the risk of
-lumping together different bacterial species into the same bin.
+are now widely used (i.e., 3\%) were based on DNA-DNA hybridization
+approaches that are not as precise as genome sequencing (2, 3). Instead,
+genome sequencing technologies have suggested that the widely used 3\%
+distance threshold to operationally define bacterial taxa is too coarse
+(4--6). As an alternative to OTUs, amplicon sequencing variants (ASVs)
+have been proposed as a way to adopt the thresholds suggested by genome
+sequencing to microbial community analysis using 16S rRNA gene sequences
+(7--10). ASVs are a unit of microbial community inference that do not
+cluster sequences based on a distance-based threshold (11). However,
+most bacterial genomes have more than 1 copy of the \emph{rrn} operon
+and those copies are not identical (12, 13). Therefore, using too fine a
+threshold to identify OTUs creates the risk of splitting a single genome
+into multiple bins and using too broad of a threshold to define OTUs
+creates the risk of lumping together bacterial species into the same
+OTU. An example of both is seen in the comparison of
+\emph{Staphylococcus aureus} (NCTC 8325) and \emph{S. epidermidis} (ATCC
+12228) where each genome has 5 copies of the 16S rRNA gene. Each of the
+10 copies of the 16S rRNA gene in these two genomes were distinct and
+would yield 10 ASVs. Conversely, if the copies were clustered using a
+3\% distance threshold, then all 10 ASVs would cluster into the same
+OTU. The goal of this study was to quantify the risk of splitting a
+single genome into multiple clusters and the risk of lumping together
+different bacterial species into the same cluster.
 
@@
 gene per genome as well as the intragenomic variation among copies of
-the 16S rRNA gene, I obtained reference 16S rRNA sequences from the rrn
-copy number database (rrnDB)(14). Among the \textbf{4,774} species
-represented in the rrnDB there were \textbf{15,614} genomes. The median
-number of rrn operson per species ranged between \textbf{1} (e.g.,
-\textbf{\emph{Mycobacterium tuberculosis}}) and \textbf{19}
-(\textbf{\emph{Metabacillus litoralis}}) copies of the rrn operon. As
-the number of copies of the operon in a genome increased, the number of
-variants of the 16S rRNA gene in each genome also increased. On average,
-there were \textbf{0.60} variants per copy of the full length 16S rRNA
-gene and an average of \textbf{0.26}, \textbf{0.33}, and \textbf{0.27}
-variants when considering the V4, V3-V4, and V4-V5 regions of the gene,
-respectively. Although a species tended to have a consistent number of
-16S rRNA gene copies per genome, the number of total variants increased
-with the number of genomes that were sampled (\textbf{Figure 1}). For
-example, \textbf{\emph{Mycobacterium tuberculosis}} generally only had
-\textbf{1} copy of the gene per genome, but across the \textbf{180}
-genomes that have been sequenced there were \textbf{11} versions of the
-gene. Similarly, a \emph{E. coli} genome typically had \textbf{7} copies
-of the 16S rRNA gene with between \textbf{6} and \textbf{10} distinct
-full length sequences per genome. Across the \textbf{958} \emph{E. coli}
-genomes that have been sequenced, there were \textbf{1,013} different
-variants of the gene. These observations highlight the risk of selecting
-a threshold for defining units of inference that is too narrow because
-it is possible to split a single genome into multiple units.
+the 16S rRNA gene, I obtained reference 16S rRNA sequences from the
+\emph{rrn} copy number database (\emph{rrn}DB)(14). Among the
+\textbf{4,774} species represented in the \emph{rrn}DB there were
+\textbf{15,614} genomes. The median number of \emph{rrn} operons per
+species ranged between \textbf{1} (e.g., \textbf{\emph{Mycobacterium
+tuberculosis}}) and \textbf{19} (\textbf{\emph{Metabacillus litoralis}})
+copies of the \emph{rrn} operon. As the number of copies of the operon
+in a genome increased, the number of variants of the 16S rRNA gene in
+each genome also increased. On average, there were \textbf{0.60}
+variants per copy of the full length 16S rRNA gene and an average of
+\textbf{0.26}, \textbf{0.33}, and \textbf{0.27} variants when
+considering the V4, V3-V4, and V4-V5 regions of the gene, respectively.
+Although a species tended to have a consistent number of 16S rRNA gene
+copies per genome, the number of total variants increased with the
+number of genomes that were sampled (\textbf{Figure 1}). For example,
+\textbf{\emph{Mycobacterium tuberculosis}} generally only had \textbf{1}
+copy of the gene per genome, but across the \textbf{180} genomes in the
+\emph{rrn}DB there were \textbf{11} versions of the gene. Similarly, a
+\emph{E. coli} genome typically had \textbf{7} copies of the 16S rRNA
+gene with between \textbf{6} and \textbf{10} distinct full length
+sequences per genome. Across the \textbf{958} \emph{E. coli} genomes in
+the \emph{rrn}DB, there were \textbf{1,013} different variants of the
+gene. These observations highlight the risk of selecting a threshold for
+defining clusters that is too narrow because it is possible to split a
+single genome into multiple clusters.
 
-A method to avoid splitting a single genome into multiple units of
-inference is to cluster together similar 16S rRNA gene sequences.
-Therefore, I assessed the impact of the distance threshold used to
-define clusters of 16S rRNA genes on the propensity to split a genome
-into separate clusters. I observed that as the number of copies of the
-\emph{rrn} operon increased, the distance threshold required to reduce
-the ASVs in each genome to a single OTU increased (Figure 1). Among
-species with 7 copies of the \emph{rrn} operon (e.g., \emph{E. coli}), I
-found that a threshold of \textbf{5.5}\% was required to reduce full
-length ASVs to a single OTU in 95\% of the species. Similarly,
-thresholds of \textbf{2.5}, \textbf{4.0}, and \textbf{3.5}\% were
-required for the V4, V3-V4, and V4-V5 regions, respectively. But, if a
-3\% distance threshold was used, then ASVs from genomes containing fewer
-than \textbf{5}, \textbf{8}, \textbf{6}, and \textbf{6} copies of the
-\emph{rrn} operon would reliably be clustered into a single OTU for ASVs
-from the V1-V9, V4, V3-V4, and V4-V5 regions, respectively.
+A method to avoid splitting a single genome into multiple clusters is to
+cluster 16S rRNA gene sequences together based on their similarity to
+each other. Therefore, I assessed the impact of the distance threshold
+used to define clusters of 16S rRNA genes on the propensity to split a
+genome into separate clusters. I observed that as the number of copies
+of the \emph{rrn} operon increased, the distance threshold required to
+reduce the ASVs in each genome to a single OTU increased (Figure 1).
+Among species with 7 copies of the \emph{rrn} operon (e.g., \emph{E.
+coli}), I found that a threshold of \textbf{5.5}\% was required to
+reduce full length ASVs to a single OTU in 95\% of the species.
+Similarly, thresholds of \textbf{2.5}, \textbf{4.0}, and \textbf{3.5}\%
+were required for the V4, V3-V4, and V4-V5 regions, respectively. But,
+if a 3\% distance threshold was used, then ASVs from genomes containing
+fewer than \textbf{5}, \textbf{8}, \textbf{6}, and \textbf{6} copies of
+the \emph{rrn} operon would reliably be clustered into a single OTU for
+ASVs from the V1-V9, V4, V3-V4, and V4-V5 regions, respectively.
 Consequently, these results demonstrate that broad thresholds must be
 used to avoid splitting different operons from the same genome into
-separate clusters.
-
-At broad thresholds multiple species could be represented by the same
-OTU (\textbf{Figure 2}). Using ASVs, \textbf{3.6}\% of the species
-shared a 16S rRNA gene sequence variant with another species when
-considering full length sequences and \textbf{14.9}, \textbf{10.2}, and
-\textbf{12.0}\% when considering the V4, V3-V4, and V4-V5 regions,
-respectively. At the commonly used 3\% threshold, \textbf{25.2}\% of the
-species shared an OTU when considering full length sequences and
-\textbf{33.0}, \textbf{29.4}, and \textbf{32.2}\% when considering the
-V4, V3-V4, and V4-V5 regions, respectively. Considering that species
-designations are unevenly applied and reflect multiple biases, the risk
-of splitting a genome into multiple OTUs more problematic than
-clustering species together. Therefore, larger thresholds are advisable.
+separate clusters. At broad thresholds multiple species could be
+represented by the same OTU (\textbf{Figure 2}). Using ASVs,
+\textbf{3.6}\% of the species shared a 16S rRNA gene sequence variant
+with another species when considering full length sequences and
+\textbf{14.9}, \textbf{10.2}, and \textbf{12.0}\% when considering the
+V4, V3-V4, and V4-V5 regions, respectively. At the commonly used 3\%
+threshold, \textbf{25.2}\% of the species shared an OTU with another
+speecies when considering full length sequences and \textbf{33.0},
+\textbf{29.4}, and \textbf{32.2}\% when considering the V4, V3-V4, and
+V4-V5 regions, respectively. Considering that species designations are
+unevenly applied and reflect multiple human-imposed biases, the risk of
+splitting a genome into multiple OTUs more problematic than clustering
+species together. Therefore, larger thresholds are advisable.
 
 The results of this analysis demonstrate that there is a significant
-risk of splitting single genomes into multiple bins if too fine of a
-threshold is applied to defining an OTU. An ongoing problem for
-amplicon-based studies is defining a meaningful taxonomic unit of
-inference (11, 15, 16). Since there is no consensus definition for a
-biological species concept (17, 18), microbiologists must accept that
-how we have named bacterial species is biased and that taxonomic rules
-are not applied in a consistent manner (e.g., (19)). This makes it more
-challenging to attempt to fit a distance threshold to define an OTU
-definition that matches a set of species names (20). Furthermore, the
-16S rRNA gene does not evolve at the same rate across all bacterial
-lineages (15), which limits the biological interpretation of a common
-OTU definition. A distance-based definition of a taxonomic unit based on
-16S rRNA gene or full genome sequences is, at best, operational and not
-grounded in biological theory (15, 21--23). There is general agreement
-in bacterial systematics that to classify something to a bacterial
-species, you need phenotypic and genome sequence data (17--19). We are
-asking too much of a short section of a bacterial genome to be able to
-differentiate between species. It is difficult to defend a unit of
-inference that would split a single genome into multiple taxonomic
-units. It is not biologically plausible to entertain the possibility
-that parts of a genome would have different ecologies. Although there
-are multiple reasons that proponents of ASVs encourage their use, the
-significant risk of splitting genomes is too high to warrant their use.
+risk of splitting a single genome into multiple clusters if too fine of
+a threshold is applied to defining an OTU. An ongoing problem for
+amplicon-based studies is defining a meaningful taxonomic unit (11, 15,
+16). Since there is no consensus definition for a biological species
+concept (17, 18), microbiologists must accept that how we have named
+bacterial species is biased and that taxonomic rules are not applied in
+a consistent manner (e.g., (19)). This makes it impossible to fit a
+distance threshold to define an OTU definition that matches a set of
+species names (20). Furthermore, the 16S rRNA gene does not evolve at
+the same rate across all bacterial lineages (15), which limits the
+biological interpretation of a common OTU definition. A distance-based
+definition of a taxonomic unit based on 16S rRNA gene or full genome
+sequences is, at best, operational and not grounded in biological theory
+(15, 21--23). There is general agreement in bacterial systematics that
+to classify something to a bacterial species, phenotypic and genome
+sequence data are needed (17--19). A short section of a bacterial genome
+simply cannot differentiate between species. Moreover, it is difficult
+to defend a clustering approach that would split a single genome into
+multiple taxonomic units. It is not biologically plausible to entertain
+the possibility that parts of a genome would have different ecologies.
+Although there are multiple reasons that proponents favor ASVs, the
+significant risk of artificially splitting genomes into separate
+clusters is too high to warrant their use.
 
@@
 (\url{https://rrndb.umms.med.umich.edu}; version 5.6, released November
-8, 2019) (14). At the time of submission, this is the most current
+8, 2019) (14). At the time of submission, this was the most current
 version of the database. The \emph{rrn}DB obtained the curated 16S rRNA
 gene sequences from the KEGG database, which ultimately obtained them
-from NCBI's non-redundant RefSeq database. The \emph{rrn}DB provides
+from NCBI's non-redundant RefSeq database. The \emph{rrn}DB provided
 downloadable versions of the sequences with their taxonomy as determined
@@
 length 16S rRNA gene sequences were aligned to a SILVA reference
-alignment of the 16S rRNA gene (v138) using the mothur software package
-(v. 1.XX) (24, 25). Regions of the 16S rRNA gene were selected because
-of their use in the microbial ecology literature. Full length sequences
-corresponded to \emph{E. coli} positions XX through XXXX, V4 to
-positions XXX through XXX, V3-V4 to positions XXX through XXX, and V4-V5
-to positions XXX through XXX.
+alignment of the 16S rRNA gene (v. 138) using the mothur software
+package (v. 1.44.2) (24, 25). Regions of the 16S rRNA gene were selected
+because of their use in the microbial ecology literature. Full length
+sequences corresponded to \emph{E. coli} str. K-12 substr. MG1655
+(NC\_000913) positions 28 through 1491, V4 to positions 534 through 786,
+V3-V4 to positions 358 through 786, and V4-V5 to positions 534 through
+908.
 
@@
 selected one genome for each species. The random selection was repeated
-100 times. Analyses based on this randomization report the median of the
-100 randomizations. The intraquartile range between randomizations was
-typically less than XXXX. Because it was so small, confidence intervals
-are not included in Figure 2.
+100 times. Analyses based on this randomization reported the median of
+the 100 randomizations. The intraquartile range between randomizations
+was less than XXXX. Because the range was so small, the confidence
+intervals were smaller than the thickness of the lines in Figures 1 and
+2 and were not included.
 
 \textbf{(iv) Reproducible data analysis.} The code to perform the
-analysis in this manuscript and its hisotry are available as a git-based
+analysis in this manuscript and its history are available as a git-based
 version control repository on GitHub
-(\url{https://github.com/pschloss/Schloss_rrnAnalysis_XXXX_2020}). The
-analysis can be regenerated using a GNU Make-based workflow that made
-use of built-in bash tools (v. 3.2.57), mothur (v. 1.XX), and R (v.
-4.X.X). Within R, I used the tidyverse (v. 4.X.X), data.table (v.
-4.X.X), Rcpp (v. 4.X.X), furrr (v. 4.X.X), and rmarkdown (v. 4.X.X)
-packages. The conception and development of this analysis is available
-as a playlist on the Riffomonas YouTube channel
-(\url{https://www.youtube.com/playlist?list=PLmNrK_nkqBpKY3SZiivlIGvcLX-KHmfR8}).
+(\url{https://github.com/SchlossLab/Schloss_rrnAnalysis_mSphere_2021}).
+The analysis can be regenerated using a GNU Make-based workflow that
+made use of built-in bash tools (v. 3.2.57), mothur (v. 1.44.2), and R
+(v. \textbf{4.0.3}). Within R, I used the tidyverse (v. \textbf{1.3.0}),
+data.table (v. \textbf{1.13.2}), Rcpp (v. packageVersion(``Rcpp'')),
+furrr (v. packageVersion(``furrr'')), and rmarkdown (v.
+packageVersion(``rmarkdown'')) packages. The conception and development
+of this analysis is available as a playlist on the Riffomonas YouTube
+channel
+(\url{https://youtube.com/playlist?list=PLmNrK_nkqBpL7m_tyWdQgdyurerttCsPY}).
 
 \textbf{Acknowledgements.} I am grateful to Robert Hein and Thomas
-Schmidt who maintain the rrnDB for their help in understanding the
-curation of the database and for making the 16S rRNA gene sequences and
-related metadata publicly available. I am also grateful to community
+Schmidt, who maintain the \emph{rrn}DB, for their help in understanding
+the curation of the database and for making the 16S rRNA gene sequences
+and related metadata publicly available. I am also grateful to community
 members who watched the serialized version of this analysis on YouTube
-and provided their suggestions and questions.
+and provided suggestions and questions over the course of the
+development of this project.
 
@@
 \textbf{Figure 1. The distance threshold required to prevent the
-splitting of genomes into multiple OTUs increases as the number of
-\emph{rrn} operons in the genome increases.} Each line represents the
+splitting of genomes into multiple OTUs increased as the number of
+\emph{rrn} operons in the genome increased.} Each line represents the
 median distance threshold for each region of the 16S rRNA gene that is
@@
 \textbf{Figure 2. As the distance threshold used to define an OTU
-increases, the fraction of genomes split into separate OTUs decreases
-while the fraction of species that are merged into the same OTU
+increased, the fraction of genomes split into separate OTUs decreased
+while the fraction of species that were merged into the same OTU
 increases.} These data represent the median fractions for both
@@
 \textbf{Figure S1. The ratio of number of distinct ASVs per copy of the
-\emph{rrn} operon increases for a species as the number of genomes
-sampled increases.} Each point represents a different species and is
+\emph{rrn} operon increased for a species as the number of genomes
+sampled increased.} Each point represents a different species and was
 shaded to be 80\% transparent so that when points overlap they become

### 2021-02-19 code/plot_copy_number_threshold.R

@@ -12,7 +12,7 @@ read_tsv("data/processed/thresholds_for_single_otu.tsv") %>%
 	scale_x_continuous(breaks=seq(1,9,2), labels=seq(1,9,2)) +
 	scale_y_continuous(breaks=seq(0,0.06,0.02), labels=seq(0,6,2), limits =c(0,0.06)) +
 	labs(x="Number of copies of *rrn* operon",
-			 y="Distance threshold where ASVs from\n95% of species have one OTU (%)") +
+			 y="Distance threshold where ASVs from\n95% of species had one OTU (%)") +
 	theme_classic() +
 	theme(
 		legend.position = c(0.8, 0.2),

### 2021-02-19 code/plot_esv_rate.R

@@ -36,7 +36,7 @@ species_esvs <- metadata_esv %>%
 facet_label <- species_esvs %>%
 	distinct(region) %>%
 	mutate(x = 1, y=100)
-	
+
 
 species_esvs %>%
 	ggplot(aes(x=n_genomes, y=esv_rate)) +
@@ -58,5 +58,5 @@ species_esvs %>%
 		axis.title.y = element_markdown()
 	)
 
-ggsave("figures/esv_rate.tiff", width=3.5, height=8, compression="lzw")
-ggsave("figures/esv_rate.pdf", width=3.5, height=8)
+ggsave("figures/esv_rate.tiff", width=3.5, height=7.5, compression="lzw")
+ggsave("figures/esv_rate.pdf", width=3.5, height=7.5)

### 2021-02-19 code/plot_lump_split.R

@@ -9,16 +9,16 @@ read_tsv("data/processed/lumped_split_rate.tsv",
 	select(-ends_with("iqr")) %>%
 	pivot_longer(cols=c(split_rate, lump_rate), names_to="method", values_to="fraction") %>%
 	mutate(method = ifelse(method == "lump_rate",
-												 "Fraction of species\nmerged together",
-												 "Fraction of genomes\nsplit apart")) %>%
-	ggplot(aes(x=threshold, y=fraction, color=region)) + 
+												 "Fraction of genomes from\ndifferent species whose ASVs\nwere merged into the same cluster",
+												 "Fraction of genomes split\ninto separate clusters")) %>%
+	ggplot(aes(x=threshold, y=fraction, color=region)) +
 	geom_line(size=1) +
 	facet_wrap(~method, nrow=2, strip.position = "left") +
 	scale_x_continuous(
 		breaks = c(0, 0.025, 0.05, 0.075, 0.1),
 		labels=c("0", "2.5", "5", "7.5", "10")
-		) + 
-	custom_color_scale() + 
+		) +
+	custom_color_scale() +
 	labs(x="Distance theshold\nto define ASV/OTU (%)", y=NULL) +
 	theme_classic() +
 	theme(

### 2021-02-19 submission/manuscript.Rmd
Second round of revisions

@@
 	group_by(species, genome_id) %>%
-	summarize(n_rrn = sum(count), .groups="drop") %>%
+	summarize(n_rrn = sum(count), n_asv = n(), .groups="drop") %>%
 	group_by(species) %>%
-	summarize(min = min(n_rrn), max = max(n_rrn), median = median(n_rrn), N=n(), .groups="drop")
+	summarize(min_rrn = min(n_rrn),
+            l_quartile_rrn = quantile(n_rrn, prob=0.25),
+            median_rrn = median(n_rrn),
+            u_quartile_rrn = quantile(n_rrn, prob=0.75),
+            max_rrn = max(n_rrn),
+            min_asv = min(n_asv),
+            l_quartile_asv = quantile(n_asv, prob=0.25),
+            median_asv = median(n_asv),
+            u_quartile_asv = quantile(n_asv, prob=0.75),
+            max_asv = max(n_asv),
+            N=n(),
+            .groups="drop")
 
-single_copy <- rrn_copy_number %>% filter(median == 1) %>% top_n(n=1, N)
-max_copy <- rrn_copy_number %>% top_n(n=1, median)
+single_copy <- rrn_copy_number %>% filter(median_rrn == 1) %>% top_n(n=1, N)
+max_copy <- rrn_copy_number %>% top_n(n=1, median_rrn)
 
-single_copy_N <- rrn_copy_number %>% filter(species == single_copy$species) %>% pull(N)
-max_copy_N <- rrn_copy_number %>% filter(species == max_copy$species) %>% pull(N)
+single_copy_N <- single_copy %>% pull(N)
+#max_copy_N <- max_copy %>% pull(N)
+
 ecoli_copy <- rrn_copy_number %>% filter(species == "Escherichia coli")
@@
 
+#below are total number of ASVs across all genomes (above was within a genome)
 single_nasvs <- asvs_per_species %>% filter(species == single_copy$species) %>% pull(n_asvs)
@@
 lump3_v45 <- lump_split %>% filter(threshold ==0.03 & region == "v45") %>% pull(lump_rate) * 100
+
+lump_split_iqr <- lump_split %>% pivot_longer(cols=c(split_iqr, lump_iqr), names_to="test", values_to="iqr") %>% summarize(max(iqr)) %>% pull
+
 ```
@@
 
-Amplicon sequencing variants (ASVs) have been proposed as an alternative to operational taxonomic units (OTUs) for analyzing microbiomes.
-ASVs have grown in popularity, in part, because of a desire to reflect a more refined level of taxonomy because they do not cluster sequences based on a distance-based threshold.
+Amplicon sequencing variants (ASVs) have been proposed as an alternative to operational taxonomic units (OTUs) for analyzing microbial communities.
+ASVs have grown in popularity, in part, because of a desire to reflect a more refined level of taxonomy since they do not cluster sequences based on a distance-based threshold.
 However, ASVs and the use of overly narrow thresholds to identify OTUs increases the risk of splitting a single genome into separate clusters.
@@
 There was an average of `r rate_v19` ASVs per copy of the 16S rRNA gene, for full length 16S rRNA genes.
-It was necessary to use a distance threshold of `r format_pct(threshold_v19)`% to cluster the ASVs from the same genome into a single OTU, for genomes with 7 copies of the 16S rRNA, such as *E.
+It was necessary to use a distance threshold of `r format_pct(threshold_v19)`% to cluster full length ASVs from the same genome into a single OTU with **95% confidence**, for genomes with 7 copies of the 16S rRNA, such as *E.
 coli*.
 This research highlights the risk of splitting a single bacterial genome into separate clusters when ASVs are used to analyze 16S rRNA gene sequence data.
-Although there is also a risk of clustering different species into the same OTU, the effects of those risks on biological inferences are less than those from artificially splitting a genome into separate ASVs and OTUs.
+Although there is also a risk of clustering ASVs from different species into the same OTU when using broad distance thresholds, those risks are of less concern than artificially splitting a genome into separate ASVs and OTUs.
 
@@
 
-16S rRNA gene sequencing has propelled significant interest into host-associated and environmental microbiomes.
-There is a tension between trying to classify 16S rRNA gene sequences to increasingly lower taxonomic levels and the reality that those levels were defined using more sequence and physiological information than is available from a fragment of the 16S rRNA gene.
+16S rRNA gene sequencing has engendered significant interest in studying microbial communities.
+There has been a tension between trying to classify 16S rRNA gene sequences to increasingly lower taxonomic levels and the reality that those levels were defined using more sequence and physiological information than is available from a fragment of the 16S rRNA gene.
 Furthermore, naming of bacterial taxa reflects the biases of those who name them.
-One motivation for the recent push to adopt ASVs in place of OTUs in microbiome analyses is to allow researchers to perform their analyes at the finest possible level that reflects species-level taxonomy.
+One motivation for the recent push to adopt ASVs in place of OTUs in microbial community analyses is to allow researchers to perform their analyes at the finest possible level that reflects species-level taxonomy.
 The current research is significant because it quantifies the risk of artificially splitting bacterial genomes into separate clusters.
-Far from providing a better represenation of microbiomes, the ASV approach could lead to conflicting inferences about the ecology of different clusters from the same genome.
+Far from providing a better represenation of bacterial taxonomy and biology, the ASV approach can lead to conflicting inferences about the ecology of different ASVs from the same genome.
 
@@
 As an alternative to OTUs, amplicon sequencing variants (ASVs) have been proposed as a way to adopt the thresholds suggested by genome sequencing to microbial community analysis using 16S rRNA gene sequences [@Edgar2016; @Amir2017; @Callahan2016; @Eren2014].
-ASVs are a unit of microbial community inference that do not cluster sequences based on a distance-based threshold [@Callahan2017].
-However, most bacterial genomes have more than 1 copy of the *rrn* operon and those copies are not identical [@Pei2010; @Sun2013].
-Therefore, using too fine a threshold to identify OTUs creates the risk of splitting a single genome into multiple bins and using too broad of a threshold to define OTUs creates the risk of lumping together bacterial species into the same OTU.
+Approaches for identifying ASVs do not cluster sequences based on a distance-based threshold [@Callahan2017].
+Proponents of ASVs are largely dissmissive of concerns that most bacterial genomes have more than 1 copy of the *rrn* operon and those copies are not identical [@Pei2010; @Sun2013].
+Yet, using too fine a threshold to identify OTUs creates the risk of splitting a single genome into multiple bins.
+Conversely, using too broad of a threshold to define OTUs creates the risk of clustering together bacterial species into the same OTU.
 An example of both is seen in the comparison of *Staphylococcus aureus* (NCTC 8325) and *S.
 epidermidis* (ATCC 12228) where each genome has 5 copies of the 16S rRNA gene.
-Each of the 10 copies of the 16S rRNA gene in these two genomes were distinct and would yield 10 ASVs.
+Each of the 10 copies of the 16S rRNA gene in these two genomes is distinct and represent 10 ASVs.
 Conversely, if the copies were clustered using a 3% distance threshold, then all 10 ASVs would cluster into the same OTU.
-The goal of this study was to quantify the risk of splitting a single genome into multiple clusters and the risk of lumping together different bacterial species into the same cluster.
+The goal of this study was to quantify the risk of splitting a single genome into multiple clusters and the risk of clustering different bacterial species into the same cluster when using ASVs and various OTU definitions.
 
-To investigate the variation in the number of copies of the 16S rRNA gene per genome as well as the intragenomic variation among copies of the 16S rRNA gene, I obtained reference 16S rRNA sequences from the *rrn* copy number database (*rrn*DB)[@Stoddard2014].
+To investigate the variation in the number of copies of the 16S rRNA gene per genome as well as the intragenomic variation among copies of the 16S rRNA gene, I obtained 16S rRNA sequences from the *rrn* copy number database (*rrn*DB)[@Stoddard2014].
 Among the `r n_distinct(metadata$species)` species represented in the *rrn*DB there were `r nrow(metadata)` genomes.
-The median number of *rrn* operons per species ranged between `r single_copy$median` (e.g., *`r single_copy$species`*) and `r max_copy$median` (*`r max_copy$species`*) copies of the *rrn* operon.
+The median *rrn* copy number per species ranged between `r single_copy$median_rrn` (e.g., *`r single_copy$species`*) and `r max_copy$median_rrn` (*`r max_copy$species`*).
 As the number of copies of the operon in a genome increased, the number of variants of the 16S rRNA gene in each genome also increased.
-On average, there were `r rate_v19` variants per copy of the full length 16S rRNA gene and an average of `r rate_v4`, `r rate_v34`, and `r rate_v45` variants when considering the V4, V3-V4, and V4-V5 regions of the gene, respectively.
+On average, there were `r rate_v19` variants per copy of the full length 16S rRNA gene and an average of `r rate_v34`, `r rate_v4`, and `r rate_v45` variants when considering the V3-V4, V4, and V4-V5 regions of the gene, respectively.
 Although a species tended to have a consistent number of 16S rRNA gene copies per genome, the number of total variants increased with the number of genomes that were sampled (**Figure 1**).
-For example, *`r single_copy$species`* generally only had `r single_copy$median` copy of the gene per genome, but across the `r single_copy_N` genomes in the *rrn*DB there were `r single_nasvs` versions of the gene.
-Similarly, a *E.
-coli* genome typically had `r ecoli_copy$median` copies of the 16S rRNA gene with between `r ecoli_copy$min` and `r ecoli_copy$max` distinct full length sequences per genome.
+For example, the `r single_copy_N` genome accessions of *`r single_copy$species`* in the *rrn*DB only had `r single_copy$median_rrn` copy of the gene per genome.
+However, across those accessions, there were `r single_nasvs` versions of the gene.
+An *E.
+coli* genome typically had `r ecoli_copy$median_rrn` copies of the 16S rRNA gene with a median `r ecoli_copy$median_asv` distinct full length ASVs per genome (intraquartile range between `r ecoli_copy$l_quartile_asv` and `r ecoli_copy$u_quartile_asv`).
 Across the `r ecoli_copy$N` *E.
-coli* genomes in the *rrn*DB, there were `r ecoli_nasvs` different variants of the gene.
+coli* genomes in the *rrn*DB, there were `r ecoli_nasvs` versions of the gene.
 These observations highlight the risk of selecting a threshold for defining clusters that is too narrow because it is possible to split a single genome into multiple clusters.
 
-A method to avoid splitting a single genome into multiple clusters is to cluster 16S rRNA gene sequences together based on their similarity to each other.
+A method to avoid splitting a single genome into multiple clusters is to cluster 16S rRNA gene sequences together based on their distances between each other.
 Therefore, I assessed the impact of the distance threshold used to define clusters of 16S rRNA genes on the propensity to split a genome into separate clusters.
-I observed that as the number of copies of the *rrn* operon increased, the distance threshold required to reduce the ASVs in each genome to a single OTU increased (Figure 1).
+I observed that as the *rrn* copy number increased, the distance threshold required to reduce the ASVs in each genome to a single OTU increased (Figure 1).
 Among species with 7 copies of the *rrn* operon (e.g., *E.
-coli*), I found that a threshold of `r format_pct(threshold_v19)`% was required to reduce full length ASVs to a single OTU in 95% of the species.
-Similarly, thresholds of `r format_pct(threshold_v4)`, `r format_pct(threshold_v34)`, and `r format_pct(threshold_v45)`% were required for the V4, V3-V4, and V4-V5 regions, respectively.
-But, if a 3% distance threshold was used, then ASVs from genomes containing fewer than `r three_v19`, `r three_v4`, `r three_v34`, and `r three_v45` copies of the *rrn* operon would reliably be clustered into a single OTU for ASVs from the V1-V9, V4, V3-V4, and V4-V5 regions, respectively.
+coli*), a distance threshold of `r format_pct(threshold_v19)`% was required to reduce full length ASVs into a single OTU for 95% of the species.
+Similarly, thresholds of `r format_pct(threshold_v34)`, `r format_pct(threshold_v4)`, and `r format_pct(threshold_v45)`% were required for the V3-V4, V4, and V4-V5 regions, respectively.
+But, if a 3% distance threshold was used, then ASVs from genomes containing fewer than `r three_v19`, `r three_v34`, `r three_v4`, and `r three_v45` copies of the *rrn* operon would reliably be clustered into a single OTU for ASVs from the V1-V9, V3-V4, V4, and V4-V5 regions, respectively.
 Consequently, these results demonstrate that broad thresholds must be used to avoid splitting different operons from the same genome into separate clusters.
-At broad thresholds multiple species could be represented by the same OTU (**Figure 2**).
-Using ASVs, `r format_pct(lump_v19)`% of the species shared a 16S rRNA gene sequence variant with another species when considering full length sequences and `r format_pct(lump_v4)`, `r format_pct(lump_v34)`, and `r format_pct(lump_v45)`% when considering the V4, V3-V4, and V4-V5 regions, respectively.
-At the commonly used 3% threshold, `r format_pct(lump3_v19)`% of the species shared an OTU with another speecies when considering full length sequences and `r format_pct(lump3_v4)`, `r format_pct(lump3_v34)`, and `r format_pct(lump3_v45)`% when considering the V4, V3-V4, and V4-V5 regions, respectively.
+At broad thresholds ASVs from multiple species could be clustered into the same OTU (**Figure 2**).
+Using ASVs, `r format_pct(lump_v19)`% of the species shared a 16S rRNA gene sequence variant with another species when considering full length sequences and `r format_pct(lump_v34)`, `r format_pct(lump_v4)`, and `r format_pct(lump_v45)`% when considering the V3-V4, V4, and V4-V5 regions, respectively.
+At the commonly used 3% threshold, `r format_pct(lump3_v19)`% of the species shared an OTU with another speecies when considering full length sequences and `r format_pct(lump3_v34)`, `r format_pct(lump3_v4)`, and `r format_pct(lump3_v45)`% when considering the V3-V4, V4, and V4-V5 regions, respectively.
 Considering that species designations are unevenly applied and reflect multiple human-imposed biases, the risk of splitting a genome into multiple OTUs more problematic than clustering species together.
@@
 An ongoing problem for amplicon-based studies is defining a meaningful taxonomic unit [@Callahan2017; @Schloss2011; @Johnson2019].
-Since there is no consensus definition for a biological species concept [@Staley2006; @Oren2013], microbiologists must accept that how we have named bacterial species is biased and that taxonomic rules are not applied in a consistent manner (e.g., [@Baltrus2016]).
+Since there is no consensus definition for a biological species concept [@Staley2006; @Oren2013; @Sanford2021], microbiologists must accept that how we have named bacterial species is biased and that taxonomic rules are not applied in a consistent manner (e.g., [@Baltrus2016; @Sanford2021]).
 This makes it impossible to fit a distance threshold to define an OTU definition that matches a set of species names [@Konstantinidis2005].
@@
 A distance-based definition of a taxonomic unit based on 16S rRNA gene or full genome sequences is, at best, operational and not grounded in biological theory [@Schloss2011; @Barco2020; @Parks2020; @Yarza2014].
-There is general agreement in bacterial systematics that to classify something to a bacterial species, phenotypic and genome sequence data are needed [@Staley2006; @Oren2013; @Baltrus2016].
+There is general agreement in bacterial systematics that to classify something to a bacterial species, phenotypic and genome sequence data are needed [@Staley2006; @Oren2013; @Baltrus2016; @Sanford2021].
 A short section of a bacterial genome simply cannot differentiate between species.
@@
 Although there are multiple reasons that proponents favor ASVs, the significant risk of artificially splitting genomes into separate clusters is too high to warrant their use.
+
+\vspace{20mm}
 
@@
 Instead, I mapped the RefSeq accession number for each genome in the database to obtain a single taxonomy for each genome.
-Because strain names were not consistently given to genomes across bacterial species, the strain level designations were ignored.
+Because strain names were not consistently given to genomes across bacterial species, the strain level designations were disregarded.
 
-**(ii) Definition of regions within 16S rRNA gene.** The full length 16S rRNA gene sequences were aligned to a SILVA reference alignment of the 16S rRNA gene (v.
+**(ii) Definition of regions within the 16S rRNA gene.** The full length 16S rRNA gene sequences were aligned to a SILVA reference alignment of the 16S rRNA gene (v.
 138) using the mothur software package (v.
@@
 MG1655 (NC_000913) positions 28 through 1491, V4 to positions 534 through 786, V3-V4 to positions 358 through 786, and V4-V5 to positions 534 through 908.
+The positions between these coordinates reflect the fragments that would be amplified using commonly used PCR primers.
 
-**(iii) Controlling for uneven sampling of genomes by species.** Because of the uneven distribution of genome sequences across species, for the analysis of splitting genomes and lumping species I randomly selected one genome for each species.
+**(iii) Controlling for uneven sampling of genomes by species.** Because of the uneven distribution of genome sequences across species, for the analysis of splitting genomes and clustering ASVs from different species, I randomly selected one genome from each species.
 The random selection was repeated 100 times.
 Analyses based on this randomization reported the median of the 100 randomizations.
-The intraquartile range between randomizations was less than XXXX.
-Because the range was so small, the confidence intervals were smaller than the thickness of the lines in Figures 1 and 2 and were not included.
+The intraquartile range between randomizations was less than `r lump_split_iqr`.
+Because the range was so small, the confidence intervals were more narrow than the thickness of the lines in Figures 1 and 2 and were not included.
 
@@
 `r packageVersion("data.table")`), Rcpp (v.
-packageVersion("Rcpp")), furrr (v.
-packageVersion("furrr")), and rmarkdown (v.
-packageVersion("rmarkdown")) packages.
+`r packageVersion("Rcpp")`), furrr (v.
+`r packageVersion("furrr")`), and rmarkdown (v.
+`r packageVersion("rmarkdown")`) packages.
 The conception and development of this analysis is available as a playlist on the Riffomonas YouTube channel (https://youtube.com/playlist?list=PLmNrK_nkqBpL7m_tyWdQgdyurerttCsPY).
+
+**(v) Note on usage of ASV, OTU, and cluster.** I used "ASV" to denote the cluster of true 16S rRNA gene sequences that were identical to each other and "OTU" to denote the product of distance-based clustering of sequences.
+Although ASVs do represent a type of operational defition of a taxonomic unit and can be thought of as an OTU formed using a distance of zero, proponents of the ASV approach prefer to avoid the term OTU given the long history of OTUs being formed by distance-based clustering (https://github.com/benjjneb/dada2/issues/62; accessed 2021-02-19).
+For this reason, when an ASV splits a genome into different units those units are called clusters rather than OTUs.
+
+\vspace{20mm}
 
@@
 I am also grateful to community members who watched the serialized version of this analysis on YouTube and provided suggestions and questions over the course of the development of this project.
-
 This work was supported, in part, through grants from the NIH to PDS (P30DK034933, U01AI124255, and R01CA215574).
@@
 **Figure 1.
-The distance threshold required to prevent the splitting of genomes into multiple OTUs increased as the number of *rrn* operons in the genome increased.** Each line represents the median distance threshold for each region of the 16S rRNA gene that is required for 95% of the species with the indicated numbrer of *rrn* operons to cluster their ASVs to a single OTU.
+The distance threshold required to prevent the splitting of genomes into multiple OTUs increased as the number of *rrn* operons in the genome increased.** Each line represents the median distance threshold for each region of the 16S rRNA gene that is required for 95% of the genomes with the indicated numbrer of *rrn* operons to cluster their ASVs to a single OTU.
 The median distance threshold was calculated across 100 randomizations in which one genome was sampled from each species.
@@
 **Figure 2.
-As the distance threshold used to define an OTU increased, the fraction of genomes split into separate OTUs decreased while the fraction of species that were merged into the same OTU increases.** These data represent the median fractions for both measurements across 100 randomizations.
+As the distance threshold used to define an OTU increased, the fraction of genomes split into separate OTUs decreased while the fraction of ASVs from genomes in different species that were merged into the same OTU increased.** These data represent the median fractions for both measurements across 100 randomizations.
 In each randomization, one genome was sampled from each species.
@@
 **Figure S1.
-The ratio of number of distinct ASVs per copy of the *rrn* operon increased for a species as the number of genomes sampled increased.** Each point represents a different species and was shaded to be 80% transparent so that when points overlap they become darker.
+The ratio of number of distinct ASVs per copy of the *rrn* operon increased for a species as the number of genomes in the *rrn*DB for that species increased.** Each point represents a different species and was shaded to be 80% transparent so that when points overlap they become darker.
 The blue line represents a smoothed fit through the data.

### 2021-02-19 submission/manuscript.tex
Second round of revisions

@@
 Amplicon sequencing variants (ASVs) have been proposed as an alternative
-to operational taxonomic units (OTUs) for analyzing microbiomes. ASVs
-have grown in popularity, in part, because of a desire to reflect a more
-refined level of taxonomy because they do not cluster sequences based on
-a distance-based threshold. However, ASVs and the use of overly narrow
-thresholds to identify OTUs increases the risk of splitting a single
-genome into separate clusters. I assessed this problem by analyzing the
-intragenomic variation of 16S rRNA genes from the bacterial genomes
-represented in a \emph{rrn} copy number database, which contained
-\textbf{15,614} genomes from \textbf{4,774} species. As the number of
-copies of the 16S rRNA gene increased in a genome, the number of ASVs
-also increased. There was an average of \textbf{0.60} ASVs per copy of
-the 16S rRNA gene, for full length 16S rRNA genes. It was necessary to
-use a distance threshold of \textbf{5.5}\% to cluster the ASVs from the
-same genome into a single OTU, for genomes with 7 copies of the 16S
-rRNA, such as \emph{E. coli}. This research highlights the risk of
-splitting a single bacterial genome into separate clusters when ASVs are
-used to analyze 16S rRNA gene sequence data. Although there is also a
-risk of clustering different species into the same OTU, the effects of
-those risks on biological inferences are less than those from
+to operational taxonomic units (OTUs) for analyzing microbial
+communities. ASVs have grown in popularity, in part, because of a desire
+to reflect a more refined level of taxonomy since they do not cluster
+sequences based on a distance-based threshold. However, ASVs and the use
+of overly narrow thresholds to identify OTUs increases the risk of
+splitting a single genome into separate clusters. I assessed this
+problem by analyzing the intragenomic variation of 16S rRNA genes from
+the bacterial genomes represented in a \emph{rrn} copy number database,
+which contained \textbf{15,614} genomes from \textbf{4,774} species. As
+the number of copies of the 16S rRNA gene increased in a genome, the
+number of ASVs also increased. There was an average of \textbf{0.60}
+ASVs per copy of the 16S rRNA gene, for full length 16S rRNA genes. It
+was necessary to use a distance threshold of \textbf{5.5}\% to cluster
+full length ASVs from the same genome into a single OTU with
+\textbf{95\% confidence}, for genomes with 7 copies of the 16S rRNA,
+such as \emph{E. coli}. This research highlights the risk of splitting a
+single bacterial genome into separate clusters when ASVs are used to
+analyze 16S rRNA gene sequence data. Although there is also a risk of
+clustering ASVs from different species into the same OTU when using
+broad distance thresholds, those risks are of less concern than
 artificially splitting a genome into separate ASVs and OTUs.
@@
 
-16S rRNA gene sequencing has propelled significant interest into
-host-associated and environmental microbiomes. There is a tension
-between trying to classify 16S rRNA gene sequences to increasingly lower
-taxonomic levels and the reality that those levels were defined using
-more sequence and physiological information than is available from a
-fragment of the 16S rRNA gene. Furthermore, naming of bacterial taxa
-reflects the biases of those who name them. One motivation for the
-recent push to adopt ASVs in place of OTUs in microbiome analyses is to
-allow researchers to perform their analyes at the finest possible level
-that reflects species-level taxonomy. The current research is
-significant because it quantifies the risk of artificially splitting
-bacterial genomes into separate clusters. Far from providing a better
-represenation of microbiomes, the ASV approach could lead to conflicting
-inferences about the ecology of different clusters from the same genome.
+16S rRNA gene sequencing has engendered significant interest in studying
+microbial communities. There has been a tension between trying to
+classify 16S rRNA gene sequences to increasingly lower taxonomic levels
+and the reality that those levels were defined using more sequence and
+physiological information than is available from a fragment of the 16S
+rRNA gene. Furthermore, naming of bacterial taxa reflects the biases of
+those who name them. One motivation for the recent push to adopt ASVs in
+place of OTUs in microbial community analyses is to allow researchers to
+perform their analyes at the finest possible level that reflects
+species-level taxonomy. The current research is significant because it
+quantifies the risk of artificially splitting bacterial genomes into
+separate clusters. Far from providing a better represenation of
+bacterial taxonomy and biology, the ASV approach can lead to conflicting
+inferences about the ecology of different ASVs from the same genome.
 
@@
 sequencing to microbial community analysis using 16S rRNA gene sequences
-(7--10). ASVs are a unit of microbial community inference that do not
-cluster sequences based on a distance-based threshold (11). However,
-most bacterial genomes have more than 1 copy of the \emph{rrn} operon
-and those copies are not identical (12, 13). Therefore, using too fine a
-threshold to identify OTUs creates the risk of splitting a single genome
-into multiple bins and using too broad of a threshold to define OTUs
-creates the risk of lumping together bacterial species into the same
-OTU. An example of both is seen in the comparison of
-\emph{Staphylococcus aureus} (NCTC 8325) and \emph{S. epidermidis} (ATCC
-12228) where each genome has 5 copies of the 16S rRNA gene. Each of the
-10 copies of the 16S rRNA gene in these two genomes were distinct and
-would yield 10 ASVs. Conversely, if the copies were clustered using a
-3\% distance threshold, then all 10 ASVs would cluster into the same
-OTU. The goal of this study was to quantify the risk of splitting a
-single genome into multiple clusters and the risk of lumping together
-different bacterial species into the same cluster.
+(7--10). Approaches for identifying ASVs do not cluster sequences based
+on a distance-based threshold (11). Proponents of ASVs are largely
+dissmissive of concerns that most bacterial genomes have more than 1
+copy of the \emph{rrn} operon and those copies are not identical (12,
+13). Yet, using too fine a threshold to identify OTUs creates the risk
+of splitting a single genome into multiple bins. Conversely, using too
+broad of a threshold to define OTUs creates the risk of clustering
+together bacterial species into the same OTU. An example of both is seen
+in the comparison of \emph{Staphylococcus aureus} (NCTC 8325) and
+\emph{S. epidermidis} (ATCC 12228) where each genome has 5 copies of the
+16S rRNA gene. Each of the 10 copies of the 16S rRNA gene in these two
+genomes is distinct and represent 10 ASVs. Conversely, if the copies
+were clustered using a 3\% distance threshold, then all 10 ASVs would
+cluster into the same OTU. The goal of this study was to quantify the
+risk of splitting a single genome into multiple clusters and the risk of
+clustering different bacterial species into the same cluster when using
+ASVs and various OTU definitions.
 
@@
 gene per genome as well as the intragenomic variation among copies of
-the 16S rRNA gene, I obtained reference 16S rRNA sequences from the
-\emph{rrn} copy number database (\emph{rrn}DB)(14). Among the
-\textbf{4,774} species represented in the \emph{rrn}DB there were
-\textbf{15,614} genomes. The median number of \emph{rrn} operons per
-species ranged between \textbf{1} (e.g., \textbf{\emph{Mycobacterium
-tuberculosis}}) and \textbf{19} (\textbf{\emph{Metabacillus litoralis}})
-copies of the \emph{rrn} operon. As the number of copies of the operon
-in a genome increased, the number of variants of the 16S rRNA gene in
-each genome also increased. On average, there were \textbf{0.60}
-variants per copy of the full length 16S rRNA gene and an average of
-\textbf{0.26}, \textbf{0.33}, and \textbf{0.27} variants when
-considering the V4, V3-V4, and V4-V5 regions of the gene, respectively.
+the 16S rRNA gene, I obtained 16S rRNA sequences from the \emph{rrn}
+copy number database (\emph{rrn}DB)(14). Among the \textbf{4,774}
+species represented in the \emph{rrn}DB there were \textbf{15,614}
+genomes. The median \emph{rrn} copy number per species ranged between
+\textbf{1} (e.g., \textbf{\emph{Mycobacterium tuberculosis}}) and
+\textbf{19} (\textbf{\emph{Metabacillus litoralis}}). As the number of
+copies of the operon in a genome increased, the number of variants of
+the 16S rRNA gene in each genome also increased. On average, there were
+\textbf{0.60} variants per copy of the full length 16S rRNA gene and an
+average of \textbf{0.33}, \textbf{0.26}, and \textbf{0.27} variants when
+considering the V3-V4, V4, and V4-V5 regions of the gene, respectively.
 Although a species tended to have a consistent number of 16S rRNA gene
@@
 number of genomes that were sampled (\textbf{Figure 1}). For example,
-\textbf{\emph{Mycobacterium tuberculosis}} generally only had \textbf{1}
-copy of the gene per genome, but across the \textbf{180} genomes in the
-\emph{rrn}DB there were \textbf{11} versions of the gene. Similarly, a
-\emph{E. coli} genome typically had \textbf{7} copies of the 16S rRNA
-gene with between \textbf{6} and \textbf{10} distinct full length
-sequences per genome. Across the \textbf{958} \emph{E. coli} genomes in
-the \emph{rrn}DB, there were \textbf{1,013} different variants of the
-gene. These observations highlight the risk of selecting a threshold for
-defining clusters that is too narrow because it is possible to split a
-single genome into multiple clusters.
+the \textbf{180} genome accessions of \textbf{\emph{Mycobacterium
+tuberculosis}} in the \emph{rrn}DB only had \textbf{1} copy of the gene
+per genome. However, across those accessions, there were \textbf{11}
+versions of the gene. An \emph{E. coli} genome typically had \textbf{7}
+copies of the 16S rRNA gene with a median \textbf{5} distinct full
+length ASVs per genome (intraquartile range between \textbf{4} and
+\textbf{6}). Across the \textbf{958} \emph{E. coli} genomes in the
+\emph{rrn}DB, there were \textbf{1,013} versions of the gene. These
+observations highlight the risk of selecting a threshold for defining
+clusters that is too narrow because it is possible to split a single
+genome into multiple clusters.
 
 A method to avoid splitting a single genome into multiple clusters is to
-cluster 16S rRNA gene sequences together based on their similarity to
-each other. Therefore, I assessed the impact of the distance threshold
-used to define clusters of 16S rRNA genes on the propensity to split a
-genome into separate clusters. I observed that as the number of copies
-of the \emph{rrn} operon increased, the distance threshold required to
-reduce the ASVs in each genome to a single OTU increased (Figure 1).
-Among species with 7 copies of the \emph{rrn} operon (e.g., \emph{E.
-coli}), I found that a threshold of \textbf{5.5}\% was required to
-reduce full length ASVs to a single OTU in 95\% of the species.
-Similarly, thresholds of \textbf{2.5}, \textbf{4.0}, and \textbf{3.5}\%
-were required for the V4, V3-V4, and V4-V5 regions, respectively. But,
-if a 3\% distance threshold was used, then ASVs from genomes containing
-fewer than \textbf{5}, \textbf{8}, \textbf{6}, and \textbf{6} copies of
-the \emph{rrn} operon would reliably be clustered into a single OTU for
-ASVs from the V1-V9, V4, V3-V4, and V4-V5 regions, respectively.
+cluster 16S rRNA gene sequences together based on their distances
+between each other. Therefore, I assessed the impact of the distance
+threshold used to define clusters of 16S rRNA genes on the propensity to
+split a genome into separate clusters. I observed that as the \emph{rrn}
+copy number increased, the distance threshold required to reduce the
+ASVs in each genome to a single OTU increased (Figure 1). Among species
+with 7 copies of the \emph{rrn} operon (e.g., \emph{E. coli}), a
+distance threshold of \textbf{5.5}\% was required to reduce full length
+ASVs into a single OTU for 95\% of the species. Similarly, thresholds of
+\textbf{4.0}, \textbf{2.5}, and \textbf{3.5}\% were required for the
+V3-V4, V4, and V4-V5 regions, respectively. But, if a 3\% distance
+threshold was used, then ASVs from genomes containing fewer than
+\textbf{5}, \textbf{6}, \textbf{8}, and \textbf{6} copies of the
+\emph{rrn} operon would reliably be clustered into a single OTU for ASVs
+from the V1-V9, V3-V4, V4, and V4-V5 regions, respectively.
 Consequently, these results demonstrate that broad thresholds must be
 used to avoid splitting different operons from the same genome into
-separate clusters. At broad thresholds multiple species could be
-represented by the same OTU (\textbf{Figure 2}). Using ASVs,
+separate clusters. At broad thresholds ASVs from multiple species could
+be clustered into the same OTU (\textbf{Figure 2}). Using ASVs,
 \textbf{3.6}\% of the species shared a 16S rRNA gene sequence variant
 with another species when considering full length sequences and
-\textbf{14.9}, \textbf{10.2}, and \textbf{12.0}\% when considering the
-V4, V3-V4, and V4-V5 regions, respectively. At the commonly used 3\%
+\textbf{10.2}, \textbf{14.9}, and \textbf{12.0}\% when considering the
+V3-V4, V4, and V4-V5 regions, respectively. At the commonly used 3\%
 threshold, \textbf{25.2}\% of the species shared an OTU with another
-speecies when considering full length sequences and \textbf{33.0},
-\textbf{29.4}, and \textbf{32.2}\% when considering the V4, V3-V4, and
+speecies when considering full length sequences and \textbf{29.4},
+\textbf{33.0}, and \textbf{32.2}\% when considering the V3-V4, V4, and
 V4-V5 regions, respectively. Considering that species designations are
@@
 16). Since there is no consensus definition for a biological species
-concept (17, 18), microbiologists must accept that how we have named
+concept (17--19), microbiologists must accept that how we have named
 bacterial species is biased and that taxonomic rules are not applied in
-a consistent manner (e.g., (19)). This makes it impossible to fit a
+a consistent manner (e.g., (19, 20)). This makes it impossible to fit a
 distance threshold to define an OTU definition that matches a set of
-species names (20). Furthermore, the 16S rRNA gene does not evolve at
+species names (21). Furthermore, the 16S rRNA gene does not evolve at
 the same rate across all bacterial lineages (15), which limits the
@@
 sequences is, at best, operational and not grounded in biological theory
-(15, 21--23). There is general agreement in bacterial systematics that
+(15, 22--24). There is general agreement in bacterial systematics that
 to classify something to a bacterial species, phenotypic and genome
-sequence data are needed (17--19). A short section of a bacterial genome
+sequence data are needed (17--20). A short section of a bacterial genome
 simply cannot differentiate between species. Moreover, it is difficult
@@
 clusters is too high to warrant their use.
+
+\vspace{20mm}
 
@@
 consistently given to genomes across bacterial species, the strain level
-designations were ignored.
+designations were disregarded.
 
-\textbf{(ii) Definition of regions within 16S rRNA gene.} The full
+\textbf{(ii) Definition of regions within the 16S rRNA gene.} The full
 length 16S rRNA gene sequences were aligned to a SILVA reference
 alignment of the 16S rRNA gene (v. 138) using the mothur software
-package (v. 1.44.2) (24, 25). Regions of the 16S rRNA gene were selected
+package (v. 1.44.2) (25, 26). Regions of the 16S rRNA gene were selected
 because of their use in the microbial ecology literature. Full length
@@
 V3-V4 to positions 358 through 786, and V4-V5 to positions 534 through
-908.
+908. The positions between these coordinates reflect the fragments that
+would be amplified using commonly used PCR primers.
 
@@
 Because of the uneven distribution of genome sequences across species,
-for the analysis of splitting genomes and lumping species I randomly
-selected one genome for each species. The random selection was repeated
-100 times. Analyses based on this randomization reported the median of
-the 100 randomizations. The intraquartile range between randomizations
-was less than XXXX. Because the range was so small, the confidence
-intervals were smaller than the thickness of the lines in Figures 1 and
-2 and were not included.
+for the analysis of splitting genomes and clustering ASVs from different
+species, I randomly selected one genome from each species. The random
+selection was repeated 100 times. Analyses based on this randomization
+reported the median of the 100 randomizations. The intraquartile range
+between randomizations was less than \textbf{0.0029}. Because the range
+was so small, the confidence intervals were more narrow than the
+thickness of the lines in Figures 1 and 2 and were not included.
 
@@
 (v. \textbf{4.0.3}). Within R, I used the tidyverse (v. \textbf{1.3.0}),
-data.table (v. \textbf{1.13.2}), Rcpp (v. packageVersion(``Rcpp'')),
-furrr (v. packageVersion(``furrr'')), and rmarkdown (v.
-packageVersion(``rmarkdown'')) packages. The conception and development
-of this analysis is available as a playlist on the Riffomonas YouTube
-channel
+data.table (v. \textbf{1.13.2}), Rcpp (v. \textbf{1.0.5}), furrr (v.
+\textbf{0.2.1}), and rmarkdown (v. \textbf{2.5}) packages. The
+conception and development of this analysis is available as a playlist
+on the Riffomonas YouTube channel
 (\url{https://youtube.com/playlist?list=PLmNrK_nkqBpL7m_tyWdQgdyurerttCsPY}).
+
+\textbf{(v) Note on usage of ASV, OTU, and cluster.} I used ``ASV'' to
+denote the cluster of true 16S rRNA gene sequences that were identical
+to each other and ``OTU'' to denote the product of distance-based
+clustering of sequences. Although ASVs do represent a type of
+operational defition of a taxonomic unit and can be thought of as an OTU
+formed using a distance of zero, proponents of the ASV approach prefer
+to avoid the term OTU given the long history of OTUs being formed by
+distance-based clustering
+(\url{https://github.com/benjjneb/dada2/issues/62}; accessed
+2021-02-19). For this reason, when an ASV splits a genome into different
+units those units are called clusters rather than OTUs.
+
+\vspace{20mm}
 
@@
 and provided suggestions and questions over the course of the
-development of this project.
-
-This work was supported, in part, through grants from the NIH to PDS
-(P30DK034933, U01AI124255, and R01CA215574).
+development of this project. This work was supported, in part, through
+grants from the NIH to PDS (P30DK034933, U01AI124255, and R01CA215574).
 
@@
 determination of 16S ribosomal RNA sequences for phylogenetic analyses.
-Proceedings of the National Academy of Sciences 82:6955--6959
-\url{https://doi.org/10.1073/pnas.82.20.6955}.
+\emph{Proceedings of the National Academy of Sciences} 82:6955--6959.
+doi:
+\href{https://doi.org/10.1073/pnas.82.20.6955}{10.1073/pnas.82.20.6955}.
 
@@
 reassociation and 16S rRNA sequence analysis in the present species
-definition in bacteriology. International Journal of Systematic and
-Evolutionary Microbiology 44:846--849
-\url{https://doi.org/10.1099/00207713-44-4-846}.
+definition in bacteriology. \emph{International Journal of Systematic
+and Evolutionary Microbiology} 44:846--849. doi:
+\href{https://doi.org/10.1099/00207713-44-4-846}{10.1099/00207713-44-4-846}.
 
@@
 Tiedje JM. 2007. DNA-DNA hybridization values and their relationship to
-whole-genome sequence similarities. International Journal of Systematic
-and Evolutionary Microbiology 57:81--91
-\url{https://doi.org/10.1099/ijs.0.64483-0}.
+whole-genome sequence similarities. \emph{International Journal of
+Systematic and Evolutionary Microbiology} 57:81--91. doi:
+\href{https://doi.org/10.1099/ijs.0.64483-0}{10.1099/ijs.0.64483-0}.
 
@@
 Konstantinidis KT. 2018. How much do rRNA gene surveys underestimate
-extant bacterial diversity? Applied and Environmental Microbiology
-84:e00014--18 \url{https://doi.org/10.1128/aem.00014-18}.
+extant bacterial diversity? \emph{Applied and Environmental
+Microbiology} 84:e00014--18. doi:
+\href{https://doi.org/10.1128/aem.00014-18}{10.1128/aem.00014-18}.
 
@@
 5. Stackebrandt E, Ebers J. 2006. Taxonomic parameters revisited:
-Tarnished gold standards. Microbiol Today 33:152--155.
+Tarnished gold standards. \emph{Microbiol Today} 33:152--155.
 
@@
 6. Edgar RC. 2018. Updating the 97\% identity threshold for 16S
-ribosomal RNA OTUs. Bioinformatics 34:2371--2375
-\url{https://doi.org/10.1093/bioinformatics/bty113}.
+ribosomal RNA OTUs. \emph{Bioinformatics} 34:2371--2375. doi:
+\href{https://doi.org/10.1093/bioinformatics/bty113}{10.1093/bioinformatics/bty113}.
 
@@
 7. Edgar RC. 2016. UNOISE2: Improved error-correction for illumina 16S
-and its amplicon sequencing. bioRxiv
-\url{https://doi.org/10.1101/081257}.
+and its amplicon sequencing. \emph{bioRxiv}. doi:
+\href{https://doi.org/10.1101/081257}{10.1101/081257}.
 
@@
 Z, Kightley EP, Thompson LR, Hyde ER, Gonzalez A, Knight R. 2017. Deblur
-rapidly resolves single-nucleotide community sequence patterns. mSystems
-2:e00191--16 \url{https://doi.org/10.1128/mSystems.00191-16}.
+rapidly resolves single-nucleotide community sequence patterns.
+\emph{mSystems} 2:e00191--16. doi:
+\href{https://doi.org/10.1128/mSystems.00191-16}{10.1128/mSystems.00191-16}.
 
@@
 2016. DADA2: High-resolution sample inference from illumina amplicon
-data. Nature Methods 13:581--583
-\url{https://doi.org/10.1038/nmeth.3869}.
+data. \emph{Nature Methods} 13:581--583. doi:
+\href{https://doi.org/10.1038/nmeth.3869}{10.1038/nmeth.3869}.
 
@@
 ML. 2014. Minimum entropy decomposition: Unsupervised oligotyping for
-sensitive partitioning of high-throughput marker gene sequences. The
-ISME Journal 9:968--979 \url{https://doi.org/10.1038/ismej.2014.195}.
+sensitive partitioning of high-throughput marker gene sequences.
+\emph{The ISME Journal} 9:968--979. doi:
+\href{https://doi.org/10.1038/ismej.2014.195}{10.1038/ismej.2014.195}.
 
@@
 should replace operational taxonomic units in marker-gene data analysis.
-The ISME Journal 11:2639--2643
-\url{https://doi.org/10.1038/ismej.2017.119}.
+\emph{The ISME Journal} 11:2639--2643. doi:
+\href{https://doi.org/10.1038/ismej.2017.119}{10.1038/ismej.2017.119}.
 
@@
 K, Pei Z. 2010. Diversity of 16S rRNA genes within individual
-prokaryotic genomes. Applied and Environmental Microbiology
-76:3886--3897 \url{https://doi.org/10.1128/aem.02953-09}.
+prokaryotic genomes. \emph{Applied and Environmental Microbiology}
+76:3886--3897. doi:
+\href{https://doi.org/10.1128/aem.02953-09}{10.1128/aem.02953-09}.
 
@@
 of 16S rRNA genes causes overestimation of prokaryotic diversity.
-Applied and Environmental Microbiology 79:5962--5969
-\url{https://doi.org/10.1128/aem.01282-13}.
+\emph{Applied and Environmental Microbiology} 79:5962--5969. doi:
+\href{https://doi.org/10.1128/aem.01282-13}{10.1128/aem.01282-13}.
 
 \leavevmode\hypertarget{ref-Stoddard2014}{}%
-14. Stoddard SF, Smith BJ, Hein R, Roller BRK, Schmidt TM. 2014. rrnDB:
-Improved tools for interpreting rRNA gene abundance in bacteria and
-archaea and a new foundation for future development. Nucleic Acids
-Research 43:D593--D598 \url{https://doi.org/10.1093/nar/gku1201}.
+14. Stoddard SF, Smith BJ, Hein R, Roller BRK, Schmidt TM. 2014.
+\emph{rrn}DB: Improved tools for interpreting rRNA gene abundance in
+bacteria and archaea and a new foundation for future development.
+\emph{Nucleic Acids Research} 43:D593--D598. doi:
+\href{https://doi.org/10.1093/nar/gku1201}{10.1093/nar/gku1201}.
 
@@
 in operational taxonomic unit-based approaches for 16S rRNA gene
-sequence analysis. Applied and Environmental Microbiology 77:3219--3226
-\url{https://doi.org/10.1128/aem.02810-10}.
+sequence analysis. \emph{Applied and Environmental Microbiology}
+77:3219--3226. doi:
+\href{https://doi.org/10.1128/aem.02810-10}{10.1128/aem.02810-10}.
 
@@
 GM. 2019. Evaluation of 16S rRNA gene sequencing for species and
-strain-level microbiome analysis. Nature Communications 10:5029
-\url{https://doi.org/10.1038/s41467-019-13036-1}.
+strain-level microbiome analysis. \emph{Nature Communications} 10:5029.
+doi:
+\href{https://doi.org/10.1038/s41467-019-13036-1}{10.1038/s41467-019-13036-1}.
 
@@
 17. Staley JT. 2006. The bacterial species dilemma and the
-genomicphylogenetic species concept. Philosophical Transactions of the
-Royal Society B: Biological Sciences 361:1899--1909
-\url{https://doi.org/10.1098/rstb.2006.1914}.
+genomicphylogenetic species concept. \emph{Philosophical Transactions of
+the Royal Society B: Biological Sciences} 361:1899--1909. doi:
+\href{https://doi.org/10.1098/rstb.2006.1914}{10.1098/rstb.2006.1914}.
 
@@
 18. Oren A, Garrity GM. 2013. Then and now: A systematic review of the
-systematics of prokaryotes in the last 80~years. Antonie van Leeuwenhoek
-106:43--56 \url{https://doi.org/10.1007/s10482-013-0084-1}.
+systematics of prokaryotes in the last 80~years. \emph{Antonie van
+Leeuwenhoek} 106:43--56. doi:
+\href{https://doi.org/10.1007/s10482-013-0084-1}{10.1007/s10482-013-0084-1}.
+
+\leavevmode\hypertarget{ref-Sanford2021}{}%
+19. Sanford RA, Lloyd KG, Konstantinidis KT, Löffler FE. 2021. Microbial
+taxonomy run amok. \emph{Trends in Microbiology}. doi:
+\href{https://doi.org/10.1016/j.tim.2020.12.010}{10.1016/j.tim.2020.12.010}.
 
 \leavevmode\hypertarget{ref-Baltrus2016}{}%
-19. Baltrus DA, McCann HC, Guttman DS. 2016. Evolution, genomics and
-epidemiology ofPseudomonas syringae. Molecular Plant Pathology
-18:152--168 \url{https://doi.org/10.1111/mpp.12506}.
+20. Baltrus DA, McCann HC, Guttman DS. 2016. Evolution, genomics and
+epidemiology of \emph{Pseudomonas syringae}. \emph{Molecular Plant
+Pathology} 18:152--168. doi:
+\href{https://doi.org/10.1111/mpp.12506}{10.1111/mpp.12506}.
 
 \leavevmode\hypertarget{ref-Konstantinidis2005}{}%
-20. Konstantinidis KT, Tiedje JM. 2005. Towards a genome-based taxonomy
-for prokaryotes. Journal of Bacteriology 187:6258--6264
-\url{https://doi.org/10.1128/jb.187.18.6258-6264.2005}.
+21. Konstantinidis KT, Tiedje JM. 2005. Towards a genome-based taxonomy
+for prokaryotes. \emph{Journal of Bacteriology} 187:6258--6264. doi:
+\href{https://doi.org/10.1128/jb.187.18.6258-6264.2005}{10.1128/jb.187.18.6258-6264.2005}.
 
 \leavevmode\hypertarget{ref-Barco2020}{}%
-21. Barco RA, Garrity GM, Scott JJ, Amend JP, Nealson KH, Emerson D.
+22. Barco RA, Garrity GM, Scott JJ, Amend JP, Nealson KH, Emerson D.
 2020. A genus definition for bacteria and archaea based on a standard
-genome relatedness index. mBio 11:02475--19
-\url{https://doi.org/10.1128/mbio.02475-19}.
+genome relatedness index. \emph{mBio} 11:02475--19. doi:
+\href{https://doi.org/10.1128/mbio.02475-19}{10.1128/mbio.02475-19}.
 
 \leavevmode\hypertarget{ref-Parks2020}{}%
-22. Parks DH, Chuvochina M, Chaumeil P-A, Rinke C, Mussig AJ, Hugenholtz
+23. Parks DH, Chuvochina M, Chaumeil P-A, Rinke C, Mussig AJ, Hugenholtz
 P. 2020. A complete domain-to-species taxonomy for bacteria and archaea.
-Nature Biotechnology 38:1079--1086
-\url{https://doi.org/10.1038/s41587-020-0501-8}.
+\emph{Nature Biotechnology} 38:1079--1086. doi:
+\href{https://doi.org/10.1038/s41587-020-0501-8}{10.1038/s41587-020-0501-8}.
 
 \leavevmode\hypertarget{ref-Yarza2014}{}%
-23. Yarza P, Yilmaz P, Pruesse E, Glöckner FO, Ludwig W, Schleifer K-H,
+24. Yarza P, Yilmaz P, Pruesse E, Glöckner FO, Ludwig W, Schleifer K-H,
 Whitman WB, Euzéby J, Amann R, Rosselló-Móra R. 2014. Uniting the
 classification of cultured and uncultured bacteria and archaea using 16S
-rRNA gene sequences. Nature Reviews Microbiology 12:635--645
-\url{https://doi.org/10.1038/nrmicro3330}.
+rRNA gene sequences. \emph{Nature Reviews Microbiology} 12:635--645.
+doi: \href{https://doi.org/10.1038/nrmicro3330}{10.1038/nrmicro3330}.
 
 \leavevmode\hypertarget{ref-Schloss2009}{}%
-24. Schloss PD, Westcott SL, Ryabin T, Hall JR, Hartmann M, Hollister
+25. Schloss PD, Westcott SL, Ryabin T, Hall JR, Hartmann M, Hollister
 EB, Lesniewski RA, Oakley BB, Parks DH, Robinson CJ, Sahl JW, Stres B,
@@
 Open-source, platform-independent, community-supported software for
-describing and comparing microbial communities. Applied and
-Environmental Microbiology 75:7537--7541
-\url{https://doi.org/10.1128/aem.01541-09}.
+describing and comparing microbial communities. \emph{Applied and
+Environmental Microbiology} 75:7537--7541. doi:
+\href{https://doi.org/10.1128/aem.01541-09}{10.1128/aem.01541-09}.
 
 \leavevmode\hypertarget{ref-Quast2012}{}%
-25. Quast C, Pruesse E, Yilmaz P, Gerken J, Schweer T, Yarza P, Peplies
+26. Quast C, Pruesse E, Yilmaz P, Gerken J, Schweer T, Yarza P, Peplies
 J, Glöckner FO. 2012. The SILVA ribosomal RNA gene database project:
-Improved data processing and web-based tools. Nucleic Acids Research
-41:D590--D596 \url{https://doi.org/10.1093/nar/gks1219}.
+Improved data processing and web-based tools. \emph{Nucleic Acids
+Research} 41:D590--D596. doi:
+\href{https://doi.org/10.1093/nar/gks1219}{10.1093/nar/gks1219}.
 \end{cslreferences}
@@
 median distance threshold for each region of the 16S rRNA gene that is
-required for 95\% of the species with the indicated numbrer of
+required for 95\% of the genomes with the indicated numbrer of
 \emph{rrn} operons to cluster their ASVs to a single OTU. The median
@@
 increased, the fraction of genomes split into separate OTUs decreased
-while the fraction of species that were merged into the same OTU
-increases.} These data represent the median fractions for both
-measurements across 100 randomizations. In each randomization, one
-genome was sampled from each species.
+while the fraction of ASVs from genomes in different species that were
+merged into the same OTU increased.} These data represent the median
+fractions for both measurements across 100 randomizations. In each
+randomization, one genome was sampled from each species.
 
@@
 \textbf{Figure S1. The ratio of number of distinct ASVs per copy of the
-\emph{rrn} operon increased for a species as the number of genomes
-sampled increased.} Each point represents a different species and was
-shaded to be 80\% transparent so that when points overlap they become
-darker. The blue line represents a smoothed fit through the data.
+\emph{rrn} operon increased for a species as the number of genomes in
+the \emph{rrn}DB for that species increased.} Each point represents a
+different species and was shaded to be 80\% transparent so that when
+points overlap they become darker. The blue line represents a smoothed
+fit through the data.
 

### 2021-02-19 code/plot_lump_split.R

@@ -9,8 +9,8 @@ read_tsv("data/processed/lumped_split_rate.tsv",
 	select(-ends_with("iqr")) %>%
 	pivot_longer(cols=c(split_rate, lump_rate), names_to="method", values_to="fraction") %>%
 	mutate(method = ifelse(method == "lump_rate",
-												 "Fraction of genomes from\ndifferent species whose ASVs\nwere merged into the same cluster",
-												 "Fraction of genomes split\ninto separate clusters")) %>%
+												 "Fraction of genomes from\ndifferent species whose ASVs\nwere clustered together",
+												 "Fraction of genomes\nsplit into separate OTUs")) %>%
 	ggplot(aes(x=threshold, y=fraction, color=region)) +
 	geom_line(size=1) +
 	facet_wrap(~method, nrow=2, strip.position = "left") +
@@ -19,7 +19,7 @@ read_tsv("data/processed/lumped_split_rate.tsv",
 		labels=c("0", "2.5", "5", "7.5", "10")
 		) +
 	custom_color_scale() +
-	labs(x="Distance theshold\nto define ASV/OTU (%)", y=NULL) +
+	labs(x="Distance theshold\nused to define OTUs (%)", y=NULL) +
 	theme_classic() +
 	theme(
 		strip.placement="outside",

### 2021-02-22 submission/manuscript.Rmd
Third round of revisions

@@
 ```{r, echo=FALSE}
-library(tidyverse)
-library(knitr)
-library(here)
+suppressPackageStartupMessages(library(tidyverse))
+suppressPackageStartupMessages(library(knitr))
+suppressPackageStartupMessages(library(here))
 
@@
 ASVs have grown in popularity, in part, because of a desire to reflect a more refined level of taxonomy since they do not cluster sequences based on a distance-based threshold.
-However, ASVs and the use of overly narrow thresholds to identify OTUs increases the risk of splitting a single genome into separate clusters.
-I assessed this problem by analyzing the intragenomic variation of 16S rRNA genes from the bacterial genomes represented in a *rrn* copy number database, which contained `r nrow(metadata)` genomes from `r n_distinct(metadata$species)` species.
+However, ASVs and the use of overly narrow thresholds to identify OTUs increase the risk of splitting a single genome into separate clusters.
+To assess this risk, I analyzed the intragenomic variation of 16S rRNA genes from the bacterial genomes represented in a *rrn* copy number database, which contained `r nrow(metadata)` genomes from `r n_distinct(metadata$species)` species.
 As the number of copies of the 16S rRNA gene increased in a genome, the number of ASVs also increased.
-There was an average of `r rate_v19` ASVs per copy of the 16S rRNA gene, for full length 16S rRNA genes.
-It was necessary to use a distance threshold of `r format_pct(threshold_v19)`% to cluster full length ASVs from the same genome into a single OTU with **95% confidence**, for genomes with 7 copies of the 16S rRNA, such as *E.
+There was an average of `r rate_v19` ASVs per copy of the 16S rRNA gene for full length 16S rRNA genes.
+It was necessary to use a distance threshold of `r format_pct(threshold_v19)`% to cluster full length ASVs from the same genome into a single OTU with 95% confidence for genomes with 7 copies of the 16S rRNA, such as *E.
 coli*.
@@
 16S rRNA gene sequencing is a powerful technique for describing and comparing microbial communities [@Lane1985].
-Efforts to link 16S rRNA gene sequences to taxonomic levels based on distance thresholds go back to at least the 1990s.
+Efforts to link 16S rRNA gene sequences to taxonomic levels based on distance thresholds date to at least the 1990s.
 The distance-based thresholds that were developed and are now widely used (i.e., 3%) were based on DNA-DNA hybridization approaches that are not as precise as genome sequencing [@Stackebrandt1994; @Goris2007].
@@
 Approaches for identifying ASVs do not cluster sequences based on a distance-based threshold [@Callahan2017].
-Proponents of ASVs are largely dissmissive of concerns that most bacterial genomes have more than 1 copy of the *rrn* operon and those copies are not identical [@Pei2010; @Sun2013].
-Yet, using too fine a threshold to identify OTUs creates the risk of splitting a single genome into multiple bins.
+Proponents of ASVs are largely dissmissive of concerns that most bacterial genomes have more than one copy of the *rrn* operon and those copies are not identical [@Pei2010; @Sun2013].
+Yet, using too fine a threshold to identify OTUs creates the risk of splitting a single genome into multiple clusters.
 Conversely, using too broad of a threshold to define OTUs creates the risk of clustering together bacterial species into the same OTU.
@@
 
-To investigate the variation in the number of copies of the 16S rRNA gene per genome as well as the intragenomic variation among copies of the 16S rRNA gene, I obtained 16S rRNA sequences from the *rrn* copy number database (*rrn*DB)[@Stoddard2014].
+To investigate the variation in the number of copies of the 16S rRNA gene per genome and the intragenomic variation among copies of the 16S rRNA gene, I obtained 16S rRNA sequences from the *rrn* copy number database (*rrn*DB)[@Stoddard2014].
 Among the `r n_distinct(metadata$species)` species represented in the *rrn*DB there were `r nrow(metadata)` genomes.
@@
 Although a species tended to have a consistent number of 16S rRNA gene copies per genome, the number of total variants increased with the number of genomes that were sampled (**Figure 1**).
-For example, the `r single_copy_N` genome accessions of *`r single_copy$species`* in the *rrn*DB only had `r single_copy$median_rrn` copy of the gene per genome.
+For example, the `r single_copy_N` genome accessions of *`r single_copy$species`* in the *rrn*DB each had `r single_copy$median_rrn` copy of the gene per genome.
 However, across those accessions, there were `r single_nasvs` versions of the gene.
 An *E.
-coli* genome typically had `r ecoli_copy$median_rrn` copies of the 16S rRNA gene with a median `r ecoli_copy$median_asv` distinct full length ASVs per genome (intraquartile range between `r ecoli_copy$l_quartile_asv` and `r ecoli_copy$u_quartile_asv`).
+coli* genome typically had `r ecoli_copy$median_rrn` copies of the 16S rRNA gene with a median of `r ecoli_copy$median_asv` distinct full length ASVs per genome (intraquartile range between `r ecoli_copy$l_quartile_asv` and `r ecoli_copy$u_quartile_asv`).
 Across the `r ecoli_copy$N` *E.
@@
 Therefore, I assessed the impact of the distance threshold used to define clusters of 16S rRNA genes on the propensity to split a genome into separate clusters.
+To control for uneven representation of genomes across species, I randomly selected one genome from each species for the following analyses and repeated each randomization 100 times.
 I observed that as the *rrn* copy number increased, the distance threshold required to reduce the ASVs in each genome to a single OTU increased (Figure 1).
@@
 Consequently, these results demonstrate that broad thresholds must be used to avoid splitting different operons from the same genome into separate clusters.
-At broad thresholds ASVs from multiple species could be clustered into the same OTU (**Figure 2**).
-Using ASVs, `r format_pct(lump_v19)`% of the species shared a 16S rRNA gene sequence variant with another species when considering full length sequences and `r format_pct(lump_v34)`, `r format_pct(lump_v4)`, and `r format_pct(lump_v45)`% when considering the V3-V4, V4, and V4-V5 regions, respectively.
-At the commonly used 3% threshold, `r format_pct(lump3_v19)`% of the species shared an OTU with another speecies when considering full length sequences and `r format_pct(lump3_v34)`, `r format_pct(lump3_v4)`, and `r format_pct(lump3_v45)`% when considering the V3-V4, V4, and V4-V5 regions, respectively.
+At broad thresholds, 16S rRNA gene sequences from multiple species could be clustered into the same ASV or OTU (**Figure 2**).
+Using ASVs, `r format_pct(lump_v19)`% of the ASVs contained sequences from multiple species when considering full length sequences and `r format_pct(lump_v34)`, `r format_pct(lump_v4)`, and `r format_pct(lump_v45)`% when considering the V3-V4, V4, and V4-V5 regions, respectively.
+At the commonly used 3% threshold for defining OTUs, `r format_pct(lump3_v19)`% of the OTUs contained 16S rRNA gene sequences from multiple species when considering full length sequences and `r format_pct(lump3_v34)`, `r format_pct(lump3_v4)`, and `r format_pct(lump3_v45)`% when considering the V3-V4, V4, and V4-V5 regions, respectively.
 Considering that species designations are unevenly applied and reflect multiple human-imposed biases, the risk of splitting a genome into multiple OTUs more problematic than clustering species together.
@@
 
-The results of this analysis demonstrate that there is a significant risk of splitting a single genome into multiple clusters if too fine of a threshold is applied to defining an OTU.
+The results of this analysis demonstrate that there is a significant risk of splitting a single genome into multiple clusters if using ASVs or too fine of a threshold to define OTUs.
 An ongoing problem for amplicon-based studies is defining a meaningful taxonomic unit [@Callahan2017; @Schloss2011; @Johnson2019].
-Since there is no consensus definition for a biological species concept [@Staley2006; @Oren2013; @Sanford2021], microbiologists must accept that how we have named bacterial species is biased and that taxonomic rules are not applied in a consistent manner (e.g., [@Baltrus2016; @Sanford2021]).
+Since there is no consensus for a biologicaly definition of a bacterial species [@Staley2006; @Oren2013; @Sanford2021], microbiologists must accept that how bacterial species are named is biased and that taxonomic rules are not applied in a consistent manner (e.g., [@Baltrus2016; @Sanford2021]).
 This makes it impossible to fit a distance threshold to define an OTU definition that matches a set of species names [@Konstantinidis2005].
@@
 A distance-based definition of a taxonomic unit based on 16S rRNA gene or full genome sequences is, at best, operational and not grounded in biological theory [@Schloss2011; @Barco2020; @Parks2020; @Yarza2014].
-There is general agreement in bacterial systematics that to classify something to a bacterial species, phenotypic and genome sequence data are needed [@Staley2006; @Oren2013; @Baltrus2016; @Sanford2021].
-A short section of a bacterial genome simply cannot differentiate between species.
+There is general agreement in bacterial systematics that to classify an organism to a bacterial species, phenotypic and genome sequence data are needed [@Staley2006; @Oren2013; @Baltrus2016; @Sanford2021].
+A short sequence from a bacterial genome simply cannot differentiate between species.
 Moreover, it is difficult to defend a clustering approach that would split a single genome into multiple taxonomic units.
@@
 Instead, I mapped the RefSeq accession number for each genome in the database to obtain a single taxonomy for each genome.
-Because strain names were not consistently given to genomes across bacterial species, the strain level designations were disregarded.
+Because strain names were not consistently given to genomes across bacterial species, I disregarded the strain level designations.
 
@@
 
-**(iii) Controlling for uneven sampling of genomes by species.** Because of the uneven distribution of genome sequences across species, for the analysis of splitting genomes and clustering ASVs from different species, I randomly selected one genome from each species.
+**(iii) Clustering sequences into OTUs.** Pairwise distances between sequences were calculated using the dist.seqs command from mothur.
+The OptiClust algorithm as implemented in mothur was used to assign 16S rRNA gene sequences to OTUs [@Westcott2017].
+Distance thresholds between 0.5 and 10.0% in 0.5 percentage point increments were used to assign sequences to OTUs.
+
+**(iv) Controlling for uneven sampling of genomes by species.** Because of the uneven distribution of genome sequences across species, for the analysis of splitting genomes and clustering ASVs from different species, I randomly selected one genome from each species.
 The random selection was repeated 100 times.
@@
 
-**(iv) Reproducible data analysis.** The code to perform the analysis in this manuscript and its history are available as a git-based version control repository on GitHub (https://github.com/SchlossLab/Schloss_rrnAnalysis_mSphere_2021).
+**(v) Reproducible data analysis.** The code to perform the analysis in this manuscript and its history are available as a git-based version control repository on GitHub (https://github.com/SchlossLab/Schloss_rrnAnalysis_mSphere_2021).
 The analysis can be regenerated using a GNU Make-based workflow that made use of built-in bash tools (v.
@@
 
-**(v) Note on usage of ASV, OTU, and cluster.** I used "ASV" to denote the cluster of true 16S rRNA gene sequences that were identical to each other and "OTU" to denote the product of distance-based clustering of sequences.
+**(vi) Note on usage of ASV, OTU, and cluster.** I used "ASV" to denote the cluster of true 16S rRNA gene sequences that were identical to each other and "OTU" to denote the product of distance-based clustering of sequences.
 Although ASVs do represent a type of operational defition of a taxonomic unit and can be thought of as an OTU formed using a distance of zero, proponents of the ASV approach prefer to avoid the term OTU given the long history of OTUs being formed by distance-based clustering (https://github.com/benjjneb/dada2/issues/62; accessed 2021-02-19).
@@
 **Figure 2.
-As the distance threshold used to define an OTU increased, the fraction of genomes split into separate OTUs decreased while the fraction of ASVs from genomes in different species that were merged into the same OTU increased.** These data represent the median fractions for both measurements across 100 randomizations.
+As the distance threshold used to define an OTU increased, the fraction of genomes split into separate OTUs decreased while the fraction of ASVs and OTUs representing multiple species increased.** These data represent the median fractions for both measurements across 100 randomizations.
 In each randomization, one genome was sampled from each species.
@@
 The blue line represents a smoothed fit through the data.
+The axes both use a logarithmic scale (base 10).

### 2021-02-22 submission/manuscript.tex
Third round of revisions

@@
 sequences based on a distance-based threshold. However, ASVs and the use
-of overly narrow thresholds to identify OTUs increases the risk of
-splitting a single genome into separate clusters. I assessed this
-problem by analyzing the intragenomic variation of 16S rRNA genes from
-the bacterial genomes represented in a \emph{rrn} copy number database,
-which contained \textbf{15,614} genomes from \textbf{4,774} species. As
-the number of copies of the 16S rRNA gene increased in a genome, the
-number of ASVs also increased. There was an average of \textbf{0.60}
-ASVs per copy of the 16S rRNA gene, for full length 16S rRNA genes. It
-was necessary to use a distance threshold of \textbf{5.5}\% to cluster
-full length ASVs from the same genome into a single OTU with
-\textbf{95\% confidence}, for genomes with 7 copies of the 16S rRNA,
-such as \emph{E. coli}. This research highlights the risk of splitting a
-single bacterial genome into separate clusters when ASVs are used to
-analyze 16S rRNA gene sequence data. Although there is also a risk of
-clustering ASVs from different species into the same OTU when using
-broad distance thresholds, those risks are of less concern than
-artificially splitting a genome into separate ASVs and OTUs.
+of overly narrow thresholds to identify OTUs increase the risk of
+splitting a single genome into separate clusters. To assess this risk, I
+analyzed the intragenomic variation of 16S rRNA genes from the bacterial
+genomes represented in a \emph{rrn} copy number database, which
+contained \textbf{15,614} genomes from \textbf{4,774} species. As the
+number of copies of the 16S rRNA gene increased in a genome, the number
+of ASVs also increased. There was an average of \textbf{0.60} ASVs per
+copy of the 16S rRNA gene for full length 16S rRNA genes. It was
+necessary to use a distance threshold of \textbf{5.5}\% to cluster full
+length ASVs from the same genome into a single OTU with 95\% confidence
+for genomes with 7 copies of the 16S rRNA, such as \emph{E. coli}. This
+research highlights the risk of splitting a single bacterial genome into
+separate clusters when ASVs are used to analyze 16S rRNA gene sequence
+data. Although there is also a risk of clustering ASVs from different
+species into the same OTU when using broad distance thresholds, those
+risks are of less concern than artificially splitting a genome into
+separate ASVs and OTUs.
 
@@
 comparing microbial communities (1). Efforts to link 16S rRNA gene
-sequences to taxonomic levels based on distance thresholds go back to at
+sequences to taxonomic levels based on distance thresholds date to at
 least the 1990s. The distance-based thresholds that were developed and
@@
 on a distance-based threshold (11). Proponents of ASVs are largely
-dissmissive of concerns that most bacterial genomes have more than 1
+dissmissive of concerns that most bacterial genomes have more than one
 copy of the \emph{rrn} operon and those copies are not identical (12,
 13). Yet, using too fine a threshold to identify OTUs creates the risk
-of splitting a single genome into multiple bins. Conversely, using too
-broad of a threshold to define OTUs creates the risk of clustering
+of splitting a single genome into multiple clusters. Conversely, using
+too broad of a threshold to define OTUs creates the risk of clustering
 together bacterial species into the same OTU. An example of both is seen
@@
 To investigate the variation in the number of copies of the 16S rRNA
-gene per genome as well as the intragenomic variation among copies of
-the 16S rRNA gene, I obtained 16S rRNA sequences from the \emph{rrn}
-copy number database (\emph{rrn}DB)(14). Among the \textbf{4,774}
-species represented in the \emph{rrn}DB there were \textbf{15,614}
-genomes. The median \emph{rrn} copy number per species ranged between
-\textbf{1} (e.g., \textbf{\emph{Mycobacterium tuberculosis}}) and
-\textbf{19} (\textbf{\emph{Metabacillus litoralis}}). As the number of
-copies of the operon in a genome increased, the number of variants of
-the 16S rRNA gene in each genome also increased. On average, there were
-\textbf{0.60} variants per copy of the full length 16S rRNA gene and an
-average of \textbf{0.33}, \textbf{0.26}, and \textbf{0.27} variants when
+gene per genome and the intragenomic variation among copies of the 16S
+rRNA gene, I obtained 16S rRNA sequences from the \emph{rrn} copy number
+database (\emph{rrn}DB)(14). Among the \textbf{4,774} species
+represented in the \emph{rrn}DB there were \textbf{15,614} genomes. The
+median \emph{rrn} copy number per species ranged between \textbf{1}
+(e.g., \textbf{\emph{Mycobacterium tuberculosis}}) and \textbf{19}
+(\textbf{\emph{Metabacillus litoralis}}). As the number of copies of the
+operon in a genome increased, the number of variants of the 16S rRNA
+gene in each genome also increased. On average, there were \textbf{0.60}
+variants per copy of the full length 16S rRNA gene and an average of
+\textbf{0.33}, \textbf{0.26}, and \textbf{0.27} variants when
 considering the V3-V4, V4, and V4-V5 regions of the gene, respectively.
@@
 the \textbf{180} genome accessions of \textbf{\emph{Mycobacterium
-tuberculosis}} in the \emph{rrn}DB only had \textbf{1} copy of the gene
+tuberculosis}} in the \emph{rrn}DB each had \textbf{1} copy of the gene
 per genome. However, across those accessions, there were \textbf{11}
 versions of the gene. An \emph{E. coli} genome typically had \textbf{7}
-copies of the 16S rRNA gene with a median \textbf{5} distinct full
+copies of the 16S rRNA gene with a median of \textbf{5} distinct full
 length ASVs per genome (intraquartile range between \textbf{4} and
@@
 threshold used to define clusters of 16S rRNA genes on the propensity to
-split a genome into separate clusters. I observed that as the \emph{rrn}
-copy number increased, the distance threshold required to reduce the
-ASVs in each genome to a single OTU increased (Figure 1). Among species
-with 7 copies of the \emph{rrn} operon (e.g., \emph{E. coli}), a
-distance threshold of \textbf{5.5}\% was required to reduce full length
-ASVs into a single OTU for 95\% of the species. Similarly, thresholds of
-\textbf{4.0}, \textbf{2.5}, and \textbf{3.5}\% were required for the
-V3-V4, V4, and V4-V5 regions, respectively. But, if a 3\% distance
-threshold was used, then ASVs from genomes containing fewer than
-\textbf{5}, \textbf{6}, \textbf{8}, and \textbf{6} copies of the
-\emph{rrn} operon would reliably be clustered into a single OTU for ASVs
-from the V1-V9, V3-V4, V4, and V4-V5 regions, respectively.
-Consequently, these results demonstrate that broad thresholds must be
-used to avoid splitting different operons from the same genome into
-separate clusters. At broad thresholds ASVs from multiple species could
-be clustered into the same OTU (\textbf{Figure 2}). Using ASVs,
-\textbf{3.6}\% of the species shared a 16S rRNA gene sequence variant
-with another species when considering full length sequences and
-\textbf{10.2}, \textbf{14.9}, and \textbf{12.0}\% when considering the
-V3-V4, V4, and V4-V5 regions, respectively. At the commonly used 3\%
-threshold, \textbf{25.2}\% of the species shared an OTU with another
-speecies when considering full length sequences and \textbf{29.4},
-\textbf{33.0}, and \textbf{32.2}\% when considering the V3-V4, V4, and
-V4-V5 regions, respectively. Considering that species designations are
-unevenly applied and reflect multiple human-imposed biases, the risk of
-splitting a genome into multiple OTUs more problematic than clustering
-species together. Therefore, larger thresholds are advisable.
+split a genome into separate clusters. To control for uneven
+representation of genomes across species, I randomly selected one genome
+from each species for the following analyses and repeated each
+randomization 100 times. I observed that as the \emph{rrn} copy number
+increased, the distance threshold required to reduce the ASVs in each
+genome to a single OTU increased (Figure 1). Among species with 7 copies
+of the \emph{rrn} operon (e.g., \emph{E. coli}), a distance threshold of
+\textbf{5.5}\% was required to reduce full length ASVs into a single OTU
+for 95\% of the species. Similarly, thresholds of \textbf{4.0},
+\textbf{2.5}, and \textbf{3.5}\% were required for the V3-V4, V4, and
+V4-V5 regions, respectively. But, if a 3\% distance threshold was used,
+then ASVs from genomes containing fewer than \textbf{5}, \textbf{6},
+\textbf{8}, and \textbf{6} copies of the \emph{rrn} operon would
+reliably be clustered into a single OTU for ASVs from the V1-V9, V3-V4,
+V4, and V4-V5 regions, respectively. Consequently, these results
+demonstrate that broad thresholds must be used to avoid splitting
+different operons from the same genome into separate clusters. At broad
+thresholds, 16S rRNA gene sequences from multiple species could be
+clustered into the same ASV or OTU (\textbf{Figure 2}). Using ASVs,
+\textbf{3.6}\% of the ASVs contained sequences from multiple species
+when considering full length sequences and \textbf{10.2}, \textbf{14.9},
+and \textbf{12.0}\% when considering the V3-V4, V4, and V4-V5 regions,
+respectively. At the commonly used 3\% threshold for defining OTUs,
+\textbf{25.2}\% of the OTUs contained 16S rRNA gene sequences from
+multiple species when considering full length sequences and
+\textbf{29.4}, \textbf{33.0}, and \textbf{32.2}\% when considering the
+V3-V4, V4, and V4-V5 regions, respectively. Considering that species
+designations are unevenly applied and reflect multiple human-imposed
+biases, the risk of splitting a genome into multiple OTUs more
+problematic than clustering species together. Therefore, larger
+thresholds are advisable.
 
 The results of this analysis demonstrate that there is a significant
-risk of splitting a single genome into multiple clusters if too fine of
-a threshold is applied to defining an OTU. An ongoing problem for
+risk of splitting a single genome into multiple clusters if using ASVs
+or too fine of a threshold to define OTUs. An ongoing problem for
 amplicon-based studies is defining a meaningful taxonomic unit (11, 15,
-16). Since there is no consensus definition for a biological species
-concept (17--19), microbiologists must accept that how we have named
-bacterial species is biased and that taxonomic rules are not applied in
-a consistent manner (e.g., (19, 20)). This makes it impossible to fit a
-distance threshold to define an OTU definition that matches a set of
-species names (21). Furthermore, the 16S rRNA gene does not evolve at
-the same rate across all bacterial lineages (15), which limits the
-biological interpretation of a common OTU definition. A distance-based
-definition of a taxonomic unit based on 16S rRNA gene or full genome
-sequences is, at best, operational and not grounded in biological theory
-(15, 22--24). There is general agreement in bacterial systematics that
-to classify something to a bacterial species, phenotypic and genome
-sequence data are needed (17--20). A short section of a bacterial genome
-simply cannot differentiate between species. Moreover, it is difficult
-to defend a clustering approach that would split a single genome into
-multiple taxonomic units. It is not biologically plausible to entertain
-the possibility that parts of a genome would have different ecologies.
-Although there are multiple reasons that proponents favor ASVs, the
-significant risk of artificially splitting genomes into separate
-clusters is too high to warrant their use.
+16). Since there is no consensus for a biologicaly definition of a
+bacterial species (17--19), microbiologists must accept that how
+bacterial species are named is biased and that taxonomic rules are not
+applied in a consistent manner (e.g., (19, 20)). This makes it
+impossible to fit a distance threshold to define an OTU definition that
+matches a set of species names (21). Furthermore, the 16S rRNA gene does
+not evolve at the same rate across all bacterial lineages (15), which
+limits the biological interpretation of a common OTU definition. A
+distance-based definition of a taxonomic unit based on 16S rRNA gene or
+full genome sequences is, at best, operational and not grounded in
+biological theory (15, 22--24). There is general agreement in bacterial
+systematics that to classify an organism to a bacterial species,
+phenotypic and genome sequence data are needed (17--20). A short
+sequence from a bacterial genome simply cannot differentiate between
+species. Moreover, it is difficult to defend a clustering approach that
+would split a single genome into multiple taxonomic units. It is not
+biologically plausible to entertain the possibility that parts of a
+genome would have different ecologies. Although there are multiple
+reasons that proponents favor ASVs, the significant risk of artificially
+splitting genomes into separate clusters is too high to warrant their
+use.
 
@@
 obtain a single taxonomy for each genome. Because strain names were not
-consistently given to genomes across bacterial species, the strain level
-designations were disregarded.
+consistently given to genomes across bacterial species, I disregarded
+the strain level designations.
 
@@
 
-\textbf{(iii) Controlling for uneven sampling of genomes by species.}
+\textbf{(iii) Clustering sequences into OTUs.} Pairwise distances
+between sequences were calculated using the dist.seqs command from
+mothur. The OptiClust algorithm as implemented in mothur was used to
+assign 16S rRNA gene sequences to OTUs (27). Distance thresholds between
+0.5 and 10.0\% in 0.5 percentage point increments were used to assign
+sequences to OTUs.
+
+\textbf{(iv) Controlling for uneven sampling of genomes by species.}
 Because of the uneven distribution of genome sequences across species,
@@
 
-\textbf{(iv) Reproducible data analysis.} The code to perform the
+\textbf{(v) Reproducible data analysis.} The code to perform the
 analysis in this manuscript and its history are available as a git-based
@@
 made use of built-in bash tools (v. 3.2.57), mothur (v. 1.44.2), and R
-(v. \textbf{4.0.3}). Within R, I used the tidyverse (v. \textbf{1.3.0}),
+(v. \textbf{4.0.4}). Within R, I used the tidyverse (v. \textbf{1.3.0}),
 data.table (v. \textbf{1.13.2}), Rcpp (v. \textbf{1.0.5}), furrr (v.
@@
 
-\textbf{(v) Note on usage of ASV, OTU, and cluster.} I used ``ASV'' to
+\textbf{(vi) Note on usage of ASV, OTU, and cluster.} I used ``ASV'' to
 denote the cluster of true 16S rRNA gene sequences that were identical
@@
 \href{https://doi.org/10.1093/nar/gks1219}{10.1093/nar/gks1219}.
+
+\leavevmode\hypertarget{ref-Westcott2017}{}%
+27. Westcott SL, Schloss PD. 2017. OptiClust, an improved method for
+assigning amplicon-based sequence data to operational taxonomic units.
+\emph{mSphere} 2:00073--17. doi:
+\href{https://doi.org/10.1128/mspheredirect.00073-17}{10.1128/mspheredirect.00073-17}.
 \end{cslreferences}
@@
 increased, the fraction of genomes split into separate OTUs decreased
-while the fraction of ASVs from genomes in different species that were
-merged into the same OTU increased.} These data represent the median
-fractions for both measurements across 100 randomizations. In each
-randomization, one genome was sampled from each species.
+while the fraction of ASVs and OTUs representing multiple species
+increased.} These data represent the median fractions for both
+measurements across 100 randomizations. In each randomization, one
+genome was sampled from each species.
 
@@
 points overlap they become darker. The blue line represents a smoothed
-fit through the data.
+fit through the data. The axes both use a logarithmic scale (base 10).
 

### 2021-02-22 code/plot_lump_split.R

@@ -8,9 +8,13 @@ read_tsv("data/processed/lumped_split_rate.tsv",
 				 								 .default = col_double())) %>%
 	select(-ends_with("iqr")) %>%
 	pivot_longer(cols=c(split_rate, lump_rate), names_to="method", values_to="fraction") %>%
-	mutate(method = ifelse(method == "lump_rate",
-												 "Fraction of genomes from\ndifferent species whose ASVs\nwere clustered together",
-												 "Fraction of genomes\nsplit into separate OTUs")) %>%
+	mutate(method = factor(ifelse(method == "lump_rate",
+												 "Fraction of ASVs or OTUs\nthat included sequences\nfrom multiple species",
+												 "Fraction of genomes\nsplit into separate OTUs"),
+											 levels=c("Fraction of genomes\nsplit into separate OTUs",
+											 "Fraction of ASVs or OTUs\nthat included sequences\nfrom multiple species")
+									 )
+								 ) %>%
 	ggplot(aes(x=threshold, y=fraction, color=region)) +
 	geom_line(size=1) +
 	facet_wrap(~method, nrow=2, strip.position = "left") +

### 2021-02-22 submission/manuscript.Rmd
Fourth round of revisions

@@
 single_copy <- rrn_copy_number %>% filter(median_rrn == 1) %>% top_n(n=1, N)
+
+stopifnot(single_copy$min_rrn == single_copy$max_rrn)
+
 max_copy <- rrn_copy_number %>% top_n(n=1, median_rrn)
@@
 Efforts to link 16S rRNA gene sequences to taxonomic levels based on distance thresholds date to at least the 1990s.
-The distance-based thresholds that were developed and are now widely used (i.e., 3%) were based on DNA-DNA hybridization approaches that are not as precise as genome sequencing [@Stackebrandt1994; @Goris2007].
+The distance-based threshold that was developed and is now widely used was based on DNA-DNA hybridization approaches that are not as precise as genome sequencing [@Stackebrandt1994; @Goris2007].
 Instead, genome sequencing technologies have suggested that the widely used 3% distance threshold to operationally define bacterial taxa is too coarse [@RodriguezR2018; @Stackebrandt2006; @Edgar2018].
@@
 Approaches for identifying ASVs do not cluster sequences based on a distance-based threshold [@Callahan2017].
-Proponents of ASVs are largely dissmissive of concerns that most bacterial genomes have more than one copy of the *rrn* operon and those copies are not identical [@Pei2010; @Sun2013].
+Proponents of ASVs are largely dissmissive of concerns that most bacterial genomes have more than one copy of the *rrn* operon and that those copies are not identical [@Pei2010; @Sun2013].
 Yet, using too fine a threshold to identify OTUs creates the risk of splitting a single genome into multiple clusters.
@@
 Conversely, if the copies were clustered using a 3% distance threshold, then all 10 ASVs would cluster into the same OTU.
-The goal of this study was to quantify the risk of splitting a single genome into multiple clusters and the risk of clustering different bacterial species into the same cluster when using ASVs and various OTU definitions.
+The goal of this study was to quantify the tradeoff of splitting a single genome into multiple clusters and the risk of clustering different bacterial species into the same cluster when using ASVs and various OTU definitions.
 
@@
 The median *rrn* copy number per species ranged between `r single_copy$median_rrn` (e.g., *`r single_copy$species`*) and `r max_copy$median_rrn` (*`r max_copy$species`*).
-As the number of copies of the operon in a genome increased, the number of variants of the 16S rRNA gene in each genome also increased.
+As **rrn** copy number for a genome increased, the number of variants of the 16S rRNA gene in each genome also increased.
 On average, there were `r rate_v19` variants per copy of the full length 16S rRNA gene and an average of `r rate_v34`, `r rate_v4`, and `r rate_v45` variants when considering the V3-V4, V4, and V4-V5 regions of the gene, respectively.
@@
 Among species with 7 copies of the *rrn* operon (e.g., *E.
-coli*), a distance threshold of `r format_pct(threshold_v19)`% was required to reduce full length ASVs into a single OTU for 95% of the species.
+coli*), a distance threshold of `r format_pct(threshold_v19)`% was required to reduce full length ASVs into a single OTU for 95% of the species (Figure 2).
 Similarly, thresholds of `r format_pct(threshold_v34)`, `r format_pct(threshold_v4)`, and `r format_pct(threshold_v45)`% were required for the V3-V4, V4, and V4-V5 regions, respectively.
-But, if a 3% distance threshold was used, then ASVs from genomes containing fewer than `r three_v19`, `r three_v34`, `r three_v4`, and `r three_v45` copies of the *rrn* operon would reliably be clustered into a single OTU for ASVs from the V1-V9, V3-V4, V4, and V4-V5 regions, respectively.
+But, if a 3% distance threshold was used, then ASVs from genomes containing fewer than `r three_v19`, `r three_v34`, `r three_v4`, and `r three_v45` copies of the *rrn* operon would reliably be clustered into a single OTU for ASVs from the V1-V9, V3-V4, V4, and V4-V5 regions, respectively (Figure 2).
 Consequently, these results demonstrate that broad thresholds must be used to avoid splitting different operons from the same genome into separate clusters.
@@
 At the commonly used 3% threshold for defining OTUs, `r format_pct(lump3_v19)`% of the OTUs contained 16S rRNA gene sequences from multiple species when considering full length sequences and `r format_pct(lump3_v34)`, `r format_pct(lump3_v4)`, and `r format_pct(lump3_v45)`% when considering the V3-V4, V4, and V4-V5 regions, respectively.
-Considering that species designations are unevenly applied and reflect multiple human-imposed biases, the risk of splitting a genome into multiple OTUs more problematic than clustering species together.
+Considering that species designations are unevenly applied and reflect multiple human-imposed biases, the risk of splitting a genome into multiple OTUs is more problematic than clustering species together.
 Therefore, larger thresholds are advisable.
@@
 Although ASVs do represent a type of operational defition of a taxonomic unit and can be thought of as an OTU formed using a distance of zero, proponents of the ASV approach prefer to avoid the term OTU given the long history of OTUs being formed by distance-based clustering (https://github.com/benjjneb/dada2/issues/62; accessed 2021-02-19).
-For this reason, when an ASV splits a genome into different units those units are called clusters rather than OTUs.
+For this reason, when an ASV splits a genome into different units, those units were called clusters rather than OTUs.
 
@@
 I am also grateful to community members who watched the serialized version of this analysis on YouTube and provided suggestions and questions over the course of the development of this project.
-This work was supported, in part, through grants from the NIH to PDS (P30DK034933, U01AI124255, and R01CA215574).
+This work was supported, in part, through grants from the NIH (P30DK034933, U01AI124255, and R01CA215574).
 
@@
 The blue line represents a smoothed fit through the data.
-The axes both use a logarithmic scale (base 10).
+Both axes use a logarithmic scale (base 10).

### 2021-02-22 submission/manuscript.tex
Fourth round of revisions

@@
 sequences to taxonomic levels based on distance thresholds date to at
-least the 1990s. The distance-based thresholds that were developed and
-are now widely used (i.e., 3\%) were based on DNA-DNA hybridization
-approaches that are not as precise as genome sequencing (2, 3). Instead,
-genome sequencing technologies have suggested that the widely used 3\%
-distance threshold to operationally define bacterial taxa is too coarse
-(4--6). As an alternative to OTUs, amplicon sequencing variants (ASVs)
-have been proposed as a way to adopt the thresholds suggested by genome
-sequencing to microbial community analysis using 16S rRNA gene sequences
-(7--10). Approaches for identifying ASVs do not cluster sequences based
-on a distance-based threshold (11). Proponents of ASVs are largely
+least the 1990s. The distance-based threshold that was developed and is
+now widely used was based on DNA-DNA hybridization approaches that are
+not as precise as genome sequencing (2, 3). Instead, genome sequencing
+technologies have suggested that the widely used 3\% distance threshold
+to operationally define bacterial taxa is too coarse (4--6). As an
+alternative to OTUs, amplicon sequencing variants (ASVs) have been
+proposed as a way to adopt the thresholds suggested by genome sequencing
+to microbial community analysis using 16S rRNA gene sequences (7--10).
+Approaches for identifying ASVs do not cluster sequences based on a
+distance-based threshold (11). Proponents of ASVs are largely
 dissmissive of concerns that most bacterial genomes have more than one
-copy of the \emph{rrn} operon and those copies are not identical (12,
-13). Yet, using too fine a threshold to identify OTUs creates the risk
-of splitting a single genome into multiple clusters. Conversely, using
-too broad of a threshold to define OTUs creates the risk of clustering
-together bacterial species into the same OTU. An example of both is seen
-in the comparison of \emph{Staphylococcus aureus} (NCTC 8325) and
-\emph{S. epidermidis} (ATCC 12228) where each genome has 5 copies of the
-16S rRNA gene. Each of the 10 copies of the 16S rRNA gene in these two
-genomes is distinct and represent 10 ASVs. Conversely, if the copies
-were clustered using a 3\% distance threshold, then all 10 ASVs would
-cluster into the same OTU. The goal of this study was to quantify the
-risk of splitting a single genome into multiple clusters and the risk of
-clustering different bacterial species into the same cluster when using
-ASVs and various OTU definitions.
+copy of the \emph{rrn} operon and that those copies are not identical
+(12, 13). Yet, using too fine a threshold to identify OTUs creates the
+risk of splitting a single genome into multiple clusters. Conversely,
+using too broad of a threshold to define OTUs creates the risk of
+clustering together bacterial species into the same OTU. An example of
+both is seen in the comparison of \emph{Staphylococcus aureus} (NCTC
+8325) and \emph{S. epidermidis} (ATCC 12228) where each genome has 5
+copies of the 16S rRNA gene. Each of the 10 copies of the 16S rRNA gene
+in these two genomes is distinct and represent 10 ASVs. Conversely, if
+the copies were clustered using a 3\% distance threshold, then all 10
+ASVs would cluster into the same OTU. The goal of this study was to
+quantify the tradeoff of splitting a single genome into multiple
+clusters and the risk of clustering different bacterial species into the
+same cluster when using ASVs and various OTU definitions.
 
@@
 (e.g., \textbf{\emph{Mycobacterium tuberculosis}}) and \textbf{19}
-(\textbf{\emph{Metabacillus litoralis}}). As the number of copies of the
-operon in a genome increased, the number of variants of the 16S rRNA
-gene in each genome also increased. On average, there were \textbf{0.60}
+(\textbf{\emph{Metabacillus litoralis}}). As \textbf{rrn} copy number
+for a genome increased, the number of variants of the 16S rRNA gene in
+each genome also increased. On average, there were \textbf{0.60}
 variants per copy of the full length 16S rRNA gene and an average of
@@
 \textbf{5.5}\% was required to reduce full length ASVs into a single OTU
-for 95\% of the species. Similarly, thresholds of \textbf{4.0},
-\textbf{2.5}, and \textbf{3.5}\% were required for the V3-V4, V4, and
-V4-V5 regions, respectively. But, if a 3\% distance threshold was used,
-then ASVs from genomes containing fewer than \textbf{5}, \textbf{6},
-\textbf{8}, and \textbf{6} copies of the \emph{rrn} operon would
-reliably be clustered into a single OTU for ASVs from the V1-V9, V3-V4,
-V4, and V4-V5 regions, respectively. Consequently, these results
-demonstrate that broad thresholds must be used to avoid splitting
-different operons from the same genome into separate clusters. At broad
-thresholds, 16S rRNA gene sequences from multiple species could be
-clustered into the same ASV or OTU (\textbf{Figure 2}). Using ASVs,
-\textbf{3.6}\% of the ASVs contained sequences from multiple species
-when considering full length sequences and \textbf{10.2}, \textbf{14.9},
-and \textbf{12.0}\% when considering the V3-V4, V4, and V4-V5 regions,
-respectively. At the commonly used 3\% threshold for defining OTUs,
-\textbf{25.2}\% of the OTUs contained 16S rRNA gene sequences from
-multiple species when considering full length sequences and
-\textbf{29.4}, \textbf{33.0}, and \textbf{32.2}\% when considering the
-V3-V4, V4, and V4-V5 regions, respectively. Considering that species
-designations are unevenly applied and reflect multiple human-imposed
-biases, the risk of splitting a genome into multiple OTUs more
-problematic than clustering species together. Therefore, larger
+for 95\% of the species (Figure 2). Similarly, thresholds of
+\textbf{4.0}, \textbf{2.5}, and \textbf{3.5}\% were required for the
+V3-V4, V4, and V4-V5 regions, respectively. But, if a 3\% distance
+threshold was used, then ASVs from genomes containing fewer than
+\textbf{5}, \textbf{6}, \textbf{8}, and \textbf{6} copies of the
+\emph{rrn} operon would reliably be clustered into a single OTU for ASVs
+from the V1-V9, V3-V4, V4, and V4-V5 regions, respectively (Figure 2).
+Consequently, these results demonstrate that broad thresholds must be
+used to avoid splitting different operons from the same genome into
+separate clusters. At broad thresholds, 16S rRNA gene sequences from
+multiple species could be clustered into the same ASV or OTU
+(\textbf{Figure 2}). Using ASVs, \textbf{3.6}\% of the ASVs contained
+sequences from multiple species when considering full length sequences
+and \textbf{10.2}, \textbf{14.9}, and \textbf{12.0}\% when considering
+the V3-V4, V4, and V4-V5 regions, respectively. At the commonly used 3\%
+threshold for defining OTUs, \textbf{25.2}\% of the OTUs contained 16S
+rRNA gene sequences from multiple species when considering full length
+sequences and \textbf{29.4}, \textbf{33.0}, and \textbf{32.2}\% when
+considering the V3-V4, V4, and V4-V5 regions, respectively. Considering
+that species designations are unevenly applied and reflect multiple
+human-imposed biases, the risk of splitting a genome into multiple OTUs
+is more problematic than clustering species together. Therefore, larger
 thresholds are advisable.
@@
 2021-02-19). For this reason, when an ASV splits a genome into different
-units those units are called clusters rather than OTUs.
+units, those units were called clusters rather than OTUs.
 
@@
 development of this project. This work was supported, in part, through
-grants from the NIH to PDS (P30DK034933, U01AI124255, and R01CA215574).
+grants from the NIH (P30DK034933, U01AI124255, and R01CA215574).
 
@@
 points overlap they become darker. The blue line represents a smoothed
-fit through the data. The axes both use a logarithmic scale (base 10).
+fit through the data. Both axes use a logarithmic scale (base 10).
 

### 2021-02-23 README.Rmd
Use Rcpp to speed up all v all comparison
Deploy generation of confusion matrix data for 100 iterations
Update rrnDB database and NCBI taxonomy

@@
 ---
-title: "Code Club Project: Assessing whether intra and inter-genomic variation hinder utility of ESVs and ASVs"
+title: "Amplicon sequence variants artificially split bacterial genomes into separate clusters"
 author: "Pat Schloss"
-date: "9/11/2020"
+date: "2/22/2021"
 output:
@@
 
-Developed over a series of *Code Club* episodes led by Pat Schloss to answer an important question in microbiology and develop comfort using tools to develop reproducible research practices.
+```{r, echo=FALSE}
+suppressPackageStartupMessages(library(tidyverse))
+suppressPackageStartupMessages(library(knitr))
+suppressPackageStartupMessages(library(here))
 
-## Questions
-* Within a genome, how many distinct sequences of the 16S rRNA gene are present relative to the number of copies per genome?
-How far apart are these sequences from each other?
-How does this scale from a genome to kingdoms?
-* Within a taxa (any level), how many ESVs or ASVs from that taxa are shared with sister taxa? How does this change with taxonomic level? Variable region?
+opts_chunk$set(echo = FALSE)
+
+inline_hook <- function(x){
+
+	if(is.numeric(x)){
+		if(abs(x - round(x)) < .Machine$double.eps){
+			# treat as integer
+			formatted <- format(x, digits=0, big.mark = ",")
+
+		} else {
+			# treat as floating point number
+			formatted <- format(x, digits=2, nsmall=2, big.mark = ",")
+		}
+	} else {
+		formatted <-x
+	}
+
+	paste0("**", formatted, "**")
+
+}
+
+knit_hooks$set(inline=inline_hook)
+
+format_pct <- function(x, my_digits=1){
+
+	format(x, digits=my_digits, nsmall=my_digits)
+
+}
+```
+
+```{r}
+## Paragraph 1
+metadata <- read_tsv(here("data/references/genome_id_taxonomy.tsv"),
+										 col_types = cols(.default = col_character()))
+
+easv <- read_tsv(here("data/processed/rrnDB.easv.count_tibble"),
+								col_types = cols(.default = col_character(),
+																 count = col_integer()))
+
+metadata_easv <- inner_join(metadata, easv, by=c("genome_id" = "genome")) %>%
+	mutate(threshold = recode(threshold, "esv" = "0.000"),
+				 threshold = as.numeric(threshold))
+
+rrn_copy_number <- metadata_easv %>%
+	filter(region == "v19", threshold == 0) %>%
+	group_by(species, genome_id) %>%
+	summarize(n_rrn = sum(count), n_asv = n(), .groups="drop") %>%
+	group_by(species) %>%
+	summarize(min_rrn = min(n_rrn),
+            l_quartile_rrn = quantile(n_rrn, prob=0.25),
+            median_rrn = median(n_rrn),
+            u_quartile_rrn = quantile(n_rrn, prob=0.75),
+            max_rrn = max(n_rrn),
+            min_asv = min(n_asv),
+            l_quartile_asv = quantile(n_asv, prob=0.25),
+            median_asv = median(n_asv),
+            u_quartile_asv = quantile(n_asv, prob=0.75),
+            max_asv = max(n_asv),
+            N=n(),
+            .groups="drop")
+
+single_copy <- rrn_copy_number %>% filter(median_rrn == 1) %>% top_n(n=1, N)
+
+stopifnot(single_copy$min_rrn == single_copy$max_rrn)
+
+max_copy <- rrn_copy_number %>% top_n(n=1, median_rrn)
+
+single_copy_N <- single_copy %>% pull(N)
+#max_copy_N <- max_copy %>% pull(N)
+
+ecoli_copy <- rrn_copy_number %>% filter(species == "Escherichia coli")
+
+rates <- metadata_easv %>%
+	filter(threshold == 0) %>%
+	group_by(region, genome_id) %>%
+	summarize(n_rrns = sum(count), n_asvs = n_distinct(easv), .groups="drop") %>%
+	group_by(region) %>%
+	summarize(rate = sum(n_asvs)/sum(n_rrns), .groups="drop")
+
+rate_v19 <- rates %>% filter(region == "v19") %>% pull(rate)
+rate_v4 <- rates %>% filter(region == "v4") %>% pull(rate)
+rate_v34 <- rates %>% filter(region == "v34") %>% pull(rate)
+rate_v45 <- rates %>% filter(region == "v45") %>% pull(rate)
+
+asvs_per_species <- metadata_easv %>%
+	filter(threshold == 0 & region == "v19") %>%
+	group_by(species) %>%
+	summarize(n_asvs = n_distinct(easv), .groups="drop")
+
+#below are total number of ASVs across all genomes (above was within a genome)
+single_nasvs <- asvs_per_species %>% filter(species == single_copy$species) %>% pull(n_asvs)
+ecoli_nasvs <- asvs_per_species %>% filter(species == "Escherichia coli") %>% pull(n_asvs)
+
+## Paragraph 2
+thresholds <- read_tsv(here("data/processed/thresholds_for_single_otu.tsv"),
+											 col_types=cols(region = col_character(),
+											 							 .default = col_double())) %>%
+								filter(n_genomes > 100)
+
+threshold_v19 <- thresholds %>% filter(n_rrns == 7 & region == "v19") %>% pull(threshold) * 100
+threshold_v4 <- thresholds %>% filter(n_rrns == 7 & region == "v4") %>% pull(threshold) * 100
+threshold_v34 <- thresholds %>% filter(n_rrns == 7 & region == "v34") %>% pull(threshold) * 100
+threshold_v45 <- thresholds %>% filter(n_rrns == 7 & region == "v45") %>% pull(threshold) * 100
+
+three_v19 <- thresholds %>% filter(threshold <= 0.03 & region == "v19") %>% tail(n=1) %>% pull(n_rrns)
+three_v4 <- thresholds %>% filter(threshold <= 0.03 & region == "v4") %>% tail(n=1) %>% pull(n_rrns)
+three_v34 <- thresholds %>% filter(threshold <= 0.03 & region == "v34") %>% tail(n=1) %>% pull(n_rrns)
+three_v45 <- thresholds %>% filter(threshold <= 0.03 & region == "v45") %>% tail(n=1) %>% pull(n_rrns)
+
+## Paragraph 3
+lump_split <- read_tsv(here("data/processed/lumped_split_rate.tsv"),
+											 col_types = cols(region = col_character(),
+											 								 .default = col_double()))
+
+lump_v19 <- lump_split %>% filter(threshold ==0 &region == "v19") %>% pull(lump_rate) * 100
+lump_v4 <- lump_split %>% filter(threshold ==0 &region == "v4") %>% pull(lump_rate) * 100
+lump_v34 <- lump_split %>% filter(threshold ==0 &region == "v34") %>% pull(lump_rate) * 100
+lump_v45 <- lump_split %>% filter(threshold ==0 &region == "v45") %>% pull(lump_rate) * 100
+
+lump3_v19 <- lump_split %>% filter(threshold ==0.03 & region == "v19") %>% pull(lump_rate) * 100
+lump3_v4 <- lump_split %>% filter(threshold ==0.03 & region == "v4") %>% pull(lump_rate) * 100
+lump3_v34 <- lump_split %>% filter(threshold ==0.03 & region == "v34") %>% pull(lump_rate) * 100
+lump3_v45 <- lump_split %>% filter(threshold ==0.03 & region == "v45") %>% pull(lump_rate) * 100
+
+lump_split_iqr <- lump_split %>% pivot_longer(cols=c(split_iqr, lump_iqr), names_to="test", values_to="iqr") %>% summarize(max(iqr)) %>% pull
+
+```
+
+## Abstract
+
+Amplicon sequencing variants (ASVs) have been proposed as an alternative to operational taxonomic units (OTUs) for analyzing microbial communities.
+ASVs have grown in popularity, in part, because of a desire to reflect a more refined level of taxonomy since they do not cluster sequences based on a distance-based threshold.
+However, ASVs and the use of overly narrow thresholds to identify OTUs increase the risk of splitting a single genome into separate clusters.
+To assess this risk, I analyzed the intragenomic variation of 16S rRNA genes from the bacterial genomes represented in a *rrn* copy number database, which contained `r nrow(metadata)` genomes from `r n_distinct(metadata$species)` species.
+As the number of copies of the 16S rRNA gene increased in a genome, the number of ASVs also increased.
+There was an average of `r rate_v19` ASVs per copy of the 16S rRNA gene for full length 16S rRNA genes.
+It was necessary to use a distance threshold of `r format_pct(threshold_v19)`% to cluster full length ASVs from the same genome into a single OTU with 95% confidence for genomes with 7 copies of the 16S rRNA, such as *E.
+coli*.
+This research highlights the risk of splitting a single bacterial genome into separate clusters when ASVs are used to analyze 16S rRNA gene sequence data.
+Although there is also a risk of clustering ASVs from different species into the same OTU when using broad distance thresholds, those risks are of less concern than artificially splitting a genome into separate ASVs and OTUs.
+
+## Importance
+
+16S rRNA gene sequencing has engendered significant interest in studying microbial communities.
+There has been a tension between trying to classify 16S rRNA gene sequences to increasingly lower taxonomic levels and the reality that those levels were defined using more sequence and physiological information than is available from a fragment of the 16S rRNA gene.
+Furthermore, naming of bacterial taxa reflects the biases of those who name them.
+One motivation for the recent push to adopt ASVs in place of OTUs in microbial community analyses is to allow researchers to perform their analyes at the finest possible level that reflects species-level taxonomy.
+The current research is significant because it quantifies the risk of artificially splitting bacterial genomes into separate clusters.
+Far from providing a better represenation of bacterial taxonomy and biology, the ASV approach can lead to conflicting inferences about the ecology of different ASVs from the same genome.
 
@@
 library(rmarkdown)
+library(Rcpp)
+library(furrr)
 ```
@@
   * `tidyverse` (v. `r packageVersion("tidyverse")`)
+  * `Rcpp` (v. `r packageVersion("Rcpp")`)
+  * `furrr` (v. `r packageVersion("furrr")`)
   * `data.table` (v. `r packageVersion("data.table")`)

### 2021-02-23 submission/manuscript.Rmd
Update rrnDB database and NCBI taxonomy

@@
 
-format_pct <- function(x, my_digits=1){
+format_pct <- function(x, my_digits=2){
 
@@
 **Materials and Methods.
-(i) Data availability.** The 16S rRNA gene sequences used in this study were obtained from the *rrn*DB (https://rrndb.umms.med.umich.edu; version 5.6, released November 8, 2019) [@Stoddard2014].
+(i) Data availability.** The 16S rRNA gene sequences used in this study were obtained from the *rrn*DB (https://rrndb.umms.med.umich.edu; version 5.7, released January 18, 2021) [@Stoddard2014].
 At the time of submission, this was the most current version of the database.
@@
 The OptiClust algorithm as implemented in mothur was used to assign 16S rRNA gene sequences to OTUs [@Westcott2017].
-Distance thresholds between 0.5 and 10.0% in 0.5 percentage point increments were used to assign sequences to OTUs.
+Distance thresholds between 0.25 and 10.0% in 0.25 percentage point increments were used to assign sequences to OTUs.
 

### 2021-02-23 submission/manuscript.tex
Update rrnDB database and NCBI taxonomy

@@
 genomes represented in a \emph{rrn} copy number database, which
-contained \textbf{15,614} genomes from \textbf{4,774} species. As the
+contained \textbf{20,427} genomes from \textbf{5,972} species. As the
 number of copies of the 16S rRNA gene increased in a genome, the number
-of ASVs also increased. There was an average of \textbf{0.60} ASVs per
+of ASVs also increased. There was an average of \textbf{0.58} ASVs per
 copy of the 16S rRNA gene for full length 16S rRNA genes. It was
-necessary to use a distance threshold of \textbf{5.5}\% to cluster full
+necessary to use a distance threshold of \textbf{5.25}\% to cluster full
 length ASVs from the same genome into a single OTU with 95\% confidence
@@
 rRNA gene, I obtained 16S rRNA sequences from the \emph{rrn} copy number
-database (\emph{rrn}DB)(14). Among the \textbf{4,774} species
-represented in the \emph{rrn}DB there were \textbf{15,614} genomes. The
+database (\emph{rrn}DB)(14). Among the \textbf{5,972} species
+represented in the \emph{rrn}DB there were \textbf{20,427} genomes. The
 median \emph{rrn} copy number per species ranged between \textbf{1}
@@
 for a genome increased, the number of variants of the 16S rRNA gene in
-each genome also increased. On average, there were \textbf{0.60}
+each genome also increased. On average, there were \textbf{0.58}
 variants per copy of the full length 16S rRNA gene and an average of
-\textbf{0.33}, \textbf{0.26}, and \textbf{0.27} variants when
+\textbf{0.32}, \textbf{0.25}, and \textbf{0.27} variants when
 considering the V3-V4, V4, and V4-V5 regions of the gene, respectively.
@@
 number of genomes that were sampled (\textbf{Figure 1}). For example,
-the \textbf{180} genome accessions of \textbf{\emph{Mycobacterium
+the \textbf{271} genome accessions of \textbf{\emph{Mycobacterium
 tuberculosis}} in the \emph{rrn}DB each had \textbf{1} copy of the gene
-per genome. However, across those accessions, there were \textbf{11}
+per genome. However, across those accessions, there were \textbf{17}
 versions of the gene. An \emph{E. coli} genome typically had \textbf{7}
 copies of the 16S rRNA gene with a median of \textbf{5} distinct full
-length ASVs per genome (intraquartile range between \textbf{4} and
-\textbf{6}). Across the \textbf{958} \emph{E. coli} genomes in the
-\emph{rrn}DB, there were \textbf{1,013} versions of the gene. These
+length ASVs per genome (intraquartile range between \textbf{3} and
+\textbf{6}). Across the \textbf{1,390} \emph{E. coli} genomes in the
+\emph{rrn}DB, there were \textbf{1,402} versions of the gene. These
 observations highlight the risk of selecting a threshold for defining
@@
 of the \emph{rrn} operon (e.g., \emph{E. coli}), a distance threshold of
-\textbf{5.5}\% was required to reduce full length ASVs into a single OTU
-for 95\% of the species (Figure 2). Similarly, thresholds of
-\textbf{4.0}, \textbf{2.5}, and \textbf{3.5}\% were required for the
+\textbf{5.25}\% was required to reduce full length ASVs into a single
+OTU for 95\% of the species (Figure 2). Similarly, thresholds of
+\textbf{5.25}, \textbf{2.50}, and \textbf{3.75}\% were required for the
 V3-V4, V4, and V4-V5 regions, respectively. But, if a 3\% distance
 threshold was used, then ASVs from genomes containing fewer than
-\textbf{5}, \textbf{6}, \textbf{8}, and \textbf{6} copies of the
+\textbf{10}, \textbf{10}, \textbf{8}, and \textbf{10} copies of the
 \emph{rrn} operon would reliably be clustered into a single OTU for ASVs
@@
 multiple species could be clustered into the same ASV or OTU
-(\textbf{Figure 2}). Using ASVs, \textbf{3.6}\% of the ASVs contained
+(\textbf{Figure 2}). Using ASVs, \textbf{4.11}\% of the ASVs contained
 sequences from multiple species when considering full length sequences
-and \textbf{10.2}, \textbf{14.9}, and \textbf{12.0}\% when considering
-the V3-V4, V4, and V4-V5 regions, respectively. At the commonly used 3\%
-threshold for defining OTUs, \textbf{25.2}\% of the OTUs contained 16S
-rRNA gene sequences from multiple species when considering full length
-sequences and \textbf{29.4}, \textbf{33.0}, and \textbf{32.2}\% when
-considering the V3-V4, V4, and V4-V5 regions, respectively. Considering
-that species designations are unevenly applied and reflect multiple
-human-imposed biases, the risk of splitting a genome into multiple OTUs
-is more problematic than clustering species together. Therefore, larger
-thresholds are advisable.
+and \textbf{10.87}, \textbf{16.20}, and \textbf{13.06}\% when
+considering the V3-V4, V4, and V4-V5 regions, respectively. At the
+commonly used 3\% threshold for defining OTUs, \textbf{27.41}\% of the
+OTUs contained 16S rRNA gene sequences from multiple species when
+considering full length sequences and \textbf{31.71}, \textbf{34.31},
+and \textbf{34.78}\% when considering the V3-V4, V4, and V4-V5 regions,
+respectively. Considering that species designations are unevenly applied
+and reflect multiple human-imposed biases, the risk of splitting a
+genome into multiple OTUs is more problematic than clustering species
+together. Therefore, larger thresholds are advisable.
 
@@
 sequences used in this study were obtained from the \emph{rrn}DB
-(\url{https://rrndb.umms.med.umich.edu}; version 5.6, released November
-8, 2019) (14). At the time of submission, this was the most current
+(\url{https://rrndb.umms.med.umich.edu}; version 5.7, released January
+18, 2021) (14). At the time of submission, this was the most current
 version of the database. The \emph{rrn}DB obtained the curated 16S rRNA
@@
 assign 16S rRNA gene sequences to OTUs (27). Distance thresholds between
-0.5 and 10.0\% in 0.5 percentage point increments were used to assign
+0.25 and 10.0\% in 0.25 percentage point increments were used to assign
 sequences to OTUs.
@@
 reported the median of the 100 randomizations. The intraquartile range
-between randomizations was less than \textbf{0.0029}. Because the range
+between randomizations was less than \textbf{0.0024}. Because the range
 was so small, the confidence intervals were more narrow than the

### 2021-02-23 code/align_sequences.sh

@@ -1,22 +1,22 @@
 #!/usr/bin/env bash
 
 # author: pat schloss
-# inputs: data/raw/rrnDB-5.6_16S_rRNA.fasta
+# inputs: data/raw/rrnDB-5.7_16S_rRNA.fasta
 #					data/references/silva_seed/silva.seed_v138.align
-# outputs: data/raw/rrnDB-5.6_16S_rRNA.align
+# outputs: data/raw/rrnDB-5.7_16S_rRNA.align
 #
 # We need to include flip=T to make sure all sequences are pointed in the same
 # direction
 
-sed "s/ /_/g" data/raw/rrnDB-5.6_16S_rRNA.fasta > data/raw/rrnDB-5.6_16S_rRNA.temp.fasta
+sed "s/ /_/g" data/raw/rrnDB-5.7_16S_rRNA.fasta > data/raw/rrnDB-5.7_16S_rRNA.temp.fasta
 
-code/mothur/mothur '#align.seqs(fasta=data/raw/rrnDB-5.6_16S_rRNA.temp.fasta,
+code/mothur/mothur '#align.seqs(fasta=data/raw/rrnDB-5.7_16S_rRNA.temp.fasta,
 reference=data/references/silva_seed/silva.seed_v138.align, flip=T)'
 
 if [[ $? -eq 0 ]]
 then
-mv data/raw/rrnDB-5.6_16S_rRNA.temp.align data/raw/rrnDB-5.6_16S_rRNA.align
-rm data/raw/rrnDB-5.6_16S_rRNA.temp.fasta
+mv data/raw/rrnDB-5.7_16S_rRNA.temp.align data/raw/rrnDB-5.7_16S_rRNA.align
+rm data/raw/rrnDB-5.7_16S_rRNA.temp.fasta
 else
 echo "FAIL: mothur failed to align sequences"
 exit 1

### 2021-02-23 code/extract_region.sh

@@ -1,7 +1,7 @@
 #!/usr/bin/env bash
 
 # author: pat schloss
-# dependency: data/raw/rrnDB-5.6_16S_rRNA.align
+# dependency: data/raw/rrnDB-5.7_16S_rRNA.align
 # input: the target name, which is really the ouput
 # 				data/<region>/rrnDB.align
 #
@@ -38,9 +38,9 @@ fi
 
 mkdir -p $path
 
-code/mothur/mothur "#pcr.seqs(fasta=data/raw/rrnDB-5.6_16S_rRNA.align, start=$start, end=$end, outputdir=$path); filter.seqs(vertical=TRUE)"
+code/mothur/mothur "#pcr.seqs(fasta=data/raw/rrnDB-5.7_16S_rRNA.align, start=$start, end=$end, outputdir=$path); filter.seqs(vertical=TRUE)"
 
-# grep -v "^>" $path/rrnDB-5.6_16S_rRNA.pcr.filter.fasta | grep "\."
+# grep -v "^>" $path/rrnDB-5.7_16S_rRNA.pcr.filter.fasta | grep "\."
 
 # if mothur executed successfully, then touch the files that might not have been
 # generated in pcr.seqs because the sequences spanned the deisred region's
@@ -48,16 +48,16 @@ code/mothur/mothur "#pcr.seqs(fasta=data/raw/rrnDB-5.6_16S_rRNA.align, start=$st
 
 if [[ $? -eq 0 ]]
 then
-	sed "/^[^>]/ s/\./-/g" $path/rrnDB-5.6_16S_rRNA.pcr.filter.fasta > $path/rrnDB-5.6_16S_rRNA.pcr.filter.test.fasta
+	sed "/^[^>]/ s/\./-/g" $path/rrnDB-5.7_16S_rRNA.pcr.filter.fasta > $path/rrnDB-5.7_16S_rRNA.pcr.filter.test.fasta
 
-	touch $path/rrnDB-5.6_16S_rRNA.bad.accnos
-	touch $path/rrnDB-5.6_16S_rRNA.scrap.pcr.align
+	touch $path/rrnDB-5.7_16S_rRNA.bad.accnos
+	touch $path/rrnDB-5.7_16S_rRNA.scrap.pcr.align
 else
 	echo "FAIL: mothur ran into a problem"
 	exit 1
 fi
[…]

### 2021-02-23 code/get_genome_id_taxonomy.R

@@ -3,9 +3,9 @@
 # name: get_genome_id_taxonomy.R
 #
 # input: 
-# 	- data/raw/rrnDB-5.6.tsv
+# 	- data/raw/rrnDB-5.7.tsv
 #		- data/references/sp_spp_lookup.tsv
-#		- data/raw/rrnDB-5.6_pantaxa_stats_NCBI.tsv
+#		- data/raw/rrnDB-5.7_pantaxa_stats_NCBI.tsv
 # output: tsv containing the genome id along with taxonomic information
 # 				data/references/genome_id_taxonomy.tsv
 
@@ -17,7 +17,7 @@ merged <- read_delim("data/references/ncbi_merged_lookup.tsv", delim="|",
 	select(-blank)
 
 	
-metadata <- read_tsv("data/raw/rrnDB-5.6.tsv") %>%
+metadata <- read_tsv("data/raw/rrnDB-5.7.tsv") %>%
 		rename(genome_id = `Data source record id`, 
 					 tax_id = `NCBI tax id`,
 					 rdp = `RDP taxonomic lineage`,

### 2021-02-23 code/plot_copy_number_threshold.R

@@ -10,7 +10,7 @@ read_tsv("data/processed/thresholds_for_single_otu.tsv") %>%
 	geom_line(size=1) +
 	custom_color_scale() +
 	scale_x_continuous(breaks=seq(1,9,2), labels=seq(1,9,2)) +
-	scale_y_continuous(breaks=seq(0,0.06,0.02), labels=seq(0,6,2), limits =c(0,0.06)) +
+	scale_y_continuous(breaks=seq(0,0.08,0.02), labels=seq(0,8,2), limits =c(0,0.08)) +
 	labs(x="Number of copies of *rrn* operon",
 			 y="Distance threshold where ASVs from\n95% of species had one OTU (%)") +
 	theme_classic() +

### 2021-02-26 submission/manuscript.Rmd
Fifth round of revisions

@@
   - \linenumbers
+  - \raggedright
 editor_options:
@@
 
-format_pct <- function(x, my_digits=2){
+format_pct <- function(x, my_digits=1){
 
@@
 Proponents of ASVs are largely dissmissive of concerns that most bacterial genomes have more than one copy of the *rrn* operon and that those copies are not identical [@Pei2010; @Sun2013].
-Yet, using too fine a threshold to identify OTUs creates the risk of splitting a single genome into multiple clusters.
-Conversely, using too broad of a threshold to define OTUs creates the risk of clustering together bacterial species into the same OTU.
+Yet, ASVs and using too fine a threshold to identify OTUs could split a single genome into multiple clusters.
+Conversely, using too broad of a threshold to define OTUs could cluster together multiple bacterial species into the same OTU.
 An example of both is seen in the comparison of *Staphylococcus aureus* (NCTC 8325) and *S.
@@
 The median *rrn* copy number per species ranged between `r single_copy$median_rrn` (e.g., *`r single_copy$species`*) and `r max_copy$median_rrn` (*`r max_copy$species`*).
-As **rrn** copy number for a genome increased, the number of variants of the 16S rRNA gene in each genome also increased.
+As *rrn* copy number for a genome increased, the number of variants of the 16S rRNA gene in each genome also increased.
 On average, there were `r rate_v19` variants per copy of the full length 16S rRNA gene and an average of `r rate_v34`, `r rate_v4`, and `r rate_v45` variants when considering the V3-V4, V4, and V4-V5 regions of the gene, respectively.
@@
 Therefore, I assessed the impact of the distance threshold used to define clusters of 16S rRNA genes on the propensity to split a genome into separate clusters.
-To control for uneven representation of genomes across species, I randomly selected one genome from each species for the following analyses and repeated each randomization 100 times.
+To control for uneven representation of genomes across species, I randomly selected one genome from each species and repeated each randomization 100 times.
 I observed that as the *rrn* copy number increased, the distance threshold required to reduce the ASVs in each genome to a single OTU increased (Figure 1).
 Among species with 7 copies of the *rrn* operon (e.g., *E.
-coli*), a distance threshold of `r format_pct(threshold_v19)`% was required to reduce full length ASVs into a single OTU for 95% of the species (Figure 2).
-Similarly, thresholds of `r format_pct(threshold_v34)`, `r format_pct(threshold_v4)`, and `r format_pct(threshold_v45)`% were required for the V3-V4, V4, and V4-V5 regions, respectively.
-But, if a 3% distance threshold was used, then ASVs from genomes containing fewer than `r three_v19`, `r three_v34`, `r three_v4`, and `r three_v45` copies of the *rrn* operon would reliably be clustered into a single OTU for ASVs from the V1-V9, V3-V4, V4, and V4-V5 regions, respectively (Figure 2).
+coli*), a distance threshold of `r format_pct(threshold_v19, 2)`% was required to reduce full length ASVs into a single OTU for 95% of the species (Figure 1).
+Similarly, thresholds of `r format_pct(threshold_v34, 2)`, `r format_pct(threshold_v4, 2)`, and `r format_pct(threshold_v45, 2)`% were required for the V3-V4, V4, and V4-V5 regions, respectively.
+But, if a 3% distance threshold was used, then ASVs from genomes containing fewer than `r three_v19`, `r three_v34`, `r three_v4`, and `r three_v45` copies of the *rrn* operon would reliably be clustered into a single OTU for ASVs from the V1-V9, V3-V4, V4, and V4-V5 regions, respectively (Figure 1).
 Consequently, these results demonstrate that broad thresholds must be used to avoid splitting different operons from the same genome into separate clusters.
-At broad thresholds, 16S rRNA gene sequences from multiple species could be clustered into the same ASV or OTU (**Figure 2**).
-Using ASVs, `r format_pct(lump_v19)`% of the ASVs contained sequences from multiple species when considering full length sequences and `r format_pct(lump_v34)`, `r format_pct(lump_v4)`, and `r format_pct(lump_v45)`% when considering the V3-V4, V4, and V4-V5 regions, respectively.
+
+At broad thresholds, 16S rRNA gene sequences from multiple species could be clustered into the same ASV or OTU (Figure 2).
+For this analysis, I again randomly selected one genome from each species to control for uneven representation of genomes across species and repeated each randomization 100 times.
+Without using distance-based thresholds, `r format_pct(lump_v19)`% of the ASVs contained sequences from multiple species when considering full length sequences and `r format_pct(lump_v34)`, `r format_pct(lump_v4)`, and `r format_pct(lump_v45)`% when considering the V3-V4, V4, and V4-V5 regions, respectively.
 At the commonly used 3% threshold for defining OTUs, `r format_pct(lump3_v19)`% of the OTUs contained 16S rRNA gene sequences from multiple species when considering full length sequences and `r format_pct(lump3_v34)`, `r format_pct(lump3_v4)`, and `r format_pct(lump3_v45)`% when considering the V3-V4, V4, and V4-V5 regions, respectively.
@@
 A short sequence from a bacterial genome simply cannot differentiate between species.
-Moreover, it is difficult to defend a clustering approach that would split a single genome into multiple taxonomic units.
-It is not biologically plausible to entertain the possibility that parts of a genome would have different ecologies.
+Moreover, it is difficult to defend a clustering threshold that would split a single genome into multiple taxonomic units.
+It is not biologically plausible to entertain the possibility that different *rrn* operons from the same genome would have different ecologies.
 Although there are multiple reasons that proponents favor ASVs, the significant risk of artificially splitting genomes into separate clusters is too high to warrant their use.
@@
 **(iii) Clustering sequences into OTUs.** Pairwise distances between sequences were calculated using the dist.seqs command from mothur.
-The OptiClust algorithm as implemented in mothur was used to assign 16S rRNA gene sequences to OTUs [@Westcott2017].
-Distance thresholds between 0.25 and 10.0% in 0.25 percentage point increments were used to assign sequences to OTUs.
+The OptiClust algorithm, as implemented in mothur, was used to assign 16S rRNA gene sequences to OTUs [@Westcott2017].
+Distance thresholds between 0.25 and 10.00% in 0.25 percentage point increments were used to assign sequences to OTUs.
 
-**(iv) Controlling for uneven sampling of genomes by species.** Because of the uneven distribution of genome sequences across species, for the analysis of splitting genomes and clustering ASVs from different species, I randomly selected one genome from each species.
+**(iv) Controlling for uneven sampling of genomes by species.** Because of the uneven distribution of genome sequences across species, for the analysis of splitting genomes and clustering ASVs from different species (Figures 1 and 2), I randomly selected one genome from each species.
 The random selection was repeated 100 times.
@@
 **(vi) Note on usage of ASV, OTU, and cluster.** I used "ASV" to denote the cluster of true 16S rRNA gene sequences that were identical to each other and "OTU" to denote the product of distance-based clustering of sequences.
-Although ASVs do represent a type of operational defition of a taxonomic unit and can be thought of as an OTU formed using a distance of zero, proponents of the ASV approach prefer to avoid the term OTU given the long history of OTUs being formed by distance-based clustering (https://github.com/benjjneb/dada2/issues/62; accessed 2021-02-19).
-For this reason, when an ASV splits a genome into different units, those units were called clusters rather than OTUs.
+Although ASVs do represent a type of operational defition of a taxonomic unit and can be thought of as an OTU formed using a distance of zero, proponents of the ASV approach prefer to avoid the term OTU given the long history of OTUs being formed by distance-based clustering (https://github.com/benjjneb/dada2/issues/62; accessed 2021-02-26).
+For this reason, when an ASV split a genome into different units, those units were called clusters rather than OTUs.
 
@@
 **Figure 2.
-As the distance threshold used to define an OTU increased, the fraction of genomes split into separate OTUs decreased while the fraction of ASVs and OTUs representing multiple species increased.** These data represent the median fractions for both measurements across 100 randomizations.
+As the distance threshold used to define an OTU increased, the percentage of ASVs and OTUs representing multiple species increased.** These data represent the median fractions for both measurements across 100 randomizations.
 In each randomization, one genome was sampled from each species.

### 2021-02-26 submission/manuscript.tex
Fifth round of revisions

@@
 \linenumbers
+\raggedright
 \ifluatex
@@
 copy of the 16S rRNA gene for full length 16S rRNA genes. It was
-necessary to use a distance threshold of \textbf{5.25}\% to cluster full
+necessary to use a distance threshold of \textbf{5.2}\% to cluster full
 length ASVs from the same genome into a single OTU with 95\% confidence
@@
 copy of the \emph{rrn} operon and that those copies are not identical
-(12, 13). Yet, using too fine a threshold to identify OTUs creates the
-risk of splitting a single genome into multiple clusters. Conversely,
-using too broad of a threshold to define OTUs creates the risk of
-clustering together bacterial species into the same OTU. An example of
-both is seen in the comparison of \emph{Staphylococcus aureus} (NCTC
-8325) and \emph{S. epidermidis} (ATCC 12228) where each genome has 5
-copies of the 16S rRNA gene. Each of the 10 copies of the 16S rRNA gene
-in these two genomes is distinct and represent 10 ASVs. Conversely, if
-the copies were clustered using a 3\% distance threshold, then all 10
-ASVs would cluster into the same OTU. The goal of this study was to
-quantify the tradeoff of splitting a single genome into multiple
-clusters and the risk of clustering different bacterial species into the
-same cluster when using ASVs and various OTU definitions.
+(12, 13). Yet, ASVs and using too fine a threshold to identify OTUs
+could split a single genome into multiple clusters. Conversely, using
+too broad of a threshold to define OTUs could cluster together multiple
+bacterial species into the same OTU. An example of both is seen in the
+comparison of \emph{Staphylococcus aureus} (NCTC 8325) and \emph{S.
+epidermidis} (ATCC 12228) where each genome has 5 copies of the 16S rRNA
+gene. Each of the 10 copies of the 16S rRNA gene in these two genomes is
+distinct and represent 10 ASVs. Conversely, if the copies were clustered
+using a 3\% distance threshold, then all 10 ASVs would cluster into the
+same OTU. The goal of this study was to quantify the tradeoff of
+splitting a single genome into multiple clusters and the risk of
+clustering different bacterial species into the same cluster when using
+ASVs and various OTU definitions.
 
@@
 (e.g., \textbf{\emph{Mycobacterium tuberculosis}}) and \textbf{19}
-(\textbf{\emph{Metabacillus litoralis}}). As \textbf{rrn} copy number
-for a genome increased, the number of variants of the 16S rRNA gene in
-each genome also increased. On average, there were \textbf{0.58}
-variants per copy of the full length 16S rRNA gene and an average of
-\textbf{0.32}, \textbf{0.25}, and \textbf{0.27} variants when
-considering the V3-V4, V4, and V4-V5 regions of the gene, respectively.
-Although a species tended to have a consistent number of 16S rRNA gene
-copies per genome, the number of total variants increased with the
-number of genomes that were sampled (\textbf{Figure 1}). For example,
-the \textbf{271} genome accessions of \textbf{\emph{Mycobacterium
-tuberculosis}} in the \emph{rrn}DB each had \textbf{1} copy of the gene
-per genome. However, across those accessions, there were \textbf{17}
-versions of the gene. An \emph{E. coli} genome typically had \textbf{7}
-copies of the 16S rRNA gene with a median of \textbf{5} distinct full
-length ASVs per genome (intraquartile range between \textbf{3} and
-\textbf{6}). Across the \textbf{1,390} \emph{E. coli} genomes in the
-\emph{rrn}DB, there were \textbf{1,402} versions of the gene. These
-observations highlight the risk of selecting a threshold for defining
-clusters that is too narrow because it is possible to split a single
-genome into multiple clusters.
+(\textbf{\emph{Metabacillus litoralis}}). As \emph{rrn} copy number for
+a genome increased, the number of variants of the 16S rRNA gene in each
+genome also increased. On average, there were \textbf{0.58} variants per
+copy of the full length 16S rRNA gene and an average of \textbf{0.32},
+\textbf{0.25}, and \textbf{0.27} variants when considering the V3-V4,
+V4, and V4-V5 regions of the gene, respectively. Although a species
+tended to have a consistent number of 16S rRNA gene copies per genome,
+the number of total variants increased with the number of genomes that
+were sampled (\textbf{Figure 1}). For example, the \textbf{271} genome
+accessions of \textbf{\emph{Mycobacterium tuberculosis}} in the
+\emph{rrn}DB each had \textbf{1} copy of the gene per genome. However,
+across those accessions, there were \textbf{17} versions of the gene. An
+\emph{E. coli} genome typically had \textbf{7} copies of the 16S rRNA
+gene with a median of \textbf{5} distinct full length ASVs per genome
+(intraquartile range between \textbf{3} and \textbf{6}). Across the
+\textbf{1,390} \emph{E. coli} genomes in the \emph{rrn}DB, there were
+\textbf{1,402} versions of the gene. These observations highlight the
+risk of selecting a threshold for defining clusters that is too narrow
+because it is possible to split a single genome into multiple clusters.
 
@@
 representation of genomes across species, I randomly selected one genome
-from each species for the following analyses and repeated each
-randomization 100 times. I observed that as the \emph{rrn} copy number
-increased, the distance threshold required to reduce the ASVs in each
-genome to a single OTU increased (Figure 1). Among species with 7 copies
-of the \emph{rrn} operon (e.g., \emph{E. coli}), a distance threshold of
-\textbf{5.25}\% was required to reduce full length ASVs into a single
-OTU for 95\% of the species (Figure 2). Similarly, thresholds of
-\textbf{5.25}, \textbf{2.50}, and \textbf{3.75}\% were required for the
-V3-V4, V4, and V4-V5 regions, respectively. But, if a 3\% distance
-threshold was used, then ASVs from genomes containing fewer than
-\textbf{10}, \textbf{10}, \textbf{8}, and \textbf{10} copies of the
-\emph{rrn} operon would reliably be clustered into a single OTU for ASVs
-from the V1-V9, V3-V4, V4, and V4-V5 regions, respectively (Figure 2).
-Consequently, these results demonstrate that broad thresholds must be
-used to avoid splitting different operons from the same genome into
-separate clusters. At broad thresholds, 16S rRNA gene sequences from
-multiple species could be clustered into the same ASV or OTU
-(\textbf{Figure 2}). Using ASVs, \textbf{4.11}\% of the ASVs contained
-sequences from multiple species when considering full length sequences
-and \textbf{10.87}, \textbf{16.20}, and \textbf{13.06}\% when
-considering the V3-V4, V4, and V4-V5 regions, respectively. At the
-commonly used 3\% threshold for defining OTUs, \textbf{27.41}\% of the
-OTUs contained 16S rRNA gene sequences from multiple species when
-considering full length sequences and \textbf{31.71}, \textbf{34.31},
-and \textbf{34.78}\% when considering the V3-V4, V4, and V4-V5 regions,
-respectively. Considering that species designations are unevenly applied
-and reflect multiple human-imposed biases, the risk of splitting a
-genome into multiple OTUs is more problematic than clustering species
-together. Therefore, larger thresholds are advisable.
+from each species and repeated each randomization 100 times. I observed
+that as the \emph{rrn} copy number increased, the distance threshold
+required to reduce the ASVs in each genome to a single OTU increased
+(Figure 1). Among species with 7 copies of the \emph{rrn} operon (e.g.,
+\emph{E. coli}), a distance threshold of \textbf{5.25}\% was required to
+reduce full length ASVs into a single OTU for 95\% of the species
+(Figure 1). Similarly, thresholds of \textbf{5.25}, \textbf{2.50}, and
+\textbf{3.75}\% were required for the V3-V4, V4, and V4-V5 regions,
+respectively. But, if a 3\% distance threshold was used, then ASVs from
+genomes containing fewer than \textbf{10}, \textbf{10}, \textbf{8}, and
+\textbf{10} copies of the \emph{rrn} operon would reliably be clustered
+into a single OTU for ASVs from the V1-V9, V3-V4, V4, and V4-V5 regions,
+respectively (Figure 1). Consequently, these results demonstrate that
+broad thresholds must be used to avoid splitting different operons from
+the same genome into separate clusters.
+
+At broad thresholds, 16S rRNA gene sequences from multiple species could
+be clustered into the same ASV or OTU (Figure 2). For this analysis, I
+again randomly selected one genome from each species to control for
+uneven representation of genomes across species and repeated each
+randomization 100 times. Without using distance-based thresholds,
+\textbf{4.1}\% of the ASVs contained sequences from multiple species
+when considering full length sequences and \textbf{10.9}, \textbf{16.2},
+and \textbf{13.1}\% when considering the V3-V4, V4, and V4-V5 regions,
+respectively. At the commonly used 3\% threshold for defining OTUs,
+\textbf{27.4}\% of the OTUs contained 16S rRNA gene sequences from
+multiple species when considering full length sequences and
+\textbf{31.7}, \textbf{34.3}, and \textbf{34.8}\% when considering the
+V3-V4, V4, and V4-V5 regions, respectively. Considering that species
+designations are unevenly applied and reflect multiple human-imposed
+biases, the risk of splitting a genome into multiple OTUs is more
+problematic than clustering species together. Therefore, larger
+thresholds are advisable.
 
@@
 sequence from a bacterial genome simply cannot differentiate between
-species. Moreover, it is difficult to defend a clustering approach that
+species. Moreover, it is difficult to defend a clustering threshold that
 would split a single genome into multiple taxonomic units. It is not
-biologically plausible to entertain the possibility that parts of a
-genome would have different ecologies. Although there are multiple
-reasons that proponents favor ASVs, the significant risk of artificially
-splitting genomes into separate clusters is too high to warrant their
-use.
+biologically plausible to entertain the possibility that different
+\emph{rrn} operons from the same genome would have different ecologies.
+Although there are multiple reasons that proponents favor ASVs, the
+significant risk of artificially splitting genomes into separate
+clusters is too high to warrant their use.
 
@@
 between sequences were calculated using the dist.seqs command from
-mothur. The OptiClust algorithm as implemented in mothur was used to
+mothur. The OptiClust algorithm, as implemented in mothur, was used to
 assign 16S rRNA gene sequences to OTUs (27). Distance thresholds between
-0.25 and 10.0\% in 0.25 percentage point increments were used to assign
+0.25 and 10.00\% in 0.25 percentage point increments were used to assign
 sequences to OTUs.
@@
 for the analysis of splitting genomes and clustering ASVs from different
-species, I randomly selected one genome from each species. The random
-selection was repeated 100 times. Analyses based on this randomization
-reported the median of the 100 randomizations. The intraquartile range
-between randomizations was less than \textbf{0.0024}. Because the range
-was so small, the confidence intervals were more narrow than the
-thickness of the lines in Figures 1 and 2 and were not included.
+species (Figures 1 and 2), I randomly selected one genome from each
+species. The random selection was repeated 100 times. Analyses based on
+this randomization reported the median of the 100 randomizations. The
+intraquartile range between randomizations was less than
+\textbf{0.0024}. Because the range was so small, the confidence
+intervals were more narrow than the thickness of the lines in Figures 1
+and 2 and were not included.
 
@@
 (\url{https://github.com/benjjneb/dada2/issues/62}; accessed
-2021-02-19). For this reason, when an ASV splits a genome into different
+2021-02-26). For this reason, when an ASV split a genome into different
 units, those units were called clusters rather than OTUs.
@@
 \textbf{Figure 2. As the distance threshold used to define an OTU
-increased, the fraction of genomes split into separate OTUs decreased
-while the fraction of ASVs and OTUs representing multiple species
+increased, the percentage of ASVs and OTUs representing multiple species
 increased.} These data represent the median fractions for both

### 2021-02-26 code/plot_copy_number_threshold.R

@@ -15,7 +15,7 @@ read_tsv("data/processed/thresholds_for_single_otu.tsv") %>%
 			 y="Distance threshold where ASVs from\n95% of species had one OTU (%)") +
 	theme_classic() +
 	theme(
-		legend.position = c(0.8, 0.2),
+		legend.position = c(0.85, 0.15),
 		legend.key.height = unit(0.4, "cm"),
 		axis.title.x = element_markdown(),
 		axis.title.y = element_text(margin=margin(t=0, r=10, b=0, l=0))

### 2021-02-26 code/plot_lump_split.R

@@ -6,32 +6,26 @@ source('code/colors.R')
 read_tsv("data/processed/lumped_split_rate.tsv",
 				 col_types = cols(region = col_character(),
 				 								 .default = col_double())) %>%
-	select(-ends_with("iqr")) %>%
-	pivot_longer(cols=c(split_rate, lump_rate), names_to="method", values_to="fraction") %>%
-	mutate(method = factor(ifelse(method == "lump_rate",
-												 "Fraction of ASVs or OTUs\nthat included sequences\nfrom multiple species",
-												 "Fraction of genomes\nsplit into separate OTUs"),
-											 levels=c("Fraction of genomes\nsplit into separate OTUs",
-											 "Fraction of ASVs or OTUs\nthat included sequences\nfrom multiple species")
-									 )
-								 ) %>%
-	ggplot(aes(x=threshold, y=fraction, color=region)) +
+	select(region, threshold, lump_rate) %>%
+	ggplot(aes(x=threshold, y=lump_rate, color=region)) +
 	geom_line(size=1) +
-	facet_wrap(~method, nrow=2, strip.position = "left") +
 	scale_x_continuous(
-		breaks = c(0, 0.025, 0.05, 0.075, 0.1),
-		labels=c("0", "2.5", "5", "7.5", "10")
+		breaks = seq(0,0.1,0.02),
+		labels = seq(0,10,2)
+		) +
+	scale_y_continuous(
+		breaks = seq(0,0.5,0.1),
+		labels = seq(0,50,10)
 		) +
 	custom_color_scale() +
-	labs(x="Distance theshold\nused to define OTUs (%)", y=NULL) +
+	labs(
+		x="Distance theshold\nused to define OTUs (%)",
+		y="Percentage of ASVs or OTUs that included\nsequences from multiple species") +
 	theme_classic() +
 	theme(
-		strip.placement="outside",
-		strip.background =element_rect(color=NA),
-		strip.text = element_text(size=11),
-		legend.position = c(0.8, 0.9),
+		legend.position = c(0.8, 0.2),
 		legend.key.height = unit(0.4, "cm")
 	)
 
-ggsave("figures/lump_split.tiff", height =5, width =4, compression="lzw")
-ggsave("figures/lump_split.pdf", height =5, width =4)
+ggsave("figures/lump_split.tiff", height =5, width =5, compression="lzw")
+ggsave("figures/lump_split.pdf", height =5, width =5)

### 2021-02-26 README.Rmd
Sixth round of revisions

@@
 
-	paste0("**", formatted, "**")
+	formatted
 
@@
 There was an average of `r rate_v19` ASVs per copy of the 16S rRNA gene for full length 16S rRNA genes.
-It was necessary to use a distance threshold of `r format_pct(threshold_v19)`% to cluster full length ASVs from the same genome into a single OTU with 95% confidence for genomes with 7 copies of the 16S rRNA, such as *E.
+It was necessary to use a distance threshold of `r format_pct(threshold_v19, 2)`% to cluster full length ASVs from the same genome into a single OTU with 95% confidence for genomes with 7 copies of the 16S rRNA, such as *E.
 coli*.
@@
   * `data.table` (v. `r packageVersion("data.table")`)
+  * `here` (v. `r packageVersion("here")`)
   * `rmarkdown` (v. `r packageVersion("rmarkdown")`)

### 2021-02-26 submission/manuscript.Rmd
Sixth round of revisions

@@
 
-	paste0("**", formatted, "**")
+  formatted
 
@@
 
-three_v19 <- thresholds %>% filter(threshold <= 0.03 & region == "v19") %>% tail(n=1) %>% pull(n_rrns)
-three_v4 <- thresholds %>% filter(threshold <= 0.03 & region == "v4") %>% tail(n=1) %>% pull(n_rrns)
-three_v34 <- thresholds %>% filter(threshold <= 0.03 & region == "v34") %>% tail(n=1) %>% pull(n_rrns)
-three_v45 <- thresholds %>% filter(threshold <= 0.03 & region == "v45") %>% tail(n=1) %>% pull(n_rrns)
+three_v19 <- thresholds %>% filter(threshold <= 0.03 & region == "v19" & n_rrns < 10) %>% tail(n=1) %>% pull(n_rrns)
+three_v4 <- thresholds %>% filter(threshold <= 0.03 & region == "v4" & n_rrns < 10) %>% tail(n=1) %>% pull(n_rrns)
+three_v34 <- thresholds %>% filter(threshold <= 0.03 & region == "v34" & n_rrns < 10) %>% tail(n=1) %>% pull(n_rrns)
+three_v45 <- thresholds %>% filter(threshold <= 0.03 & region == "v45" & n_rrns < 10) %>% tail(n=1) %>% pull(n_rrns)
 
@@
 Instead, genome sequencing technologies have suggested that the widely used 3% distance threshold to operationally define bacterial taxa is too coarse [@RodriguezR2018; @Stackebrandt2006; @Edgar2018].
-As an alternative to OTUs, amplicon sequencing variants (ASVs) have been proposed as a way to adopt the thresholds suggested by genome sequencing to microbial community analysis using 16S rRNA gene sequences [@Edgar2016; @Amir2017; @Callahan2016; @Eren2014].
+As an alternative to operational taxonomic units (OTUs), amplicon sequencing variants (ASVs) have been proposed as a way to adopt the thresholds suggested by genome sequencing to microbial community analysis using 16S rRNA gene sequences [@Edgar2016; @Amir2017; @Callahan2016; @Eren2014].
 Approaches for identifying ASVs do not cluster sequences based on a distance-based threshold [@Callahan2017].
@@
 The median *rrn* copy number per species ranged between `r single_copy$median_rrn` (e.g., *`r single_copy$species`*) and `r max_copy$median_rrn` (*`r max_copy$species`*).
-As *rrn* copy number for a genome increased, the number of variants of the 16S rRNA gene in each genome also increased.
+As the *rrn* copy number for a genome increased, the number of variants of the 16S rRNA gene in each genome also increased.
 On average, there were `r rate_v19` variants per copy of the full length 16S rRNA gene and an average of `r rate_v34`, `r rate_v4`, and `r rate_v45` variants when considering the V3-V4, V4, and V4-V5 regions of the gene, respectively.
-Although a species tended to have a consistent number of 16S rRNA gene copies per genome, the number of total variants increased with the number of genomes that were sampled (**Figure 1**).
+Although a species tended to have a consistent number of 16S rRNA gene copies per genome, the number of total variants increased with the number of genomes that were sampled (Figure S1).
 For example, the `r single_copy_N` genome accessions of *`r single_copy$species`* in the *rrn*DB each had `r single_copy$median_rrn` copy of the gene per genome.
@@
 Among species with 7 copies of the *rrn* operon (e.g., *E.
-coli*), a distance threshold of `r format_pct(threshold_v19, 2)`% was required to reduce full length ASVs into a single OTU for 95% of the species (Figure 1).
+coli*), a distance threshold of `r format_pct(threshold_v19, 2)`% was required to reduce full length ASVs into a single OTU for 95% of the species.
 Similarly, thresholds of `r format_pct(threshold_v34, 2)`, `r format_pct(threshold_v4, 2)`, and `r format_pct(threshold_v45, 2)`% were required for the V3-V4, V4, and V4-V5 regions, respectively.
-But, if a 3% distance threshold was used, then ASVs from genomes containing fewer than `r three_v19`, `r three_v34`, `r three_v4`, and `r three_v45` copies of the *rrn* operon would reliably be clustered into a single OTU for ASVs from the V1-V9, V3-V4, V4, and V4-V5 regions, respectively (Figure 1).
+But, if a 3% distance threshold was used, then ASVs from genomes containing fewer than `r three_v19`, `r three_v34`, `r three_v4`, and `r three_v45` copies of the *rrn* operon would reliably be clustered into a single OTU for ASVs from the V1-V9, V3-V4, V4, and V4-V5 regions, respectively.
 Consequently, these results demonstrate that broad thresholds must be used to avoid splitting different operons from the same genome into separate clusters.
 
-At broad thresholds, 16S rRNA gene sequences from multiple species could be clustered into the same ASV or OTU (Figure 2).
-For this analysis, I again randomly selected one genome from each species to control for uneven representation of genomes across species and repeated each randomization 100 times.
+At broad thresholds, 16S rRNA gene sequences from multiple species could be clustered into the same ASV or OTU.
+I again randomly selected one genome from each species to control for uneven representation of genomes across species and for this analysis I measured the percentage of ASVs and OTUs that contained 16S rRNA gene sequences from multiple species (Figure 2).
 Without using distance-based thresholds, `r format_pct(lump_v19)`% of the ASVs contained sequences from multiple species when considering full length sequences and `r format_pct(lump_v34)`, `r format_pct(lump_v4)`, and `r format_pct(lump_v45)`% when considering the V3-V4, V4, and V4-V5 regions, respectively.
 At the commonly used 3% threshold for defining OTUs, `r format_pct(lump3_v19)`% of the OTUs contained 16S rRNA gene sequences from multiple species when considering full length sequences and `r format_pct(lump3_v34)`, `r format_pct(lump3_v4)`, and `r format_pct(lump3_v45)`% when considering the V3-V4, V4, and V4-V5 regions, respectively.
-Considering that species designations are unevenly applied and reflect multiple human-imposed biases, the risk of splitting a genome into multiple OTUs is more problematic than clustering species together.
+Considering that species designations are inconsistently applied and reflect multiple human-imposed biases, the risk of splitting a genome into multiple OTUs is more problematic than clustering species together.
 Therefore, larger thresholds are advisable.
@@
 
-\vspace{20mm}
+\vspace{10mm}
 
@@
 
-**(iv) Controlling for uneven sampling of genomes by species.** Because of the uneven distribution of genome sequences across species, for the analysis of splitting genomes and clustering ASVs from different species (Figures 1 and 2), I randomly selected one genome from each species.
+**(iv) Controlling for uneven sampling of genomes by species.** Because of the uneven distribution of genome sequences across species I randomly selected one genome from each species for the analysis of splitting genomes and clustering ASVs from different species (Figures 1 and 2).
 The random selection was repeated 100 times.
@@
 `r packageVersion("Rcpp")`), furrr (v.
-`r packageVersion("furrr")`), and rmarkdown (v.
+`r packageVersion("furrr")`), here (v.
+`r packageVersion("here")`) and rmarkdown (v.
 `r packageVersion("rmarkdown")`) packages.

### 2021-02-26 submission/manuscript.tex
Sixth round of revisions

@@
 genomes represented in a \emph{rrn} copy number database, which
-contained \textbf{20,427} genomes from \textbf{5,972} species. As the
-number of copies of the 16S rRNA gene increased in a genome, the number
-of ASVs also increased. There was an average of \textbf{0.58} ASVs per
-copy of the 16S rRNA gene for full length 16S rRNA genes. It was
-necessary to use a distance threshold of \textbf{5.2}\% to cluster full
-length ASVs from the same genome into a single OTU with 95\% confidence
-for genomes with 7 copies of the 16S rRNA, such as \emph{E. coli}. This
-research highlights the risk of splitting a single bacterial genome into
-separate clusters when ASVs are used to analyze 16S rRNA gene sequence
-data. Although there is also a risk of clustering ASVs from different
-species into the same OTU when using broad distance thresholds, those
-risks are of less concern than artificially splitting a genome into
-separate ASVs and OTUs.
+contained 20,427 genomes from 5,972 species. As the number of copies of
+the 16S rRNA gene increased in a genome, the number of ASVs also
+increased. There was an average of 0.58 ASVs per copy of the 16S rRNA
+gene for full length 16S rRNA genes. It was necessary to use a distance
+threshold of 5.25\% to cluster full length ASVs from the same genome
+into a single OTU with 95\% confidence for genomes with 7 copies of the
+16S rRNA, such as \emph{E. coli}. This research highlights the risk of
+splitting a single bacterial genome into separate clusters when ASVs are
+used to analyze 16S rRNA gene sequence data. Although there is also a
+risk of clustering ASVs from different species into the same OTU when
+using broad distance thresholds, those risks are of less concern than
+artificially splitting a genome into separate ASVs and OTUs.
 
@@
 to operationally define bacterial taxa is too coarse (4--6). As an
-alternative to OTUs, amplicon sequencing variants (ASVs) have been
-proposed as a way to adopt the thresholds suggested by genome sequencing
-to microbial community analysis using 16S rRNA gene sequences (7--10).
-Approaches for identifying ASVs do not cluster sequences based on a
-distance-based threshold (11). Proponents of ASVs are largely
-dissmissive of concerns that most bacterial genomes have more than one
-copy of the \emph{rrn} operon and that those copies are not identical
-(12, 13). Yet, ASVs and using too fine a threshold to identify OTUs
-could split a single genome into multiple clusters. Conversely, using
-too broad of a threshold to define OTUs could cluster together multiple
-bacterial species into the same OTU. An example of both is seen in the
-comparison of \emph{Staphylococcus aureus} (NCTC 8325) and \emph{S.
-epidermidis} (ATCC 12228) where each genome has 5 copies of the 16S rRNA
-gene. Each of the 10 copies of the 16S rRNA gene in these two genomes is
-distinct and represent 10 ASVs. Conversely, if the copies were clustered
-using a 3\% distance threshold, then all 10 ASVs would cluster into the
-same OTU. The goal of this study was to quantify the tradeoff of
-splitting a single genome into multiple clusters and the risk of
-clustering different bacterial species into the same cluster when using
-ASVs and various OTU definitions.
+alternative to operational taxonomic units (OTUs), amplicon sequencing
+variants (ASVs) have been proposed as a way to adopt the thresholds
+suggested by genome sequencing to microbial community analysis using 16S
+rRNA gene sequences (7--10). Approaches for identifying ASVs do not
+cluster sequences based on a distance-based threshold (11). Proponents
+of ASVs are largely dissmissive of concerns that most bacterial genomes
+have more than one copy of the \emph{rrn} operon and that those copies
+are not identical (12, 13). Yet, ASVs and using too fine a threshold to
+identify OTUs could split a single genome into multiple clusters.
+Conversely, using too broad of a threshold to define OTUs could cluster
+together multiple bacterial species into the same OTU. An example of
+both is seen in the comparison of \emph{Staphylococcus aureus} (NCTC
+8325) and \emph{S. epidermidis} (ATCC 12228) where each genome has 5
+copies of the 16S rRNA gene. Each of the 10 copies of the 16S rRNA gene
+in these two genomes is distinct and represent 10 ASVs. Conversely, if
+the copies were clustered using a 3\% distance threshold, then all 10
+ASVs would cluster into the same OTU. The goal of this study was to
+quantify the tradeoff of splitting a single genome into multiple
+clusters and the risk of clustering different bacterial species into the
+same cluster when using ASVs and various OTU definitions.
 
@@
 rRNA gene, I obtained 16S rRNA sequences from the \emph{rrn} copy number
-database (\emph{rrn}DB)(14). Among the \textbf{5,972} species
-represented in the \emph{rrn}DB there were \textbf{20,427} genomes. The
-median \emph{rrn} copy number per species ranged between \textbf{1}
-(e.g., \textbf{\emph{Mycobacterium tuberculosis}}) and \textbf{19}
-(\textbf{\emph{Metabacillus litoralis}}). As \emph{rrn} copy number for
-a genome increased, the number of variants of the 16S rRNA gene in each
-genome also increased. On average, there were \textbf{0.58} variants per
-copy of the full length 16S rRNA gene and an average of \textbf{0.32},
-\textbf{0.25}, and \textbf{0.27} variants when considering the V3-V4,
-V4, and V4-V5 regions of the gene, respectively. Although a species
-tended to have a consistent number of 16S rRNA gene copies per genome,
-the number of total variants increased with the number of genomes that
-were sampled (\textbf{Figure 1}). For example, the \textbf{271} genome
-accessions of \textbf{\emph{Mycobacterium tuberculosis}} in the
-\emph{rrn}DB each had \textbf{1} copy of the gene per genome. However,
-across those accessions, there were \textbf{17} versions of the gene. An
-\emph{E. coli} genome typically had \textbf{7} copies of the 16S rRNA
-gene with a median of \textbf{5} distinct full length ASVs per genome
-(intraquartile range between \textbf{3} and \textbf{6}). Across the
-\textbf{1,390} \emph{E. coli} genomes in the \emph{rrn}DB, there were
-\textbf{1,402} versions of the gene. These observations highlight the
-risk of selecting a threshold for defining clusters that is too narrow
-because it is possible to split a single genome into multiple clusters.
+database (\emph{rrn}DB)(14). Among the 5,972 species represented in the
+\emph{rrn}DB there were 20,427 genomes. The median \emph{rrn} copy
+number per species ranged between 1 (e.g., \emph{Mycobacterium
+tuberculosis}) and 19 (\emph{Metabacillus litoralis}). As the \emph{rrn}
+copy number for a genome increased, the number of variants of the 16S
+rRNA gene in each genome also increased. On average, there were 0.58
+variants per copy of the full length 16S rRNA gene and an average of
+0.32, 0.25, and 0.27 variants when considering the V3-V4, V4, and V4-V5
+regions of the gene, respectively. Although a species tended to have a
+consistent number of 16S rRNA gene copies per genome, the number of
+total variants increased with the number of genomes that were sampled
+(Figure S1). For example, the 271 genome accessions of
+\emph{Mycobacterium tuberculosis} in the \emph{rrn}DB each had 1 copy of
+the gene per genome. However, across those accessions, there were 17
+versions of the gene. An \emph{E. coli} genome typically had 7 copies of
+the 16S rRNA gene with a median of 5 distinct full length ASVs per
+genome (intraquartile range between 3 and 6). Across the 1,390 \emph{E.
+coli} genomes in the \emph{rrn}DB, there were 1,402 versions of the
+gene. These observations highlight the risk of selecting a threshold for
+defining clusters that is too narrow because it is possible to split a
+single genome into multiple clusters.
 
@@
 (Figure 1). Among species with 7 copies of the \emph{rrn} operon (e.g.,
-\emph{E. coli}), a distance threshold of \textbf{5.25}\% was required to
-reduce full length ASVs into a single OTU for 95\% of the species
-(Figure 1). Similarly, thresholds of \textbf{5.25}, \textbf{2.50}, and
-\textbf{3.75}\% were required for the V3-V4, V4, and V4-V5 regions,
-respectively. But, if a 3\% distance threshold was used, then ASVs from
-genomes containing fewer than \textbf{10}, \textbf{10}, \textbf{8}, and
-\textbf{10} copies of the \emph{rrn} operon would reliably be clustered
-into a single OTU for ASVs from the V1-V9, V3-V4, V4, and V4-V5 regions,
-respectively (Figure 1). Consequently, these results demonstrate that
-broad thresholds must be used to avoid splitting different operons from
-the same genome into separate clusters.
+\emph{E. coli}), a distance threshold of 5.25\% was required to reduce
+full length ASVs into a single OTU for 95\% of the species. Similarly,
+thresholds of 5.25, 2.50, and 3.75\% were required for the V3-V4, V4,
+and V4-V5 regions, respectively. But, if a 3\% distance threshold was
+used, then ASVs from genomes containing fewer than 6, 6, 8, and 6 copies
+of the \emph{rrn} operon would reliably be clustered into a single OTU
+for ASVs from the V1-V9, V3-V4, V4, and V4-V5 regions, respectively.
+Consequently, these results demonstrate that broad thresholds must be
+used to avoid splitting different operons from the same genome into
+separate clusters.
 
 At broad thresholds, 16S rRNA gene sequences from multiple species could
-be clustered into the same ASV or OTU (Figure 2). For this analysis, I
-again randomly selected one genome from each species to control for
-uneven representation of genomes across species and repeated each
-randomization 100 times. Without using distance-based thresholds,
-\textbf{4.1}\% of the ASVs contained sequences from multiple species
-when considering full length sequences and \textbf{10.9}, \textbf{16.2},
-and \textbf{13.1}\% when considering the V3-V4, V4, and V4-V5 regions,
-respectively. At the commonly used 3\% threshold for defining OTUs,
-\textbf{27.4}\% of the OTUs contained 16S rRNA gene sequences from
-multiple species when considering full length sequences and
-\textbf{31.7}, \textbf{34.3}, and \textbf{34.8}\% when considering the
-V3-V4, V4, and V4-V5 regions, respectively. Considering that species
-designations are unevenly applied and reflect multiple human-imposed
-biases, the risk of splitting a genome into multiple OTUs is more
-problematic than clustering species together. Therefore, larger
-thresholds are advisable.
+be clustered into the same ASV or OTU. I again randomly selected one
+genome from each species to control for uneven representation of genomes
+across species and for this analysis I measured the percentage of ASVs
+and OTUs that contained 16S rRNA gene sequences from multiple species
+(Figure 2). Without using distance-based thresholds, 4.1\% of the ASVs
+contained sequences from multiple species when considering full length
+sequences and 10.9, 16.2, and 13.1\% when considering the V3-V4, V4, and
+V4-V5 regions, respectively. At the commonly used 3\% threshold for
+defining OTUs, 27.4\% of the OTUs contained 16S rRNA gene sequences from
+multiple species when considering full length sequences and 31.7, 34.3,
+and 34.8\% when considering the V3-V4, V4, and V4-V5 regions,
+respectively. Considering that species designations are inconsistently
+applied and reflect multiple human-imposed biases, the risk of splitting
+a genome into multiple OTUs is more problematic than clustering species
+together. Therefore, larger thresholds are advisable.
 
@@
 \textbf{(iv) Controlling for uneven sampling of genomes by species.}
-Because of the uneven distribution of genome sequences across species,
-for the analysis of splitting genomes and clustering ASVs from different
-species (Figures 1 and 2), I randomly selected one genome from each
-species. The random selection was repeated 100 times. Analyses based on
+Because of the uneven distribution of genome sequences across species I
+randomly selected one genome from each species for the analysis of
+splitting genomes and clustering ASVs from different species (Figures 1
+and 2). The random selection was repeated 100 times. Analyses based on
 this randomization reported the median of the 100 randomizations. The
-intraquartile range between randomizations was less than
-\textbf{0.0024}. Because the range was so small, the confidence
-intervals were more narrow than the thickness of the lines in Figures 1
-and 2 and were not included.
+intraquartile range between randomizations was less than 0.0024. Because
+the range was so small, the confidence intervals were more narrow than
+the thickness of the lines in Figures 1 and 2 and were not included.
 
@@
 made use of built-in bash tools (v. 3.2.57), mothur (v. 1.44.2), and R
-(v. \textbf{4.0.4}). Within R, I used the tidyverse (v. \textbf{1.3.0}),
-data.table (v. \textbf{1.13.2}), Rcpp (v. \textbf{1.0.5}), furrr (v.
-\textbf{0.2.1}), and rmarkdown (v. \textbf{2.5}) packages. The
-conception and development of this analysis is available as a playlist
-on the Riffomonas YouTube channel
+(v. 4.0.4). Within R, I used the tidyverse (v. 1.3.0), data.table (v.
+1.13.2), Rcpp (v. 1.0.5), furrr (v. 0.2.1), here (v. 1.0.1) and
+rmarkdown (v. 2.5) packages. The conception and development of this
+analysis is available as a playlist on the Riffomonas YouTube channel
 (\url{https://youtube.com/playlist?list=PLmNrK_nkqBpL7m_tyWdQgdyurerttCsPY}).

### 2021-06-23 submission/manuscript.Rmd
Respond to reviewers' comments

@@
 As an alternative to operational taxonomic units (OTUs), amplicon sequencing variants (ASVs) have been proposed as a way to adopt the thresholds suggested by genome sequencing to microbial community analysis using 16S rRNA gene sequences [@Edgar2016; @Amir2017; @Callahan2016; @Eren2014].
-Approaches for identifying ASVs do not cluster sequences based on a distance-based threshold [@Callahan2017].
-Proponents of ASVs are largely dissmissive of concerns that most bacterial genomes have more than one copy of the *rrn* operon and that those copies are not identical [@Pei2010; @Sun2013].
-Yet, ASVs and using too fine a threshold to identify OTUs could split a single genome into multiple clusters.
+It is widely understood that individual bacterial genomes often have multiple 16S rRNA genes that are not identical and that a 16S rRNA gene sequence could be found with different versions of the sequence in different genomes [@Pei2010; @Sun2013].
+This could lead to the problem that ASVs and using too fine a threshold to identify OTUs could split a single genome into multiple clusters.
+Proponents of ASVs minimize concerns that most bacterial genomes have more than one copy of the *rrn* operon and that those copies are not identical [@Callahan2017; @Edgar2018].
 Conversely, using too broad of a threshold to define OTUs could cluster together multiple bacterial species into the same OTU.
@@
 At broad thresholds, 16S rRNA gene sequences from multiple species could be clustered into the same ASV or OTU.
-I again randomly selected one genome from each species to control for uneven representation of genomes across species and for this analysis I measured the percentage of ASVs and OTUs that contained 16S rRNA gene sequences from multiple species (Figure 2).
+I again randomly selected one genome from each species to control for uneven representation of genomes across species.
+For this analysis I measured the percentage of ASVs and OTUs that contained 16S rRNA gene sequences from multiple species (Figure 2).
 Without using distance-based thresholds, `r format_pct(lump_v19)`% of the ASVs contained sequences from multiple species when considering full length sequences and `r format_pct(lump_v34)`, `r format_pct(lump_v4)`, and `r format_pct(lump_v45)`% when considering the V3-V4, V4, and V4-V5 regions, respectively.
 At the commonly used 3% threshold for defining OTUs, `r format_pct(lump3_v19)`% of the OTUs contained 16S rRNA gene sequences from multiple species when considering full length sequences and `r format_pct(lump3_v34)`, `r format_pct(lump3_v4)`, and `r format_pct(lump3_v45)`% when considering the V3-V4, V4, and V4-V5 regions, respectively.
-Considering that species designations are inconsistently applied and reflect multiple human-imposed biases, the risk of splitting a genome into multiple OTUs is more problematic than clustering species together.
-Therefore, larger thresholds are advisable.
+Although the actual fraction of ASVs and OTUs that contain sequences from multiple species is dependent on the taxonomic composition of the sequences in the *rrn*DB, this analysis highlights the tradeoffs of using distance-based thresholds.
 
@@
 Furthermore, the 16S rRNA gene does not evolve at the same rate across all bacterial lineages [@Schloss2011], which limits the biological interpretation of a common OTU definition.
-A distance-based definition of a taxonomic unit based on 16S rRNA gene or full genome sequences is, at best, operational and not grounded in biological theory [@Schloss2011; @Barco2020; @Parks2020; @Yarza2014].
+A distance-based definition of a taxonomic unit based on 16S rRNA gene or full genome sequences is operational and not necessarily grounded in biological theory [@Schloss2011; @Barco2020; @Parks2020; @Yarza2014].
+One benefit of a distance-based OTU definition is the ability to mask residual sequencing error.
+The analysis in this study was conducted using ideal sequences from assembled genomes whereas sequences generated in microbiome studies would harbor PCR and sequencing errors.
+These errors would only exacerbate the inflated number of ASVs.
 There is general agreement in bacterial systematics that to classify an organism to a bacterial species, phenotypic and genome sequence data are needed [@Staley2006; @Oren2013; @Baltrus2016; @Sanford2021].
@@
 It is not biologically plausible to entertain the possibility that different *rrn* operons from the same genome would have different ecologies.
+Individual bacteria are defined at the cellular or chromosomal level and not at the gene level.
+One could argue that, in practice, communities are compared on a relative rather than absolute basis.
+However, communities harboring populations that tend to have more copies of the *rrn* operon would appear to have higher richness and diversity than those with fewer copies purely due to the propensity for populations with more *rrn* operons to generate more ASVs.
 Although there are multiple reasons that proponents favor ASVs, the significant risk of artificially splitting genomes into separate clusters is too high to warrant their use.
@@
 
-\includegraphics{../figures/copy_number_threshold_plot.pdf}
-
 **Figure 1.
@@
 
-\newpage
-
-\includegraphics{../figures/lump_split.pdf}
-
 **Figure 2.
@@
 In each randomization, one genome was sampled from each species.
-
-\newpage
-
-\includegraphics{../figures/esv_rate.pdf}
 

### 2021-06-23 submission/manuscript.tex
Respond to reviewers' comments

@@
 suggested by genome sequencing to microbial community analysis using 16S
-rRNA gene sequences (7--10). Approaches for identifying ASVs do not
-cluster sequences based on a distance-based threshold (11). Proponents
-of ASVs are largely dissmissive of concerns that most bacterial genomes
-have more than one copy of the \emph{rrn} operon and that those copies
-are not identical (12, 13). Yet, ASVs and using too fine a threshold to
+rRNA gene sequences (7--10). It is widely understood that individual
+bacterial genomes often have multiple 16S rRNA genes that are not
+identical and that a 16S rRNA gene sequence could be found with
+different versions of the sequence in different genomes (11, 12). This
+could lead to the problem that ASVs and using too fine a threshold to
 identify OTUs could split a single genome into multiple clusters.
-Conversely, using too broad of a threshold to define OTUs could cluster
-together multiple bacterial species into the same OTU. An example of
-both is seen in the comparison of \emph{Staphylococcus aureus} (NCTC
-8325) and \emph{S. epidermidis} (ATCC 12228) where each genome has 5
-copies of the 16S rRNA gene. Each of the 10 copies of the 16S rRNA gene
-in these two genomes is distinct and represent 10 ASVs. Conversely, if
-the copies were clustered using a 3\% distance threshold, then all 10
-ASVs would cluster into the same OTU. The goal of this study was to
-quantify the tradeoff of splitting a single genome into multiple
-clusters and the risk of clustering different bacterial species into the
-same cluster when using ASVs and various OTU definitions.
+Proponents of ASVs minimize concerns that most bacterial genomes have
+more than one copy of the \emph{rrn} operon and that those copies are
+not identical (6, 13). Conversely, using too broad of a threshold to
+define OTUs could cluster together multiple bacterial species into the
+same OTU. An example of both is seen in the comparison of
+\emph{Staphylococcus aureus} (NCTC 8325) and \emph{S. epidermidis} (ATCC
+12228) where each genome has 5 copies of the 16S rRNA gene. Each of the
+10 copies of the 16S rRNA gene in these two genomes is distinct and
+represent 10 ASVs. Conversely, if the copies were clustered using a 3\%
+distance threshold, then all 10 ASVs would cluster into the same OTU.
+The goal of this study was to quantify the tradeoff of splitting a
+single genome into multiple clusters and the risk of clustering
+different bacterial species into the same cluster when using ASVs and
+various OTU definitions.
 
@@
 genome from each species to control for uneven representation of genomes
-across species and for this analysis I measured the percentage of ASVs
-and OTUs that contained 16S rRNA gene sequences from multiple species
+across species. For this analysis I measured the percentage of ASVs and
+OTUs that contained 16S rRNA gene sequences from multiple species
 (Figure 2). Without using distance-based thresholds, 4.1\% of the ASVs
@@
 and 34.8\% when considering the V3-V4, V4, and V4-V5 regions,
-respectively. Considering that species designations are inconsistently
-applied and reflect multiple human-imposed biases, the risk of splitting
-a genome into multiple OTUs is more problematic than clustering species
-together. Therefore, larger thresholds are advisable.
+respectively. Although the actual fraction of ASVs and OTUs that contain
+sequences from multiple species is dependent on the taxonomic
+composition of the sequences in the \emph{rrn}DB, this analysis
+highlights the tradeoffs of using distance-based thresholds.
 
@@
 or too fine of a threshold to define OTUs. An ongoing problem for
-amplicon-based studies is defining a meaningful taxonomic unit (11, 15,
+amplicon-based studies is defining a meaningful taxonomic unit (13, 15,
 16). Since there is no consensus for a biologicaly definition of a
@@
 distance-based definition of a taxonomic unit based on 16S rRNA gene or
-full genome sequences is, at best, operational and not grounded in
-biological theory (15, 22--24). There is general agreement in bacterial
-systematics that to classify an organism to a bacterial species,
-phenotypic and genome sequence data are needed (17--20). A short
-sequence from a bacterial genome simply cannot differentiate between
-species. Moreover, it is difficult to defend a clustering threshold that
-would split a single genome into multiple taxonomic units. It is not
-biologically plausible to entertain the possibility that different
-\emph{rrn} operons from the same genome would have different ecologies.
-Although there are multiple reasons that proponents favor ASVs, the
-significant risk of artificially splitting genomes into separate
-clusters is too high to warrant their use.
+full genome sequences is operational and not necessarily grounded in
+biological theory (15, 22--24). One benefit of a distance-based OTU
+definition is the ability to mask residual sequencing error. The
+analysis in this study was conducted using ideal sequences from
+assembled genomes whereas sequences generated in microbiome studies
+would harbor PCR and sequencing errors. These errors would only
+exacerbate the inflated number of ASVs. There is general agreement in
+bacterial systematics that to classify an organism to a bacterial
+species, phenotypic and genome sequence data are needed (17--20). A
+short sequence from a bacterial genome simply cannot differentiate
+between species. Moreover, it is difficult to defend a clustering
+threshold that would split a single genome into multiple taxonomic
+units. It is not biologically plausible to entertain the possibility
+that different \emph{rrn} operons from the same genome would have
+different ecologies. Individual bacteria are defined at the cellular or
+chromosomal level and not at the gene level. One could argue that, in
+practice, communities are compared on a relative rather than absolute
+basis. However, communities harboring populations that tend to have more
+copies of the \emph{rrn} operon would appear to have higher richness and
+diversity than those with fewer copies purely due to the propensity for
+populations with more \emph{rrn} operons to generate more ASVs. Although
+there are multiple reasons that proponents favor ASVs, the significant
+risk of artificially splitting genomes into separate clusters is too
+high to warrant their use.
 
@@
 made use of built-in bash tools (v. 3.2.57), mothur (v. 1.44.2), and R
-(v. 4.0.4). Within R, I used the tidyverse (v. 1.3.0), data.table (v.
-1.13.2), Rcpp (v. 1.0.5), furrr (v. 0.2.1), here (v. 1.0.1) and
-rmarkdown (v. 2.5) packages. The conception and development of this
+(v. 4.1.0). Within R, I used the tidyverse (v. 1.3.1), data.table (v.
+1.14.0), Rcpp (v. 1.0.6), furrr (v. 0.2.2), here (v. 1.0.1) and
+rmarkdown (v. 2.8) packages. The conception and development of this
 analysis is available as a playlist on the Riffomonas YouTube channel
@@
 
-\leavevmode\hypertarget{ref-Callahan2017}{}%
-11. Callahan BJ, McMurdie PJ, Holmes SP. 2017. Exact sequence variants
-should replace operational taxonomic units in marker-gene data analysis.
-\emph{The ISME Journal} 11:2639--2643. doi:
-\href{https://doi.org/10.1038/ismej.2017.119}{10.1038/ismej.2017.119}.
-
 \leavevmode\hypertarget{ref-Pei2010}{}%
-12. Pei AY, Oberdorf WE, Nossa CW, Agarwal A, Chokshi P, Gerz EA, Jin Z,
+11. Pei AY, Oberdorf WE, Nossa CW, Agarwal A, Chokshi P, Gerz EA, Jin Z,
 Lee P, Yang L, Poles M, Brown SM, Sotero S, DeSantis T, Brodie E, Nelson
@@
 \leavevmode\hypertarget{ref-Sun2013}{}%
-13. Sun D-L, Jiang X, Wu QL, Zhou N-Y. 2013. Intragenomic heterogeneity
+12. Sun D-L, Jiang X, Wu QL, Zhou N-Y. 2013. Intragenomic heterogeneity
 of 16S rRNA genes causes overestimation of prokaryotic diversity.
@@
 \href{https://doi.org/10.1128/aem.01282-13}{10.1128/aem.01282-13}.
+
+\leavevmode\hypertarget{ref-Callahan2017}{}%
+13. Callahan BJ, McMurdie PJ, Holmes SP. 2017. Exact sequence variants
+should replace operational taxonomic units in marker-gene data analysis.
+\emph{The ISME Journal} 11:2639--2643. doi:
+\href{https://doi.org/10.1038/ismej.2017.119}{10.1038/ismej.2017.119}.
 
@@
 
-\includegraphics{../figures/copy_number_threshold_plot.pdf}
-
 \textbf{Figure 1. The distance threshold required to prevent the
@@
 
-\newpage
-
-\includegraphics{../figures/lump_split.pdf}
-
 \textbf{Figure 2. As the distance threshold used to define an OTU
@@
 genome was sampled from each species.
-
-\newpage
-
-\includegraphics{../figures/esv_rate.pdf}
 

### 2021-06-23 code/plot_lump_split.R

@@ -19,7 +19,7 @@ read_tsv("data/processed/lumped_split_rate.tsv",
 		) +
 	custom_color_scale() +
 	labs(
-		x="Distance theshold\nused to define OTUs (%)",
+		x="Distance threshold\nused to define OTUs (%)",
 		y="Percentage of ASVs or OTUs that included\nsequences from multiple species") +
 	theme_classic() +
 	theme(
